---
title: "Separating balsam woolly adelgid from bark beetle damage in Pacific silver fir on Vancouver Island: a multi-sensor attribution test"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: >
  Balsam woolly adelgid (*Adelges piceae*) and western balsam bark beetle
  (*Dryocoetes confusus*) are both mapped on Pacific silver fir (*Abies
  amabilis*) in the British Columbia aerial overview survey, so a canopy-damage
  detector that cannot separate them measures neither. In this study, we examine
  whether the addition of lidar and radar predictors to a spectral baseline improves
  separation of the two agents, and by how much.
keywords:
  - balsam woolly adelgid
  - Adelges piceae
  - Abies amabilis
  - forest health monitoring
  - lidar
  - synthetic aperture radar
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
library(sf)
library(terra)
library(flextable)

root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
outputs <- file.path(root, "03.outputs")
derived <- file.path(inputs, "derived")

BC_ALBERS <- 3005

# Figures are written to 03.outputs/figures as well as embedded, so the README
# and any submission package reference committed files rather than extracting
# images from a rendered document.
#
# The path must be relative to this document. Pandoc resolves an image src
# against the document's directory, so an absolute path arrives as
# "./Volumes/..." which resolves to nothing, and embed-resources then drops
# every figure without failing the render. The files still land in the same
# place; only the reference changes.
knitr::opts_chunk$set(fig.path = "../03.outputs/figures/",
                      dpi = 300, dev = "png")

fmt_tbl <- function(df, right = NULL) {
  right_cols <- union(names(df)[vapply(df, is.numeric, TRUE)], right)
  left_cols <- setdiff(names(df), right_cols)
  ft <- flextable::flextable(df)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "header")
  if (length(right_cols))
    ft <- flextable::align(ft, j = right_cols, align = "right", part = "body")
  if (length(left_cols))
    ft <- flextable::align(ft, j = left_cols, align = "left", part = "body")
  ft <- flextable::line_spacing(ft, space = 1, part = "all")
  ft <- flextable::padding(ft, padding.top = 1, padding.bottom = 1,
                           part = "all")
  flextable::set_table_properties(ft, layout = "autofit", width = 1)
}
```

```{r}
#| label: load-labels
#| echo: true
#| output: false
iab <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                   quiet = TRUE)
vi <- sf::st_read(file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)

# The analysis set, built by 05.scripts/14-build-analysis-set.R: adelgid and bark
# beetle polygons on Pacific silver fir that fall under lidar coverage acquired
# within five years of the survey, on Vancouver Island. This file is the sample;
# nothing below filters it again.
aset <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
n_set <- nrow(aset)
n_set_iab <- sum(aset$PEST_SPECIES_CODE == "IAB")
n_set_ibb <- sum(aset$PEST_SPECIES_CODE == "IBB")
# Reported in the direction of the majority class, so the prose cannot name the
# wrong one if the sample is rebuilt.
set_imbalance <- round(max(n_set_iab, n_set_ibb) / min(n_set_iab, n_set_ibb), 1)
set_majority <- if (n_set_iab >= n_set_ibb) "adelgid" else "bark beetle"
set_years <- range(aset$CAPTURE_YEAR)
set_offset_med <- median(aset$year_offset)
set_offset_unit <- if (set_offset_med == 1) "year" else "years"
set_ha <- round(sum(aset$AREA_HA))
# Counts behind the landmass restriction, saved by the same script, so the
# justification for excluding the mainland is recomputable rather than quoted.
restr <- read.csv(file.path(derived, "analysis-set-restriction.csv"))
off <- restr[restr$landmass != "Vancouver Island", ]
n_off <- sum(off$n)
n_off_ibb <- sum(off$n[off$agent == "IBB"])
n_off_campaign <- max(tapply(off$n, off$lidar_year, sum))
yr_off_campaign <- names(which.max(tapply(off$n, off$lidar_year, sum)))

host_label <- c(B = "True fir, unspecified", BA = "Pacific silver fir",
                BG = "Grand fir", BL = "Subalpine fir")
agent_label <- c(IAB = "Balsam woolly adelgid",
                 IBB = "Western balsam bark beetle")
sev_label <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
               V = "Very severe")

# The severity classes actually present in the sample, named rather than
# counted, so the sentence reporting them cannot go stale or read "0 polygons".
sev_set_txt <- {
  s <- tolower(sev_label[intersect(names(sev_label),
                                   unique(aset$PEST_SEVERITY_CODE))])
  if (length(s) == 1) s else
    paste(paste(s[-length(s)], collapse = ", "), "and", s[length(s)])
}

n_iab_bc <- nrow(iab)
yr_iab <- range(iab$CAPTURE_YEAR)
n_vi_iab <- sum(vi$PEST_SPECIES_CODE == "IAB")
n_vi_ibb <- sum(vi$PEST_SPECIES_CODE == "IBB")
ha_vi_iab <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IAB"]))
ha_vi_ibb <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IBB"]))

ba <- vi[vi$TREE_SPECIES_CODE == "BA", ]
n_ba_iab <- sum(ba$PEST_SPECIES_CODE == "IAB")
n_ba_ibb <- sum(ba$PEST_SPECIES_CODE == "IBB")
imbalance <- round(n_ba_ibb / n_ba_iab, 1)

sev_iab <- table(factor(vi$PEST_SEVERITY_CODE[vi$PEST_SPECIES_CODE == "IAB"],
                        levels = names(sev_label)))
n_iab_ms <- sum(sev_iab[c("M", "S", "V")])
```

# Introduction

Balsam woolly adelgid, *Adelges piceae* (Ratzeburg), is an introduced insect that damages and kills North American true firs of all age classes, having entered the continent in the early twentieth century through separate introductions on the east and west coasts [@Campbell_2023]. Its congener the hemlock woolly adelgid, *Adelges tsugae* Annand, is the better-studied insect, but the two behave differently on this coast. In eastern North America *A. tsugae* is invasive and lethal, and infested eastern hemlock die within four to fifteen years with within-stand mortality that can exceed ninety percent [@MacQuarrie_2025]. In western North America the same species is a separate and native lineage that "exists at low densities in forests and rarely kills trees" [@MacQuarrie_2025]. Damage to western hemlock in British Columbia has accordingly been minor, and the provincial forest health survey reflects the asymmetry directly: it carries a damage-agent code for balsam woolly adelgid and none for hemlock woolly adelgid. In British Columbia the woolly adelgid of management concern is therefore *A. piceae* on true fir, and its principal host on the coast is Pacific silver fir, *Abies amabilis* (Douglas) Forbes.

The two agents of interest here attack the same host and are recorded by the same survey, so a manager reading a damage map of Pacific silver fir cannot tell from the map which insect produced it. Their damage differs in kind, however, and that difference is the basis of this study; it is set out in the study system below.

Remote detection of adelgid damage has a long record. @FRANKLIN_1995 separated damage classes on individual balsam fir from airborne multispectral imagery at 40 to 76 percent accuracy. Landsat change detection was applied to adelgid-induced hemlock decline from the late 1990s [@Royle_1997; @Bonneau_1999], and later work pushed detection towards lower infestation levels [@Williams_2017]. @Cook_2010 discriminated infested from non-infested subalpine fir at 94 percent from branch-level spectra. Most recently @Campbell_2023 mapped infestation severity in subalpine fir from satellite imagery with terrain and climate, and their result frames the present study: spectral data alone "successfully detected forest degradation" but "failed to distinguish between causal agents of change", returning a mean coefficient of determination of 0.482, while terrain and climate reached 0.746 but "lacked the ability to characterize local-scale tree damage", and only the combination reached 0.836.

Attribution has since become a distinct problem with its own methods, addressed mainly with optical time series. @Oeser_2017 attributed windthrow, cleared windthrow, bark beetle and harvest in Central Europe with overall accuracies of 76 to 86 percent; @Zhang_2022 mapped causal agents across boreal and arctic North America; @Sebald_2021 improved separation of human from natural disturbance using landscape context; and @Morin_Bernard_2023 reached 82.9 percent attributing agent before estimating severity, finding that agent-specific severity models required different predictors. Structure and radar have entered the field on the detection side. Terrestrial lidar detects *Ips typographus* infestation early [@Junttila_2019], simulated spaceborne lidar recovers the understorey and midstorey defoliation that hemlock woolly adelgid causes before any overstorey signal appears [@Boucher_2020], and L-band synthetic aperture radar delineates windthrow from insect outbreak at 69 to 88 percent [@Tanase_2018; @Hollaus_2019]. Most directly, @Atkins_2020 used terrestrial lidar canopy structural traits in a machine learning framework across six moderate-severity agents including hemlock woolly adelgid, and found that fire and ice storms changed the amount and position of canopy vegetation while insect and pathogen infestation altered canopy arrangement and complexity instead.

Two limits define what remains unresolved, and both emerged from a structured search of this literature. We searched Semantic Scholar and Crossref for work on adelgid remote sensing, on attribution of damage to specific agents, on lidar and radar applied to insect damage, and on the accuracy of the aerial surveys that supply agent labels, screening 5,663 records against criteria fixed before searching and carrying 721 forward for reading. The searched literature is deep on detection and comparatively shallow on attribution, and the two limits below are what that asymmetry exposes.

First, attribution between insect agents has been demonstrated, but only where the agents attack different hosts. @Meigs_2011 first characterised the spectral trajectories of a defoliator and a bark beetle separately against field plots, finding that western spruce budworm showed long-duration decline followed by recovery while mountain pine beetle showed both short and long-duration declines with variable recovery, but stopped short of classifying them. @Senf_2015 took that step, separating mountain pine beetle from western spruce budworm in British Columbia using LandTrendr spectral trajectories and random forests, mapping insect disturbance at 76.8 percent overall accuracy and attributing agent at "overall accuracies ranging from 75.3% to 88.0%, depending on whether only pure host-stands or mixed stands with both insect hosts were considered". That conditional is the important result for the present study. Where host composition disambiguated the agents, attribution succeeded; where it did not, it degraded sharply, and the authors report that "in the case of mixed host stands, nearly 45% of the western spruce budworm disturbances were falsely attributed to mountain pine beetle". Host is therefore doing much of the work that appears to be done by damage. Balsam woolly adelgid and western balsam bark beetle attack the same host species, which removes that cue entirely rather than diluting it, and we found no study that attributes between insect agents under that condition. Second, the structural evidence that agents leave distinguishable signatures rests on plot-scale instrumented sites; @Atkins_2020 state their findings are "limited in their generalizability due to lack of replication among disturbances". Whether canopy structure carries agent-diagnostic information at the coarse support of an operational survey, across a landscape, is untested.

That distinction matters operationally because the survey record is itself the working map of which insect is where. @Coleman_2018 assessed aerial detection surveys against ground observation across multiple ecological regions and reported overall accuracies above 70 percent, with commission errors of 2 and 5 percent for mortality and defoliation damage types, but found that "accuracy declined as the specificity for observations went from genera to species level for tree species and damaging agents". Species-level agent attribution is precisely the level at which balsam woolly adelgid and western balsam bark beetle are distinguished, and independent evaluations of survey severity estimates have returned overall accuracies as low as 35 to 56 percent depending on the class scheme [@Bright_2020b; @Housman_2018].

Each of our objectives follows from one of these findings. Because @Campbell_2023 established that spectral predictors detect adelgid degradation but fail to identify its cause, and because @Senf_2015 showed that spectral attribution between insect agents weakens when host stops disambiguating them, our first objective is to quantify how far a spectral-only classifier separates balsam woolly adelgid from western balsam bark beetle damage in Pacific silver fir on Vancouver Island, establishing that baseline rather than assuming it. Because the structural work of @Atkins_2020 and @Boucher_2020 shows that agents alter canopy arrangement in ways that colour alone does not register, but shows it at plot scale and without replication, our second objective is to test whether adding lidar-derived canopy structure and Sentinel-1 backscatter improves that separation at operational support. And because @Oeser_2017, @Sebald_2021 and @Morin_Bernard_2023 demonstrate that attribution accuracy depends heavily on how the evaluation is constructed, our third objective is to report any improvement against a validation design that respects the spatial dependence of the labels and the support at which they were drawn, rather than against a random partition that would inflate it.

We hypothesise that the multi-sensor model achieves higher balanced accuracy than the spectral baseline, because gout and progressive top-kill alter canopy geometry in a way that rapid whole-crown mortality does not, and because the spectral baseline is expected to be weak here for the reason @Senf_2015 identify: with host held constant it cannot fall back on host composition. The null hypothesis is that structural and radar predictors add nothing beyond the spectral baseline.

# Materials and methods

## Study system

Balsam woolly adelgid is a small, wingless, parthenogenetic insect that feeds on the bark of true firs through a fine stylet. In North America it reproduces without males and disperses passively, as first-instar crawlers carried by wind, birds and human movement of infested material, so spread is slow, local and irregular rather than following a front. Feeding is concentrated on the bole and in the crown, and its damage is chemically mediated rather than mechanical: saliva injected during feeding is toxic to the host, inhibiting bud formation and inducing abnormal xylem. The visible consequences are swellings at nodes and branch tips, known as gouting, whose incidence varies systematically with crown class and plant hardiness zone [@Guillet_2010], followed by crown thinning and progressive dieback from the top of the tree downward. Trees may persist in this deformed and declining condition for years, and death, where it occurs, follows a long structural decline rather than preceding it.

Western balsam bark beetle, *Dryocoetes confusus* Swaine, is the other agent recorded on the same host in this landscape. It is a phloem-feeding beetle that mass-attacks the bole, girdling the tree by disrupting phloem transport and introducing a pathogenic fungus. Its damage is not preceded by structural deformation of branches or by top-down crown thinning: an attacked tree retains its architecture and fades whole-crown, from green through yellow to red, over roughly one to two years.

The beetle does not act alone, and the comparison class is defined accordingly. On true fir it operates within a mortality complex alongside *Armillaria* root disease, the associated fungus *Ophiostoma dryocoetidis* and, at some sites, spruce beetle. Interactions of this kind alter canopy and stand characteristics jointly rather than separately: @Klutsch_2014 found on a single ponderosa pine host that "fuel bed and canopy characteristics were dependent on the presence of both dwarf mistletoe and mountain pine beetle-caused mortality". The class contrasted with the adelgid here is therefore best understood as bark-beetle-associated mortality rather than as the effect of one insect in isolation, and it is described that way throughout. This is a property of the system, not of the labels, and it would persist under any labelling scheme. The two agents therefore differ in the geometry of the damage, deformation and progressive top-kill against retained architecture, and in its tempo, years of decline against a rapid whole-crown fade.

That contrast is the mechanism this study tests. An instrument sensitive only to foliage colour observes both agents at the end state, when crowns of either kind are dead or dying and spectrally similar. An instrument sensitive to canopy geometry observes the difference in how the crown arrived there. Whether that difference survives at the coarse support of an operational survey polygon, rather than at the plot scale where it has been examined, is the open question. The plot-scale evidence is itself mixed, and both sides of it concern the congener. @Atkins_2020 found that insect and pathogen infestation altered canopy arrangement and complexity, and @Alveshere_2025 report that hemlock woolly adelgid infestation influences canopy structural complexity, while @Choi_2023 found for press disturbances including the same insect that "the impact ... on other dimensions of canopy structure could not be clearly detected, likely because of compensatory growth". A structural signal is therefore plausible but not established, and compensatory growth is a specific mechanism by which it could be absent.

Pacific silver fir, *Abies amabilis*, is the host that carries both agents on the coast. It is a shade-tolerant, late-seral true fir of the wetter and cooler coastal forests, occurring from near sea level to the subalpine, and it is a commercially and ecologically significant component of the stands in which it grows. Subalpine fir, *Abies lasiocarpa*, carries western balsam bark beetle at higher elevations in the same region but is not recorded with adelgid damage here, so the analysis is confined to Pacific silver fir in order to hold host constant.

## Study area

The study area is Vancouver Island, British Columbia, Canada. Survey records are shown over a wider envelope from 128.5 to 123.0 degrees west and 48.3 to 51.0 degrees north (@fig-studyarea), which takes in the adjacent mainland coast, but the analysis is confined to the island itself for the reason set out with the sample below. Vancouver Island is roughly 460 km long and separated from the mainland by the Strait of Georgia and Johnstone Strait. A mountain spine runs its length, rising above 2000 m in the Vancouver Island Ranges, and it divides a wet, exposed west coast from a drier eastern lowland in the rain shadow of the same mountains.

Forest composition follows that gradient. The eastern lowlands carry the Coastal Douglas-fir zone, the wetter and more extensive middle elevations the Coastal Western Hemlock zone, and the upper slopes the Mountain Hemlock zone before passing into subalpine parkland. Pacific silver fir occurs through the Coastal Western Hemlock and Mountain Hemlock zones, generally above the Coastal Douglas-fir lowlands, growing with western hemlock, western redcedar and yellow cedar. Subalpine fir is largely confined to the highest elevations. The climate is maritime, with mild wet winters and comparatively dry summers, and it moderates the winter cold that constrains adelgid survival at the northern and continental edges of the insect's range.

Both damage agents are recorded across this area by the provincial aerial overview survey, and their distributions overlap on the shared host without coinciding: the adelgid record concentrates at lower and middle elevations on Pacific silver fir, while western balsam bark beetle extends to higher elevations where subalpine fir dominates (@fig-studyarea).

All spatial data are handled in NAD83 / BC Albers, EPSG 3005, which is the projection of the source survey archive, and every map is drawn in that projection.

```{r}
#| label: fig-studyarea
#| fig-cap: "Study area on Vancouver Island and the adjacent mainland coast of British Columbia, Canada, showing aerial overview survey polygons attributed to balsam woolly adelgid and to western balsam bark beetle on true fir hosts. Relief is shaded from a 1 km digital elevation model illuminated from the northwest. Projection NAD83 / BC Albers (EPSG 3005); axis labels in degrees."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
bm <- file.path(derived, "basemap.gpkg")
land   <- sf::st_read(bm, "land",   quiet = TRUE)
admin  <- sf::st_read(bm, "admin",  quiet = TRUE)
rivers <- sf::st_read(bm, "rivers", quiet = TRUE)
lakes  <- sf::st_read(bm, "lakes",  quiet = TRUE)
places <- sf::st_read(bm, "places", quiet = TRUE)
hs <- terra::rast(file.path(derived, "hillshade.tif"))
# The view rectangle, not the extent of the land. Land and relief are prepared
# wider than the view so that the fixed aspect ratio, which expands the shorter
# axis, does not open a band of sea where the mainland belongs.
frame_box <- sf::st_read(bm, "frame", quiet = TRUE)

vi_p <- sf::st_transform(vi, BC_ALBERS)
fr <- sf::st_bbox(frame_box)

op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")

# Sea, then relief-shaded land, then hydrography, then boundaries. The sea is
# painted across the whole plot region rather than the land bounding box, since
# the fixed aspect ratio leaves the region taller than the data.
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.30, end = 1),
            legend = FALSE, add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey30", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = "#6f9ab5", lwd = 0.3,
     add = TRUE)
plot(sf::st_geometry(rivers), col = "#5b8db0", lwd = 0.8, add = TRUE)
# Only the international boundary, drawn dashed. The provincial outline is not
# drawn because within this frame it coincides with the coastline.
plot(sf::st_geometry(admin), col = "grey20", lwd = 1.1, lty = 2, add = TRUE)

# Bark beetle first: the far less numerous adelgid class must stay legible.
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IBB", ]),
     col = "#2c6f9e", border = "#2c6f9e", add = TRUE)
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IAB", ]),
     col = "#c1372f", border = "#c1372f", add = TRUE)

pxy <- sf::st_coordinates(places)
points(pxy, pch = 21, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Label side chosen per settlement so no label runs off the frame or across a
# neighbouring one.
lab_pos <- c(Comox = 4, `Campbell River` = 2, `Port Alberni` = 2,
             Nanaimo = 4, Victoria = 2, Vancouver = 4, Tofino = 2,
             `Port Hardy` = 4)
text(pxy[, 1], pxy[, 2], places$name, pos = lab_pos[places$name],
     offset = 0.4, cex = 0.6, col = "grey10")

# Axes labelled in degrees although the plot is projected, so the reader gets
# geographic coordinates without the distortion of plotting in degrees.
xt <- pretty(fr[c("xmin", "xmax")], 5)
yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

# Scale bar drawn in projected metres, so the stated length is metrically true.
sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")

# North arrow.
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.62, pt.cex = 1.1, pch = 22,
       pt.bg = c("#c1372f", "#2c6f9e"), col = c("#c1372f", "#2c6f9e"),
       legend = c("Balsam woolly adelgid", "Western balsam bark beetle"))
par(op)
```

## Data

@tbl-inputs lists each dataset and its access terms. The custodian of the survey archive is the BC Ministry of Forests; GeoBC operates the lidar programme and the two are separate agencies.

```{r}
#| label: tbl-inputs
#| tbl-cap: "Datasets used in the analysis, with their role and access terms."
inputs_tbl <- data.frame(
  Dataset = c("Pest Infestation Polygons", "Sentinel-2 L2A", "Sentinel-1 GRD",
              "LidarBC open lidar"),
  Role = c("Damage agent labels", "Optical predictors", "Radar predictors",
           "Canopy structure predictors"),
  Source = c("BC Data Catalogue", "Google Earth Engine",
             "Google Earth Engine", "LidarBC open data portal"),
  Licence = c("Open Government Licence - British Columbia",
              "Copernicus open access", "Copernicus open access",
              "Open Government Licence - British Columbia")
)
fmt_tbl(inputs_tbl)
```

Damage agents in the delivered geodatabase are recorded as three-letter codes without names. Codes were resolved against the *Forest Health Aerial Overview Survey Standards for British Columbia*, the Ministry of Forests adaptation of Canadian Forest Service FHN Report 97-1, which gives IAB for balsam woolly adelgid and IBB for western balsam bark beetle. The same document lists IAB against sub-alpine true firs, amabilis fir and grand fir, matching the host codes present in the extract. That code list contains no entry for hemlock woolly adelgid, so the absence of that species from the archive is a property of the survey design rather than of the years sampled.

### Survey data

Damage labels are polygons from the British Columbia aerial overview survey. Surveyors sketch-map visible forest damage from a fixed-wing aircraft, recording host species, damage agent and an ordinal severity class for each polygon. A minimum of eighty percent of the province is flown annually, with gaps caused by weather and smoke.

Survey polygons carry limitations that constrain the analysis and that the custodian states explicitly: errors of omission and commission occur, spatial accuracy can be poor, and the survey is not intended to be accurate enough to direct operational activities. Its stated purpose is to record general trends in disturbance size and frequency. Polygon area is in part a product of the digitising rule rather than a measurement, since a clump of one to thirty dead or dying trees is delineated as 0.25 ha and a clump of thirty-one to fifty as 0.5 ha. Severity is likewise an ordinal judgement: the standards define light, moderate and severe as 1 to 10, 11 to 29 and 30 percent or more of trees recently killed, and classify all point sources as severe on digitising. The unit of analysis here is therefore the survey polygon rather than the pixel, and severity is not modelled.

Across the study area the survey records `r n_vi_iab` adelgid polygons covering `r format(ha_vi_iab, big.mark = ",")` ha and `r n_vi_ibb` bark beetle polygons covering `r format(ha_vi_ibb, big.mark = ",")` ha on true fir hosts (@tbl-labels). Province-wide the adelgid record comprises `r n_iab_bc` polygons spanning `r yr_iab[1]` to `r yr_iab[2]`. Both agents are recorded on Pacific silver fir, `r n_ba_iab` adelgid against `r n_ba_ibb` bark beetle polygons, and it is this shared host that makes the separation an attribution problem rather than a detection problem. Analysis is restricted to that host so that agent is not confounded with differences in host reflectance.

```{r}
#| label: tbl-labels
#| tbl-cap: "Aerial overview survey polygons in the study area by damage agent and host tree species. Area is the survey-recorded polygon area."
lab <- as.data.frame(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Host  = host_label[vi$TREE_SPECIES_CODE]
), stringsAsFactors = FALSE)
names(lab)[3] <- "Polygons"
lab <- lab[lab$Polygons > 0, ]

area <- tapply(vi$AREA_HA,
               list(agent_label[vi$PEST_SPECIES_CODE],
                    host_label[vi$TREE_SPECIES_CODE]),
               function(v) round(sum(v)))
lab$`Area (ha)` <- mapply(function(a, h) area[a, h], lab$Agent, lab$Host)
lab <- lab[order(lab$Agent, -lab$Polygons), ]
fmt_tbl(lab)
```

### Analysis sample

Not every polygon on the shared host can enter the analysis, because the structural predictors require lidar and lidar coverage is neither universal nor contemporaneous with the survey. The sample is therefore defined by data availability and is stated before any result, so that its restrictions are read as a property of the design rather than as a post hoc exclusion.

A polygon enters the analysis when it is attributed to either agent, lies on Pacific silver fir, falls within LidarBC point cloud coverage, has a nearest lidar acquisition within five years of its survey year, and lies on Vancouver Island.

The last condition guards against confounding rather than saving effort. Of the `r n_off` polygons that satisfy the first four conditions but lie off the island, `r n_off_ibb` are bark beetle and `r n_off_campaign` fall under a single `r yr_off_campaign` lidar acquisition. Landmass would therefore predict the agent almost perfectly, and so would acquisition campaign, to which every structural metric used here responds through pulse density, scan angle and flight geometry. Spatially blocked validation prevents adjacent polygons leaking between folds but does nothing to stop a model learning region in place of damage, because the association holds within folds as well as across them. Region is accordingly neutralised by restriction, the same instrument that holds host constant.

The five conditions together yield `r n_set` polygons, `r n_set_iab` adelgid and `r n_set_ibb` bark beetle, covering `r format(set_ha, big.mark = ",")` ha, surveyed between `r set_years[1]` and `r set_years[2]`, with a median label-to-acquisition offset of `r set_offset_med` `r set_offset_unit`. The classes are imbalanced by `r set_imbalance` to 1 in favour of the `r set_majority`, considerably more even than the `r imbalance` to 1 of the unrestricted host-matched record.

The cost of that definition is severity, and it is severe. The only severity classes present in the analysis sample are `r sev_set_txt`. This is not an artefact of the filter thresholds. Across the province no adelgid polygon rated moderate or worse falls under lidar at a one, three, five or ten year offset, and only one does at twenty years, by which point stand condition cannot be assumed to have persisted. The moderate and severe adelgid record and the lidar programme do not overlap in British Columbia, and no choice of study area within the province changes that. The study accepts a low-severity sample in exchange for the structural arm, and the consequences for interpretation are set out in the limitations.

The severity distribution is strongly weighted towards the two lowest classes (@tbl-severity). Only `r n_iab_ms` adelgid polygons in the study area are rated moderate or worse, too few to support the severity regression used by @Campbell_2023, and a further reason the response here is the two-class agent label.

```{r}
#| label: tbl-severity
#| tbl-cap: "Severity class of survey polygons by damage agent. Severity is the surveyor's ordinal rating."
sev <- as.data.frame.matrix(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Severity = factor(sev_label[vi$PEST_SEVERITY_CODE],
                    levels = unname(sev_label))
))
sev <- sev[, colSums(sev) > 0, drop = FALSE]
sev <- cbind(Agent = rownames(sev), sev)
fmt_tbl(sev)
```

The two agents also peak in different survey years (@fig-timeseries). The adelgid record concentrates in 2019 to 2021 while the bark beetle record is distributed more evenly across the past decade. A classifier trained on imagery drawn from each polygon's own survey year would therefore risk learning the year rather than the agent, and a single common imagery epoch is used for both classes.

```{r}
#| label: fig-timeseries
#| fig-cap: "Survey polygons mapped per year on true fir hosts in the study area, by damage agent."
#| fig-width: 7
#| fig-height: 3.6
tt <- table(vi$CAPTURE_YEAR, agent_label[vi$PEST_SPECIES_CODE])
op <- par(mar = c(4, 4.2, 0.6, 0.6))
matplot(as.numeric(rownames(tt)), tt, type = "l", lty = 1, lwd = 2,
        col = c("#c1372f", "#2c6f9e"), las = 1,
        xlab = "Survey year", ylab = "Polygons mapped")
legend("topleft", colnames(tt), col = c("#c1372f", "#2c6f9e"),
       lty = 1, lwd = 2, bty = "n", cex = 0.75)
par(op)
```

### Remote sensing data

Optical predictors are drawn from Sentinel-2 Level 2A surface reflectance and radar predictors from Sentinel-1 Ground Range Detected imagery in interferometric wide swath mode, both accessed through Google Earth Engine. Scenes are restricted to the leaf-on window of the epoch in which the adelgid record concentrates, 1 June 2019 to 30 September 2021, and optical scenes to those with less than forty percent cloud cover. Scene counts available over the study area are given in @tbl-imagery.

```{r}
#| label: gee-availability
#| echo: true
#| output: false
gee_status <- tryCatch({
  library(rgee)
  ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  aoi <- ee$Geometry$Rectangle(c(-128.5, 48.3, -123.0, 51.0))

  s2 <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))

  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$eq("instrumentMode", "IW"))

  data.frame(
    Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
    Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
    Period = "2019-06-01 to 2021-09-30",
    Scenes = c(s2$size()$getInfo(), s1$size()$getInfo())
  )
}, error = function(e) {
  data.frame(Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
             Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
             Period = "2019-06-01 to 2021-09-30",
             Scenes = NA_integer_)
})
```

```{r}
#| label: tbl-imagery
#| tbl-cap: "Satellite scenes available over the study area for the analysis period."
fmt_tbl(gee_status)
```

### Lidar data

Canopy structure is derived from the LidarBC open lidar programme, which publishes classified point clouds for public download. Coverage over the study area comprises 18,840 tiles at a nominal density of eight points per square metre (@tbl-lidar), including acquisitions in 2019 that coincide with the concentration of the adelgid survey record. Tiles are classified to ground, low point and overlap but carry no vegetation classes, so canopy height is computed as height above a ground surface interpolated from ground returns rather than read from an existing classification.

```{r}
#| label: tbl-lidar
#| tbl-cap: "Lidar point cloud coverage over the study area and over the Comox field window. Acquisition years and point classes are summarised from a sample of up to 1000 tiles."
lidar <- read.csv(file.path(derived, "lidar-coverage.csv"), check.names = FALSE)
names(lidar) <- c("Extent", "Tiles", "Full index", "Acquisition years",
                  "Density (pts/m2)", "Point classes")
fmt_tbl(lidar)
```

### Field site selection

Ground confirmation requires polygons that carry both an adelgid label and lidar coverage, and that can be reached. @fig-fieldsites shows where those conditions coincide, and the answer constrains the design rather than merely illustrating it.

```{r}
#| label: fig-fieldsites
#| fig-cap: "Candidate field sites. Points are aerial overview survey polygons attributed to balsam woolly adelgid, sized by area and coloured by severity class; open circles lack lidar coverage and filled circles have it. Shaded envelopes show LidarBC point cloud coverage. Rings mark 50, 100 and 150 km from Comox and are shown for travel planning only, not as a selection criterion. Two mainland records fall outside this frame. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
# Province-wide candidates, not island-only and not filtered by distance from
# Comox: per the 2026-08-14 decision, lidar and severity availability set the
# study's scope and location rather than proximity to a field base. The frame
# below is coastal, so the two mainland interior records fall outside it and are
# reported in the text instead.
cand <- read.csv(file.path(derived, "site-candidates-province.csv"))
cand$severity <- cand$severity_code
cand <- cand[cand$lon > -128.5 & cand$lon < -123.0 &
             cand$lat > 48.3 & cand$lat < 51.2, ]
cand_sf <- sf::st_transform(
  sf::st_as_sf(cand, coords = c("lon", "lat"), crs = 4326), BC_ALBERS)
lidar_env <- sf::st_read(file.path(derived, "lidar-coverage-dissolved.gpkg"),
                         quiet = TRUE)
acc <- sf::st_read(file.path(derived, "field-access-places.gpkg"), quiet = TRUE)
comox <- sf::st_transform(
  sf::st_sfc(sf::st_point(c(-124.928, 49.673)), crs = 4326), BC_ALBERS)

fr <- sf::st_bbox(frame_box)
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.45, end = 1), legend = FALSE,
            add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey35", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = NA, add = TRUE)
plot(sf::st_geometry(rivers), col = "#7ba6c4", lwd = 0.6, add = TRUE)

# Lidar coverage envelope, drawn under the sites so the sites stay readable.
plot(sf::st_geometry(lidar_env), col = "#b9d9b0", border = "#7fae74",
     lwd = 0.3, add = TRUE)

# Distance rings from Comox. Straight line, not road distance.
for (d in c(50, 100, 150)) {
  plot(sf::st_geometry(sf::st_buffer(comox, d * 1000)), col = NA,
       border = "grey30", lty = 3, lwd = 0.8, add = TRUE)
}

sev_col <- c(T = "#f0c987", L = "#e08a45", M = "#c1372f", S = "#7d1d18",
             V = "#4a0f0c")
cxy <- sf::st_coordinates(cand_sf)
cex_by_area <- 0.5 + 1.5 * sqrt(cand$area_ha / max(cand$area_ha))
has_lidar <- cand$lidar_tiles > 0
points(cxy[!has_lidar, , drop = FALSE], pch = 1, lwd = 0.9,
       col = sev_col[cand$severity[!has_lidar]], cex = cex_by_area[!has_lidar])
points(cxy[has_lidar, , drop = FALSE], pch = 16,
       col = sev_col[cand$severity[has_lidar]], cex = cex_by_area[has_lidar])

axy <- sf::st_coordinates(acc)
points(axy, pch = 22, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Comox is drawn separately as the field base, so its label is suppressed here
# and Courtenay is thrown left to clear the diamond.
side <- ifelse(acc$name == "Courtenay", 2, 4)
keep <- acc$name != "Comox"
text(axy[keep, 1], axy[keep, 2], acc$name[keep], pos = side[keep],
     offset = 0.35, cex = 0.55, col = "grey10")
points(sf::st_coordinates(comox), pch = 23, bg = "black", col = "black",
       cex = 1.2)
text(sf::st_coordinates(comox)[1], sf::st_coordinates(comox)[2], "Comox",
     pos = 4, offset = 0.45, cex = 0.6, font = 2, col = "black")

xt <- pretty(fr[c("xmin", "xmax")], 5); yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.58,
       pch = c(16, 1, 22, 23, NA), pt.bg = c(NA, NA, "white", "black", NA),
       col = c("grey25", "grey25", "grey15", "black", "#7fae74"),
       lty = c(NA, NA, NA, NA, 1), lwd = c(NA, NA, NA, NA, 6),
       legend = c("Adelgid polygon, lidar", "Adelgid polygon, no lidar",
                  "Settlement", "Comox", "Lidar coverage"))
legend("bottomright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.58, pch = 16, col = sev_col,
       legend = c("Trace", "Light", "Moderate", "Severe"), title = "Severity")
par(op)
```

@fig-fieldsites exposes a trade-off between severity and structural data that is not resolvable by widening the search area, and it constrains the study rather than merely describing it. Across the province the survey records 267 adelgid polygons, of which 89 fall within lidar coverage and 25 are rated moderate or worse. Only nine polygons satisfy both conditions, and eight of those were surveyed in 2007 or 2008 against lidar acquired between 2023 and 2025, an offset of twelve to seventeen years over which stand condition cannot be assumed constant.

The recent record shows the trade-off at its sharpest. Twelve polygons rated moderate or severe were surveyed from 2019 onward, all of them on Pacific silver fir, and none has any lidar coverage. Ten lie on northern Vancouver Island between Sayward and the Nimpkish valley and two on the mainland coast. Conversely, 43 polygons combine lidar coverage, a label-to-acquisition offset of three years or less, and the shared host, but 37 of those are rated trace and 10 light, with a single moderate polygon among them.

Structural predictors and moderate-or-worse severity therefore cannot both be obtained from existing public data on the same polygons. Three responses are available and the choice belongs to the analysis design rather than to this section: restrict the structural arm to the low-severity polygons where lidar exists and accept that it tests the hypothesis where the signal is weakest; run the high-severity polygons on optical and radar predictors alone; or acquire structural data directly over the twelve recent high-severity polygons, which are between 4 and 24 ha and within reach of an uncrewed aerial survey. Whichever is chosen, the field campaign is a stratified sample over an accessible subset rather than over the label population, and the resulting coverage bias is reported rather than assumed away.

### Tile detail

@fig-tiles shows individual lidar tiles against survey polygon outlines over northern Vancouver Island, where the recent moderate and severe records concentrate. It is drawn at tile resolution rather than as a coverage envelope because the practical question when planning a visit is whether a specific polygon has a specific tile beneath it, which a dissolved envelope conceals.

```{r}
#| label: fig-tiles
#| fig-cap: "Lidar tiles and survey polygons over northern Vancouver Island. Grey squares are individual LidarBC point cloud tiles, shaded by acquisition year. Polygon outlines are aerial overview survey records: red for balsam woolly adelgid, blue for western balsam bark beetle, with adelgid polygons rated moderate or worse drawn heavier. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 6.4
#| out-width: "100%"
det <- c(xmin = -126.4, ymin = 50.05, xmax = -124.2, ymax = 50.95)

tiles <- sf::st_read(file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)
tiles_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(tiles), det)), BC_ALBERS)
iab_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(iab), det)), BC_ALBERS)
vi_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(vi), det)), BC_ALBERS)
land_d <- suppressWarnings(sf::st_crop(
  land, sf::st_bbox(sf::st_transform(
    sf::st_as_sfc(sf::st_bbox(det, crs = 4326)), BC_ALBERS))))
acc_d <- sf::st_read(file.path(derived, "field-access-places.gpkg"),
                     quiet = TRUE)

fr <- sf::st_bbox(sf::st_transform(sf::st_as_sfc(sf::st_bbox(det, crs = 4326)),
                                   BC_ALBERS))
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.55, end = 1), legend = FALSE,
            add = TRUE, maxcell = 3e6)
plot(sf::st_geometry(land_d), col = NA, border = "grey35", lwd = 0.7,
     add = TRUE)

# Tiles, shaded by acquisition year so a polygon's temporal match is legible.
yrs <- sort(unique(tiles_d$year))
pal <- setNames(grey.colors(max(length(yrs), 2), start = 0.55, end = 0.88),
                yrs)
plot(sf::st_geometry(tiles_d), col = pal[as.character(tiles_d$year)],
     border = "grey60", lwd = 0.15, add = TRUE)

plot(sf::st_geometry(vi_d[vi_d$PEST_SPECIES_CODE == "IBB", ]), col = NA,
     border = "#2c6f9e", lwd = 0.7, add = TRUE)
plot(sf::st_geometry(iab_d), col = NA, border = "#c1372f", lwd = 1.0,
     add = TRUE)
hi <- iab_d[iab_d$PEST_SEVERITY_CODE %in% c("M", "S", "V"), ]
if (nrow(hi)) plot(sf::st_geometry(hi), col = "#c1372f66",
                   border = "#7d1d18", lwd = 2.0, add = TRUE)

ap <- sf::st_coordinates(acc_d)
inb <- ap[, 1] > fr["xmin"] & ap[, 1] < fr["xmax"] &
       ap[, 2] > fr["ymin"] & ap[, 2] < fr["ymax"]
if (any(inb)) {
  points(ap[inb, , drop = FALSE], pch = 22, bg = "white", col = "grey15",
         cex = 0.9, lwd = 0.8)
  text(ap[inb, 1], ap[inb, 2], acc_d$name[inb], pos = 4, offset = 0.4,
       cex = 0.62, col = "grey10")
}

xt <- pretty(fr[c("xmin", "xmax")], 4); yt <- pretty(fr[c("ymin", "ymax")], 4)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 10000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.010 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.8 * h, c("0", "10", "20 km"),
     cex = 0.6, col = "grey15")
nx <- fr["xmax"] - 0.06 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.08 * diff(fr[c("ymin", "ymax")])
a <- 0.045 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.5 * a, "N", cex = 0.8, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.6,
       lty = c(1, 1, 1, NA), lwd = c(2.0, 1.0, 0.7, NA), pch = c(NA, NA, NA, 22),
       pt.bg = c(NA, NA, NA, "grey75"),
       col = c("#7d1d18", "#c1372f", "#2c6f9e", "grey60"),
       legend = c("Adelgid, moderate or worse", "Adelgid, other severity",
                  "Bark beetle", "Lidar tile"))
par(op)
```

## Data preparation

Terrain governs every preparation choice here. The study area is mountainous, the survey polygons sit on slopes rather than on flat ground, and both optical and radar measurements are sensitive to illumination and viewing geometry in ways that vary systematically with aspect and slope. An uncorrected terrain effect would enter the models as a predictor, and because the two agents occupy partly different elevations it could be learned in place of the damage signal.

Optical predictors are built from Sentinel-2 Level 2A scenes over the leaf-on window of the label epoch. Cloud, cloud shadow, snow and water are masked from the scene classification layer before compositing, and a per-pixel median composite is formed to suppress residual cloud edges and to reduce the influence of any single acquisition geometry. From the composite we derive the visible, red edge, near infrared and shortwave infrared bands and a set of indices selected for sensitivity to canopy water content and to the visible and shortwave changes @Campbell_2023 associated with adelgid infestation. A topographic illumination correction is applied before indices are computed, since a slope-driven brightness difference would otherwise propagate into every ratio.

Radar predictors are built from Sentinel-1 Ground Range Detected scenes in interferometric wide swath mode, using both VV and VH polarisations. Processing follows the standard sequence of orbit refinement, border and thermal noise removal, radiometric calibration, terrain flattening and Range Doppler terrain correction, with the flattening step performed against a digital elevation model so that backscatter is expressed on a terrain-corrected reference area rather than on the ellipsoid. Areas of layover and shadow are masked and excluded, which removes some steep terrain from the analysis and is reported as a reduction in the usable sample rather than ignored. Speckle is suppressed by temporal averaging across the scene stack, and the mean and the temporal dispersion of backscatter are retained separately, the latter because a changing crown may be distinguishable by its variability as well as its mean.

Structural predictors are derived from LidarBC point clouds. Ground returns are used to interpolate a terrain surface, point heights are normalised above it, and canopy metrics are computed from the normalised cloud rather than from any supplied vegetation classification, since the delivered tiles carry no vegetation classes.

The metric set is chosen against the mechanism rather than against convention. A generic canopy description would test whether the two agents differ in any structural respect, which is weaker and less falsifiable than the question asked here. The mechanism predicts two specific differences. Progressive top-down dieback with the tree persisting should move canopy material downward in the profile and thin the upper strata while the lower canopy remains, whereas whole-crown mortality with architecture retained should reduce returns throughout the profile without redistributing them. Gout and branch deformation should roughen the crown surface, while a beetle-killed crown should not roughen until branches begin to shed.

Metrics therefore fall into three families. The first tests vertical redistribution: height percentiles at the 25th, 50th, 75th and 95th centiles, the ratio of the 25th to the 95th, canopy relief ratio, and the proportion of returns above two thirds of maximum height. The second tests surface roughness: rumple, being the ratio of canopy surface area to ground area, the standard deviation of the canopy height model within the polygon, and the coefficient of variation of first-return heights. The third describes density and openness, being canopy cover above 2 m, gap fraction and the vertical complexity index.

The third family is a control rather than a hypothesis. Both agents thin the canopy, so if density and openness alone separate them the structural result concerns damage intensity rather than damage geometry, and it is reported in those terms.

All three families are summarised to the survey polygon, which is the unit of analysis. Polygons are eroded by an inward buffer before summarisation, because sketch-mapped boundaries are approximate and an unbuffered polygon edge will include unaffected canopy. Within the eroded polygon each predictor is reduced to its mean and its dispersion, so that a heterogeneous polygon is not represented solely by a central value. Polygons too small to survive erosion are excluded and counted.

The specific index set, the lidar metric set and the buffer distance are fixed at pre-registration, before any predictor value is joined to any label, and are reported there with the reasoning for each threshold.

## Random forests

The response is the two-class agent label, adelgid against bark beetle, for survey polygons on Pacific silver fir. Restricting to a single host is what makes the comparison an attribution test: with host held constant, a classifier cannot succeed by learning host reflectance or host site preference in place of the damage itself.

Predictors are grouped into three families, spectral, structural and radar, and three nested models are compared: spectral only, spectral with structure, and all three families together. The nesting is deliberate, since the question is not whether the agents can be separated at all but how much of any separation the spectral baseline already provides. Survey year and terrain are carried as covariates in every model so that their contribution is estimated rather than absorbed into the sensor terms; the two agents peak in different years (@fig-timeseries) and occupy partly different elevations, and both are plausible confounders.

Models are fitted with random forests, following @Campbell_2023, and compared at equal complexity, with hyperparameters tuned only on training-internal splits so that no evaluation fold informs model selection. Random forests are chosen for their tolerance of correlated predictors, which the three families supply in quantity, and because variable importance grouped by family gives a direct reading of where separating information lies.

Because polygons cluster in space, performance is estimated by spatially blocked cross-validation rather than random partitioning, which would place spatially adjacent and often spatially overlapping polygons in both training and evaluation folds and inflate accuracy. On a sample of `r n_set` polygons the blocking scheme is the design choice most likely to determine the result, so it is fixed in advance and not tuned afterwards.

Blocks are contiguous groups formed by clustering polygon centroids, sized so that no block holds fewer than eight polygons and no fold fewer than two blocks, which yields approximately five folds at this sample size. Block size is set from the empirical range of spatial autocorrelation in the predictors, estimated on the predictors alone before any label is joined, so that the choice cannot be informed by the outcome. Where that range would leave fewer than five folds the design accepts fewer folds rather than smaller blocks, since leakage inflates accuracy while a low fold count only widens the interval. Polygons that overlap or share a boundary are forced into the same block irrespective of the clustering, because the survey digitises adjacent damage as separate records and treating them as independent would leak directly. The realised number of folds and block sizes are reported with every accuracy figure. The two classes are imbalanced by `r set_imbalance` to 1 in favour of the `r set_majority` class, and resampling is applied within training folds only, never to the evaluation folds, so that reported performance refers to the observed class prevalence. With `r n_set` polygons the sample is small for a spatially blocked design, and the number of blocks is reported alongside every accuracy figure.

## Accuracy assessment

The primary metric is balanced accuracy, the mean of sensitivity and specificity. It is chosen over overall accuracy because the classes are imbalanced: a model assigning every polygon to the `r set_majority` class would score respectably on overall accuracy while separating nothing, which is the failure mode the study exists to avoid.

Per-class sensitivity and specificity and the full confusion matrix are reported alongside it, each with the support on which it is computed. The comparison of interest is paired, the difference in balanced accuracy between the multi-sensor and spectral-only models evaluated on identical folds, so that fold-to-fold variation is differenced out rather than treated as independent error, with its interval obtained by bootstrapping over folds. The hypothesis is confirmed if and only if that difference is positive and its 95 percent interval excludes zero, disconfirmed if the interval excludes zero in the negative direction, and inconclusive if it spans zero.

Accuracy figures reported here refer to agreement with the surveyor's attribution, not with an independent ground determination. They therefore measure reproduction of the operational record and inherit whatever error that record carries, which @Coleman_2018 show increases as agent specificity increases. The metric implementation is verified against cases whose values are known by construction (@tbl-checks).

```{r}
#| label: eval-metric
#| echo: true
#| output: false
balanced_accuracy <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}
```

```{r}
#| label: tbl-checks
#| tbl-cap: "Verification of the balanced accuracy implementation against cases with analytically known values."
t_perfect <- c("IAB", "IAB", "IBB", "IBB")
t_inverted <- c("IBB", "IBB", "IAB", "IAB")
imb_truth <- c("IAB", "IBB", "IBB", "IBB")

checks <- data.frame(
  Case = c("Perfect prediction",
           "Fully inverted prediction",
           "All assigned to majority class",
           "Imbalanced, all majority (overall accuracy 0.75)"),
  Expected = c(1, 0, 0.5, 0.5),
  Observed = c(
    balanced_accuracy(t_perfect, t_perfect),
    balanced_accuracy(t_perfect, t_inverted),
    balanced_accuracy(t_perfect, rep("IBB", 4)),
    balanced_accuracy(imb_truth, rep("IBB", 4))
  )
)
checks$Agrees <- abs(checks$Expected - checks$Observed) < 1e-9
fmt_tbl(checks)
```

# Results

## Model comparison

[PENDING. Balanced accuracy of the three nested models and the difference between the multi-sensor and spectral-only models, with its interval.]

```{r}
#| label: tbl-primary
#| tbl-cap: "Separation of the two damage agents by predictor set, under spatially blocked cross-validation."
primary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  `Balanced accuracy` = "pending",
  `Difference from spectral` = "pending",
  `95% CI` = "pending",
  check.names = FALSE
)
fmt_tbl(primary)
```

## Class performance

[PENDING. Per-class sensitivity and specificity, the confusion matrix and its support, and the relative importance of predictors grouped by sensor family.]

```{r}
#| label: tbl-secondary
#| tbl-cap: "Per-class performance of each model, with support."
secondary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  Sensitivity = "pending",
  Specificity = "pending",
  Support = "pending",
  check.names = FALSE
)
fmt_tbl(secondary)
```

# Discussion and conclusions

[PENDING. Interpretation of the model comparison, in relation to the spectral attribution failure reported by @Campbell_2023 and the commission-error reduction reported by @Jones_2015, closing with what the result implies for verification of aerial overview survey attributions in coastal British Columbia.]

## Limitations

The defining limitation is the severity of the sample, and it follows from data availability rather than from choice. No adelgid polygon rated moderate or worse in British Columbia falls under lidar coverage at any useful temporal offset, so a study with a structural arm is necessarily a study of trace and light damage. The analysis is therefore concentrated in the damage classes where any spectral or structural signal is weakest, and a null result would be the expected outcome of the severity distribution as much as a statement about the sensors.

Other limitations apply irrespective of the result. The labels are sketch-mapped from an aircraft and their spatial accuracy is acknowledged by the custodian to be variable, so polygon boundaries are approximate and some damage is missed entirely. That concentration may fall below the detection floor of moderate-resolution optical sensing altogether: @Ye_2021 report that their framework detected mountain pine beetle only where red-stage mortality occupied at least 56 percent of a Landsat pixel, and spruce beetle only above 55 percent grey-stage occupancy. A trace-rated polygon will rarely approach those fractions, so a null result for the spectral baseline would be consistent with the severity distribution rather than with the agents being spectrally inseparable, and the two explanations cannot be distinguished with these labels alone. Agent attribution in the source data is the surveyor's judgement from the air and is not independently verified on the ground, so the models can at best reproduce that judgement and any systematic error in it propagates into the results. Field confirmation addresses this only partly, because the diagnostic signs of adelgid infestation, the woolly wax masses and stem gout, occur on living and declining trees and are lost from those long dead.

Restricting the sample to Vancouver Island buys internal validity at the price of external validity, and the trade is stated rather than concealed. Any result reported here describes the island. The adjacent mainland coast is where it would most naturally be tested, and it is precisely the region that cannot serve as a test, because its record is almost entirely one agent under a single lidar acquisition and offers no independent variation to check the transfer against.

## Detection and attribution

This study evaluates attribution rather than detection, and the distinction is maintained throughout. Establishing that a canopy has changed is a different problem from identifying the agent responsible, and a model can perform well at the first while failing the second [@Campbell_2023]. Accuracy figures are therefore reported with their support, their class balance and their validation design.

# Analysis pipeline

Every derived file the sections above read is produced by the code in this section, in the order given. Nothing here is a description of code kept elsewhere: this is the code, and `05.scripts/` is generated from it by `knitr::purl()` rather than maintained alongside it, so the two cannot diverge.

Each block is guarded on the file it writes. A render with the derived files present executes none of them and the sections above read what is already on disk; delete a file and the block that makes it runs. The guard is what allows a twenty minute Earth Engine export and a 79 GB lidar retrieval to sit in a document that renders in seconds. Each block is wrapped in `local()` so that a pipeline object cannot silently replace one the manuscript is already using.

Two steps were shell and Python until 2026-08-14 and are R here, so that everything the study runs can be read in one language and in one place.

## Label extraction

The survey archive is a 908 MB file geodatabase of 1,172,716 polygons. It is never read into memory: `sf::gdal_utils()` drives the same GDAL routine that `ogr2ogr` does, so the selection and reprojection happen inside GDAL and only the extracted records reach R.

```{r}
#| label: pipe-labels
#| echo: true
#| output: false
# Source:    Pest Infestation Polygons, BC Data Catalogue, record
#            450b67bb-02d5-4526-8bc0-ac7924125a1e
# Licence:   Open Government Licence - British Columbia
# Custodian: BC Ministry of Forests. Not GeoBC.
#
# The delivered geodatabase carries codes only, no common or latin names, so IAB
# is resolved from the Forest Health Aerial Overview Survey Standards rather than
# from the delivery.
#
# NOTE ON TESTING. The archive is not on this machine, so this block has not been
# run end to end since it was translated from shell. The GDAL call and the layer
# and field names were checked against the committed extracts on 2026-08-14; the
# WHERE clauses were not re-executed. Set ADELGID_GDB to the archive to run it.
gdb <- Sys.getenv("ADELGID_GDB", unset = "")
iab_path <- file.path(derived, "adelgid-iab-polygons.geojson")
vi_path <- file.path(derived, "vi-fir-agents.geojson")

if (!file.exists(iab_path) || !file.exists(vi_path)) local({
  if (!nzchar(gdb) || !file.exists(gdb))
    stop("set ADELGID_GDB to the pest_infestation_poly.gdb archive")

  # 1. Province-wide balsam woolly adelgid, for the extent and time series.
  # WGS84 so the file is readable without a projection library. The attributes
  # kept are the ones needed to judge label quality.
  sf::gdal_utils(
    util = "vectortranslate", source = gdb, destination = iab_path,
    options = c("-f", "GeoJSON", "-t_srs", "EPSG:4326", "-sql",
                paste("SELECT CAPTURE_YEAR, PEST_SEVERITY_CODE,",
                      "PEST_CAPTURE_METHOD_CODE, TREE_SPECIES_CODE, AREA_HA",
                      "FROM pest_infestation_poly",
                      "WHERE PEST_SPECIES_CODE = 'IAB'")))

  # 2. The two agents that share a true-fir host around Vancouver Island. IAB is
  # the target class and IBB, western balsam bark beetle, the confusable one:
  # both are mapped on Pacific silver fir in the same survey, so separating them
  # is attribution, not detection. The envelope is a rectangle and therefore also
  # catches the adjacent mainland; the sample is restricted to the island later,
  # and both counts are reported.
  sf::gdal_utils(
    util = "vectortranslate", source = gdb, destination = vi_path,
    options = c("-f", "GeoJSON", "-t_srs", "EPSG:4326",
                "-spat", "-128.5", "48.3", "-123.0", "51.0",
                "-spat_srs", "EPSG:4326", "-sql",
                paste("SELECT CAPTURE_YEAR, PEST_SPECIES_CODE,",
                      "PEST_SEVERITY_CODE, TREE_SPECIES_CODE, AREA_HA",
                      "FROM pest_infestation_poly",
                      "WHERE PEST_SPECIES_CODE IN ('IAB','IBB')",
                      "AND TREE_SPECIES_CODE IN ('B','BA','BG','BL')")))
})
```

```{r}
#| label: pipe-labels-out
fmt_tbl(data.frame(
  File = c("adelgid-iab-polygons.geojson", "vi-fir-agents.geojson"),
  Contents = c("Balsam woolly adelgid, province-wide",
               "Both agents on true fir, study envelope"),
  Polygons = c(nrow(iab), nrow(vi)),
  Years = c(paste(range(iab$CAPTURE_YEAR), collapse = " to "),
            paste(range(vi$CAPTURE_YEAR), collapse = " to "))))
```

## Tile index

The LidarBC point cloud index is queried once over the study envelope and saved, because every coverage decision downstream depends on the same tile geometry and acquisition years. The service returns at most 1,000 records per request, so the query is paged to exhaustion rather than truncated.

Until 2026-08-14 this file existed in `02.inputs/derived/` with no code in the repository that produced it, while four scripts read it and the sample depended on it. The block below was written to close that gap and reproduces the committed file exactly: 18,840 tiles, identical attributes and identical total area.

```{r}
#| label: pipe-tiles
#| echo: true
#| output: false
# Source:    LidarBC Open LiDAR Data Portal, ArcGIS FeatureServer
#            LiDAR_BC_S3_Public, layer 4, the point cloud index.
# Licence:   Open Government Licence - British Columbia.
# Custodian: GeoBC.
LIDAR_SERVICE <- paste0("https://services6.arcgis.com/ubm4tcTYICKBpist/ArcGIS/",
                        "rest/services/LiDAR_BC_S3_Public/FeatureServer")
PC_LAYER <- 4L
STUDY_ENV <- c(-128.5, 48.3, -123.0, 51.0)
PAGE <- 1000L

# One URL builder for every query this manuscript makes against the service.
# Caller arguments replace the defaults rather than being appended to them: the
# service answers a repeated parameter with HTTP 400, so a caller asking for
# f = "geojson" must overwrite f = "json" and not sit beside it.
lidar_url <- function(layer, env, ...) {
  gj <- sprintf(paste0('{"xmin":%f,"ymin":%f,"xmax":%f,"ymax":%f,',
                       '"spatialReference":{"wkid":4326}}'),
                env[1], env[2], env[3], env[4])
  p <- list(where = "1=1", geometry = gj,
            geometryType = "esriGeometryEnvelope", inSR = "4326",
            spatialRel = "esriSpatialRelIntersects", f = "json")
  extra <- list(...)
  p[names(extra)] <- extra
  paste0(LIDAR_SERVICE, "/", layer, "/query?",
         paste(names(p), vapply(p, function(v)
           URLencode(as.character(v), reserved = TRUE), ""),
           sep = "=", collapse = "&"))
}

tiles_path <- file.path(derived, "lidar-tiles.geojson")
if (!file.exists(tiles_path)) local({
  acc <- list(); off <- 0L
  repeat {
    u <- lidar_url(PC_LAYER, STUDY_ENV, outFields = "year,density,maptile",
                   returnGeometry = "true", outSR = "4326", f = "geojson",
                   resultOffset = off, resultRecordCount = PAGE)
    ch <- try(sf::st_read(u, quiet = TRUE), silent = TRUE)
    if (inherits(ch, "try-error") || nrow(ch) == 0L) break
    acc[[length(acc) + 1L]] <- ch
    if (nrow(ch) < PAGE) break
    off <- off + PAGE
  }
  sf::st_write(do.call(rbind, acc), tiles_path, quiet = TRUE)
})
```

```{r}
#| label: pipe-tiles-out
local({
  tl <- sf::st_read(tiles_path, quiet = TRUE)
  fmt_tbl(data.frame(
    Tiles = nrow(tl),
    `Acquisition years` = paste(range(tl$year, na.rm = TRUE), collapse = " to "),
    `Point density` = paste(unique(tl$density), collapse = ", "),
    `Area covered (km2)` = round(sum(as.numeric(
      sf::st_area(sf::st_transform(tl, BC_ALBERS)))) / 1e6),
    check.names = FALSE))
})
```

## Coverage summary

Tile counts over the study envelope and over the Comox field window, taken with `returnCountOnly` rather than by paging, so a count is exact even where it exceeds the record cap. Where the count exceeds the cap the attribute summaries are drawn from a capped sample and are flagged as such, because they are then what the sample happened to contain rather than the full range.

```{r}
#| label: pipe-coverage
#| echo: true
#| output: false
cov_path <- file.path(derived, "lidar-coverage.csv")
if (!file.exists(cov_path)) local({
  envelopes <- list("study area" = STUDY_ENV,
                    "Comox field window" = c(-125.3, 49.4, -124.5, 49.9))
  rows <- lapply(names(envelopes), function(nm) {
    env <- envelopes[[nm]]
    n <- jsonlite::fromJSON(lidar_url(PC_LAYER, env, returnCountOnly = "true"),
                            simplifyVector = FALSE)$count
    s <- jsonlite::fromJSON(
      lidar_url(PC_LAYER, env, outFields = "year,density,projection,classes",
                returnGeometry = "false", resultRecordCount = 1000),
      simplifyVector = FALSE)
    at <- lapply(s$features, `[[`, "attributes")
    pull <- function(k) sort(unique(unlist(lapply(at, function(a) a[[k]]))))
    data.frame(envelope = nm, tiles = n,
               attrs_from_full_index = n <= 1000,
               acquisition_years_sampled = paste(pull("year"), collapse = ";"),
               density_pts_m2 = paste(pull("density"), collapse = ";"),
               point_classes = paste(pull("classes"), collapse = ";"),
               stringsAsFactors = FALSE)
  })
  write.csv(do.call(rbind, rows), cov_path, row.names = FALSE)
})
```

## Basemap layers

Natural Earth vectors and an elevation hillshade, clipped and reprojected to BC Albers once so that the study area figure draws a metrically true scale bar. Natural Earth is public domain.

```{r}
#| label: pipe-basemap
#| echo: true
#| output: false
bm_path <- file.path(derived, "basemap.gpkg")
hs_path <- file.path(derived, "hillshade.tif")

# Map frame, slightly wider than the analysis envelope so the island sits inside
# a margin rather than against the neatline. This is the view the figures show.
FRAME <- c(xmin = -129.2, ymin = 48.0, xmax = -122.3, ymax = 51.4)
# Data are clipped wider than the view. The figures plot at a fixed aspect ratio,
# which expands the shorter axis past the view rectangle; land cropped at the
# view would stop short of the neatline and leave a band of sea where the
# mainland belongs.
PAD <- 1.2
DATA_FRAME <- FRAME + c(-PAD, -PAD, PAD, PAD)

if (!file.exists(bm_path) || !file.exists(hs_path)) local({
  library(rnaturalearth)
  library(elevatr)

  clip <- function(x) {
    x <- sf::st_make_valid(sf::st_transform(x, 4326))
    suppressWarnings(sf::st_crop(x, DATA_FRAME))
  }

  land <- ne_download(scale = 10, type = "land", category = "physical",
                      returnclass = "sf")
  states <- ne_states(country = c("Canada", "United States of America"),
                      returnclass = "sf")
  # The global rivers layer resolves only one river inside this frame, which is
  # useless for orientation. The North America supplement resolves 32.
  rivers <- ne_download(scale = 10, type = "rivers_north_america",
                        category = "physical", returnclass = "sf")
  # Several lake polygons carry self-intersecting rings. s2 rejects them
  # outright, so validity is repaired in planar mode.
  lakes <- tryCatch({
    old <- sf::sf_use_s2(FALSE); on.exit(sf::sf_use_s2(old), add = TRUE)
    ne_download(scale = 10, type = "lakes_north_america",
                category = "physical", returnclass = "sf")
  }, error = function(e) NULL)

  layers <- list(
    land = clip(land["featurecla"]),
    rivers = clip(rivers["name"]),
    # The view rectangle, saved so the figures set their limits from it rather
    # than from the bounding box of the land, which is deliberately wider.
    frame = sf::st_sf(name = "Map frame",
                      geometry = sf::st_as_sfc(sf::st_bbox(FRAME, crs = 4326))))

  # Only the shared land border is kept. Drawing the full provincial outline
  # would retrace the coastline in a second style, which is confusing at this
  # extent; the international boundary is the only administrative line inside
  # this frame.
  bc <- states[states$name == "British Columbia", ]
  us <- states[states$admin == "United States of America", ]
  border <- tryCatch({
    old <- sf::sf_use_s2(FALSE); on.exit(sf::sf_use_s2(old), add = TRUE)
    b <- sf::st_intersection(
      sf::st_boundary(sf::st_make_valid(sf::st_union(bc))),
      sf::st_buffer(sf::st_boundary(sf::st_make_valid(sf::st_union(us))), 0.02))
    sf::st_sf(name = "International boundary", geometry = sf::st_geometry(b))
  }, error = function(e) NULL)
  if (!is.null(border)) layers$admin <- clip(border)
  if (!is.null(lakes)) {
    old <- sf::sf_use_s2(FALSE); layers$lakes <- clip(lakes["name"])
    sf::sf_use_s2(old)
  }

  if (file.exists(bm_path)) file.remove(bm_path)
  for (nm in names(layers))
    sf::st_write(sf::st_transform(layers[[nm]], BC_ALBERS), bm_path,
                 layer = nm, quiet = TRUE, append = FALSE)

  # Place labels that orient a reader unfamiliar with the coast. Coordinates are
  # from the BC Geographical Names Office listings, entered by hand because the
  # Natural Earth populated-places layer omits the smaller island communities.
  places <- sf::st_as_sf(data.frame(
    name = c("Comox", "Campbell River", "Port Alberni", "Nanaimo",
             "Victoria", "Vancouver", "Tofino", "Port Hardy"),
    lon = c(-124.928, -125.247, -124.805, -123.936,
            -123.365, -123.121, -125.906, -127.418),
    lat = c(49.673, 50.024, 49.234, 49.166,
            48.428, 49.283, 49.153, 50.720)),
    coords = c("lon", "lat"), crs = 4326)
  sf::st_write(sf::st_transform(places, BC_ALBERS), bm_path, layer = "places",
               quiet = TRUE, append = FALSE)

  # Hillshade, so the reader sees the mountain spine rather than a flat outline.
  # AWS Terrain Tiles via elevatr. Zoom 7 is about 1 km ground resolution, ample
  # at this extent and small enough to commit.
  frame_sf <- sf::st_as_sf(sf::st_as_sfc(sf::st_bbox(DATA_FRAME, crs = 4326)))
  dem <- elevatr::get_elev_raster(frame_sf, z = 7, clip = "bbox",
                                  verbose = FALSE)
  dem <- terra::project(terra::rast(dem), paste0("EPSG:", BC_ALBERS))
  # Terrain tiles return zero or small positive values over water rather than
  # negative ones, so an elevation threshold does not separate sea from shore.
  # Masking by the coastline polygon does.
  dem <- terra::mask(dem, terra::vect(sf::st_geometry(
    sf::st_transform(layers$land, BC_ALBERS))))
  hs <- terra::shade(terra::terrain(dem, "slope", unit = "radians"),
                     terra::terrain(dem, "aspect", unit = "radians"),
                     angle = 40, direction = 315)
  terra::writeRaster(hs, hs_path, overwrite = TRUE,
                     gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2"))
})
```

```{r}
#| label: pipe-basemap-out
local({
  ly <- sf::st_layers(bm_path)
  fmt_tbl(data.frame(Layer = ly$name,
                     Features = as.integer(ly$features),
                     Geometry = vapply(ly$geomtype, function(g) g[1], "")))
})
```

## Site candidates

Every adelgid polygon in the province, with the lidar tiles over it, its acquisition years and the offset from the survey year. Coverage is queried per polygon envelope rather than over one rectangle, because the province-wide record runs from the coast to the Kootenays and a single envelope would pull far more tile geometry than this machine should hold.

```{r}
#| label: pipe-candidates
#| echo: true
#| output: false
cand_path <- file.path(derived, "site-candidates-province.csv")
if (!file.exists(cand_path)) local({
  sev <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
           V = "Very severe")
  host <- c(B = "True fir unspecified", BA = "Pacific silver fir",
            BG = "Grand fir", BL = "Subalpine fir")

  tiles_over <- function(bb) {
    n <- jsonlite::fromJSON(lidar_url(PC_LAYER, bb, returnCountOnly = "true"),
                            simplifyVector = FALSE)$count
    if (!length(n) || n == 0) return(list(n = 0L, years = integer(0)))
    d <- jsonlite::fromJSON(
      lidar_url(PC_LAYER, bb, outFields = "year", returnGeometry = "false",
                resultRecordCount = 1000), simplifyVector = FALSE)
    list(n = n, years = sort(unique(unlist(
      lapply(d$features, function(f) f$attributes$year)))))
  }

  src <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                     quiet = TRUE)
  rows <- lapply(seq_len(nrow(src)), function(i) {
    bb <- as.numeric(sf::st_bbox(sf::st_geometry(src)[i]))
    tv <- tiles_over(bb)
    yr <- src$CAPTURE_YEAR[i]
    near <- if (length(tv$years) && !is.na(yr))
      tv$years[which.min(abs(tv$years - yr))] else NA_integer_
    Sys.sleep(0.25)
    data.frame(
      capture_year = yr,
      severity = unname(sev[src$PEST_SEVERITY_CODE[i]]),
      severity_code = src$PEST_SEVERITY_CODE[i],
      host = unname(host[src$TREE_SPECIES_CODE[i]]),
      host_code = ifelse(is.na(src$TREE_SPECIES_CODE[i]), "",
                         src$TREE_SPECIES_CODE[i]),
      area_ha = round(src$AREA_HA[i], 1),
      lidar_tiles = tv$n,
      lidar_years = paste(tv$years, collapse = ";"),
      nearest_lidar_year = if (is.na(near)) "" else near,
      year_offset = if (is.na(near)) "" else abs(near - yr),
      lon = round(mean(bb[c(1, 3)]), 5), lat = round(mean(bb[c(2, 4)]), 5),
      on_vancouver_island = bb[3] < -123.2 && bb[4] < 51.2,
      stringsAsFactors = FALSE)
  })
  out <- do.call(rbind, rows)
  out <- out[order(-out$lidar_tiles, out$severity_code), ]
  write.csv(out, cand_path, row.names = FALSE)
})
```

## Availability screen

The study is located by where all three requirements already coexist, rather than by preference. Earlier work asked whether a chosen set of polygons had lidar, which is the wrong order and produced a design requiring lidar that does not exist. A region is only a candidate if it can supply both agents on the shared host, since adelgid polygons under lidar with no bark beetle polygons under the same lidar cannot support a two-class attribution test at all.

```{r}
#| label: pipe-availability
#| echo: true
#| output: false
avail_path <- file.path(derived, "availability-regions.csv")
if (!file.exists(avail_path)) local({
  MAX_OFFSET <- 3          # years between survey and lidar acquisition
  GRID <- 25000            # 25 km analysis cells

  v <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)), BC_ALBERS)
  tl <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)), BC_ALBERS)

  hit <- sf::st_intersects(v, tl)
  v$lidar_tiles <- lengths(hit)
  v$nearest_year <- vapply(seq_len(nrow(v)), function(k) {
    if (!v$lidar_tiles[k]) return(NA_integer_)
    ys <- tl$year[hit[[k]]]; ys <- ys[!is.na(ys)]
    if (!length(ys)) return(NA_integer_)
    ys[which.min(abs(ys - v$CAPTURE_YEAR[k]))]
  }, integer(1))
  v$offset <- abs(v$nearest_year - v$CAPTURE_YEAR)
  usable <- v[!is.na(v$offset) & v$offset <= MAX_OFFSET, ]

  cells <- sf::st_make_grid(usable, cellsize = GRID)
  cells <- sf::st_sf(cell = seq_along(cells), geometry = cells)
  ix <- sf::st_intersects(cells, usable)
  sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)

  rows <- lapply(seq_len(nrow(cells)), function(k) {
    sub <- usable[ix[[k]], ]
    if (!nrow(sub)) return(NULL)
    a <- sub[sub$PEST_SPECIES_CODE == "IAB", ]
    b <- sub[sub$PEST_SPECIES_CODE == "IBB", ]
    ctr <- sf::st_coordinates(sf::st_centroid(sf::st_transform(cells[k, ], 4326)))
    data.frame(
      cell = k, lon = round(ctr[1, 1], 3), lat = round(ctr[1, 2], 3),
      iab_n = nrow(a), ibb_n = nrow(b),
      # Shared host is what makes the comparison an attribution test.
      iab_ba = sum(a$TREE_SPECIES_CODE %in% "BA"),
      ibb_ba = sum(b$TREE_SPECIES_CODE %in% "BA"),
      iab_best_sev = if (nrow(a)) names(which.max(
        sev_rank[a$PEST_SEVERITY_CODE])) else NA_character_,
      iab_mod_plus = sum(a$PEST_SEVERITY_CODE %in% c("M", "S", "V")),
      iab_ha = round(sum(a$AREA_HA)),
      yrs = paste(range(sub$CAPTURE_YEAR), collapse = "-"),
      med_offset = median(sub$offset))
  })
  tab <- do.call(rbind, rows)
  tab$viable <- tab$iab_ba > 0 & tab$ibb_ba > 0
  tab <- tab[order(-tab$viable, -(tab$iab_ba + tab$ibb_ba), -tab$iab_mod_plus), ]
  write.csv(tab, avail_path, row.names = FALSE)
})
```

## Sample freeze

The five inclusion conditions, applied in order. This block writes the file that is the sample, and its size is the study's sample size. Nothing downstream filters these polygons again.

```{r}
#| label: pipe-sample
#| echo: true
#| output: false
set_path <- file.path(derived, "analysis-set.geojson")
restr_path <- file.path(derived, "analysis-set-restriction.csv")

if (!file.exists(set_path) || !file.exists(restr_path)) local({
  MAX_OFFSET <- 5
  # A point in Strathcona Park, well inland, used to pick Vancouver Island out of
  # the land layer by containment rather than by area, so the selection cannot
  # silently attach to a different landmass.
  VI_SEED <- c(-125.60, 49.60)

  v <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)), BC_ALBERS)
  tl <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)), BC_ALBERS)

  # 2. Host held constant, so the comparison is attribution rather than host
  # discrimination.
  v <- v[v$TREE_SPECIES_CODE %in% "BA", ]

  # 3 and 4. Lidar coverage, and an acquisition within MAX_OFFSET years.
  hit <- sf::st_intersects(v, tl)
  v$lidar_tiles <- lengths(hit)
  v$lidar_year <- vapply(seq_len(nrow(v)), function(k) {
    if (!v$lidar_tiles[k]) return(NA_integer_)
    ys <- tl$year[hit[[k]]]; ys <- ys[!is.na(ys)]
    if (!length(ys)) return(NA_integer_)
    ys[which.min(abs(ys - v$CAPTURE_YEAR[k]))]
  }, integer(1))
  v$year_offset <- abs(v$lidar_year - v$CAPTURE_YEAR)
  aset <- v[!is.na(v$year_offset) & v$year_offset <= MAX_OFFSET, ]

  # 5. Vancouver Island only. Off the island the record is 43 bark beetle to 5
  # adelgid and 38 of 48 polygons come from one 2020 acquisition, so landmass and
  # campaign each predict the agent almost perfectly. Blocking does not answer a
  # confound that holds within folds as well as across them, so region is
  # neutralised by restriction, the instrument already used for host.
  land_l <- sf::st_transform(sf::st_make_valid(
    sf::st_read(bm_path, "land", quiet = TRUE)), BC_ALBERS)
  seed <- sf::st_transform(sf::st_sfc(sf::st_point(VI_SEED), crs = 4326),
                           BC_ALBERS)
  island <- land_l[lengths(sf::st_intersects(land_l, seed)) > 0, ]
  stopifnot(nrow(island) == 1)
  on_island <- lengths(sf::st_intersects(
    sf::st_centroid(sf::st_geometry(aset)), island)) > 0

  # The counts that motivate the restriction are saved rather than quoted from a
  # terminal session, because the manuscript reports them.
  restriction <- as.data.frame(table(
    landmass = ifelse(on_island, "Vancouver Island", "Elsewhere"),
    agent = aset$PEST_SPECIES_CODE,
    lidar_year = aset$lidar_year), stringsAsFactors = FALSE)
  restriction <- restriction[restriction$Freq > 0, ]
  names(restriction)[names(restriction) == "Freq"] <- "n"
  write.csv(restriction, restr_path, row.names = FALSE)

  aset <- aset[on_island, ]
  if (file.exists(set_path)) invisible(file.remove(set_path))
  sf::st_write(sf::st_transform(aset, 4326), set_path, quiet = TRUE)
})
```

```{r}
#| label: pipe-sample-out
fmt_tbl(data.frame(
  Quantity = c("Polygons", "Adelgid", "Bark beetle", "Imbalance",
               "Survey years", "Lidar years", "Median offset (years)",
               "Total area (ha)"),
  Value = c(as.character(n_set), as.character(n_set_iab),
            as.character(n_set_ibb), paste(set_imbalance, "to 1"),
            paste(set_years, collapse = " to "),
            paste(sort(unique(aset$lidar_year)), collapse = ", "),
            as.character(set_offset_med),
            format(set_ha, big.mark = ","))))
```

## Optical predictors

Sentinel-2 Level 2A over the leaf-on window of the label epoch, masked from the scene classification layer, corrected scene by scene for topographic illumination by SCS+C, reduced to a median composite and summarised to the inward-buffered polygon. No label enters this block: the analysis set is stripped to geometry and a positional identifier before anything is sent to Earth Engine, and the output is asserted to carry no agent, severity or host column.

The empirical `c` parameter is fitted per scene and per band by regression of reflectance on the illumination condition, over the sample polygons padded outward, so the correction is calibrated on the terrain it is applied to. Pixels facing away from the sun are masked rather than corrected, since the correction is undefined there. A tasselled cap transform is deliberately absent: the Sentinel-2 coefficients have not been read from their source publication, and unverified coefficients are not written into code.

```{r}
#| label: pipe-optical
#| echo: true
#| output: false
opt_path <- file.path(derived, "predictors-optical.csv")
opt_log <- file.path(derived, "predictors-optical-log.csv")

EE_PROJECT <- "murphys-deforisk"
S2_START <- "2019-06-01"
S2_END <- "2021-09-30"
BUFFER_M <- 20            # inward erosion of the survey polygon, provisional
SUMM_SCALE <- 20          # metres, scale of the polygon summary

if (!file.exists(opt_path)) local({
  library(rgee)
  MAX_CLOUD  <- 40        # scene cloud cover, percent
  FIT_PAD_M  <- 5000      # outward pad of the region the c parameter is fitted on
  FIT_SCALE  <- 100       # metres, scale of the c parameter fit
  FIT_MIN_PX <- 500       # minimum valid pixels in the fit region to keep a scene

  BANDS <- c("B2", "B3", "B4", "B5", "B6", "B7", "B8", "B8A", "B11", "B12")
  # Scene classification classes retained: 4 vegetation, 5 not vegetated,
  # 7 unclassified. Masked: 0 no data, 1 saturated, 2 dark, 3 cloud shadow,
  # 6 water, 8 and 9 cloud, 10 cirrus, 11 snow.
  SCL_KEEP <- c(4L, 5L, 7L)

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  # poly_id is positional in the frozen file, and every predictor block derives
  # it the same way, so the three tables join on a common key.
  poly_id <- sprintf("P%03d", seq_len(n_all))
  # Everything except geometry is dropped here. No label reaches Earth Engine.
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)

  # Sketch-mapped boundaries are approximate, so the polygon is eroded before it
  # is summarised. Polygons too small to survive erosion are dropped and counted.
  eroded <- sf::st_buffer(geom, -BUFFER_M)
  alive <- !sf::st_is_empty(eroded)
  n_dropped <- sum(!alive)
  eroded <- eroded[alive]; poly_id <- poly_id[alive]
  polys <- sf::st_sf(poly_id = poly_id,
                     geometry = sf::st_transform(eroded, 4326))
  stopifnot(identical(setdiff(names(polys), "geometry"), "poly_id"))

  ee_Initialize(project = EE_PROJECT, quiet = TRUE)
  fc <- sf_as_ee(polys)
  # The c parameter is fitted over the sample padded outward, rather than over
  # the whole scene, so the regression describes the terrain the correction is
  # applied to and the computation stays bounded.
  fit_region <- sf_as_ee(sf::st_transform(
    sf::st_union(sf::st_buffer(sf::st_union(geom), FIT_PAD_M)), 4326))

  dem <- ee$ImageCollection("COPERNICUS/DEM/GLO30")$select("DEM")$mosaic()$
    setDefaultProjection("EPSG:3857", NULL, 30)
  terrain <- ee$Terrain$products(dem)
  slope_rad <- terrain$select("slope")$multiply(pi / 180)
  aspect_rad <- terrain$select("aspect")$multiply(pi / 180)

  mask_scl <- function(img) {
    scl <- img$select("SCL")
    keep <- scl$eq(SCL_KEEP[1])
    for (k in SCL_KEEP[-1]) keep <- keep$Or(scl$eq(k))
    img$updateMask(keep)
  }

  # Illumination condition, the cosine of the solar incidence angle on the local
  # surface. A non-positive value faces away from the sun and is masked.
  illumination <- function(img) {
    sz <- ee$Number(img$get("MEAN_SOLAR_ZENITH_ANGLE"))$multiply(pi / 180)
    sa <- ee$Number(img$get("MEAN_SOLAR_AZIMUTH_ANGLE"))$multiply(pi / 180)
    il <- ee$Image$constant(sz$cos())$multiply(slope_rad$cos())$add(
      ee$Image$constant(sz$sin())$multiply(slope_rad$sin())$
        multiply(aspect_rad$subtract(sa)$cos()))
    il$rename("IL")
  }

  prepare <- function(img) {
    refl <- mask_scl(img)$select(BANDS)$divide(10000)
    il <- illumination(img)
    refl$updateMask(il$gt(0))$addBands(il)$
      copyProperties(img, list("system:time_start",
                               "MEAN_SOLAR_ZENITH_ANGLE",
                               "MEAN_SOLAR_AZIMUTH_ANGLE"))
  }

  # SCS+C. rho_corrected = rho * (cos(slope) cos(zenith) + c) / (IL + c), with c
  # the ratio of intercept to slope from the per-band regression of reflectance
  # on IL. A non-positive fitted slope is physically wrong, so c is forced large
  # and the correction collapses to the identity for that band.
  correct_scsc <- function(img) {
    sz <- ee$Number(img$get("MEAN_SOLAR_ZENITH_ANGLE"))$multiply(pi / 180)
    il <- img$select("IL")
    design <- ee$Image$constant(1)$rename("constant")$addBands(il)$
      addBands(img$select(BANDS))$updateMask(il$mask())
    fit <- design$reduceRegion(
      reducer = ee$Reducer$linearRegression(numX = 2L, numY = length(BANDS)),
      geometry = fit_region, scale = FIT_SCALE, maxPixels = 1e9, tileScale = 4L)
    coefs <- ee$Array(fit$get("coefficients"))
    intercept <- ee$Image$constant(
      coefs$slice(0L, 0L, 1L)$project(list(1L))$toList())$rename(BANDS)
    slope <- ee$Image$constant(
      coefs$slice(0L, 1L, 2L)$project(list(1L))$toList())$rename(BANDS)
    cimg <- intercept$divide(slope)
    cimg <- cimg$where(slope$lte(0), 1e6)
    cimg <- cimg$where(cimg$lte(0), 1e6)
    numer <- ee$Image$constant(sz$cos())$multiply(slope_rad$cos())$add(cimg)
    img$select(BANDS)$multiply(numer$divide(il$add(cimg)))$
      copyProperties(img, list("system:time_start"))
  }

  # Scenes with too few valid pixels cannot support the regression and are
  # dropped rather than corrected from a degenerate fit.
  count_fit_px <- function(img) {
    img$set("n_fit", img$select("B8")$reduceRegion(
      reducer = ee$Reducer$count(), geometry = fit_region, scale = FIT_SCALE,
      maxPixels = 1e9, tileScale = 4L)$get("B8"))
  }

  raw <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(fit_region)$filterDate(S2_START, S2_END)$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", MAX_CLOUD))
  n_raw <- raw$size()$getInfo()

  prepared <- raw$map(ee_utils_pyfunc(prepare))$
    map(ee_utils_pyfunc(count_fit_px))$
    filter(ee$Filter$gt("n_fit", FIT_MIN_PX))
  n_fit <- prepared$size()$getInfo()

  composite <- prepared$map(ee_utils_pyfunc(correct_scsc))$median()

  # Indices selected for sensitivity to canopy water content and to visible and
  # shortwave change. Provisional; the set is fixed at pre-registration.
  nd <- function(x, y, nm) composite$normalizedDifference(c(x, y))$rename(nm)
  indices <- nd("B8", "B4", "NDVI")$
    addBands(nd("B8", "B11", "NDMI"))$
    addBands(nd("B8", "B12", "NBR"))$
    addBands(nd("B8", "B5", "NDRE1"))$
    addBands(nd("B8", "B3", "GNDVI"))$
    addBands(composite$select("B11")$divide(composite$select("B8"))$rename("MSI"))$
    addBands(composite$select("B11")$divide(composite$select("B12"))$rename("SWIRR"))$
    addBands(composite$expression(
      "2.5 * (nir - red) / (nir + 6 * red - 7.5 * blue + 1)",
      list(nir = composite$select("B8"), red = composite$select("B4"),
           blue = composite$select("B2")))$rename("EVI"))

  stack <- composite$addBands(indices)
  VARS <- c(BANDS, "NDVI", "NDMI", "NBR", "NDRE1", "GNDVI", "MSI", "SWIRR", "EVI")

  reducer <- ee$Reducer$mean()$
    combine(reducer2 = ee$Reducer$stdDev(), sharedInputs = TRUE)$
    combine(reducer2 = ee$Reducer$count(), sharedInputs = TRUE)
  summaries <- stack$reduceRegions(collection = fc, reducer = reducer,
                                   scale = SUMM_SCALE, tileScale = 4L)

  # Drive credentials are not set on this machine, so the export goes to an
  # Earth Engine asset and is read back. Polling is to completion rather than
  # for a fixed number of attempts: reading the asset before the task has
  # written it fails with a misleading "not found".
  asset_id <- sprintf("projects/%s/assets/adelgid-optical-%s", EE_PROJECT,
                      format(Sys.Date(), "%Y%m%d"))
  try(ee_manage_delete(asset_id), silent = TRUE)
  task <- ee_table_to_asset(collection = summaries,
                            description = "adelgid_optical", assetId = asset_id)
  task$start()
  repeat {
    st <- task$status()
    if (st$state %in% c("COMPLETED", "FAILED", "CANCELLED", "CANCEL_REQUESTED"))
      break
    Sys.sleep(30)
  }
  if (st$state != "COMPLETED") stop("Earth Engine export ", st$state)

  got <- ee$FeatureCollection(asset_id)$getInfo()
  rows <- lapply(got$features, function(f)
    as.data.frame(f$properties, stringsAsFactors = FALSE))
  tab <- do.call(rbind, lapply(rows, function(r) {
    wanted <- c("poly_id", unlist(lapply(VARS, function(v)
      paste0(v, c("_mean", "_stdDev", "_count")))))
    for (m in setdiff(wanted, names(r))) r[[m]] <- NA_real_
    r[, wanted, drop = FALSE]
  }))
  tab <- tab[order(tab$poly_id), ]

  # The guard that matters: no label may leave this block.
  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE|lidar|offset|AREA",
                        names(tab), ignore.case = TRUE)) == 0)
  write.csv(tab, opt_path, row.names = FALSE)
  write.csv(data.frame(
    parameter = c("collection", "start", "end", "max_scene_cloud_pct",
                  "scl_kept", "scenes_matched", "scenes_corrected",
                  "topographic_correction", "dem", "c_fit_scale_m",
                  "c_fit_pad_m", "c_fit_min_pixels", "inward_buffer_m",
                  "summary_scale_m", "polygons_in", "polygons_dropped_by_buffer",
                  "polygons_out", "built"),
    value = c("COPERNICUS/S2_SR_HARMONIZED", S2_START, S2_END, MAX_CLOUD,
              paste(SCL_KEEP, collapse = " "), n_raw, n_fit, "SCS+C",
              "COPERNICUS/DEM/GLO30", FIT_SCALE, FIT_PAD_M, FIT_MIN_PX,
              BUFFER_M, SUMM_SCALE, n_all, n_dropped, nrow(tab),
              as.character(Sys.Date())), stringsAsFactors = FALSE),
    opt_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-optical-out
local({
  o <- read.csv(opt_path)
  v <- c("NDVI", "NDMI", "NBR", "EVI")
  fmt_tbl(data.frame(
    Index = v,
    Minimum = round(vapply(v, function(k) min(o[[paste0(k, "_mean")]]), 0), 3),
    Median = round(vapply(v, function(k) median(o[[paste0(k, "_mean")]]), 0), 3),
    Maximum = round(vapply(v, function(k) max(o[[paste0(k, "_mean")]]), 0), 3),
    `Within-polygon SD` = round(vapply(v, function(k)
      median(o[[paste0(k, "_stdDev")]]), 0), 3),
    check.names = FALSE))
})
```

## Radar predictors

Sentinel-1 Ground Range Detected in interferometric wide swath mode, both polarisations, over the same window as the optical build. The archive is delivered thermal-noise-removed, radiometrically calibrated and Range Doppler terrain corrected. Border noise removal and orbit refinement are not documented for the collection, so border noise is handled here by eroding each scene footprint and masking a low-intensity floor, and no orbit refinement is claimed.

Backscatter is flattened radiometrically against a digital elevation model by the angular volumetric model of @Vollrath_2020, which is appropriate to a vegetated surface. Layover and shadow are masked on the published conditions and the retained fraction of every polygon is written out, so the reduction is reported rather than absorbed. Both temporal statistics are formed within orbit pass before the passes are combined: ascending and descending views of the same slope differ by more than flattening removes, and pooling them would write that difference into the dispersion as a terrain effect. The level is averaged in power and converted to decibels afterwards, because averaging decibels averages logarithms and biases the result low; the dispersion is the standard deviation of the decibel series, since speckle is multiplicative and so additive in the log.

```{r}
#| label: pipe-radar
#| echo: true
#| output: false
rad_path <- file.path(derived, "predictors-radar.csv")
rad_log <- file.path(derived, "predictors-radar-log.csv")

if (!file.exists(rad_path)) local({
  library(rgee)
  BORDER_ERODE_M <- 1000   # inward erosion of the scene footprint, border noise
  BORDER_FLOOR_DB <- -30   # below this is border artefact, not target
  MIN_PASS_SCENES <- 5     # a pass with fewer gives no usable dispersion
  HEADING_SCALE <- 1000    # metres, scale of the look direction estimate
  POLS <- c("VV", "VH")
  PASSES <- c("ASCENDING", "DESCENDING")
  VARS <- c("VV_gamma0", "VH_gamma0", "VV_tsd", "VH_tsd")

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  poly_id <- sprintf("P%03d", seq_len(n_all))
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)
  eroded <- sf::st_buffer(geom, -BUFFER_M)
  alive <- !sf::st_is_empty(eroded)
  n_dropped <- sum(!alive)
  polys <- sf::st_sf(poly_id = poly_id[alive],
                     geometry = sf::st_transform(eroded[alive], 4326))
  stopifnot(identical(setdiff(names(polys), "geometry"), "poly_id"))

  ee_Initialize(project = EE_PROJECT, quiet = TRUE)
  fc <- sf_as_ee(polys)
  # sf_as_ee on an sfc returns an ee$Geometry, not a Feature.
  region <- sf_as_ee(sf::st_transform(sf::st_union(geom), 4326))

  dem <- ee$ImageCollection("COPERNICUS/DEM/GLO30")$select("DEM")$mosaic()$
    setDefaultProjection("EPSG:3857", NULL, 30)
  ninety_rad <- ee$Image$constant(90)$multiply(pi / 180)

  # Angular radiometric slope correction, volumetric model. The equations were
  # read from Vollrath, Mullissa and Reiche (2020) and the implementation checked
  # against the authors' own module:
  #   gamma0   = sigma0 / cos(theta_i)
  #   alpha_r  = arctan(tan(alpha_s) cos(phi_r))
  #   gamma0_f = gamma0 * tan(90 - theta_i) / tan(90 - theta_i + alpha_r)
  slope_correct <- function(img) {
    scene <- img$geometry()
    proj <- img$select("VV")$projection()
    # The look direction is the aspect of the incidence angle surface, taken
    # coarsely because it is near constant across the swath.
    heading <- ee$Terrain$aspect(img$select("angle"))$
      reduceRegion(reducer = ee$Reducer$mean(), geometry = scene,
                   scale = HEADING_SCALE, maxPixels = 1e9, bestEffort = TRUE)$
      get("aspect")
    sigma0_pow <- ee$Image$constant(10)$pow(img$select(POLS)$divide(10))
    theta_i <- img$select("angle")$multiply(pi / 180)$clip(scene)
    phi_i <- ee$Image$constant(heading)$multiply(pi / 180)
    # Terrain geometry, evaluated in the projection of the radar image.
    alpha_s <- ee$Terrain$slope(dem)$select("slope")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    phi_s <- ee$Terrain$aspect(dem)$select("aspect")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    alpha_r <- alpha_s$tan()$multiply(phi_i$subtract(phi_s)$cos())$atan()
    gamma0 <- sigma0_pow$divide(theta_i$cos())
    scf <- ninety_rad$subtract(theta_i)$add(alpha_r)$tan()$
      divide(ninety_rad$subtract(theta_i)$tan())
    g_db <- ee$Image$constant(10)$multiply(gamma0$divide(scf)$log10())$
      select(POLS)$rename(POLS)
    # Valid where neither in active layover nor in active shadow.
    valid <- alpha_r$lt(theta_i)$
      And(alpha_r$gt(ninety_rad$subtract(theta_i)$multiply(-1)))
    g_db$updateMask(valid)$updateMask(g_db$gt(BORDER_FLOOR_DB))$
      clip(scene$buffer(-BORDER_ERODE_M, 100))$
      copyProperties(img, list("system:time_start", "orbitProperties_pass"))
  }

  to_power <- function(img) {
    ee$Image$constant(10)$pow(img$select(POLS)$divide(10))$
      copyProperties(img, list("system:time_start"))
  }

  raw <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(region)$filterDate(S2_START, S2_END)$
    filter(ee$Filter$eq("instrumentMode", "IW"))$
    filter(ee$Filter$eq("resolution_meters", 10))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VV"))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VH"))
  n_raw <- raw$size()$getInfo()
  corrected <- raw$map(ee_utils_pyfunc(slope_correct))

  pow_list <- list(); var_list <- list()
  pass_n <- stats::setNames(integer(length(PASSES)), PASSES)
  for (p in PASSES) {
    cp <- corrected$filter(ee$Filter$eq("orbitProperties_pass", p))
    n_p <- cp$size()$getInfo()
    pass_n[p] <- n_p
    if (n_p < MIN_PASS_SCENES) next
    pow_list[[p]] <- cp$map(ee_utils_pyfunc(to_power))$mean()
    var_list[[p]] <- cp$select(POLS)$reduce(ee$Reducer$stdDev())$pow(2)
  }
  n_used <- sum(pass_n[names(pow_list)])
  stopifnot(length(pow_list) > 0)

  gamma_db <- ee$ImageCollection$fromImages(unname(pow_list))$mean()$
    log10()$multiply(10)$rename(c("VV_gamma0", "VH_gamma0"))
  pooled_sd <- ee$ImageCollection$fromImages(unname(var_list))$mean()$sqrt()$
    rename(c("VV_tsd", "VH_tsd"))
  # An unmasked copy of one band gives the polygon's full pixel count on the
  # same grid, so the layover and shadow loss is reported as a retained fraction
  # rather than left implicit in a smaller count.
  all_band <- gamma_db$select("VV_gamma0")$unmask(-999)$rename("ALL")
  stack <- gamma_db$addBands(pooled_sd)$addBands(all_band)

  reducer <- ee$Reducer$mean()$
    combine(reducer2 = ee$Reducer$stdDev(), sharedInputs = TRUE)$
    combine(reducer2 = ee$Reducer$count(), sharedInputs = TRUE)
  summaries <- stack$reduceRegions(collection = fc, reducer = reducer,
                                   scale = SUMM_SCALE, tileScale = 4L)

  asset_id <- sprintf("projects/%s/assets/adelgid-radar-%s", EE_PROJECT,
                      format(Sys.Date(), "%Y%m%d"))
  try(ee_manage_delete(asset_id), silent = TRUE)
  task <- ee_table_to_asset(collection = summaries,
                            description = "adelgid_radar", assetId = asset_id)
  task$start()
  repeat {
    st <- task$status()
    if (st$state %in% c("COMPLETED", "FAILED", "CANCELLED", "CANCEL_REQUESTED"))
      break
    Sys.sleep(30)
  }
  if (st$state != "COMPLETED") stop("Earth Engine export ", st$state)

  got <- ee$FeatureCollection(asset_id)$getInfo()
  rows <- lapply(got$features, function(f)
    as.data.frame(f$properties, stringsAsFactors = FALSE))
  tab <- do.call(rbind, lapply(rows, function(r) {
    wanted <- c("poly_id", unlist(lapply(VARS, function(v)
      paste0(v, c("_mean", "_stdDev", "_count")))), "ALL_count")
    for (m in setdiff(wanted, names(r))) r[[m]] <- NA_real_
    r[, wanted, drop = FALSE]
  }))
  tab <- tab[order(tab$poly_id), ]
  # The reduction the methods require to be reported.
  tab$retained_frac <- round(tab$VV_gamma0_count / tab$ALL_count, 4)

  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE|lidar|offset|AREA",
                        names(tab), ignore.case = TRUE)) == 0)
  write.csv(tab, rad_path, row.names = FALSE)
  write.csv(data.frame(
    parameter = c("collection", "start", "end", "instrument_mode",
                  "polarisations", "resolution_m", "scenes_matched",
                  "scenes_ascending", "scenes_descending", "scenes_used",
                  "collection_preprocessing", "orbit_file",
                  "border_noise_handling", "border_floor_db", "border_erode_m",
                  "terrain_flattening", "dem", "layover_shadow_mask",
                  "temporal_statistics", "inward_buffer_m", "summary_scale_m",
                  "polygons_in", "polygons_dropped_by_buffer", "polygons_out",
                  "min_retained_frac", "built"),
    value = c("COPERNICUS/S1_GRD", S2_START, S2_END, "IW",
              paste(POLS, collapse = " "), 10, n_raw,
              pass_n[["ASCENDING"]], pass_n[["DESCENDING"]], n_used,
              "thermal noise removal, calibration, Range Doppler terrain correction",
              "not documented by the Earth Engine catalogue, not asserted",
              "scene footprint eroded and low-intensity floor masked, applied here",
              BORDER_FLOOR_DB, BORDER_ERODE_M,
              "angular volumetric model, Vollrath et al. 2020",
              "COPERNICUS/DEM/GLO30",
              "active layover alpha_r > theta_i, active shadow alpha_r < -(90 - theta_i)",
              "within orbit pass, level as mean power, dispersion as dB standard deviation",
              BUFFER_M, SUMM_SCALE, n_all, n_dropped, nrow(tab),
              min(tab$retained_frac, na.rm = TRUE), as.character(Sys.Date())),
    stringsAsFactors = FALSE), rad_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-radar-out
local({
  r <- read.csv(rad_path)
  v <- c("VV_gamma0", "VH_gamma0", "VV_tsd", "VH_tsd")
  fmt_tbl(data.frame(
    Variable = c("VV backscatter (dB)", "VH backscatter (dB)",
                 "VV temporal dispersion (dB)", "VH temporal dispersion (dB)"),
    Minimum = round(vapply(v, function(k) min(r[[paste0(k, "_mean")]]), 0), 2),
    Median = round(vapply(v, function(k) median(r[[paste0(k, "_mean")]]), 0), 2),
    Maximum = round(vapply(v, function(k) max(r[[paste0(k, "_mean")]]), 0), 2)))
})
```

## Structural predictors

The tiles the sample needs are selected by intersection with the frozen analysis set **and** by acquisition year, so a polygon is measured on the campaign the sample matched it to rather than on whichever campaign happens to overlap it. Files already present at the right size are skipped, so the retrieval is restartable, and every file is verified against the size the server advertised because a truncated LAZ reads without error for a while and then fails deep inside a metric.

```{r}
#| label: pipe-lidar-fetch
#| echo: true
#| output: false
tile_map_path <- file.path(derived, "lidar-tile-map.csv")
tile_need_path <- file.path(derived, "lidar-tiles-needed.csv")
lidar_dir <- file.path(inputs, "lidar")
# Two streams, not four. Four saturates the link, and the object store then
# refuses new connections for seventy-five seconds at a time, so the run spends
# longer timing out than downloading.
FETCH_JOBS <- 2L

# The guard is on the tiles being complete, not on the manifest existing. A
# resume after a dropped connection must re-enter this block, and it would not
# if the presence of the manifest were taken as proof the retrieval finished.
lidar_complete <- function() {
  if (!file.exists(tile_need_path)) return(FALSE)
  tn <- read.csv(tile_need_path)
  p <- file.path(lidar_dir, tn$year, tn$filename)
  all(file.exists(p)) && all(file.size(p) == tn$bytes)
}

if (!lidar_complete()) local({
  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  keep <- data.frame(poly_id = sprintf("P%03d", seq_len(n_all)),
                     lidar_year = a$lidar_year, stringsAsFactors = FALSE)
  g4 <- sf::st_transform(sf::st_geometry(a), 4326)
  bb <- as.numeric(sf::st_bbox(g4))

  acc <- list(); off <- 0L
  repeat {
    u <- lidar_url(PC_LAYER, bb,
                   outFields = "maptile,year,density,filename,s3Url,projection",
                   returnGeometry = "true", outSR = "4326", f = "geojson",
                   resultOffset = off, resultRecordCount = PAGE)
    ch <- try(sf::st_read(u, quiet = TRUE), silent = TRUE)
    if (inherits(ch, "try-error") || nrow(ch) == 0L) break
    acc[[length(acc) + 1L]] <- ch
    if (nrow(ch) < PAGE) break
    off <- off + PAGE
  }
  idx <- do.call(rbind, acc)

  hits <- sf::st_intersects(sf::st_sf(geometry = g4), idx)
  rows <- lapply(seq_len(n_all), function(i) {
    k <- hits[[i]]
    k <- k[idx$year[k] == keep$lidar_year[i]]
    if (!length(k)) return(NULL)
    data.frame(poly_id = keep$poly_id[i], maptile = idx$maptile[k],
               year = idx$year[k], density = idx$density[k],
               projection = idx$projection[k], filename = idx$filename[k],
               s3Url = idx$s3Url[k], stringsAsFactors = FALSE)
  })
  if (!all(vapply(rows, Negate(is.null), TRUE)))
    stop("no matching-year tile for some polygons")

  # A polygon spanning a tile boundary must have its clouds merged before any
  # metric is computed, so the polygon-to-tile map is kept, not just the tiles.
  map <- do.call(rbind, rows)
  tiles <- unique(map[, c("maptile", "year", "density", "projection",
                          "filename", "s3Url")])
  tiles$path <- file.path(lidar_dir, tiles$year, tiles$filename)

  remote_size <- function(url) {
    h <- try(system2("curl", c("-sIL", "--max-time", "60", shQuote(url)),
                     stdout = TRUE, stderr = FALSE), silent = TRUE)
    if (inherits(h, "try-error")) return(NA_real_)
    l <- grep("^[Cc]ontent-[Ll]ength:", h, value = TRUE)
    if (!length(l)) return(NA_real_)
    as.numeric(sub("^[^:]*:\\s*", "", tail(l, 1)))
  }
  tiles$bytes <- vapply(tiles$s3Url, remote_size, numeric(1))
  stopifnot(!anyNA(tiles$bytes))

  have <- file.exists(tiles$path)
  local_sz <- ifelse(have, file.size(tiles$path), NA_real_)
  todo <- which(!(have & !is.na(local_sz) & local_sz == tiles$bytes))

  if (length(todo)) {
    for (y in unique(tiles$year))
      dir.create(file.path(lidar_dir, y), recursive = TRUE,
                 showWarnings = FALSE)
    # curl resumes with -C -, so an interrupted run costs only the partial file.
    # Parallelism is capped because the link saturates near four streams and the
    # object store is a shared public resource.
    ch <- split(todo, ceiling(seq_along(todo) / FETCH_JOBS))
    for (k in ch) {
      system(paste(c(sprintf(
        "curl -sSL -C - --retry 5 --retry-delay 5 -o %s %s &",
        shQuote(tiles$path[k]), shQuote(tiles$s3Url[k])), "wait"),
        collapse = "\n"))
    }
  }

  fsz <- ifelse(file.exists(tiles$path), file.size(tiles$path), NA_real_)
  bad <- which(is.na(fsz) | fsz != tiles$bytes)
  if (length(bad)) stop(length(bad), " tiles incomplete, rerun to resume")

  write.csv(map, tile_map_path, row.names = FALSE)
  write.csv(tiles[, c("maptile", "year", "density", "projection", "filename",
                      "bytes")], tile_need_path, row.names = FALSE)
})
```

```{r}
#| label: pipe-lidar-fetch-out
local({
  tn <- read.csv(tile_need_path)
  yrs <- sort(unique(tn$year))
  fmt_tbl(data.frame(
    Year = yrs,
    Tiles = as.integer(tapply(tn$bytes, tn$year, length)[as.character(yrs)]),
    `Size (GB)` = round(as.numeric(
      tapply(tn$bytes, tn$year, sum)[as.character(yrs)]) / 1e9, 1),
    check.names = FALSE))
})
```

The metrics are computed polygon by polygon rather than tile by tile, because the polygon is the unit of analysis and 49 of the 84 draw on more than one tile. No polygon draws on tiles from more than one UTM zone, so each is processed in its own tiles' native projection and no point cloud is ever reprojected; only the polygon outline is transformed, into the projection of the tile that will be read.

Two reads are made per tile and they are not symmetric. Ground returns are read at full density, because the terrain surface should be as good as the delivery allows and ground returns are a small fraction of the whole. Every other return is decimated at read time, because the delivered density is far higher than the analysis can use and reading it in full would not fit in memory.

Decimation is not an optimisation, it is the confound control. The index advertises 8 points per square metre for every tile in the province, and the delivered density over this sample ranges from 0.5 to 227, with a median of 41. It is stratified by campaign: the 2019 flights that carry 65 of the 84 polygons are the sparsest, and the 2024 and 2025 flights are the densest. Every metric below responds to density, so without a common density the model could separate the agents by reading which aircraft flew them. Each polygon is therefore brought to a common `TARGET_DENS` and the density before and after is written out, with the sensitivity restricted to the 2019 campaign reported separately.

```{r}
#| label: pipe-lidar-metrics
#| echo: true
#| output: false
struct_path <- file.path(derived, "predictors-structural.csv")
struct_log <- file.path(derived, "predictors-structural-log.csv")

# Provisional, and fixed at pre-registration.
CANOPY_MIN_H <- 2     # m, the canopy threshold, shared by cover and gap fraction
CHM_RES      <- 1     # m, canopy height model resolution
DTM_RES      <- 1     # m, terrain surface resolution
GAP_MIN_AREA <- 4     # m2, smallest patch of low canopy counted as a gap
TARGET_DENS  <- 8     # points per m2, the common density, the programme nominal

if (!file.exists(struct_path)) local({
  library(lidR)

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  poly_id <- sprintf("P%03d", seq_len(n_all))
  # Acquisition year is a design covariate, not a label, and is carried so the
  # campaign sensitivity can be run without rereading the sample.
  lidar_year <- a$lidar_year
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)
  eroded <- sf::st_buffer(geom, -BUFFER_M)

  map <- read.csv(tile_map_path, stringsAsFactors = FALSE)
  tn <- read.csv(tile_need_path, stringsAsFactors = FALSE)

  # A smoke-test switch, as scripts 15 and 16 carry. Unset in a render, so the
  # full sample is processed; set to a small number to check the machinery on a
  # few polygons before committing hours to all of them.
  lim <- suppressWarnings(as.integer(Sys.getenv("ADELGID_LIMIT", "")))
  if (!is.na(lim) && lim < n_all) {
    message("SMOKE TEST: first ", lim, " polygons")
    n_all <- lim
  }

  # --- the three metric families, each computed on the normalised cloud -----
  # Family 1 tests vertical redistribution, the prediction that top-down dieback
  # moves canopy material downward while the lower canopy persists.
  # Family 2 tests surface roughness, the prediction that gout and branch
  # deformation roughen the crown.
  # Family 3 describes density and openness, the response both agents share. It
  # is a control: if it alone separates the agents, the result is about damage
  # intensity rather than damage geometry.
  canopy_metrics <- function(las, chm, pg) {
    z <- las@data$Z
    z <- z[is.finite(z) & z >= CANOPY_MIN_H]
    if (length(z) < 100) return(NULL)
    first <- las@data[las@data$ReturnNumber == 1L, ]
    zf <- first$Z[is.finite(first$Z)]

    q <- stats::quantile(z, c(.25, .50, .75, .95, .99), names = FALSE)
    # The 99th percentile stands in for the maximum throughout. The delivered
    # tiles carry a low-noise class but no high-noise class, so a single bird or
    # atmospheric return tens of metres above the canopy would otherwise set the
    # ceiling for every ratio below it and drive them all towards zero.
    ztop <- q[5]

    # Cell accounting is confined to the polygon. The canopy model is
    # rectangular and its corners lie outside, where an empty cell means absence
    # of polygon rather than absence of canopy; counting those as gaps is what
    # would put gap fraction at 0.85 beside a canopy cover of 0.94.
    inside <- terra::rasterize(terra::vect(pg), chm, field = 1)
    in_cell <- !is.na(terra::values(inside, mat = FALSE))
    v <- terra::values(chm, mat = FALSE)
    n_cells <- sum(in_cell)
    v_in <- v[in_cell]
    v_in <- v_in[!is.na(v_in)]

    # Within the polygon an empty cell is counted as gap rather than as missing:
    # at the common density a 1 m cell expects several returns, so an empty one
    # is an opening. Contiguous patches only, since a single empty cell is where
    # one pulse found a hole in a crown, which is a property of the sampling
    # rather than of the stand.
    gap_r <- terra::ifel(!is.na(inside) & (is.na(chm) | chm < CANOPY_MIN_H),
                         1, NA)
    patches <- terra::patches(gap_r, directions = 8, zeroAsNA = TRUE)
    psz <- terra::freq(patches)
    keep_px <- if (is.null(psz) || !nrow(psz)) 0 else
      sum(psz$count[psz$count * (CHM_RES^2) >= GAP_MIN_AREA])
    gap_frac <- if (n_cells > 0) keep_px / n_cells else NA_real_

    data.frame(
      # Family 1, vertical redistribution
      zq25 = q[1], zq50 = q[2], zq75 = q[3], zq95 = q[4],
      zq25_zq95 = if (q[4] > 0) q[1] / q[4] else NA_real_,
      # Canopy relief ratio, (mean - min) / (top - min) over canopy returns.
      crr = if (ztop > min(z)) (mean(z) - min(z)) / (ztop - min(z)) else NA_real_,
      above_two_thirds = sum(z > (2 / 3) * ztop) / length(z),
      # Family 2, surface roughness
      rumple = tryCatch(lidR::rumple_index(chm), error = function(e) NA_real_),
      chm_sd = if (length(v_in) > 1) stats::sd(v_in) else NA_real_,
      cv_first = if (length(zf) > 1 && mean(zf) != 0)
        stats::sd(zf) / mean(zf) else NA_real_,
      # Family 3, density and openness, the control
      cover = if (length(zf)) sum(zf >= CANOPY_MIN_H) / length(zf) else NA_real_,
      gap_frac = gap_frac,
      # Vertical complexity index: Shannon entropy of the height distribution in
      # 1 m bins, normalised by the entropy of a uniform distribution over the
      # same bins, so 1 is an evenly filled profile and 0 a single layer.
      vci = tryCatch(lidR::VCI(z, zmax = ztop, by = 1), error = function(e) NA_real_),
      ztop = ztop, n_cells = n_cells,
      stringsAsFactors = FALSE)
  }

  one_polygon <- function(i) {
    tl <- map[map$poly_id == poly_id[i], ]
    if (!nrow(tl)) return("no tile matched in the index")
    paths <- file.path(lidar_dir, tl$year, tl$filename)
    area_m2 <- as.numeric(sf::st_area(eroded[i]))

    grd <- list(); veg <- list()
    for (k in seq_along(paths)) {
      hdr <- lidR::readLASheader(paths[k])
      crs_tile <- sf::st_crs(hdr)
      if (is.na(crs_tile)) return("tile carries no CRS")
      pg <- sf::st_transform(eroded[i], crs_tile)
      bb <- sf::st_bbox(pg)
      xy <- sprintf("-keep_xy %f %f %f %f",
                    bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"])

      # One read per tile, not two. A reader must decompress every point in a
      # chunk before it can discard any, so a second pass for ground costs a
      # second full decompression and buys nothing: reading once and splitting
      # in memory halves the only expensive step in this block. The full tile is
      # released before the next one is opened, so the peak is one tile.
      las <- tryCatch(lidR::readLAS(paths[k], select = "xyzrc", filter = xy),
                      error = function(e) NULL)
      if (is.null(las) || lidR::is.empty(las)) next

      # Ground at full density: the terrain surface should be as good as the
      # delivery allows, and ground returns are a small share of the whole.
      g <- lidR::filter_poi(las, Classification == 2L)
      if (!lidR::is.empty(g)) grd[[length(grd) + 1L]] <- g

      # Everything else brought down towards the common density with a margin,
      # so the merged cloud never holds the delivered density.
      dens <- tn$density_actual[match(tl$filename[k], tn$filename)]
      v <- if (is.finite(dens) && dens > TARGET_DENS * 3)
        lidR::decimate_points(las, lidR::random(TARGET_DENS * 3)) else las
      if (!lidR::is.empty(v)) veg[[length(veg) + 1L]] <- v
      rm(las); invisible(gc())
    }
    # The index is a footprint layer and its footprints claim ground the
    # delivered clouds do not cover, so a polygon can match a tile and still
    # have no points in it. That is a coverage fact and is reported, not
    # silently dropped.
    if (!length(grd) || !length(veg))
      return("no returns inside the polygon: the tile index claims coverage the point cloud does not contain")

    g <- if (length(grd) == 1) grd[[1]] else do.call(rbind, grd)
    v <- if (length(veg) == 1) veg[[1]] else do.call(rbind, veg)
    pg <- sf::st_transform(eroded[i], sf::st_crs(v))
    v <- tryCatch(lidR::clip_roi(v, pg), error = function(e) NULL)
    if (is.null(v) || lidR::is.empty(v)) return("no returns after clipping to the polygon")
    dens_raw <- nrow(v@data) / area_m2 / (if (length(veg)) 1 else 1)

    # Terrain from class 2 returns only. The delivered tiles carry no vegetation
    # classes, so ground is the one class that can be trusted.
    dtm <- tryCatch(lidR::rasterize_terrain(g, res = DTM_RES,
                                            algorithm = lidR::tin()),
                    error = function(e) NULL)
    if (is.null(dtm)) return("terrain surface could not be interpolated")
    v <- tryCatch(lidR::normalize_height(v, dtm), error = function(e) NULL)
    if (is.null(v)) return("height normalisation failed")

    # The common density. Every metric responds to density and density tracks
    # campaign, so this is what stops the model reading the aircraft.
    v <- tryCatch(lidR::decimate_points(v, lidR::homogenize(TARGET_DENS, res = 1)),
                  error = function(e) v)
    dens_used <- nrow(v@data) / area_m2

    chm <- tryCatch(lidR::rasterize_canopy(v, res = CHM_RES,
                                           algorithm = lidR::p2r(subcircle = 0.15)),
                    error = function(e) NULL)
    if (is.null(chm)) return("canopy height model could not be built")

    m <- canopy_metrics(v, chm, pg)
    if (is.null(m)) return("fewer than 100 canopy returns above the canopy threshold")
    cbind(data.frame(poly_id = poly_id[i], area_ha = round(area_m2 / 1e4, 2),
                     n_tiles = nrow(tl), lidar_year = lidar_year[i],
                     dens_raw = round(dens_raw, 1),
                     dens_used = round(dens_used, 2),
                     stringsAsFactors = FALSE), m)
  }

  # Each polygon is written as it finishes, and a restart skips whatever is
  # already recorded. The run takes hours and this machine has interrupted it
  # more than once; holding the whole table in memory until the end means an
  # interruption at the eightieth polygon discards the other seventy-nine.
  part_path <- file.path(derived, "predictors-structural-partial.csv")
  drop_path <- file.path(derived, "structural-dropped.csv")
  append_row <- function(df, path) {
    write.table(df, path, sep = ",", row.names = FALSE, qmethod = "double",
                col.names = !file.exists(path), append = file.exists(path))
  }
  done_ids <- character(0)
  for (p in c(part_path, drop_path))
    if (file.exists(p)) done_ids <- c(done_ids, read.csv(p)$poly_id)
  if (length(done_ids))
    message("resuming: ", length(done_ids), " polygons already recorded")

  for (i in seq_len(n_all)) {
    if (poly_id[i] %in% done_ids) next
    r <- tryCatch(one_polygon(i),
                  error = function(e) paste("error:", conditionMessage(e)))
    if (is.character(r) || is.null(r)) {
      why <- if (is.null(r)) "unknown" else r
      # A polygon that produced no metrics is written out with its reason, so
      # the difference between the frozen sample and this table is auditable
      # rather than a silent shrinkage.
      append_row(data.frame(poly_id = poly_id[i], reason = why,
                            stringsAsFactors = FALSE), drop_path)
      message(sprintf("  %s DROPPED (%d of %d): %s", poly_id[i], i, n_all, why))
    } else {
      append_row(r, part_path)
      message(sprintf("  %s ok (%d of %d)", poly_id[i], i, n_all))
    }
    invisible(gc())
  }
  tab <- read.csv(part_path, stringsAsFactors = FALSE)
  tab <- tab[order(tab$poly_id), ]
  drops <- if (file.exists(drop_path))
    read.csv(drop_path, stringsAsFactors = FALSE) else data.frame()
  tab <- tab[order(tab$poly_id), ]

  # The guard that matters: no label may leave this block. lidar_year is a
  # design covariate and is named so it cannot be mistaken for one.
  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE", names(tab),
                        ignore.case = TRUE)) == 0)
  write.csv(tab, struct_path, row.names = FALSE)
  # The partial file has served its purpose. Leaving it would let a
  # deliberate re-run resume instead of recomputing, which is not what
  # deleting the output asks for.
  if (file.exists(part_path)) invisible(file.remove(part_path))

  write.csv(data.frame(
    parameter = c("source", "tiles", "tile_points", "canopy_min_h_m",
                  "chm_res_m", "dtm_res_m", "gap_min_area_m2",
                  "target_density_pts_m2", "ground_class", "normalisation",
                  "chm_algorithm", "gap_rule", "inward_buffer_m",
                  "polygons_in", "polygons_out", "polygons_failed",
                  "density_before_min", "density_before_median",
                  "density_before_max", "density_after_median", "built"),
    value = c("LidarBC classified point clouds", nrow(tn),
              format(sum(as.numeric(tn$bytes)), big.mark = ","),
              CANOPY_MIN_H, CHM_RES, DTM_RES, GAP_MIN_AREA, TARGET_DENS,
              "2 (ground); tiles carry no vegetation classes",
              "TIN terrain from class 2, heights normalised above it",
              "p2r with 0.15 m subcircle",
              paste0("CHM cell below ", CANOPY_MIN_H,
                     " m or empty, in contiguous patches of at least ",
                     GAP_MIN_AREA, " m2"),
              BUFFER_M, n_all, nrow(tab), nrow(drops),
              min(tab$dens_raw), median(tab$dens_raw), max(tab$dens_raw),
              median(tab$dens_used), as.character(Sys.Date())),
    stringsAsFactors = FALSE), struct_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-lidar-metrics-out
local({
  s <- read.csv(struct_path)
  fam <- c(zq50 = "Median canopy height (m)", zq95 = "95th percentile height (m)",
           zq25_zq95 = "Ratio of 25th to 95th", crr = "Canopy relief ratio",
           rumple = "Rumple", chm_sd = "Canopy height SD (m)",
           cover = "Canopy cover above 2 m", gap_frac = "Gap fraction",
           vci = "Vertical complexity index")
  fmt_tbl(data.frame(
    Metric = unname(fam),
    Minimum = round(vapply(names(fam), function(k) min(s[[k]], na.rm = TRUE), 0), 3),
    Median = round(vapply(names(fam), function(k) median(s[[k]], na.rm = TRUE), 0), 3),
    Maximum = round(vapply(names(fam), function(k) max(s[[k]], na.rm = TRUE), 0), 3)))
})
```

## Generated scripts

`05.scripts/` is written from this document, not maintained beside it. The block below extracts every pipeline chunk with `knitr::purl()` and splits it by chunk label, so each numbered script is the code above and cannot drift from it. It is skipped on a normal render and run deliberately.

```{r}
#| label: pipe-purl
#| echo: true
#| eval: false
# Regenerate 05.scripts/ from this manuscript. Run from 01.manuscript/.
local({
  tmp <- tempfile(fileext = ".R")
  knitr::purl("manuscript.qmd", output = tmp, documentation = 1L, quiet = TRUE)
  src <- readLines(tmp)

  # Quarto keeps chunk options in "#|" comments rather than in the chunk header,
  # so purl emits anonymous separators and the label survives only in the body.
  # Split on the label lines, not on the separators.
  lab_at <- grep("^#\\|\\s*label:", src)
  labels <- trimws(sub("^#\\|\\s*label:", "", src[lab_at]))

  targets <- c(`pipe-labels` = "01-extract-adelgid-labels.R",
               `pipe-tiles` = "02-query-lidar-tiles.R",
               `pipe-coverage` = "02-query-lidar-coverage.R",
               `pipe-basemap` = "03-prepare-basemap.R",
               `pipe-candidates` = "12-site-candidates-province.R",
               `pipe-availability` = "13-data-availability-first.R",
               `pipe-sample` = "14-build-analysis-set.R",
               `pipe-optical` = "15-build-optical-predictors.R",
               `pipe-radar` = "16-build-radar-predictors.R",
               `pipe-lidar-fetch` = "17-build-lidar-metrics.R",
               `pipe-lidar-metrics` = "18-lidar-canopy-metrics.R")

  header <- c(
    "# GENERATED FILE. Do not edit.",
    "# Written from 01.manuscript/manuscript.qmd by the pipe-purl chunk.",
    "# Edit the chunk in the manuscript and regenerate; edits here are lost.",
    "# Run from the repository root.", "",
    "suppressMessages({ library(sf); library(terra) })",
    "root <- normalizePath('.')",
    "inputs <- file.path(root, '02.inputs')",
    "outputs <- file.path(root, '03.outputs')",
    "derived <- file.path(inputs, 'derived')",
    "BC_ALBERS <- 3005", "")

  # Blocks that define objects later blocks rely on. Prepended so a generated
  # script stands alone rather than assuming the manuscript ran first.
  prereq <- list(
    `pipe-candidates` = "pipe-tiles", `pipe-coverage` = "pipe-tiles",
    `pipe-lidar-fetch` = "pipe-tiles", `pipe-sample` = "pipe-basemap",
    `pipe-lidar-metrics` = c("pipe-optical", "pipe-lidar-fetch"),
    `pipe-radar` = "pipe-optical")

  body_of <- function(lab) {
    i <- match(lab, labels)
    stop_at <- if (i < length(lab_at)) lab_at[i + 1] - 1L else length(src)
    b <- src[lab_at[i]:stop_at]
    b <- b[!grepl("^#\\|", b)]          # drop the chunk options
    b[!grepl("^## -{4,}", b)]           # drop purl's separators
  }
  # A prerequisite contributes only the definitions above its guard.
  trim_to_defs <- function(b) {
    k <- grep("^if \\(", b)
    if (length(k)) b[seq_len(k[1] - 1L)] else b
  }

  for (lab in names(targets)) {
    if (!lab %in% labels) next
    pre <- character(0)
    for (q in prereq[[lab]])
      pre <- c(pre, paste0("# --- from chunk ", q,
                           ", definitions this script depends on ---"),
               trim_to_defs(body_of(q)), "")
    writeLines(c(header, pre, body_of(lab)),
               file.path("..", "05.scripts", targets[[lab]]))
  }
})
```

# Data availability

Survey label extracts and the derived basemap are available in the project repository, together with the scripts that produce them from the public sources cited above. Sentinel-2 and Sentinel-1 imagery are available through Google Earth Engine, lidar point clouds through the LidarBC open data portal, and the survey archive through the BC Data Catalogue.

# References

::: {#refs}
:::
