---
title: "Separating balsam woolly adelgid from bark beetle damage in Pacific silver fir on Vancouver Island: a multi-sensor attribution test"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: >
  Balsam woolly adelgid (*Adelges piceae*) and western balsam bark beetle
  (*Dryocoetes confusus*) are both mapped on Pacific silver fir (*Abies
  amabilis*) in the British Columbia aerial overview survey, so a canopy-damage
  detector that cannot separate them measures neither. In this study, we examine
  whether the addition of lidar and radar predictors to a spectral baseline improves
  separation of the two agents, and by how much.
keywords:
  - balsam woolly adelgid
  - Adelges piceae
  - Abies amabilis
  - forest health monitoring
  - lidar
  - synthetic aperture radar
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
library(sf)
library(terra)
library(flextable)

root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
outputs <- file.path(root, "03.outputs")
derived <- file.path(inputs, "derived")

BC_ALBERS <- 3005

# Figures are written to 03.outputs/figures as well as embedded, so the README
# and any submission package reference committed files rather than extracting
# images from a rendered document.
knitr::opts_chunk$set(fig.path = file.path(outputs, "figures/"),
                      dpi = 300, dev = "png")

fmt_tbl <- function(df, right = NULL) {
  right_cols <- union(names(df)[vapply(df, is.numeric, TRUE)], right)
  left_cols <- setdiff(names(df), right_cols)
  ft <- flextable::flextable(df)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "header")
  if (length(right_cols))
    ft <- flextable::align(ft, j = right_cols, align = "right", part = "body")
  if (length(left_cols))
    ft <- flextable::align(ft, j = left_cols, align = "left", part = "body")
  ft <- flextable::line_spacing(ft, space = 1, part = "all")
  ft <- flextable::padding(ft, padding.top = 1, padding.bottom = 1,
                           part = "all")
  flextable::set_table_properties(ft, layout = "autofit", width = 1)
}
```

```{r}
#| label: load-labels
#| echo: true
#| output: false
iab <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                   quiet = TRUE)
vi <- sf::st_read(file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)

# The analysis set, built by 05.scripts/14-build-analysis-set.R: adelgid and bark
# beetle polygons on Pacific silver fir that fall under lidar coverage acquired
# within five years of the survey. This file is the sample; nothing below filters
# it again.
aset <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
n_set <- nrow(aset)
n_set_iab <- sum(aset$PEST_SPECIES_CODE == "IAB")
n_set_ibb <- sum(aset$PEST_SPECIES_CODE == "IBB")
set_imbalance <- round(n_set_ibb / n_set_iab, 1)
set_years <- range(aset$CAPTURE_YEAR)
set_offset_med <- median(aset$year_offset)
set_ha <- round(sum(aset$AREA_HA))
n_set_ms <- sum(aset$PEST_SEVERITY_CODE %in% c("M", "S", "V"))

host_label <- c(B = "True fir, unspecified", BA = "Pacific silver fir",
                BG = "Grand fir", BL = "Subalpine fir")
agent_label <- c(IAB = "Balsam woolly adelgid",
                 IBB = "Western balsam bark beetle")
sev_label <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
               V = "Very severe")

n_iab_bc <- nrow(iab)
yr_iab <- range(iab$CAPTURE_YEAR)
n_vi_iab <- sum(vi$PEST_SPECIES_CODE == "IAB")
n_vi_ibb <- sum(vi$PEST_SPECIES_CODE == "IBB")
ha_vi_iab <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IAB"]))
ha_vi_ibb <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IBB"]))

ba <- vi[vi$TREE_SPECIES_CODE == "BA", ]
n_ba_iab <- sum(ba$PEST_SPECIES_CODE == "IAB")
n_ba_ibb <- sum(ba$PEST_SPECIES_CODE == "IBB")
imbalance <- round(n_ba_ibb / n_ba_iab, 1)

sev_iab <- table(factor(vi$PEST_SEVERITY_CODE[vi$PEST_SPECIES_CODE == "IAB"],
                        levels = names(sev_label)))
n_iab_ms <- sum(sev_iab[c("M", "S", "V")])
```

# Introduction

Balsam woolly adelgid, *Adelges piceae* (Ratzeburg), is an introduced insect that damages and kills North American true firs of all age classes, having entered the continent in the early twentieth century through separate introductions on the east and west coasts [@Campbell_2023]. Its congener the hemlock woolly adelgid, *Adelges tsugae* Annand, is the better-studied insect, but the two behave differently on this coast. In eastern North America *A. tsugae* is invasive and lethal, and infested eastern hemlock die within four to fifteen years with within-stand mortality that can exceed ninety percent [@MacQuarrie_2025]. In western North America the same species is a separate and native lineage that "exists at low densities in forests and rarely kills trees" [@MacQuarrie_2025]. Damage to western hemlock in British Columbia has accordingly been minor, and the provincial forest health survey reflects the asymmetry directly: it carries a damage-agent code for balsam woolly adelgid and none for hemlock woolly adelgid. In British Columbia the woolly adelgid of management concern is therefore *A. piceae* on true fir, and its principal host on the coast is Pacific silver fir, *Abies amabilis* (Douglas) Forbes.

The two agents of interest here attack the same host and are recorded by the same survey, so a manager reading a damage map of Pacific silver fir cannot tell from the map which insect produced it. Their damage differs in kind, however, and that difference is the basis of this study; it is set out in the study system below.

Remote detection of adelgid damage has a long record. @FRANKLIN_1995 separated damage classes on individual balsam fir from airborne multispectral imagery at 40 to 76 percent accuracy. Landsat change detection was applied to adelgid-induced hemlock decline from the late 1990s [@Royle_1997; @Bonneau_1999], and later work pushed detection towards lower infestation levels [@Williams_2017]. @Cook_2010 discriminated infested from non-infested subalpine fir at 94 percent from branch-level spectra. Most recently @Campbell_2023 mapped infestation severity in subalpine fir from satellite imagery with terrain and climate, and their result frames the present study: spectral data alone "successfully detected forest degradation" but "failed to distinguish between causal agents of change", returning a mean coefficient of determination of 0.482, while terrain and climate reached 0.746 but "lacked the ability to characterize local-scale tree damage", and only the combination reached 0.836.

Attribution has since become a distinct problem with its own methods, addressed mainly with optical time series. @Oeser_2017 attributed windthrow, cleared windthrow, bark beetle and harvest in Central Europe with overall accuracies of 76 to 86 percent; @Zhang_2022 mapped causal agents across boreal and arctic North America; @Sebald_2021 improved separation of human from natural disturbance using landscape context; and @Morin_Bernard_2023 reached 82.9 percent attributing agent before estimating severity, finding that agent-specific severity models required different predictors. Structure and radar have entered the field on the detection side. Terrestrial lidar detects *Ips typographus* infestation early [@Junttila_2019], simulated spaceborne lidar recovers the understorey and midstorey defoliation that hemlock woolly adelgid causes before any overstorey signal appears [@Boucher_2020], and L-band synthetic aperture radar delineates windthrow from insect outbreak at 69 to 88 percent [@Tanase_2018; @Hollaus_2019]. Most directly, @Atkins_2020 used terrestrial lidar canopy structural traits in a machine learning framework across six moderate-severity agents including hemlock woolly adelgid, and found that fire and ice storms changed the amount and position of canopy vegetation while insect and pathogen infestation altered canopy arrangement and complexity instead.

Two limits define what remains unresolved, and both emerged from a structured search of this literature. We searched Semantic Scholar and Crossref for work on adelgid remote sensing, on attribution of damage to specific agents, on lidar and radar applied to insect damage, and on the accuracy of the aerial surveys that supply agent labels, screening 5,663 records against criteria fixed before searching and carrying 721 forward for reading. The searched literature is deep on detection and comparatively shallow on attribution, and the two limits below are what that asymmetry exposes.

First, attribution between insect agents has been demonstrated, but only where the agents attack different hosts. @Meigs_2011 first characterised the spectral trajectories of a defoliator and a bark beetle separately against field plots, finding that western spruce budworm showed long-duration decline followed by recovery while mountain pine beetle showed both short and long-duration declines with variable recovery, but stopped short of classifying them. @Senf_2015 took that step, separating mountain pine beetle from western spruce budworm in British Columbia using LandTrendr spectral trajectories and random forests, mapping insect disturbance at 76.8 percent overall accuracy and attributing agent at "overall accuracies ranging from 75.3% to 88.0%, depending on whether only pure host-stands or mixed stands with both insect hosts were considered". That conditional is the important result for the present study. Where host composition disambiguated the agents, attribution succeeded; where it did not, it degraded sharply, and the authors report that "in the case of mixed host stands, nearly 45% of the western spruce budworm disturbances were falsely attributed to mountain pine beetle". Host is therefore doing much of the work that appears to be done by damage. Balsam woolly adelgid and western balsam bark beetle attack the same host species, which removes that cue entirely rather than diluting it, and we found no study that attributes between insect agents under that condition. Second, the structural evidence that agents leave distinguishable signatures rests on plot-scale instrumented sites; @Atkins_2020 state their findings are "limited in their generalizability due to lack of replication among disturbances". Whether canopy structure carries agent-diagnostic information at the coarse support of an operational survey, across a landscape, is untested.

That distinction matters operationally because the survey record is itself the working map of which insect is where. @Coleman_2018 assessed aerial detection surveys against ground observation across multiple ecological regions and reported overall accuracies above 70 percent, with commission errors of 2 and 5 percent for mortality and defoliation damage types, but found that "accuracy declined as the specificity for observations went from genera to species level for tree species and damaging agents". Species-level agent attribution is precisely the level at which balsam woolly adelgid and western balsam bark beetle are distinguished, and independent evaluations of survey severity estimates have returned overall accuracies as low as 35 to 56 percent depending on the class scheme [@Bright_2020b; @Housman_2018].

Each of our objectives follows from one of these findings. Because @Campbell_2023 established that spectral predictors detect adelgid degradation but fail to identify its cause, and because @Senf_2015 showed that spectral attribution between insect agents weakens when host stops disambiguating them, our first objective is to quantify how far a spectral-only classifier separates balsam woolly adelgid from western balsam bark beetle damage in Pacific silver fir on Vancouver Island, establishing that baseline rather than assuming it. Because the structural work of @Atkins_2020 and @Boucher_2020 shows that agents alter canopy arrangement in ways that colour alone does not register, but shows it at plot scale and without replication, our second objective is to test whether adding lidar-derived canopy structure and Sentinel-1 backscatter improves that separation at operational support. And because @Oeser_2017, @Sebald_2021 and @Morin_Bernard_2023 demonstrate that attribution accuracy depends heavily on how the evaluation is constructed, our third objective is to report any improvement against a validation design that respects the spatial dependence of the labels and the support at which they were drawn, rather than against a random partition that would inflate it.

We hypothesise that the multi-sensor model achieves higher balanced accuracy than the spectral baseline, because gout and progressive top-kill alter canopy geometry in a way that rapid whole-crown mortality does not, and because the spectral baseline is expected to be weak here for the reason @Senf_2015 identify: with host held constant it cannot fall back on host composition. The null hypothesis is that structural and radar predictors add nothing beyond the spectral baseline.

# Materials and methods

## Study system

Balsam woolly adelgid is a small, wingless, parthenogenetic insect that feeds on the bark of true firs through a fine stylet. In North America it reproduces without males and disperses passively, as first-instar crawlers carried by wind, birds and human movement of infested material, so spread is slow, local and irregular rather than following a front. Feeding is concentrated on the bole and in the crown, and its damage is chemically mediated rather than mechanical: saliva injected during feeding is toxic to the host, inhibiting bud formation and inducing abnormal xylem. The visible consequences are swellings at nodes and branch tips, known as gouting, whose incidence varies systematically with crown class and plant hardiness zone [@Guillet_2010], followed by crown thinning and progressive dieback from the top of the tree downward. Trees may persist in this deformed and declining condition for years, and death, where it occurs, follows a long structural decline rather than preceding it.

Western balsam bark beetle, *Dryocoetes confusus* Swaine, is the other agent recorded on the same host in this landscape. It is a phloem-feeding beetle that mass-attacks the bole, girdling the tree by disrupting phloem transport and introducing a pathogenic fungus. Its damage is not preceded by structural deformation of branches or by top-down crown thinning: an attacked tree retains its architecture and fades whole-crown, from green through yellow to red, over roughly one to two years.

The beetle does not act alone, and the comparison class is defined accordingly. On true fir it operates within a mortality complex alongside *Armillaria* root disease, the associated fungus *Ophiostoma dryocoetidis* and, at some sites, spruce beetle. Interactions of this kind alter canopy and stand characteristics jointly rather than separately: @Klutsch_2014 found on a single ponderosa pine host that "fuel bed and canopy characteristics were dependent on the presence of both dwarf mistletoe and mountain pine beetle-caused mortality". The class contrasted with the adelgid here is therefore best understood as bark-beetle-associated mortality rather than as the effect of one insect in isolation, and it is described that way throughout. This is a property of the system, not of the labels, and it would persist under any labelling scheme. The two agents therefore differ in the geometry of the damage, deformation and progressive top-kill against retained architecture, and in its tempo, years of decline against a rapid whole-crown fade.

That contrast is the mechanism this study tests. An instrument sensitive only to foliage colour observes both agents at the end state, when crowns of either kind are dead or dying and spectrally similar. An instrument sensitive to canopy geometry observes the difference in how the crown arrived there. Whether that difference survives at the coarse support of an operational survey polygon, rather than at the plot scale where it has been examined, is the open question. The plot-scale evidence is itself mixed, and both sides of it concern the congener. @Atkins_2020 found that insect and pathogen infestation altered canopy arrangement and complexity, and @Alveshere_2025 report that hemlock woolly adelgid infestation influences canopy structural complexity, while @Choi_2023 found for press disturbances including the same insect that "the impact ... on other dimensions of canopy structure could not be clearly detected, likely because of compensatory growth". A structural signal is therefore plausible but not established, and compensatory growth is a specific mechanism by which it could be absent.

Pacific silver fir, *Abies amabilis*, is the host that carries both agents on the coast. It is a shade-tolerant, late-seral true fir of the wetter and cooler coastal forests, occurring from near sea level to the subalpine, and it is a commercially and ecologically significant component of the stands in which it grows. Subalpine fir, *Abies lasiocarpa*, carries western balsam bark beetle at higher elevations in the same region but is not recorded with adelgid damage here, so the analysis is confined to Pacific silver fir in order to hold host constant.

## Study area

The study area is Vancouver Island and the immediately adjacent mainland coast of British Columbia, Canada, within an envelope from 128.5 to 123.0 degrees west and 48.3 to 51.0 degrees north (@fig-studyarea). Vancouver Island is roughly 460 km long and separated from the mainland by the Strait of Georgia and Johnstone Strait. A mountain spine runs its length, rising above 2000 m in the Vancouver Island Ranges, and it divides a wet, exposed west coast from a drier eastern lowland in the rain shadow of the same mountains.

Forest composition follows that gradient. The eastern lowlands carry the Coastal Douglas-fir zone, the wetter and more extensive middle elevations the Coastal Western Hemlock zone, and the upper slopes the Mountain Hemlock zone before passing into subalpine parkland. Pacific silver fir occurs through the Coastal Western Hemlock and Mountain Hemlock zones, generally above the Coastal Douglas-fir lowlands, growing with western hemlock, western redcedar and yellow cedar. Subalpine fir is largely confined to the highest elevations. The climate is maritime, with mild wet winters and comparatively dry summers, and it moderates the winter cold that constrains adelgid survival at the northern and continental edges of the insect's range.

Both damage agents are recorded across this area by the provincial aerial overview survey, and their distributions overlap on the shared host without coinciding: the adelgid record concentrates at lower and middle elevations on Pacific silver fir, while western balsam bark beetle extends to higher elevations where subalpine fir dominates (@fig-studyarea).

All spatial data are handled in NAD83 / BC Albers, EPSG 3005, which is the projection of the source survey archive, and every map is drawn in that projection.

```{r}
#| label: fig-studyarea
#| fig-cap: "Study area on Vancouver Island and the adjacent mainland coast of British Columbia, Canada, showing aerial overview survey polygons attributed to balsam woolly adelgid and to western balsam bark beetle on true fir hosts. Relief is shaded from a 1 km digital elevation model illuminated from the northwest. Projection NAD83 / BC Albers (EPSG 3005); axis labels in degrees."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
bm <- file.path(derived, "basemap.gpkg")
land   <- sf::st_read(bm, "land",   quiet = TRUE)
admin  <- sf::st_read(bm, "admin",  quiet = TRUE)
rivers <- sf::st_read(bm, "rivers", quiet = TRUE)
lakes  <- sf::st_read(bm, "lakes",  quiet = TRUE)
places <- sf::st_read(bm, "places", quiet = TRUE)
hs <- terra::rast(file.path(derived, "hillshade.tif"))

vi_p <- sf::st_transform(vi, BC_ALBERS)
fr <- sf::st_bbox(land)

op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")

# Sea, then relief-shaded land, then hydrography, then boundaries. The sea is
# painted across the whole plot region rather than the land bounding box, since
# the fixed aspect ratio leaves the region taller than the data.
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.30, end = 1),
            legend = FALSE, add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey30", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = "#6f9ab5", lwd = 0.3,
     add = TRUE)
plot(sf::st_geometry(rivers), col = "#5b8db0", lwd = 0.8, add = TRUE)
# Only the international boundary, drawn dashed. The provincial outline is not
# drawn because within this frame it coincides with the coastline.
plot(sf::st_geometry(admin), col = "grey20", lwd = 1.1, lty = 2, add = TRUE)

# Bark beetle first: the far less numerous adelgid class must stay legible.
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IBB", ]),
     col = "#2c6f9e", border = "#2c6f9e", add = TRUE)
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IAB", ]),
     col = "#c1372f", border = "#c1372f", add = TRUE)

pxy <- sf::st_coordinates(places)
points(pxy, pch = 21, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Label side chosen per settlement so no label runs off the frame or across a
# neighbouring one.
lab_pos <- c(Comox = 4, `Campbell River` = 2, `Port Alberni` = 2,
             Nanaimo = 4, Victoria = 2, Vancouver = 4, Tofino = 2,
             `Port Hardy` = 4)
text(pxy[, 1], pxy[, 2], places$name, pos = lab_pos[places$name],
     offset = 0.4, cex = 0.6, col = "grey10")

# Axes labelled in degrees although the plot is projected, so the reader gets
# geographic coordinates without the distortion of plotting in degrees.
xt <- pretty(fr[c("xmin", "xmax")], 5)
yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

# Scale bar drawn in projected metres, so the stated length is metrically true.
sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")

# North arrow.
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.62, pt.cex = 1.1, pch = 22,
       pt.bg = c("#c1372f", "#2c6f9e"), col = c("#c1372f", "#2c6f9e"),
       legend = c("Balsam woolly adelgid", "Western balsam bark beetle"))
par(op)
```

## Data

@tbl-inputs lists each dataset and its access terms. The custodian of the survey archive is the BC Ministry of Forests; GeoBC operates the lidar programme and the two are separate agencies.

```{r}
#| label: tbl-inputs
#| tbl-cap: "Datasets used in the analysis, with their role and access terms."
inputs_tbl <- data.frame(
  Dataset = c("Pest Infestation Polygons", "Sentinel-2 L2A", "Sentinel-1 GRD",
              "LidarBC open lidar"),
  Role = c("Damage agent labels", "Optical predictors", "Radar predictors",
           "Canopy structure predictors"),
  Source = c("BC Data Catalogue", "Google Earth Engine",
             "Google Earth Engine", "LidarBC open data portal"),
  Licence = c("Open Government Licence - British Columbia",
              "Copernicus open access", "Copernicus open access",
              "Open Government Licence - British Columbia")
)
fmt_tbl(inputs_tbl)
```

Damage agents in the delivered geodatabase are recorded as three-letter codes without names. Codes were resolved against the *Forest Health Aerial Overview Survey Standards for British Columbia*, the Ministry of Forests adaptation of Canadian Forest Service FHN Report 97-1, which gives IAB for balsam woolly adelgid and IBB for western balsam bark beetle. The same document lists IAB against sub-alpine true firs, amabilis fir and grand fir, matching the host codes present in the extract. That code list contains no entry for hemlock woolly adelgid, so the absence of that species from the archive is a property of the survey design rather than of the years sampled.

### Survey data

Damage labels are polygons from the British Columbia aerial overview survey. Surveyors sketch-map visible forest damage from a fixed-wing aircraft, recording host species, damage agent and an ordinal severity class for each polygon. A minimum of eighty percent of the province is flown annually, with gaps caused by weather and smoke.

Survey polygons carry limitations that constrain the analysis and that the custodian states explicitly: errors of omission and commission occur, spatial accuracy can be poor, and the survey is not intended to be accurate enough to direct operational activities. Its stated purpose is to record general trends in disturbance size and frequency. Polygon area is in part a product of the digitising rule rather than a measurement, since a clump of one to thirty dead or dying trees is delineated as 0.25 ha and a clump of thirty-one to fifty as 0.5 ha. Severity is likewise an ordinal judgement: the standards define light, moderate and severe as 1 to 10, 11 to 29 and 30 percent or more of trees recently killed, and classify all point sources as severe on digitising. The unit of analysis here is therefore the survey polygon rather than the pixel, and severity is not modelled.

Across the study area the survey records `r n_vi_iab` adelgid polygons covering `r format(ha_vi_iab, big.mark = ",")` ha and `r n_vi_ibb` bark beetle polygons covering `r format(ha_vi_ibb, big.mark = ",")` ha on true fir hosts (@tbl-labels). Province-wide the adelgid record comprises `r n_iab_bc` polygons spanning `r yr_iab[1]` to `r yr_iab[2]`. Both agents are recorded on Pacific silver fir, `r n_ba_iab` adelgid against `r n_ba_ibb` bark beetle polygons, and it is this shared host that makes the separation an attribution problem rather than a detection problem. Analysis is restricted to that host so that agent is not confounded with differences in host reflectance.

```{r}
#| label: tbl-labels
#| tbl-cap: "Aerial overview survey polygons in the study area by damage agent and host tree species. Area is the survey-recorded polygon area."
lab <- as.data.frame(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Host  = host_label[vi$TREE_SPECIES_CODE]
), stringsAsFactors = FALSE)
names(lab)[3] <- "Polygons"
lab <- lab[lab$Polygons > 0, ]

area <- tapply(vi$AREA_HA,
               list(agent_label[vi$PEST_SPECIES_CODE],
                    host_label[vi$TREE_SPECIES_CODE]),
               function(v) round(sum(v)))
lab$`Area (ha)` <- mapply(function(a, h) area[a, h], lab$Agent, lab$Host)
lab <- lab[order(lab$Agent, -lab$Polygons), ]
fmt_tbl(lab)
```

### Analysis sample

Not every polygon on the shared host can enter the analysis, because the structural predictors require lidar and lidar coverage is neither universal nor contemporaneous with the survey. The sample is therefore defined by data availability and is stated before any result, so that its restrictions are read as a property of the design rather than as a post hoc exclusion.

A polygon enters the analysis when it is attributed to either agent, lies on Pacific silver fir, falls within LidarBC point cloud coverage, and has a nearest lidar acquisition within five years of its survey year. That yields `r n_set` polygons, `r n_set_iab` adelgid and `r n_set_ibb` bark beetle, covering `r format(set_ha, big.mark = ",")` ha, surveyed between `r set_years[1]` and `r set_years[2]`, with a median label-to-acquisition offset of `r set_offset_med` years. The classes are imbalanced by `r set_imbalance` to 1 in favour of the bark beetle, which is considerably more even than the `r imbalance` to 1 of the unrestricted host-matched record.

The cost of that definition is severity, and it is severe. Only `r n_set_ms` polygons in the analysis sample are rated moderate or worse, none of them adelgid. This is not an artefact of the filter thresholds. Across the province no adelgid polygon rated moderate or worse falls under lidar at a one, three, five or ten year offset, and only one does at twenty years, by which point stand condition cannot be assumed to have persisted. The moderate and severe adelgid record and the lidar programme do not overlap in British Columbia, and no choice of study area within the province changes that. The study accepts a low-severity sample in exchange for the structural arm, and the consequences for interpretation are set out in the limitations.

The severity distribution is strongly weighted towards the two lowest classes (@tbl-severity). Only `r n_iab_ms` adelgid polygons in the study area are rated moderate or worse, too few to support the severity regression used by @Campbell_2023, and a further reason the response here is the two-class agent label.

```{r}
#| label: tbl-severity
#| tbl-cap: "Severity class of survey polygons by damage agent. Severity is the surveyor's ordinal rating."
sev <- as.data.frame.matrix(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Severity = factor(sev_label[vi$PEST_SEVERITY_CODE],
                    levels = unname(sev_label))
))
sev <- sev[, colSums(sev) > 0, drop = FALSE]
sev <- cbind(Agent = rownames(sev), sev)
fmt_tbl(sev)
```

The two agents also peak in different survey years (@fig-timeseries). The adelgid record concentrates in 2019 to 2021 while the bark beetle record is distributed more evenly across the past decade. A classifier trained on imagery drawn from each polygon's own survey year would therefore risk learning the year rather than the agent, and a single common imagery epoch is used for both classes.

```{r}
#| label: fig-timeseries
#| fig-cap: "Survey polygons mapped per year on true fir hosts in the study area, by damage agent."
#| fig-width: 7
#| fig-height: 3.6
tt <- table(vi$CAPTURE_YEAR, agent_label[vi$PEST_SPECIES_CODE])
op <- par(mar = c(4, 4.2, 0.6, 0.6))
matplot(as.numeric(rownames(tt)), tt, type = "l", lty = 1, lwd = 2,
        col = c("#c1372f", "#2c6f9e"), las = 1,
        xlab = "Survey year", ylab = "Polygons mapped")
legend("topleft", colnames(tt), col = c("#c1372f", "#2c6f9e"),
       lty = 1, lwd = 2, bty = "n", cex = 0.75)
par(op)
```

### Remote sensing data

Optical predictors are drawn from Sentinel-2 Level 2A surface reflectance and radar predictors from Sentinel-1 Ground Range Detected imagery in interferometric wide swath mode, both accessed through Google Earth Engine. Scenes are restricted to the leaf-on window of the epoch in which the adelgid record concentrates, 1 June 2019 to 30 September 2021, and optical scenes to those with less than forty percent cloud cover. Scene counts available over the study area are given in @tbl-imagery.

```{r}
#| label: gee-availability
#| echo: true
#| output: false
gee_status <- tryCatch({
  library(rgee)
  ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  aoi <- ee$Geometry$Rectangle(c(-128.5, 48.3, -123.0, 51.0))

  s2 <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))

  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$eq("instrumentMode", "IW"))

  data.frame(
    Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
    Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
    Period = "2019-06-01 to 2021-09-30",
    Scenes = c(s2$size()$getInfo(), s1$size()$getInfo())
  )
}, error = function(e) {
  data.frame(Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
             Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
             Period = "2019-06-01 to 2021-09-30",
             Scenes = NA_integer_)
})
```

```{r}
#| label: tbl-imagery
#| tbl-cap: "Satellite scenes available over the study area for the analysis period."
fmt_tbl(gee_status)
```

### Lidar data

Canopy structure is derived from the LidarBC open lidar programme, which publishes classified point clouds for public download. Coverage over the study area comprises 18,840 tiles at a nominal density of eight points per square metre (@tbl-lidar), including acquisitions in 2019 that coincide with the concentration of the adelgid survey record. Tiles are classified to ground, low point and overlap but carry no vegetation classes, so canopy height is computed as height above a ground surface interpolated from ground returns rather than read from an existing classification.

```{r}
#| label: tbl-lidar
#| tbl-cap: "Lidar point cloud coverage over the study area and over the Comox field window. Acquisition years and point classes are summarised from a sample of up to 1000 tiles."
lidar <- read.csv(file.path(derived, "lidar-coverage.csv"), check.names = FALSE)
names(lidar) <- c("Extent", "Tiles", "Full index", "Acquisition years",
                  "Density (pts/m2)", "Point classes")
fmt_tbl(lidar)
```

### Field site selection

Ground confirmation requires polygons that carry both an adelgid label and lidar coverage, and that can be reached. @fig-fieldsites shows where those conditions coincide, and the answer constrains the design rather than merely illustrating it.

```{r}
#| label: fig-fieldsites
#| fig-cap: "Candidate field sites. Points are aerial overview survey polygons attributed to balsam woolly adelgid, sized by area and coloured by severity class; open circles lack lidar coverage and filled circles have it. Shaded envelopes show LidarBC point cloud coverage. Rings mark 50, 100 and 150 km from Comox and are shown for travel planning only, not as a selection criterion. Two mainland records fall outside this frame. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
# Province-wide candidates, not island-only and not filtered by distance from
# Comox: per the 2026-08-14 decision, lidar and severity availability set the
# study's scope and location rather than proximity to a field base. The frame
# below is coastal, so the two mainland interior records fall outside it and are
# reported in the text instead.
cand <- read.csv(file.path(derived, "site-candidates-province.csv"))
cand$severity <- cand$severity_code
cand <- cand[cand$lon > -128.5 & cand$lon < -123.0 &
             cand$lat > 48.3 & cand$lat < 51.2, ]
cand_sf <- sf::st_transform(
  sf::st_as_sf(cand, coords = c("lon", "lat"), crs = 4326), BC_ALBERS)
lidar_env <- sf::st_read(file.path(derived, "lidar-coverage-dissolved.gpkg"),
                         quiet = TRUE)
acc <- sf::st_read(file.path(derived, "field-access-places.gpkg"), quiet = TRUE)
comox <- sf::st_transform(
  sf::st_sfc(sf::st_point(c(-124.928, 49.673)), crs = 4326), BC_ALBERS)

fr <- sf::st_bbox(land)
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.45, end = 1), legend = FALSE,
            add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey35", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = NA, add = TRUE)
plot(sf::st_geometry(rivers), col = "#7ba6c4", lwd = 0.6, add = TRUE)

# Lidar coverage envelope, drawn under the sites so the sites stay readable.
plot(sf::st_geometry(lidar_env), col = "#b9d9b0", border = "#7fae74",
     lwd = 0.3, add = TRUE)

# Distance rings from Comox. Straight line, not road distance.
for (d in c(50, 100, 150)) {
  plot(sf::st_geometry(sf::st_buffer(comox, d * 1000)), col = NA,
       border = "grey30", lty = 3, lwd = 0.8, add = TRUE)
}

sev_col <- c(T = "#f0c987", L = "#e08a45", M = "#c1372f", S = "#7d1d18",
             V = "#4a0f0c")
cxy <- sf::st_coordinates(cand_sf)
cex_by_area <- 0.5 + 1.5 * sqrt(cand$area_ha / max(cand$area_ha))
has_lidar <- cand$lidar_tiles > 0
points(cxy[!has_lidar, , drop = FALSE], pch = 1, lwd = 0.9,
       col = sev_col[cand$severity[!has_lidar]], cex = cex_by_area[!has_lidar])
points(cxy[has_lidar, , drop = FALSE], pch = 16,
       col = sev_col[cand$severity[has_lidar]], cex = cex_by_area[has_lidar])

axy <- sf::st_coordinates(acc)
points(axy, pch = 22, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Comox is drawn separately as the field base, so its label is suppressed here
# and Courtenay is thrown left to clear the diamond.
side <- ifelse(acc$name == "Courtenay", 2, 4)
keep <- acc$name != "Comox"
text(axy[keep, 1], axy[keep, 2], acc$name[keep], pos = side[keep],
     offset = 0.35, cex = 0.55, col = "grey10")
points(sf::st_coordinates(comox), pch = 23, bg = "black", col = "black",
       cex = 1.2)
text(sf::st_coordinates(comox)[1], sf::st_coordinates(comox)[2], "Comox",
     pos = 4, offset = 0.45, cex = 0.6, font = 2, col = "black")

xt <- pretty(fr[c("xmin", "xmax")], 5); yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.58,
       pch = c(16, 1, 22, 23, NA), pt.bg = c(NA, NA, "white", "black", NA),
       col = c("grey25", "grey25", "grey15", "black", "#7fae74"),
       lty = c(NA, NA, NA, NA, 1), lwd = c(NA, NA, NA, NA, 6),
       legend = c("Adelgid polygon, lidar", "Adelgid polygon, no lidar",
                  "Settlement", "Comox", "Lidar coverage"))
legend("bottomright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.58, pch = 16, col = sev_col,
       legend = c("Trace", "Light", "Moderate", "Severe"), title = "Severity")
par(op)
```

@fig-fieldsites exposes a trade-off between severity and structural data that is not resolvable by widening the search area, and it constrains the study rather than merely describing it. Across the province the survey records 267 adelgid polygons, of which 89 fall within lidar coverage and 25 are rated moderate or worse. Only nine polygons satisfy both conditions, and eight of those were surveyed in 2007 or 2008 against lidar acquired between 2023 and 2025, an offset of twelve to seventeen years over which stand condition cannot be assumed constant.

The recent record shows the trade-off at its sharpest. Twelve polygons rated moderate or severe were surveyed from 2019 onward, all of them on Pacific silver fir, and none has any lidar coverage. Ten lie on northern Vancouver Island between Sayward and the Nimpkish valley and two on the mainland coast. Conversely, 43 polygons combine lidar coverage, a label-to-acquisition offset of three years or less, and the shared host, but 37 of those are rated trace and 10 light, with a single moderate polygon among them.

Structural predictors and moderate-or-worse severity therefore cannot both be obtained from existing public data on the same polygons. Three responses are available and the choice belongs to the analysis design rather than to this section: restrict the structural arm to the low-severity polygons where lidar exists and accept that it tests the hypothesis where the signal is weakest; run the high-severity polygons on optical and radar predictors alone; or acquire structural data directly over the twelve recent high-severity polygons, which are between 4 and 24 ha and within reach of an uncrewed aerial survey. Whichever is chosen, the field campaign is a stratified sample over an accessible subset rather than over the label population, and the resulting coverage bias is reported rather than assumed away.

### Tile detail

@fig-tiles shows individual lidar tiles against survey polygon outlines over northern Vancouver Island, where the recent moderate and severe records concentrate. It is drawn at tile resolution rather than as a coverage envelope because the practical question when planning a visit is whether a specific polygon has a specific tile beneath it, which a dissolved envelope conceals.

```{r}
#| label: fig-tiles
#| fig-cap: "Lidar tiles and survey polygons over northern Vancouver Island. Grey squares are individual LidarBC point cloud tiles, shaded by acquisition year. Polygon outlines are aerial overview survey records: red for balsam woolly adelgid, blue for western balsam bark beetle, with adelgid polygons rated moderate or worse drawn heavier. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 6.4
#| out-width: "100%"
det <- c(xmin = -126.4, ymin = 50.05, xmax = -124.2, ymax = 50.95)

tiles <- sf::st_read(file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)
tiles_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(tiles), det)), BC_ALBERS)
iab_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(iab), det)), BC_ALBERS)
vi_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(vi), det)), BC_ALBERS)
land_d <- suppressWarnings(sf::st_crop(
  land, sf::st_bbox(sf::st_transform(
    sf::st_as_sfc(sf::st_bbox(det, crs = 4326)), BC_ALBERS))))
acc_d <- sf::st_read(file.path(derived, "field-access-places.gpkg"),
                     quiet = TRUE)

fr <- sf::st_bbox(sf::st_transform(sf::st_as_sfc(sf::st_bbox(det, crs = 4326)),
                                   BC_ALBERS))
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.55, end = 1), legend = FALSE,
            add = TRUE, maxcell = 3e6)
plot(sf::st_geometry(land_d), col = NA, border = "grey35", lwd = 0.7,
     add = TRUE)

# Tiles, shaded by acquisition year so a polygon's temporal match is legible.
yrs <- sort(unique(tiles_d$year))
pal <- setNames(grey.colors(max(length(yrs), 2), start = 0.55, end = 0.88),
                yrs)
plot(sf::st_geometry(tiles_d), col = pal[as.character(tiles_d$year)],
     border = "grey60", lwd = 0.15, add = TRUE)

plot(sf::st_geometry(vi_d[vi_d$PEST_SPECIES_CODE == "IBB", ]), col = NA,
     border = "#2c6f9e", lwd = 0.7, add = TRUE)
plot(sf::st_geometry(iab_d), col = NA, border = "#c1372f", lwd = 1.0,
     add = TRUE)
hi <- iab_d[iab_d$PEST_SEVERITY_CODE %in% c("M", "S", "V"), ]
if (nrow(hi)) plot(sf::st_geometry(hi), col = "#c1372f66",
                   border = "#7d1d18", lwd = 2.0, add = TRUE)

ap <- sf::st_coordinates(acc_d)
inb <- ap[, 1] > fr["xmin"] & ap[, 1] < fr["xmax"] &
       ap[, 2] > fr["ymin"] & ap[, 2] < fr["ymax"]
if (any(inb)) {
  points(ap[inb, , drop = FALSE], pch = 22, bg = "white", col = "grey15",
         cex = 0.9, lwd = 0.8)
  text(ap[inb, 1], ap[inb, 2], acc_d$name[inb], pos = 4, offset = 0.4,
       cex = 0.62, col = "grey10")
}

xt <- pretty(fr[c("xmin", "xmax")], 4); yt <- pretty(fr[c("ymin", "ymax")], 4)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 10000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.010 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.8 * h, c("0", "10", "20 km"),
     cex = 0.6, col = "grey15")
nx <- fr["xmax"] - 0.06 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.08 * diff(fr[c("ymin", "ymax")])
a <- 0.045 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.5 * a, "N", cex = 0.8, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.6,
       lty = c(1, 1, 1, NA), lwd = c(2.0, 1.0, 0.7, NA), pch = c(NA, NA, NA, 22),
       pt.bg = c(NA, NA, NA, "grey75"),
       col = c("#7d1d18", "#c1372f", "#2c6f9e", "grey60"),
       legend = c("Adelgid, moderate or worse", "Adelgid, other severity",
                  "Bark beetle", "Lidar tile"))
par(op)
```

## Data preparation

Terrain governs every preparation choice here. The study area is mountainous, the survey polygons sit on slopes rather than on flat ground, and both optical and radar measurements are sensitive to illumination and viewing geometry in ways that vary systematically with aspect and slope. An uncorrected terrain effect would enter the models as a predictor, and because the two agents occupy partly different elevations it could be learned in place of the damage signal.

Optical predictors are built from Sentinel-2 Level 2A scenes over the leaf-on window of the label epoch. Cloud, cloud shadow, snow and water are masked from the scene classification layer before compositing, and a per-pixel median composite is formed to suppress residual cloud edges and to reduce the influence of any single acquisition geometry. From the composite we derive the visible, red edge, near infrared and shortwave infrared bands and a set of indices selected for sensitivity to canopy water content and to the visible and shortwave changes @Campbell_2023 associated with adelgid infestation. A topographic illumination correction is applied before indices are computed, since a slope-driven brightness difference would otherwise propagate into every ratio.

Radar predictors are built from Sentinel-1 Ground Range Detected scenes in interferometric wide swath mode, using both VV and VH polarisations. Processing follows the standard sequence of orbit refinement, border and thermal noise removal, radiometric calibration, terrain flattening and Range Doppler terrain correction, with the flattening step performed against a digital elevation model so that backscatter is expressed on a terrain-corrected reference area rather than on the ellipsoid. Areas of layover and shadow are masked and excluded, which removes some steep terrain from the analysis and is reported as a reduction in the usable sample rather than ignored. Speckle is suppressed by temporal averaging across the scene stack, and the mean and the temporal dispersion of backscatter are retained separately, the latter because a changing crown may be distinguishable by its variability as well as its mean.

Structural predictors are derived from LidarBC point clouds. Ground returns are used to interpolate a terrain surface, point heights are normalised above it, and canopy metrics are computed from the normalised cloud rather than from any supplied vegetation classification, since the delivered tiles carry no vegetation classes.

The metric set is chosen against the mechanism rather than against convention. A generic canopy description would test whether the two agents differ in any structural respect, which is weaker and less falsifiable than the question asked here. The mechanism predicts two specific differences. Progressive top-down dieback with the tree persisting should move canopy material downward in the profile and thin the upper strata while the lower canopy remains, whereas whole-crown mortality with architecture retained should reduce returns throughout the profile without redistributing them. Gout and branch deformation should roughen the crown surface, while a beetle-killed crown should not roughen until branches begin to shed.

Metrics therefore fall into three families. The first tests vertical redistribution: height percentiles at the 25th, 50th, 75th and 95th centiles, the ratio of the 25th to the 95th, canopy relief ratio, and the proportion of returns above two thirds of maximum height. The second tests surface roughness: rumple, being the ratio of canopy surface area to ground area, the standard deviation of the canopy height model within the polygon, and the coefficient of variation of first-return heights. The third describes density and openness, being canopy cover above 2 m, gap fraction and the vertical complexity index.

The third family is a control rather than a hypothesis. Both agents thin the canopy, so if density and openness alone separate them the structural result concerns damage intensity rather than damage geometry, and it is reported in those terms.

All three families are summarised to the survey polygon, which is the unit of analysis. Polygons are eroded by an inward buffer before summarisation, because sketch-mapped boundaries are approximate and an unbuffered polygon edge will include unaffected canopy. Within the eroded polygon each predictor is reduced to its mean and its dispersion, so that a heterogeneous polygon is not represented solely by a central value. Polygons too small to survive erosion are excluded and counted.

The specific index set, the lidar metric set and the buffer distance are fixed at pre-registration, before any predictor value is joined to any label, and are reported there with the reasoning for each threshold.

## Random forests

The response is the two-class agent label, adelgid against bark beetle, for survey polygons on Pacific silver fir. Restricting to a single host is what makes the comparison an attribution test: with host held constant, a classifier cannot succeed by learning host reflectance or host site preference in place of the damage itself.

Predictors are grouped into three families, spectral, structural and radar, and three nested models are compared: spectral only, spectral with structure, and all three families together. The nesting is deliberate, since the question is not whether the agents can be separated at all but how much of any separation the spectral baseline already provides. Survey year and terrain are carried as covariates in every model so that their contribution is estimated rather than absorbed into the sensor terms; the two agents peak in different years (@fig-timeseries) and occupy partly different elevations, and both are plausible confounders.

Models are fitted with random forests, following @Campbell_2023, and compared at equal complexity, with hyperparameters tuned only on training-internal splits so that no evaluation fold informs model selection. Random forests are chosen for their tolerance of correlated predictors, which the three families supply in quantity, and because variable importance grouped by family gives a direct reading of where separating information lies.

Because polygons cluster in space, performance is estimated by spatially blocked cross-validation rather than random partitioning, which would place spatially adjacent and often spatially overlapping polygons in both training and evaluation folds and inflate accuracy. On a sample of `r n_set` polygons the blocking scheme is the design choice most likely to determine the result, so it is fixed in advance and not tuned afterwards.

Blocks are contiguous groups formed by clustering polygon centroids, sized so that no block holds fewer than eight polygons and no fold fewer than two blocks, which yields approximately five folds at this sample size. Block size is set from the empirical range of spatial autocorrelation in the predictors, estimated on the predictors alone before any label is joined, so that the choice cannot be informed by the outcome. Where that range would leave fewer than five folds the design accepts fewer folds rather than smaller blocks, since leakage inflates accuracy while a low fold count only widens the interval. Polygons that overlap or share a boundary are forced into the same block irrespective of the clustering, because the survey digitises adjacent damage as separate records and treating them as independent would leak directly. The realised number of folds and block sizes are reported with every accuracy figure. The two classes are imbalanced by `r set_imbalance` to 1 in favour of the bark beetle class, and resampling is applied within training folds only, never to the evaluation folds, so that reported performance refers to the observed class prevalence. With `r n_set` polygons the sample is small for a spatially blocked design, and the number of blocks is reported alongside every accuracy figure.

## Accuracy assessment

The primary metric is balanced accuracy, the mean of sensitivity and specificity. It is chosen over overall accuracy because the classes are imbalanced: a model assigning every polygon to the bark beetle class would score highly on overall accuracy while identifying no adelgid damage at all, which is the failure mode the study exists to avoid.

Per-class sensitivity and specificity and the full confusion matrix are reported alongside it, each with the support on which it is computed. The comparison of interest is paired, the difference in balanced accuracy between the multi-sensor and spectral-only models evaluated on identical folds, so that fold-to-fold variation is differenced out rather than treated as independent error, with its interval obtained by bootstrapping over folds. The hypothesis is confirmed if and only if that difference is positive and its 95 percent interval excludes zero, disconfirmed if the interval excludes zero in the negative direction, and inconclusive if it spans zero.

Accuracy figures reported here refer to agreement with the surveyor's attribution, not with an independent ground determination. They therefore measure reproduction of the operational record and inherit whatever error that record carries, which @Coleman_2018 show increases as agent specificity increases. The metric implementation is verified against cases whose values are known by construction (@tbl-checks).

```{r}
#| label: eval-metric
#| echo: true
#| output: false
balanced_accuracy <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}
```

```{r}
#| label: tbl-checks
#| tbl-cap: "Verification of the balanced accuracy implementation against cases with analytically known values."
t_perfect <- c("IAB", "IAB", "IBB", "IBB")
t_inverted <- c("IBB", "IBB", "IAB", "IAB")
imb_truth <- c("IAB", "IBB", "IBB", "IBB")

checks <- data.frame(
  Case = c("Perfect prediction",
           "Fully inverted prediction",
           "All assigned to majority class",
           "Imbalanced, all majority (overall accuracy 0.75)"),
  Expected = c(1, 0, 0.5, 0.5),
  Observed = c(
    balanced_accuracy(t_perfect, t_perfect),
    balanced_accuracy(t_perfect, t_inverted),
    balanced_accuracy(t_perfect, rep("IBB", 4)),
    balanced_accuracy(imb_truth, rep("IBB", 4))
  )
)
checks$Agrees <- abs(checks$Expected - checks$Observed) < 1e-9
fmt_tbl(checks)
```

# Results

## Model comparison

[PENDING. Balanced accuracy of the three nested models and the difference between the multi-sensor and spectral-only models, with its interval.]

```{r}
#| label: tbl-primary
#| tbl-cap: "Separation of the two damage agents by predictor set, under spatially blocked cross-validation."
primary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  `Balanced accuracy` = "pending",
  `Difference from spectral` = "pending",
  `95% CI` = "pending",
  check.names = FALSE
)
fmt_tbl(primary)
```

## Class performance

[PENDING. Per-class sensitivity and specificity, the confusion matrix and its support, and the relative importance of predictors grouped by sensor family.]

```{r}
#| label: tbl-secondary
#| tbl-cap: "Per-class performance of each model, with support."
secondary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  Sensitivity = "pending",
  Specificity = "pending",
  Support = "pending",
  check.names = FALSE
)
fmt_tbl(secondary)
```

# Discussion and conclusions

[PENDING. Interpretation of the model comparison, in relation to the spectral attribution failure reported by @Campbell_2023 and the commission-error reduction reported by @Jones_2015, closing with what the result implies for verification of aerial overview survey attributions in coastal British Columbia.]

## Limitations

The defining limitation is the severity of the sample, and it follows from data availability rather than from choice. No adelgid polygon rated moderate or worse in British Columbia falls under lidar coverage at any useful temporal offset, so a study with a structural arm is necessarily a study of trace and light damage. The analysis is therefore concentrated in the damage classes where any spectral or structural signal is weakest, and a null result would be the expected outcome of the severity distribution as much as a statement about the sensors.

Other limitations apply irrespective of the result. The labels are sketch-mapped from an aircraft and their spatial accuracy is acknowledged by the custodian to be variable, so polygon boundaries are approximate and some damage is missed entirely. That concentration may fall below the detection floor of moderate-resolution optical sensing altogether: @Ye_2021 report that their framework detected mountain pine beetle only where red-stage mortality occupied at least 56 percent of a Landsat pixel, and spruce beetle only above 55 percent grey-stage occupancy. A trace-rated polygon will rarely approach those fractions, so a null result for the spectral baseline would be consistent with the severity distribution rather than with the agents being spectrally inseparable, and the two explanations cannot be distinguished with these labels alone. Agent attribution in the source data is the surveyor's judgement from the air and is not independently verified on the ground, so the models can at best reproduce that judgement and any systematic error in it propagates into the results. Field confirmation addresses this only partly, because the diagnostic signs of adelgid infestation, the woolly wax masses and stem gout, occur on living and declining trees and are lost from those long dead.

## Detection and attribution

This study evaluates attribution rather than detection, and the distinction is maintained throughout. Establishing that a canopy has changed is a different problem from identifying the agent responsible, and a model can perform well at the first while failing the second [@Campbell_2023]. Accuracy figures are therefore reported with their support, their class balance and their validation design.

# Data availability

Survey label extracts and the derived basemap are available in the project repository, together with the scripts that produce them from the public sources cited above. Sentinel-2 and Sentinel-1 imagery are available through Google Earth Engine, lidar point clouds through the LidarBC open data portal, and the survey archive through the BC Data Catalogue.

# References

::: {#refs}
:::
