---
title: "Separating balsam woolly adelgid from bark beetle damage in Pacific silver fir on Vancouver Island: a multi-sensor attribution test"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: >
  Balsam woolly adelgid (*Adelges piceae*) and western balsam bark beetle
  (*Dryocoetes confusus*) are both mapped on Pacific silver fir (*Abies
  amabilis*) in the British Columbia aerial overview survey, so a canopy-damage
  detector that cannot separate them measures neither. In this study, we examine
  whether adding lidar and radar predictors to a spectral baseline improves separation of
  the two agents. On 69 survey polygons under spatially blocked cross-validation, with the
  sample, folds and decision rule fixed before any model was fitted, it does not. A
  spectral baseline reaches a balanced accuracy of 0.632; adding canopy structure lowers
  it to 0.549, a paired difference of -0.083 whose interval excludes zero; adding radar as
  well returns 0.640, indistinguishable from the baseline. The structural hypothesis is
  contradicted rather than merely unsupported, which is consistent with published evidence
  that press disturbances leave canopy structure largely unchanged. The result holds for
  trace and light damage, the only severities for which public lidar and the survey record
  coincide anywhere in British Columbia.
keywords:
  - balsam woolly adelgid
  - Adelges piceae
  - Abies amabilis
  - forest health monitoring
  - lidar
  - synthetic aperture radar
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
library(sf)
library(terra)
library(flextable)

root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
outputs <- file.path(root, "03.outputs")
derived <- file.path(inputs, "derived")

BC_ALBERS <- 3005

# Figures are written to 03.outputs/figures as well as embedded, so the README
# and any submission package reference committed files rather than extracting
# images from a rendered document.
#
# The path must be relative to this document. Pandoc resolves an image src
# against the document's directory, so an absolute path arrives as
# "./Volumes/..." which resolves to nothing, and embed-resources then drops
# every figure without failing the render. The files still land in the same
# place; only the reference changes.
knitr::opts_chunk$set(fig.path = "../03.outputs/figures/",
                      dpi = 300, dev = "png")

fmt_tbl <- function(df, right = NULL) {
  right_cols <- union(names(df)[vapply(df, is.numeric, TRUE)], right)
  left_cols <- setdiff(names(df), right_cols)
  ft <- flextable::flextable(df)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "header")
  if (length(right_cols))
    ft <- flextable::align(ft, j = right_cols, align = "right", part = "body")
  if (length(left_cols))
    ft <- flextable::align(ft, j = left_cols, align = "left", part = "body")
  ft <- flextable::line_spacing(ft, space = 1, part = "all")
  ft <- flextable::padding(ft, padding.top = 1, padding.bottom = 1,
                           part = "all")
  flextable::set_table_properties(ft, layout = "autofit", width = 1)
}
```

```{r}
#| label: load-labels
#| echo: true
#| output: false
iab <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                   quiet = TRUE)
vi <- sf::st_read(file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)

# The analysis set, built by 05.scripts/14-build-analysis-set.R: adelgid and bark
# beetle polygons on Pacific silver fir that fall under lidar coverage acquired
# within five years of the survey, on Vancouver Island. This file is the sample;
# nothing below filters it again.
aset <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
n_set <- nrow(aset)
n_set_iab <- sum(aset$PEST_SPECIES_CODE == "IAB")
n_set_ibb <- sum(aset$PEST_SPECIES_CODE == "IBB")
# Reported in the direction of the majority class, so the prose cannot name the
# wrong one if the sample is rebuilt.
set_imbalance <- round(max(n_set_iab, n_set_ibb) / min(n_set_iab, n_set_ibb), 1)
set_majority <- if (n_set_iab >= n_set_ibb) "adelgid" else "bark beetle"
set_years <- range(aset$CAPTURE_YEAR)
set_offset_med <- median(aset$year_offset)
set_offset_unit <- if (set_offset_med == 1) "year" else "years"
set_ha <- round(sum(aset$AREA_HA))
# Counts behind the landmass restriction, saved by the same script, so the
# justification for excluding the mainland is recomputable rather than quoted.
restr <- read.csv(file.path(derived, "analysis-set-restriction.csv"))
off <- restr[restr$landmass != "Vancouver Island", ]
n_off <- sum(off$n)
n_off_ibb <- sum(off$n[off$agent == "IBB"])
n_off_campaign <- max(tapply(off$n, off$lidar_year, sum))
yr_off_campaign <- names(which.max(tapply(off$n, off$lidar_year, sum)))

host_label <- c(B = "True fir, unspecified", BA = "Pacific silver fir",
                BG = "Grand fir", BL = "Subalpine fir")
agent_label <- c(IAB = "Balsam woolly adelgid",
                 IBB = "Western balsam bark beetle")
sev_label <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
               V = "Very severe")

# The severity classes actually present in the sample, named rather than
# counted, so the sentence reporting them cannot go stale or read "0 polygons".
sev_set_txt <- {
  s <- tolower(sev_label[intersect(names(sev_label),
                                   unique(aset$PEST_SEVERITY_CODE))])
  if (length(s) == 1) s else
    paste(paste(s[-length(s)], collapse = ", "), "and", s[length(s)])
}

n_iab_bc <- nrow(iab)
yr_iab <- range(iab$CAPTURE_YEAR)
n_vi_iab <- sum(vi$PEST_SPECIES_CODE == "IAB")
n_vi_ibb <- sum(vi$PEST_SPECIES_CODE == "IBB")
ha_vi_iab <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IAB"]))
ha_vi_ibb <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IBB"]))

ba <- vi[vi$TREE_SPECIES_CODE == "BA", ]
n_ba_iab <- sum(ba$PEST_SPECIES_CODE == "IAB")
n_ba_ibb <- sum(ba$PEST_SPECIES_CODE == "IBB")
imbalance <- round(n_ba_ibb / n_ba_iab, 1)

sev_iab <- table(factor(vi$PEST_SEVERITY_CODE[vi$PEST_SPECIES_CODE == "IAB"],
                        levels = names(sev_label)))
n_iab_ms <- sum(sev_iab[c("M", "S", "V")])
```

# Introduction

Balsam woolly adelgid, *Adelges piceae* (Ratzeburg), is an introduced insect that damages and kills North American true firs of all age classes, having entered the continent in the early twentieth century through separate introductions on the east and west coasts [@Campbell_2023]. Its congener the hemlock woolly adelgid, *Adelges tsugae* Annand, is the better-studied insect, but the two behave differently on this coast. In eastern North America *A. tsugae* is invasive and lethal, and infested eastern hemlock die within four to fifteen years with within-stand mortality that can exceed ninety percent [@MacQuarrie_2025]. In western North America the same species is a separate and native lineage that "exists at low densities in forests and rarely kills trees" [@MacQuarrie_2025]. Damage to western hemlock in British Columbia has accordingly been minor, and the provincial forest health survey reflects the asymmetry directly: it carries a damage-agent code for balsam woolly adelgid and none for hemlock woolly adelgid. In British Columbia the woolly adelgid of management concern is therefore *A. piceae* on true fir, and its principal host on the coast is Pacific silver fir, *Abies amabilis* (Douglas) Forbes.

The two agents of interest here attack the same host and are recorded by the same survey, so a manager reading a damage map of Pacific silver fir cannot tell from the map which insect produced it. Their damage differs in kind, however, and that difference is the basis of this study; it is set out in the study system below.

Remote detection of adelgid damage has a long record. @FRANKLIN_1995 separated damage classes on individual balsam fir from airborne multispectral imagery at 40 to 76 percent accuracy. Landsat change detection was applied to adelgid-induced hemlock decline from the late 1990s [@Royle_1997; @Bonneau_1999], and later work pushed detection towards lower infestation levels [@Williams_2017]. @Cook_2010 discriminated infested from non-infested subalpine fir at 94 percent from branch-level spectra. Most recently @Campbell_2023 mapped infestation severity in subalpine fir from satellite imagery with terrain and climate, and their result frames the present study: spectral data alone "successfully detected forest degradation" but "failed to distinguish between causal agents of change", returning a mean coefficient of determination of 0.482, while terrain and climate reached 0.746 but "lacked the ability to characterize local-scale tree damage", and only the combination reached 0.836.

Attribution has since become a distinct problem with its own methods, addressed mainly with optical time series. @Oeser_2017 attributed windthrow, cleared windthrow, bark beetle and harvest in Central Europe with overall accuracies of 76 to 86 percent; @Zhang_2022 mapped causal agents across boreal and arctic North America; @Sebald_2021 improved separation of human from natural disturbance using landscape context; and @Morin_Bernard_2023 reached 82.9 percent attributing agent before estimating severity, finding that agent-specific severity models required different predictors. Structure and radar have entered the field on the detection side. Terrestrial lidar detects *Ips typographus* infestation early [@Junttila_2019], simulated spaceborne lidar recovers the understorey and midstorey defoliation that hemlock woolly adelgid causes before any overstorey signal appears [@Boucher_2020], and L-band synthetic aperture radar delineates windthrow from insect outbreak at 69 to 88 percent [@Tanase_2018; @Hollaus_2019]. Most directly, @Atkins_2020 used terrestrial lidar canopy structural traits in a machine learning framework across six moderate-severity agents including hemlock woolly adelgid, and found that fire and ice storms changed the amount and position of canopy vegetation while insect and pathogen infestation altered canopy arrangement and complexity instead.

Two limits define what remains unresolved, and both emerged from a structured search of this literature. We searched Semantic Scholar and Crossref for work on adelgid remote sensing, on attribution of damage to specific agents, on lidar and radar applied to insect damage, and on the accuracy of the aerial surveys that supply agent labels, screening 5,663 records against criteria fixed before searching and carrying 721 forward for reading. The searched literature is deep on detection and comparatively shallow on attribution, and the two limits below are what that asymmetry exposes.

First, attribution between insect agents has been demonstrated, but only where the agents attack different hosts. @Meigs_2011 first characterised the spectral trajectories of a defoliator and a bark beetle separately against field plots, finding that western spruce budworm showed long-duration decline followed by recovery while mountain pine beetle showed both short and long-duration declines with variable recovery, but stopped short of classifying them. @Senf_2015 took that step, separating mountain pine beetle from western spruce budworm in British Columbia using LandTrendr spectral trajectories and random forests, mapping insect disturbance at 76.8 percent overall accuracy and attributing agent at "overall accuracies ranging from 75.3% to 88.0%, depending on whether only pure host-stands or mixed stands with both insect hosts were considered". That conditional is the important result for the present study. Where host composition disambiguated the agents, attribution succeeded; where it did not, it degraded sharply, and the authors report that "in the case of mixed host stands, nearly 45% of the western spruce budworm disturbances were falsely attributed to mountain pine beetle". Host is therefore doing much of the work that appears to be done by damage. Balsam woolly adelgid and western balsam bark beetle attack the same host species, which removes that cue entirely rather than diluting it, and we found no study that attributes between insect agents under that condition. Second, the structural evidence that agents leave distinguishable signatures rests on plot-scale instrumented sites; @Atkins_2020 state their findings are "limited in their generalizability due to lack of replication among disturbances". Whether canopy structure carries agent-diagnostic information at the coarse support of an operational survey, across a landscape, is untested.

That distinction matters operationally because the survey record is itself the working map of which insect is where. @Coleman_2018 assessed aerial detection surveys against ground observation across multiple ecological regions and reported overall accuracies above 70 percent, with commission errors of 2 and 5 percent for mortality and defoliation damage types, but found that "accuracy declined as the specificity for observations went from genera to species level for tree species and damaging agents". Species-level agent attribution is precisely the level at which balsam woolly adelgid and western balsam bark beetle are distinguished, and independent evaluations of survey severity estimates have returned overall accuracies as low as 35 to 56 percent depending on the class scheme [@Bright_2020b; @Housman_2018].

Each of our objectives follows from one of these findings. Because @Campbell_2023 established that spectral predictors detect adelgid degradation but fail to identify its cause, and because @Senf_2015 showed that spectral attribution between insect agents weakens when host stops disambiguating them, our first objective is to quantify how far a spectral-only classifier separates balsam woolly adelgid from western balsam bark beetle damage in Pacific silver fir on Vancouver Island, establishing that baseline rather than assuming it. Because the structural work of @Atkins_2020 and @Boucher_2020 shows that agents alter canopy arrangement in ways that colour alone does not register, but shows it at plot scale and without replication, our second objective is to test whether adding lidar-derived canopy structure and Sentinel-1 backscatter improves that separation at operational support. And because @Oeser_2017, @Sebald_2021 and @Morin_Bernard_2023 demonstrate that attribution accuracy depends heavily on how the evaluation is constructed, our third objective is to report any improvement against a validation design that respects the spatial dependence of the labels and the support at which they were drawn, rather than against a random partition that would inflate it.

We hypothesise that the multi-sensor model achieves higher balanced accuracy than the spectral baseline, because gout and progressive top-kill alter canopy geometry in a way that rapid whole-crown mortality does not, and because the spectral baseline is expected to be weak here for the reason @Senf_2015 identify: with host held constant it cannot fall back on host composition. The null hypothesis is that structural and radar predictors add nothing beyond the spectral baseline.

# Materials and methods

## Study system

Balsam woolly adelgid is a small, wingless, parthenogenetic insect that feeds on the bark of true firs through a fine stylet. In North America it reproduces without males and disperses passively, as first-instar crawlers carried by wind, birds and human movement of infested material, so spread is slow, local and irregular rather than following a front. Feeding is concentrated on the bole and in the crown, and its damage is chemically mediated rather than mechanical: saliva injected during feeding is toxic to the host, inhibiting bud formation and inducing abnormal xylem. The visible consequences are swellings at nodes and branch tips, known as gouting, whose incidence varies systematically with crown class and plant hardiness zone [@Guillet_2010], followed by crown thinning and progressive dieback from the top of the tree downward. Trees may persist in this deformed and declining condition for years, and death, where it occurs, follows a long structural decline rather than preceding it.

Western balsam bark beetle, *Dryocoetes confusus* Swaine, is the other agent recorded on the same host in this landscape. It is a phloem-feeding beetle that mass-attacks the bole, girdling the tree by disrupting phloem transport and introducing a pathogenic fungus. Its damage is not preceded by structural deformation of branches or by top-down crown thinning: an attacked tree retains its architecture and fades whole-crown, from green through yellow to red, over roughly one to two years.

The beetle does not act alone, and the comparison class is defined accordingly. On true fir it operates within a mortality complex alongside *Armillaria* root disease, the associated fungus *Ophiostoma dryocoetidis* and, at some sites, spruce beetle. Interactions of this kind alter canopy and stand characteristics jointly rather than separately: @Klutsch_2014 found on a single ponderosa pine host that "fuel bed and canopy characteristics were dependent on the presence of both dwarf mistletoe and mountain pine beetle-caused mortality". The class contrasted with the adelgid here is therefore best understood as bark-beetle-associated mortality rather than as the effect of one insect in isolation, and it is described that way throughout. This is a property of the system, not of the labels, and it would persist under any labelling scheme. The two agents therefore differ in the geometry of the damage, deformation and progressive top-kill against retained architecture, and in its tempo, years of decline against a rapid whole-crown fade.

That contrast is the mechanism this study tests. An instrument sensitive only to foliage colour observes both agents at the end state, when crowns of either kind are dead or dying and spectrally similar. An instrument sensitive to canopy geometry observes the difference in how the crown arrived there. Whether that difference survives at the coarse support of an operational survey polygon, rather than at the plot scale where it has been examined, is the open question. The plot-scale evidence is itself mixed, and both sides of it concern the congener. @Atkins_2020 found that insect and pathogen infestation altered canopy arrangement and complexity, and @Alveshere_2025 report that hemlock woolly adelgid infestation influences canopy structural complexity, while @Choi_2023 found for press disturbances including the same insect that "the impact ... on other dimensions of canopy structure could not be clearly detected, likely because of compensatory growth". A structural signal is therefore plausible but not established, and compensatory growth is a specific mechanism by which it could be absent.

Pacific silver fir, *Abies amabilis*, is the host that carries both agents on the coast. It is a shade-tolerant, late-seral true fir of the wetter and cooler coastal forests, occurring from near sea level to the subalpine, and it is a commercially and ecologically significant component of the stands in which it grows. Subalpine fir, *Abies lasiocarpa*, carries western balsam bark beetle at higher elevations in the same region but is not recorded with adelgid damage here, so the analysis is confined to Pacific silver fir in order to hold host constant.

## Study area

The study area is Vancouver Island, British Columbia, Canada. Survey records are shown over a wider envelope from 128.5 to 123.0 degrees west and 48.3 to 51.0 degrees north (@fig-studyarea), which takes in the adjacent mainland coast, but the analysis is confined to the island itself for the reason set out with the sample below. Vancouver Island is roughly 460 km long and separated from the mainland by the Strait of Georgia and Johnstone Strait. A mountain spine runs its length, rising above 2000 m in the Vancouver Island Ranges, and it divides a wet, exposed west coast from a drier eastern lowland in the rain shadow of the same mountains.

Forest composition follows that gradient. The eastern lowlands carry the Coastal Douglas-fir zone, the wetter and more extensive middle elevations the Coastal Western Hemlock zone, and the upper slopes the Mountain Hemlock zone before passing into subalpine parkland. Pacific silver fir occurs through the Coastal Western Hemlock and Mountain Hemlock zones, generally above the Coastal Douglas-fir lowlands, growing with western hemlock, western redcedar and yellow cedar. Subalpine fir is largely confined to the highest elevations. The climate is maritime, with mild wet winters and comparatively dry summers, and it moderates the winter cold that constrains adelgid survival at the northern and continental edges of the insect's range.

Both damage agents are recorded across this area by the provincial aerial overview survey, and their distributions overlap on the shared host without coinciding: the adelgid record concentrates at lower and middle elevations on Pacific silver fir, while western balsam bark beetle extends to higher elevations where subalpine fir dominates (@fig-studyarea).

All spatial data are handled in NAD83 / BC Albers, EPSG 3005, which is the projection of the source survey archive, and every map is drawn in that projection.

```{r}
#| label: fig-studyarea
#| fig-cap: "Study area on Vancouver Island and the adjacent mainland coast of British Columbia, Canada, showing aerial overview survey polygons attributed to balsam woolly adelgid and to western balsam bark beetle on true fir hosts. Relief is shaded from a 1 km digital elevation model illuminated from the northwest. Projection NAD83 / BC Albers (EPSG 3005); axis labels in degrees."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
bm <- file.path(derived, "basemap.gpkg")
land   <- sf::st_read(bm, "land",   quiet = TRUE)
admin  <- sf::st_read(bm, "admin",  quiet = TRUE)
rivers <- sf::st_read(bm, "rivers", quiet = TRUE)
lakes  <- sf::st_read(bm, "lakes",  quiet = TRUE)
places <- sf::st_read(bm, "places", quiet = TRUE)
hs <- terra::rast(file.path(derived, "hillshade.tif"))
# The view rectangle, not the extent of the land. Land and relief are prepared
# wider than the view so that the fixed aspect ratio, which expands the shorter
# axis, does not open a band of sea where the mainland belongs.
frame_box <- sf::st_read(bm, "frame", quiet = TRUE)

vi_p <- sf::st_transform(vi, BC_ALBERS)
fr <- sf::st_bbox(frame_box)

op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")

# Sea, then relief-shaded land, then hydrography, then boundaries. The sea is
# painted across the whole plot region rather than the land bounding box, since
# the fixed aspect ratio leaves the region taller than the data.
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.30, end = 1),
            legend = FALSE, add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey30", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = "#6f9ab5", lwd = 0.3,
     add = TRUE)
plot(sf::st_geometry(rivers), col = "#5b8db0", lwd = 0.8, add = TRUE)
# Only the international boundary, drawn dashed. The provincial outline is not
# drawn because within this frame it coincides with the coastline.
plot(sf::st_geometry(admin), col = "grey20", lwd = 1.1, lty = 2, add = TRUE)

# Bark beetle first: the far less numerous adelgid class must stay legible.
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IBB", ]),
     col = "#2c6f9e", border = "#2c6f9e", add = TRUE)
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IAB", ]),
     col = "#c1372f", border = "#c1372f", add = TRUE)

pxy <- sf::st_coordinates(places)
points(pxy, pch = 21, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Label side chosen per settlement so no label runs off the frame or across a
# neighbouring one.
lab_pos <- c(Comox = 4, `Campbell River` = 2, `Port Alberni` = 2,
             Nanaimo = 4, Victoria = 2, Vancouver = 4, Tofino = 2,
             `Port Hardy` = 4)
text(pxy[, 1], pxy[, 2], places$name, pos = lab_pos[places$name],
     offset = 0.4, cex = 0.6, col = "grey10")

# Axes labelled in degrees although the plot is projected, so the reader gets
# geographic coordinates without the distortion of plotting in degrees.
xt <- pretty(fr[c("xmin", "xmax")], 5)
yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

# Scale bar drawn in projected metres, so the stated length is metrically true.
sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")

# North arrow.
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.62, pt.cex = 1.1, pch = 22,
       pt.bg = c("#c1372f", "#2c6f9e"), col = c("#c1372f", "#2c6f9e"),
       legend = c("Balsam woolly adelgid", "Western balsam bark beetle"))
par(op)
```

## Data

@tbl-inputs lists each dataset and its access terms. The custodian of the survey archive is the BC Ministry of Forests; GeoBC operates the lidar programme and the two are separate agencies.

```{r}
#| label: tbl-inputs
#| tbl-cap: "Datasets used in the analysis, with their role and access terms."
inputs_tbl <- data.frame(
  Dataset = c("Pest Infestation Polygons", "Sentinel-2 L2A", "Sentinel-1 GRD",
              "LidarBC open lidar"),
  Role = c("Damage agent labels", "Optical predictors", "Radar predictors",
           "Canopy structure predictors"),
  Source = c("BC Data Catalogue", "Google Earth Engine",
             "Google Earth Engine", "LidarBC open data portal"),
  Licence = c("Open Government Licence - British Columbia",
              "Copernicus open access", "Copernicus open access",
              "Open Government Licence - British Columbia")
)
fmt_tbl(inputs_tbl)
```

Damage agents in the delivered geodatabase are recorded as three-letter codes without names. Codes were resolved against the *Forest Health Aerial Overview Survey Standards for British Columbia*, the Ministry of Forests adaptation of Canadian Forest Service FHN Report 97-1, which gives IAB for balsam woolly adelgid and IBB for western balsam bark beetle. The same document lists IAB against sub-alpine true firs, amabilis fir and grand fir, matching the host codes present in the extract. That code list contains no entry for hemlock woolly adelgid, so the absence of that species from the archive is a property of the survey design rather than of the years sampled.

### Survey data

Damage labels are polygons from the British Columbia aerial overview survey. Surveyors sketch-map visible forest damage from a fixed-wing aircraft, recording host species, damage agent and an ordinal severity class for each polygon. A minimum of eighty percent of the province is flown annually, with gaps caused by weather and smoke.

Survey polygons carry limitations that constrain the analysis and that the custodian states explicitly: errors of omission and commission occur, spatial accuracy can be poor, and the survey is not intended to be accurate enough to direct operational activities. Its stated purpose is to record general trends in disturbance size and frequency. Polygon area is in part a product of the digitising rule rather than a measurement, since a clump of one to thirty dead or dying trees is delineated as 0.25 ha and a clump of thirty-one to fifty as 0.5 ha. Severity is likewise an ordinal judgement: the standards define light, moderate and severe as 1 to 10, 11 to 29 and 30 percent or more of trees recently killed, and classify all point sources as severe on digitising. The unit of analysis here is therefore the survey polygon rather than the pixel, and severity is not modelled.

Across the study area the survey records `r n_vi_iab` adelgid polygons covering `r format(ha_vi_iab, big.mark = ",")` ha and `r n_vi_ibb` bark beetle polygons covering `r format(ha_vi_ibb, big.mark = ",")` ha on true fir hosts (@tbl-labels). Province-wide the adelgid record comprises `r n_iab_bc` polygons spanning `r yr_iab[1]` to `r yr_iab[2]`. Both agents are recorded on Pacific silver fir, `r n_ba_iab` adelgid against `r n_ba_ibb` bark beetle polygons, and it is this shared host that makes the separation an attribution problem rather than a detection problem. Analysis is restricted to that host so that agent is not confounded with differences in host reflectance.

```{r}
#| label: tbl-labels
#| tbl-cap: "Aerial overview survey polygons in the study area by damage agent and host tree species. Area is the survey-recorded polygon area."
lab <- as.data.frame(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Host  = host_label[vi$TREE_SPECIES_CODE]
), stringsAsFactors = FALSE)
names(lab)[3] <- "Polygons"
lab <- lab[lab$Polygons > 0, ]

area <- tapply(vi$AREA_HA,
               list(agent_label[vi$PEST_SPECIES_CODE],
                    host_label[vi$TREE_SPECIES_CODE]),
               function(v) round(sum(v)))
lab$`Area (ha)` <- mapply(function(a, h) area[a, h], lab$Agent, lab$Host)
lab <- lab[order(lab$Agent, -lab$Polygons), ]
fmt_tbl(lab)
```

### Analysis sample

Not every polygon on the shared host can enter the analysis, because the structural predictors require lidar and lidar coverage is neither universal nor contemporaneous with the survey. The sample is therefore defined by data availability and is stated before any result, so that its restrictions are read as a property of the design rather than as a post hoc exclusion.

A polygon enters the analysis when it is attributed to either agent, lies on Pacific silver fir, falls within LidarBC point cloud coverage, has a nearest lidar acquisition within five years of its survey year, and lies on Vancouver Island.

The last condition guards against confounding rather than saving effort. Of the `r n_off` polygons that satisfy the first four conditions but lie off the island, `r n_off_ibb` are bark beetle and `r n_off_campaign` fall under a single `r yr_off_campaign` lidar acquisition. Landmass would therefore predict the agent almost perfectly, and so would acquisition campaign, to which every structural metric used here responds through pulse density, scan angle and flight geometry. Spatially blocked validation prevents adjacent polygons leaking between folds but does nothing to stop a model learning region in place of damage, because the association holds within folds as well as across them. Region is accordingly neutralised by restriction, the same instrument that holds host constant.

The five conditions together yield `r n_set` polygons, `r n_set_iab` adelgid and `r n_set_ibb` bark beetle, covering `r format(set_ha, big.mark = ",")` ha, surveyed between `r set_years[1]` and `r set_years[2]`, with a median label-to-acquisition offset of `r set_offset_med` `r set_offset_unit`. The classes are imbalanced by `r set_imbalance` to 1 in favour of the `r set_majority`, considerably more even than the `r imbalance` to 1 of the unrestricted host-matched record.

The cost of that definition is severity, and it is severe. The only severity classes present in the analysis sample are `r sev_set_txt`. This is not an artefact of the filter thresholds. Across the province no adelgid polygon rated moderate or worse falls under lidar at a one, three, five or ten year offset, and only one does at twenty years, by which point stand condition cannot be assumed to have persisted. The moderate and severe adelgid record and the lidar programme do not overlap in British Columbia, and no choice of study area within the province changes that. The study accepts a low-severity sample in exchange for the structural arm, and the consequences for interpretation are set out in the limitations.

The severity distribution is strongly weighted towards the two lowest classes (@tbl-severity). Only `r n_iab_ms` adelgid polygons in the study area are rated moderate or worse, too few to support the severity regression used by @Campbell_2023, and a further reason the response here is the two-class agent label.

```{r}
#| label: tbl-severity
#| tbl-cap: "Severity class of survey polygons by damage agent. Severity is the surveyor's ordinal rating."
sev <- as.data.frame.matrix(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Severity = factor(sev_label[vi$PEST_SEVERITY_CODE],
                    levels = unname(sev_label))
))
sev <- sev[, colSums(sev) > 0, drop = FALSE]
sev <- cbind(Agent = rownames(sev), sev)
fmt_tbl(sev)
```

The two agents also peak in different survey years (@fig-timeseries). The adelgid record concentrates in 2019 to 2021 while the bark beetle record is distributed more evenly across the past decade. A classifier trained on imagery drawn from each polygon's own survey year would therefore risk learning the year rather than the agent, and a single common imagery epoch is used for both classes.

```{r}
#| label: fig-timeseries
#| fig-cap: "Survey polygons mapped per year on true fir hosts in the study area, by damage agent."
#| fig-width: 7
#| fig-height: 3.6
tt <- table(vi$CAPTURE_YEAR, agent_label[vi$PEST_SPECIES_CODE])
op <- par(mar = c(4, 4.2, 0.6, 0.6))
matplot(as.numeric(rownames(tt)), tt, type = "l", lty = 1, lwd = 2,
        col = c("#c1372f", "#2c6f9e"), las = 1,
        xlab = "Survey year", ylab = "Polygons mapped")
legend("topleft", colnames(tt), col = c("#c1372f", "#2c6f9e"),
       lty = 1, lwd = 2, bty = "n", cex = 0.75)
par(op)
```

### Remote sensing data

Optical predictors are drawn from Sentinel-2 Level 2A surface reflectance and radar predictors from Sentinel-1 Ground Range Detected imagery in interferometric wide swath mode, both accessed through Google Earth Engine. Scenes are restricted to the leaf-on window of the epoch in which the adelgid record concentrates, 1 June 2019 to 30 September 2021, and optical scenes to those with less than forty percent cloud cover. Scene counts available over the study area are given in @tbl-imagery.

```{r}
#| label: gee-availability
#| echo: true
#| output: false
gee_status <- tryCatch({
  library(rgee)
  ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  aoi <- ee$Geometry$Rectangle(c(-128.5, 48.3, -123.0, 51.0))

  s2 <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))

  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$eq("instrumentMode", "IW"))

  data.frame(
    Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
    Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
    Period = "2019-06-01 to 2021-09-30",
    Scenes = c(s2$size()$getInfo(), s1$size()$getInfo())
  )
}, error = function(e) {
  data.frame(Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
             Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
             Period = "2019-06-01 to 2021-09-30",
             Scenes = NA_integer_)
})
```

```{r}
#| label: tbl-imagery
#| tbl-cap: "Satellite scenes available over the study area for the analysis period."
fmt_tbl(gee_status)
```

### Lidar data

Canopy structure is derived from the LidarBC open lidar programme, which publishes classified point clouds for public download. Coverage over the study area comprises 18,840 tiles at a nominal density of eight points per square metre (@tbl-lidar), including acquisitions in 2019 that coincide with the concentration of the adelgid survey record. Tiles are classified to ground, low point and overlap but carry no vegetation classes, so canopy height is computed as height above a ground surface interpolated from ground returns rather than read from an existing classification.

```{r}
#| label: tbl-lidar
#| tbl-cap: "Lidar point cloud coverage over the study area and over the Comox field window. Acquisition years and point classes are summarised from a sample of up to 1000 tiles."
lidar <- read.csv(file.path(derived, "lidar-coverage.csv"), check.names = FALSE)
names(lidar) <- c("Extent", "Tiles", "Full index", "Acquisition years",
                  "Density (pts/m2)", "Point classes")
fmt_tbl(lidar)
```

### Field site selection

Ground confirmation requires polygons that carry both an adelgid label and lidar coverage, and that can be reached. @fig-fieldsites shows where those conditions coincide, and the answer constrains the design rather than merely illustrating it.

```{r}
#| label: fig-fieldsites
#| fig-cap: "Candidate field sites. Points are aerial overview survey polygons attributed to balsam woolly adelgid, sized by area and coloured by severity class; open circles lack lidar coverage and filled circles have it. Shaded envelopes show LidarBC point cloud coverage. Rings mark 50, 100 and 150 km from Comox and are shown for travel planning only, not as a selection criterion. Two mainland records fall outside this frame. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
# Province-wide candidates, not island-only and not filtered by distance from
# Comox: per the 2026-08-14 decision, lidar and severity availability set the
# study's scope and location rather than proximity to a field base. The frame
# below is coastal, so the two mainland interior records fall outside it and are
# reported in the text instead.
cand <- read.csv(file.path(derived, "site-candidates-province.csv"))
cand$severity <- cand$severity_code
cand <- cand[cand$lon > -128.5 & cand$lon < -123.0 &
             cand$lat > 48.3 & cand$lat < 51.2, ]
cand_sf <- sf::st_transform(
  sf::st_as_sf(cand, coords = c("lon", "lat"), crs = 4326), BC_ALBERS)
lidar_env <- sf::st_read(file.path(derived, "lidar-coverage-dissolved.gpkg"),
                         quiet = TRUE)
acc <- sf::st_read(file.path(derived, "field-access-places.gpkg"), quiet = TRUE)
comox <- sf::st_transform(
  sf::st_sfc(sf::st_point(c(-124.928, 49.673)), crs = 4326), BC_ALBERS)

fr <- sf::st_bbox(frame_box)
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.45, end = 1), legend = FALSE,
            add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey35", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = NA, add = TRUE)
plot(sf::st_geometry(rivers), col = "#7ba6c4", lwd = 0.6, add = TRUE)

# Lidar coverage envelope, drawn under the sites so the sites stay readable.
plot(sf::st_geometry(lidar_env), col = "#b9d9b0", border = "#7fae74",
     lwd = 0.3, add = TRUE)

# Distance rings from Comox. Straight line, not road distance.
for (d in c(50, 100, 150)) {
  plot(sf::st_geometry(sf::st_buffer(comox, d * 1000)), col = NA,
       border = "grey30", lty = 3, lwd = 0.8, add = TRUE)
}

sev_col <- c(T = "#f0c987", L = "#e08a45", M = "#c1372f", S = "#7d1d18",
             V = "#4a0f0c")
cxy <- sf::st_coordinates(cand_sf)
cex_by_area <- 0.5 + 1.5 * sqrt(cand$area_ha / max(cand$area_ha))
has_lidar <- cand$lidar_tiles > 0
points(cxy[!has_lidar, , drop = FALSE], pch = 1, lwd = 0.9,
       col = sev_col[cand$severity[!has_lidar]], cex = cex_by_area[!has_lidar])
points(cxy[has_lidar, , drop = FALSE], pch = 16,
       col = sev_col[cand$severity[has_lidar]], cex = cex_by_area[has_lidar])

axy <- sf::st_coordinates(acc)
points(axy, pch = 22, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Comox is drawn separately as the field base, so its label is suppressed here
# and Courtenay is thrown left to clear the diamond.
side <- ifelse(acc$name == "Courtenay", 2, 4)
keep <- acc$name != "Comox"
text(axy[keep, 1], axy[keep, 2], acc$name[keep], pos = side[keep],
     offset = 0.35, cex = 0.55, col = "grey10")
points(sf::st_coordinates(comox), pch = 23, bg = "black", col = "black",
       cex = 1.2)
text(sf::st_coordinates(comox)[1], sf::st_coordinates(comox)[2], "Comox",
     pos = 4, offset = 0.45, cex = 0.6, font = 2, col = "black")

xt <- pretty(fr[c("xmin", "xmax")], 5); yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.58,
       pch = c(16, 1, 22, 23, NA), pt.bg = c(NA, NA, "white", "black", NA),
       col = c("grey25", "grey25", "grey15", "black", "#7fae74"),
       lty = c(NA, NA, NA, NA, 1), lwd = c(NA, NA, NA, NA, 6),
       legend = c("Adelgid polygon, lidar", "Adelgid polygon, no lidar",
                  "Settlement", "Comox", "Lidar coverage"))
legend("bottomright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.58, pch = 16, col = sev_col,
       legend = c("Trace", "Light", "Moderate", "Severe"), title = "Severity")
par(op)
```

@fig-fieldsites exposes a trade-off between severity and structural data that is not resolvable by widening the search area, and it constrains the study rather than merely describing it. Across the province the survey records 267 adelgid polygons, of which 89 fall within lidar coverage and 25 are rated moderate or worse. Only nine polygons satisfy both conditions, and eight of those were surveyed in 2007 or 2008 against lidar acquired between 2023 and 2025, an offset of twelve to seventeen years over which stand condition cannot be assumed constant.

The recent record shows the trade-off at its sharpest. Twelve polygons rated moderate or severe were surveyed from 2019 onward, all of them on Pacific silver fir, and none has any lidar coverage. Ten lie on northern Vancouver Island between Sayward and the Nimpkish valley and two on the mainland coast. Conversely, 43 polygons combine lidar coverage, a label-to-acquisition offset of three years or less, and the shared host, but 37 of those are rated trace and 10 light, with a single moderate polygon among them.

Structural predictors and moderate-or-worse severity therefore cannot both be obtained from existing public data on the same polygons. Three responses are available and the choice belongs to the analysis design rather than to this section: restrict the structural arm to the low-severity polygons where lidar exists and accept that it tests the hypothesis where the signal is weakest; run the high-severity polygons on optical and radar predictors alone; or acquire structural data directly over the twelve recent high-severity polygons, which are between 4 and 24 ha and within reach of an uncrewed aerial survey. Whichever is chosen, the field campaign is a stratified sample over an accessible subset rather than over the label population, and the resulting coverage bias is reported rather than assumed away.

### Tile detail

@fig-tiles shows individual lidar tiles against survey polygon outlines over northern Vancouver Island, where the recent moderate and severe records concentrate. It is drawn at tile resolution rather than as a coverage envelope because the practical question when planning a visit is whether a specific polygon has a specific tile beneath it, which a dissolved envelope conceals.

```{r}
#| label: fig-tiles
#| fig-cap: "Lidar tiles and survey polygons over northern Vancouver Island. Grey squares are individual LidarBC point cloud tiles, shaded by acquisition year. Polygon outlines are aerial overview survey records: red for balsam woolly adelgid, blue for western balsam bark beetle, with adelgid polygons rated moderate or worse drawn heavier. Projection NAD83 / BC Albers (EPSG 3005)."
#| fig-width: 7.2
#| fig-height: 6.4
#| out-width: "100%"
det <- c(xmin = -126.4, ymin = 50.05, xmax = -124.2, ymax = 50.95)

tiles <- sf::st_read(file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)
tiles_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(tiles), det)), BC_ALBERS)
iab_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(iab), det)), BC_ALBERS)
vi_d <- sf::st_transform(
  suppressWarnings(sf::st_crop(sf::st_make_valid(vi), det)), BC_ALBERS)
land_d <- suppressWarnings(sf::st_crop(
  land, sf::st_bbox(sf::st_transform(
    sf::st_as_sfc(sf::st_bbox(det, crs = 4326)), BC_ALBERS))))
acc_d <- sf::st_read(file.path(derived, "field-access-places.gpkg"),
                     quiet = TRUE)

fr <- sf::st_bbox(sf::st_transform(sf::st_as_sfc(sf::st_bbox(det, crs = 4326)),
                                   BC_ALBERS))
op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.55, end = 1), legend = FALSE,
            add = TRUE, maxcell = 3e6)
plot(sf::st_geometry(land_d), col = NA, border = "grey35", lwd = 0.7,
     add = TRUE)

# Tiles, shaded by acquisition year so a polygon's temporal match is legible.
yrs <- sort(unique(tiles_d$year))
pal <- setNames(grey.colors(max(length(yrs), 2), start = 0.55, end = 0.88),
                yrs)
plot(sf::st_geometry(tiles_d), col = pal[as.character(tiles_d$year)],
     border = "grey60", lwd = 0.15, add = TRUE)

plot(sf::st_geometry(vi_d[vi_d$PEST_SPECIES_CODE == "IBB", ]), col = NA,
     border = "#2c6f9e", lwd = 0.7, add = TRUE)
plot(sf::st_geometry(iab_d), col = NA, border = "#c1372f", lwd = 1.0,
     add = TRUE)
hi <- iab_d[iab_d$PEST_SEVERITY_CODE %in% c("M", "S", "V"), ]
if (nrow(hi)) plot(sf::st_geometry(hi), col = "#c1372f66",
                   border = "#7d1d18", lwd = 2.0, add = TRUE)

ap <- sf::st_coordinates(acc_d)
inb <- ap[, 1] > fr["xmin"] & ap[, 1] < fr["xmax"] &
       ap[, 2] > fr["ymin"] & ap[, 2] < fr["ymax"]
if (any(inb)) {
  points(ap[inb, , drop = FALSE], pch = 22, bg = "white", col = "grey15",
         cex = 0.9, lwd = 0.8)
  text(ap[inb, 1], ap[inb, 2], acc_d$name[inb], pos = 4, offset = 0.4,
       cex = 0.62, col = "grey10")
}

xt <- pretty(fr[c("xmin", "xmax")], 4); yt <- pretty(fr[c("ymin", "ymax")], 4)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

sb <- 10000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.010 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.8 * h, c("0", "10", "20 km"),
     cex = 0.6, col = "grey15")
nx <- fr["xmax"] - 0.06 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.08 * diff(fr[c("ymin", "ymax")])
a <- 0.045 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.5 * a, "N", cex = 0.8, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55", cex = 0.6,
       lty = c(1, 1, 1, NA), lwd = c(2.0, 1.0, 0.7, NA), pch = c(NA, NA, NA, 22),
       pt.bg = c(NA, NA, NA, "grey75"),
       col = c("#7d1d18", "#c1372f", "#2c6f9e", "grey60"),
       legend = c("Adelgid, moderate or worse", "Adelgid, other severity",
                  "Bark beetle", "Lidar tile"))
par(op)
```

## Data preparation

Terrain governs every preparation choice here. The study area is mountainous, the survey polygons sit on slopes rather than on flat ground, and both optical and radar measurements are sensitive to illumination and viewing geometry in ways that vary systematically with aspect and slope. An uncorrected terrain effect would enter the models as a predictor, and because the two agents occupy partly different elevations it could be learned in place of the damage signal.

Optical predictors are built from Sentinel-2 Level 2A scenes over the leaf-on window of the label epoch. Cloud, cloud shadow, snow and water are masked from the scene classification layer before compositing, and a per-pixel median composite is formed to suppress residual cloud edges and to reduce the influence of any single acquisition geometry. From the composite we derive the visible, red edge, near infrared and shortwave infrared bands and a set of indices selected for sensitivity to canopy water content and to the visible and shortwave changes @Campbell_2023 associated with adelgid infestation. A topographic illumination correction is applied before indices are computed, since a slope-driven brightness difference would otherwise propagate into every ratio.

Radar predictors are built from Sentinel-1 Ground Range Detected scenes in interferometric wide swath mode, using both VV and VH polarisations. Processing follows the standard sequence of orbit refinement, border and thermal noise removal, radiometric calibration, terrain flattening and Range Doppler terrain correction, with the flattening step performed against a digital elevation model so that backscatter is expressed on a terrain-corrected reference area rather than on the ellipsoid. Areas of layover and shadow are masked and excluded, which removes some steep terrain from the analysis and is reported as a reduction in the usable sample rather than ignored. Speckle is suppressed by temporal averaging across the scene stack, and the mean and the temporal dispersion of backscatter are retained separately, the latter because a changing crown may be distinguishable by its variability as well as its mean.

Structural predictors are derived from LidarBC point clouds. Ground returns are used to interpolate a terrain surface, point heights are normalised above it, and canopy metrics are computed from the normalised cloud rather than from any supplied vegetation classification, since the delivered tiles carry no vegetation classes.

The metric set is chosen against the mechanism and against what the prior literature found, rather than against convention. A generic canopy description would test whether the two agents differ in any structural respect, which is weaker and less falsifiable than the question asked here.

Two cautions govern the choice, and both come from work on the congener, since no structural study of balsam woolly adelgid exists. The first is that the change to be detected is small. @Pontius_2005 and @Hanavan_2015 record fine twig dieback rising from 9 to 15 percent and live crown ratio falling from 67 to 56 percent across a decade of hemlock decline, so a metric set tuned to gross canopy loss will not see it. The second is that the change is not where a canopy-height study would look for it. @Boucher_2020 simulated waveforms over two airborne acquisitions at Harvard Forest and found that "changes in midstory plant area (PAI 11 to 12 m above ground) and overall canopy permeability (indicated by RH10) accounted for 60% of the variation in hemlock mortality", reasoning that the insect "often causes its host species... to defoliate in the midstory and understory before showing impacts in the overstory" and that active sensing can "penetrate canopy gaps and record lower canopy structural changes". @Garris_2019 place the signal in the same stratum from the optical side, reporting that adelgid presence changes lower canopy leaf optical properties.

That evidence points away from the overstorey, and an earlier version of this metric set was weighted towards it. A fourth family is therefore included and the hypothesis is stated in terms of it: if adelgid damage in *Abies amabilis* proceeds as it does in *Tsuga canadensis*, defoliating from beneath before the overstorey registers it, then the separating information sits in the midstorey and in how far pulses penetrate, not in canopy height.

Metrics therefore fall into four families, each tied to a prediction. The first tests vertical redistribution: height percentiles at the 25th, 50th, 75th and 95th centiles, the ratio of the 25th to the 95th, canopy relief ratio, and the proportion of returns above two thirds of the canopy top. The second tests surface roughness: rumple, being the ratio of canopy surface area to ground area, the standard deviation of the canopy height model within the polygon, and the coefficient of variation of first-return heights. The third describes density and openness, being canopy cover above 2 m, gap fraction and the vertical complexity index. The fourth describes the lower profile, being canopy permeability as the height below which a tenth of all returns fall, which is the discrete-return analogue of the RH10 that @Boucher_2020 used, together with the proportion of returns in a midstorey stratum from 5 to 15 m, which spans the 11 to 12 m band in which that study located the signal, and the proportion below 5 m.

The third family is a control rather than a hypothesis. Both agents thin the canopy, so if density and openness alone separate them the structural result concerns damage intensity rather than damage geometry, and it is reported in those terms.

Statistical outlier removal is not applied, and because it is standard practice the deviation was measured rather than argued. Both arms were run through one code path on the three densest small polygons in the sample, where the even-density assumption the filter rests on comes closest to being met (@tbl-sor). Against the canopy metrics the filter changes almost nothing: the 99th percentile of height moves by 0.04 to 0.05 m, the 95th by at most 0.07 m, and rumple and canopy height model dispersion by under half a percent. It does what it is advertised to do at the extreme, pulling a raw maximum of 195.1 m down to 114.1 m on one polygon, but the metric set never uses the raw maximum for that reason and takes the 99th percentile instead.

The filter is declined because of what it removes and where. Of the 115,808 to 236,740 returns it flags per polygon, between 205 and 968 lie above the 95th percentile of height, and 84,706 to 205,479 lie inside the canopy between 2 m and that percentile. It is therefore deleting canopy returns rather than points floating above the canopy, by a factor of roughly a hundred to one, which follows from a global neighbour-distance statistic applied to a cloud whose delivered density spans two orders of magnitude. The consequence falls on exactly the variables the hypothesis now rests on: canopy permeability rises by up to 0.8 m and the midstorey return fraction falls by up to 8.6 percent, both in the direction of erasing the penetration signal that @Boucher_2020 identified, while every overstorey metric is unmoved. Applying the filter would cost 13 hours of computation to alter the lower-profile variables in a direction that cannot be verified without ground measurement, so it is not applied and this paragraph is the record of that choice.

Two limits on this design are stated here rather than discovered later. Airborne lidar at any operational density cannot resolve gout, which deforms branches at the scale of centimetres; @Atkins_2020 characterised crown architecture at that scale with terrestrial lidar, not airborne. The roughness family therefore tests metre-scale crown surface irregularity and is a coarse proxy for the deformation mechanism, not a measurement of it. And the structural hypothesis has published evidence against it: @Choi_2023 examined six moderate-severity agents with repeat lidar and found that the impact of press disturbances on canopy structure "could not be clearly detected, likely because of compensatory growth", while pulse disturbances were more readily apparent. Adelgid damage is a press disturbance and bark beetle mortality is closer to a pulse, so the design is being run against a documented expectation of a weak or absent structural signal for the target class.

```{r}
#| label: tbl-sor
#| tbl-cap: "Effect of statistical outlier removal on the structural metrics, measured on the three densest small polygons. Overstorey metrics are unmoved; the lower-profile metrics shift, and almost everything the filter removes lies inside the canopy rather than above it."
local({
  d <- read.csv(file.path(derived, "sor-check.csv"), stringsAsFactors = FALSE)
  w <- function(v, a) d[[v]][d$arm == a]
  fmt_tbl(data.frame(
    Quantity = c("Raw maximum height (m)", "99th percentile height (m)",
                 "95th percentile height (m)", "Canopy permeability rh10 (m)",
                 "Midstorey return fraction", "Rumple",
                 "Returns flagged", "Of those, above the 95th percentile",
                 "Of those, inside the canopy"),
    `As delivered` = c(
      paste(w("z_max", "as delivered"), collapse = ", "),
      paste(w("ztop_p99", "as delivered"), collapse = ", "),
      paste(w("zq95", "as delivered"), collapse = ", "),
      paste(w("rh10", "as delivered"), collapse = ", "),
      paste(w("p_mid", "as delivered"), collapse = ", "),
      paste(w("rumple", "as delivered"), collapse = ", "), "", "", ""),
    `After the filter` = c(
      paste(w("z_max", "after SOR"), collapse = ", "),
      paste(w("ztop_p99", "after SOR"), collapse = ", "),
      paste(w("zq95", "after SOR"), collapse = ", "),
      paste(w("rh10", "after SOR"), collapse = ", "),
      paste(w("p_mid", "after SOR"), collapse = ", "),
      paste(w("rumple", "after SOR"), collapse = ", "),
      paste(format(w("flagged_n", "after SOR"), big.mark = ","), collapse = ", "),
      paste(w("flagged_above_zq95", "after SOR"), collapse = ", "),
      paste(format(w("flagged_in_canopy", "after SOR"), big.mark = ","),
            collapse = ", ")),
    check.names = FALSE))
})
```

All three families are summarised to the survey polygon, which is the unit of analysis. Polygons are eroded by an inward buffer before summarisation, because sketch-mapped boundaries are approximate and an unbuffered polygon edge will include unaffected canopy. Within the eroded polygon each predictor is reduced to its mean and its dispersion, so that a heterogeneous polygon is not represented solely by a central value. Polygons too small to survive erosion are excluded and counted.

The specific index set, the lidar metric set and the buffer distance are fixed at pre-registration, before any predictor value is joined to any label, and are reported there with the reasoning for each threshold.

## Random forests

The response is the two-class agent label, adelgid against bark beetle, for survey polygons on Pacific silver fir. Restricting to a single host is what makes the comparison an attribution test: with host held constant, a classifier cannot succeed by learning host reflectance or host site preference in place of the damage itself.

Predictors are grouped into three families, spectral, structural and radar, and three nested models are compared: spectral only, spectral with structure, and all three families together. The nesting is deliberate, since the question is not whether the agents can be separated at all but how much of any separation the spectral baseline already provides. Survey year and terrain are carried as covariates in every model so that their contribution is estimated rather than absorbed into the sensor terms; the two agents peak in different years (@fig-timeseries) and occupy partly different elevations, and both are plausible confounders.

Models are fitted with random forests, following @Campbell_2023, and compared at equal complexity, with hyperparameters tuned only on training-internal splits so that no evaluation fold informs model selection. Random forests are chosen for their tolerance of correlated predictors, which the three families supply in quantity, and because variable importance grouped by family gives a direct reading of where separating information lies.

Because polygons cluster in space, performance is estimated by spatially blocked cross-validation rather than random partitioning, which would place spatially adjacent and often spatially overlapping polygons in both training and evaluation folds and inflate accuracy. On a sample of `r n_set` polygons the blocking scheme is the design choice most likely to determine the result, so it is fixed in advance and not tuned afterwards.

Blocks are contiguous groups formed by clustering polygon centroids, sized so that no block holds fewer than eight polygons and no fold fewer than two blocks, which yields approximately five folds at this sample size. Block size is set from the empirical range of spatial autocorrelation in the predictors, estimated on the predictors alone before any label is joined, so that the choice cannot be informed by the outcome. Where that range would leave fewer than five folds the design accepts fewer folds rather than smaller blocks, since leakage inflates accuracy while a low fold count only widens the interval. Polygons that overlap or share a boundary are forced into the same block irrespective of the clustering, because the survey digitises adjacent damage as separate records and treating them as independent would leak directly. The realised number of folds and block sizes are reported with every accuracy figure. The two classes are imbalanced by `r set_imbalance` to 1 in favour of the `r set_majority` class, and resampling is applied within training folds only, never to the evaluation folds, so that reported performance refers to the observed class prevalence. With `r n_set` polygons the sample is small for a spatially blocked design, and the number of blocks is reported alongside every accuracy figure.

## Accuracy assessment

The primary metric is balanced accuracy, the mean of sensitivity and specificity. It is chosen over overall accuracy because the classes are imbalanced: a model assigning every polygon to the `r set_majority` class would score respectably on overall accuracy while separating nothing, which is the failure mode the study exists to avoid.

Per-class sensitivity and specificity and the full confusion matrix are reported alongside it, each with the support on which it is computed. The comparison of interest is paired, the difference in balanced accuracy between the multi-sensor and spectral-only models evaluated on identical folds, so that fold-to-fold variation is differenced out rather than treated as independent error, with its interval obtained by bootstrapping over folds. The hypothesis is confirmed if and only if that difference is positive and its 95 percent interval excludes zero, disconfirmed if the interval excludes zero in the negative direction, and inconclusive if it spans zero.

Accuracy figures reported here refer to agreement with the surveyor's attribution, not with an independent ground determination. They therefore measure reproduction of the operational record and inherit whatever error that record carries, which @Coleman_2018 show increases as agent specificity increases. The metric implementation is verified against cases whose values are known by construction (@tbl-checks).

```{r}
#| label: eval-metric
#| echo: true
#| output: false
balanced_accuracy <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}
```

```{r}
#| label: tbl-checks
#| tbl-cap: "Verification of the balanced accuracy implementation against cases with analytically known values."
t_perfect <- c("IAB", "IAB", "IBB", "IBB")
t_inverted <- c("IBB", "IBB", "IAB", "IAB")
imb_truth <- c("IAB", "IBB", "IBB", "IBB")

checks <- data.frame(
  Case = c("Perfect prediction",
           "Fully inverted prediction",
           "All assigned to majority class",
           "Imbalanced, all majority (overall accuracy 0.75)"),
  Expected = c(1, 0, 0.5, 0.5),
  Observed = c(
    balanced_accuracy(t_perfect, t_perfect),
    balanced_accuracy(t_perfect, t_inverted),
    balanced_accuracy(t_perfect, rep("IBB", 4)),
    balanced_accuracy(imb_truth, rep("IBB", 4))
  )
)
checks$Agrees <- abs(checks$Expected - checks$Observed) < 1e-9
fmt_tbl(checks)
```

```{r}
#| label: results-inputs
#| echo: true
#| output: false
# Everything the results below report, read from the tables the pipeline wrote.
# No predictor value is joined to any label here: the only quantity crossed with
# the agent is whether a polygon carries lidar returns, which is an inclusion
# criterion of the same kind as host and landmass, not an outcome.
suppressMessages(library(lidR))
opt <- read.csv(file.path(derived, "predictors-optical.csv"))
rad <- read.csv(file.path(derived, "predictors-radar.csv"))
str_tab <- read.csv(file.path(derived, "predictors-structural.csv"))
opt_log <- read.csv(file.path(derived, "predictors-optical-log.csv"))
rad_log <- read.csv(file.path(derived, "predictors-radar-log.csv"))
# The structural log is absent while the metric set is being rebuilt, so the
# chunk reads it if present and falls back to the table itself. A missing
# parameter log should not stop the document rendering.
str_log_path <- file.path(derived, "predictors-structural-log.csv")
str_log <- if (file.exists(str_log_path)) read.csv(str_log_path) else
  data.frame(parameter = character(), value = character())
dropped <- read.csv(file.path(derived, "structural-dropped.csv"))
tiles_need <- read.csv(file.path(derived, "lidar-tiles-needed.csv"))
flat <- read.csv(file.path(derived, "radar-flattening-check.csv"))

lg <- function(tbl, p) tbl$value[tbl$parameter == p]

# The positional identifier is the same one every predictor block derives, so
# the frozen sample can be matched to the structural table row for row.
aset_pid <- sprintf("P%03d", seq_len(nrow(aset)))
has_struct <- aset_pid %in% str_tab$poly_id

n_struct <- nrow(str_tab)
# Derived from the two counts rather than read from the drop file, so the
# column cannot disagree with its own total if that file is mid-rebuild.
n_dropped_struct <- n_set - n_struct
n_struct_iab <- sum(aset$PEST_SPECIES_CODE[has_struct] == "IAB")
n_struct_ibb <- sum(aset$PEST_SPECIES_CODE[has_struct] == "IBB")
n_drop_iab <- sum(aset$PEST_SPECIES_CODE[!has_struct] == "IAB")
n_drop_ibb <- sum(aset$PEST_SPECIES_CODE[!has_struct] == "IBB")
struct_imbalance <- round(max(n_struct_iab, n_struct_ibb) /
                          min(n_struct_iab, n_struct_ibb), 2)
drop_year <- paste(sort(unique(aset$lidar_year[!has_struct])), collapse = " and ")

opt_scenes <- as.integer(lg(opt_log, "scenes_matched"))
opt_corrected <- as.integer(lg(opt_log, "scenes_corrected"))
rad_scenes <- as.integer(lg(rad_log, "scenes_matched"))
rad_asc <- as.integer(lg(rad_log, "scenes_ascending"))
rad_desc <- as.integer(lg(rad_log, "scenes_descending"))
opt_scenes_txt <- format(opt_scenes, big.mark = ",")
opt_corr_txt <- format(opt_corrected, big.mark = ",")
rad_scenes_txt <- format(rad_scenes, big.mark = ",")
rad_ret_min <- sprintf("%.3f", min(rad$retained_frac))
rad_ret_med <- sprintf("%.3f", median(rad$retained_frac))

n_tiles_used <- nrow(tiles_need)
tile_gb <- round(sum(as.numeric(tiles_need$bytes)) / 1e9, 1)
lgv <- function(p, alt) { v <- lg(str_log, p); if (length(v)) v else alt }
# Counted from the tile headers rather than from the log, which records the
# delivered volume in bytes.
tile_pts <- sum(vapply(file.path(inputs, "lidar", tiles_need$year,
                                 tiles_need$filename),
                       function(f) as.numeric(
                         lidR::readLASheader(f)@PHB[["Number of point records"]]),
                       numeric(1)))
tile_pts_txt <- paste(round(tile_pts / 1e9, 1), "billion")

# Terrain flattening, measured rather than asserted.
flat_r_sigma <- sprintf("%.3f", cor(flat$VV_sigma0, flat$alpha_r_deg))
flat_r_gamma <- sprintf("%.3f", cor(flat$VV_flat, flat$alpha_r_deg))
flat_r_sigma_slope <- sprintf("%.3f", cor(flat$VV_sigma0, flat$slope_deg))
flat_r_gamma_slope <- sprintf("%.3f", cor(flat$VV_flat, flat$slope_deg))
flat_sd_sigma <- sprintf("%.2f", sd(flat$VV_sigma0))
flat_sd_gamma <- sprintf("%.2f", sd(flat$VV_flat))

adv_dens <- paste(unique(tiles_need$density), collapse = ", ")
dens_tile_min <- sprintf("%.1f", min(tiles_need$density_actual))
dens_tile_max <- sprintf("%.1f", max(tiles_need$density_actual))
dens_tile_med <- sprintf("%.1f", median(tiles_need$density_actual))

target_dens <- suppressWarnings(as.numeric(lg(str_log, "target_density_pts_m2")))
if (!length(target_dens) || is.na(target_dens))
  target_dens <- round(max(str_tab$dens_used))
n_at_target <- sum(str_tab$dens_used >= target_dens * 0.94)
n_below_target <- n_struct - n_at_target
n_below_four <- sum(str_tab$dens_used < 4)
n_below_two <- sum(str_tab$dens_used < 2)
gap_dens_r <- sprintf("%.2f", cor(str_tab$dens_used, str_tab$gap_frac))

yrs <- sort(unique(tiles_need$year))
dens_tab <- data.frame(
  Year = yrs,
  Tiles = as.integer(tapply(tiles_need$density_actual, tiles_need$year,
                            length)[as.character(yrs)]),
  `Advertised` = as.integer(tapply(tiles_need$density, tiles_need$year,
                                   median)[as.character(yrs)]),
  `Delivered median` = round(as.numeric(tapply(tiles_need$density_actual,
                                               tiles_need$year,
                                               median)[as.character(yrs)]), 1),
  `Delivered range` = vapply(yrs, function(y) {
    v <- tiles_need$density_actual[tiles_need$year == y]
    sprintf("%.1f to %.1f", min(v), max(v)) }, ""),
  `Achieved median` = round(as.numeric(tapply(str_tab$dens_used,
                                              str_tab$lidar_year,
                                              median)[as.character(yrs)]), 2),
  check.names = FALSE)
```

# Results

The results below divide at the pre-registration gate. Everything in the first four sections concerns the construction of the sample and the predictors and is reported here because it is established, reproducible and, in two cases, consequential for anyone attempting this design elsewhere. The model comparison that the study exists to make lies beyond that gate: no predictor value has been joined to any label, and the tables that will carry it are pre-registered shells whose cells read pending until the analysis is registered and run.

## Sample realised

The frozen sample of `r n_set` polygons was defined by intersecting the survey record with the LidarBC point cloud index. That index is a footprint layer, and its footprints claim ground the delivered point clouds do not contain. Reading the clouds themselves shows that `r n_struct` of the `r n_set` polygons hold lidar returns and `r n_dropped_struct` hold none at all, a result confirmed by reading every one of the empty polygons in full rather than by sampling (@tbl-coverage).

The loss is not concentrated in one agent. It falls `r n_drop_iab` to `r n_drop_ibb` between adelgid and bark beetle, so the class ratio moves from `r set_imbalance` to 1 to `r struct_imbalance` to 1 and the sample available to the structural arm becomes `r n_struct_iab` adelgid against `r n_struct_ibb` bark beetle. Every empty polygon was matched to a `r drop_year` acquisition; the later campaigns are covered without exception.

```{r}
#| label: tbl-coverage
#| tbl-cap: "Polygons of the frozen sample that carry lidar returns, against those the tile index claimed but the point clouds do not cover."
cov_tab <- data.frame(
  Class = c("Balsam woolly adelgid", "Western balsam bark beetle", "Total"),
  `Frozen sample` = c(n_set_iab, n_set_ibb, n_set),
  `With returns` = c(n_struct_iab, n_struct_ibb, n_struct),
  `No returns` = c(n_drop_iab, n_drop_ibb, n_dropped_struct),
  check.names = FALSE)
fmt_tbl(cov_tab)
```

## Predictor sets

**Rebuild in progress.** The structural figures in this section and the next describe the metric set as first built, twelve variables at a common density of 8 points per square metre. Amendment 1 to the pre-registration, dated 2026-08-15 and made before any model was fitted, adds canopy permeability and the midstorey and understorey return fractions and raises the common density to 16, following @Boucher_2020. That rebuild is running and these numbers will be restated from it. The methods above describe the amended set; the structural counts below do not yet reflect it.

All three predictor families are built and none has a missing value (@tbl-predictors). The optical set summarises a median composite of `r opt_scenes_txt` Sentinel-2 scenes, of which `r opt_corr_txt` carried enough valid pixels over the sample to support the per-scene topographic correction. The radar set summarises `r rad_scenes_txt` Sentinel-1 scenes, `r rad_asc` ascending and `r rad_desc` descending. The structural set draws on `r n_tiles_used` lidar tiles holding `r tile_pts_txt` returns, `r tile_gb` GB as delivered.

Layover and shadow cost almost nothing at these viewing geometries. The retained fraction of each polygon after masking runs from `r rad_ret_min` to 1.000 with a median of `r rad_ret_med`, so the reduction the method promises to report is real but negligible, and no polygon was lost to it.

```{r}
#| label: tbl-predictors
#| tbl-cap: "The three predictor families as built, with the imagery and point cloud volumes behind them."
pred_tab <- data.frame(
  Family = c("Spectral", "Radar", "Structural"),
  Source = c("Sentinel-2 L2A", "Sentinel-1 GRD", "LidarBC point clouds"),
  Polygons = c(nrow(opt), nrow(rad), nrow(str_tab)),
  Variables = c(36L, 8L, 14L),
  Observations = c(paste(format(opt_scenes, big.mark = ","), "scenes"),
                   paste(format(rad_scenes, big.mark = ","), "scenes"),
                   paste(n_tiles_used, "tiles")))
fmt_tbl(pred_tab)
```

## Terrain correction

Over mountainous ground the uncorrected radar product is very largely a picture of the terrain rather than of the canopy. Across the sample the correlation between backscatter and the slope in the radar range direction, which is the geometry that governs the scattering area, is `r flat_r_sigma` before radiometric flattening and `r flat_r_gamma` after it, while the spread between polygons falls from `r flat_sd_sigma` dB to `r flat_sd_gamma` dB (@tbl-flattening). Rather more than half of the between-polygon variance in the uncorrected product is therefore geometry, and the correction removes it.

Flattening does not neutralise terrain entirely, and the residue runs in a direction worth stating. Uncorrected backscatter carries no association with slope itself, `r flat_r_sigma_slope`, because slopes tilted toward and away from the sensor cancel when aspect is not held fixed. The flattened product carries `r flat_r_gamma_slope`. This is a property of the volumetric model rather than an error in its application: the tangent ratio penalises a slope tilted toward the radar more than it rewards one tilted away, so the net effect grows with steepness whatever the aspect. Terrain is accordingly carried as a covariate in every model, and it is doing work rather than serving as insurance.

```{r}
#| label: tbl-flattening
#| tbl-cap: "Association between Sentinel-1 backscatter and terrain geometry, before and after radiometric terrain flattening."
flat_tab <- data.frame(
  Quantity = c("Correlation with range slope",
               "Correlation with slope",
               "Between-polygon standard deviation (dB)"),
  `Uncorrected` = c(flat_r_sigma, flat_r_sigma_slope, flat_sd_sigma),
  `Flattened` = c(flat_r_gamma, flat_r_gamma_slope, flat_sd_gamma),
  check.names = FALSE)
fmt_tbl(flat_tab)
```

## Point density

The point cloud index advertises `r adv_dens` points per square metre for every tile in the province. The delivered density over this sample runs from `r dens_tile_min` to `r dens_tile_max`, with a median of `r dens_tile_med`, and it is stratified by acquisition (@tbl-density). Because every structural metric responds to density, and because the campaigns are unevenly distributed across the sample, a model given the metrics as delivered could separate the agents by reading which aircraft flew them.

Thinning to a common density is the control for that, and it is a partial control rather than a complete one, because thinning can discard points but cannot create them. Of the `r n_struct` polygons with returns, `r n_at_target` reach the `r target_dens` points per square metre target and `r n_below_target` fall short, `r n_below_four` of them below four and `r n_below_two` below two. The achieved density is therefore bimodal rather than common.

That shortfall reaches the metrics. Gap fraction correlates with achieved density at `r gap_dens_r` across the polygons with returns, because the rule that an empty canopy cell marks an opening holds where a one metre cell expects several returns and fails where it expects less than one, in which case an empty cell records only that no pulse was fired there. The polygons carrying the highest gap fractions are all in the sparse tail. An inclusion floor on achieved density is therefore required before the control family can be trusted, and the threshold belongs in the pre-registration rather than here.

```{r}
#| label: tbl-density
#| tbl-cap: "Delivered point density by acquisition year against the density advertised for every tile, and the density achieved after thinning."
fmt_tbl(dens_tab)
```

## Model comparison

This is the only block in the document that joins a predictor value to a label. Everything that governs it was fixed before it ran: the hypotheses, the three nested predictor sets, the hyperparameters, the fold assignment, the interval method and the decision rule are in the pre-registration frozen at commit `5d69fc2` and amended at `8f8ca60`, and the fold structure was written by `pipe-blocks` from geometry alone.

```{r}
#| label: pipe-fit
#| echo: true
#| output: false
fit_path <- file.path(derived, "model-results.csv")
folds_path <- file.path(derived, "model-folds.csv")
BOOT_DRAWS <- 10000L

if (!file.exists(fit_path)) local({
  library(ranger)
  b <- read.csv(blocks_path, stringsAsFactors = FALSE)
  opt <- read.csv(opt_path, stringsAsFactors = FALSE)
  rad <- read.csv(rad_path, stringsAsFactors = FALSE)
  st <- read.csv(struct_path, stringsAsFactors = FALSE)
  flat <- read.csv(file.path(derived, "radar-flattening-check.csv"),
                   stringsAsFactors = FALSE)

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  lab <- data.frame(poly_id = sprintf("P%03d", seq_len(nrow(a))),
                    y = a$PEST_SPECIES_CODE,
                    survey_year = a$CAPTURE_YEAR, stringsAsFactors = FALSE)
  rm(a)

  # Terrain covariate from the flattening check, which already carries mean
  # polygon slope from the same digital elevation model the radar used.
  slope <- flat[, c("poly_id", "slope_deg")]

  # The predictor sets, exactly as the pre-registration lists them.
  spec_v <- grep("_(mean|stdDev)$", names(opt), value = TRUE)
  rad_v <- grep("_(mean|stdDev)$", names(rad), value = TRUE)
  struct_v <- c("zq25", "zq50", "zq75", "zq95", "zq25_zq95", "crr",
                "above_two_thirds", "rumple", "chm_sd", "cv_first",
                "cover", "gap_frac", "vci", "rh10", "p_mid", "p_under")
  cov_v <- c("survey_year", "slope_deg", "dens_used")

  d <- merge(b[, c("poly_id", "fold")], lab, by = "poly_id")
  d <- merge(d, slope, by = "poly_id")
  d <- merge(d, opt[, c("poly_id", spec_v)], by = "poly_id")
  d <- merge(d, rad[, c("poly_id", rad_v)], by = "poly_id")
  d <- merge(d, st[, c("poly_id", struct_v, "dens_used")], by = "poly_id")
  stopifnot(nrow(d) == nrow(b), !anyNA(d))

  sets <- list(spectral = c(spec_v, cov_v),
               structure = c(spec_v, struct_v, cov_v),
               all = c(spec_v, struct_v, rad_v, cov_v))

  ba <- run_folds(d, sets, d$fold)

  # Per-fold values are kept, because the interval is taken over folds and a
  # reader should be able to see the spread the interval is built from.
  fold_tab <- do.call(rbind, lapply(names(ba), function(nm)
    data.frame(model = nm, fold = sort(unique(d$fold)),
               balanced_accuracy = round(ba[[nm]], 4),
               stringsAsFactors = FALSE)))
  write.csv(fold_tab, folds_path, row.names = FALSE)

  # Bias-corrected percentile bootstrap over folds, the interval named in the
  # pre-registration. With five folds it is wide by construction, which is the
  # honest consequence of a spatially blocked design on this sample.
  set.seed(SEED)
  bc_ci <- function(v) {
    v <- v[is.finite(v)]
    if (length(v) < 2) return(c(NA_real_, NA_real_))
    th <- mean(v)
    bs <- replicate(BOOT_DRAWS, mean(sample(v, length(v), replace = TRUE)))
    z0 <- stats::qnorm(mean(bs < th))
    if (!is.finite(z0)) z0 <- 0
    p <- stats::pnorm(2 * z0 + stats::qnorm(c(0.025, 0.975)))
    unname(stats::quantile(bs, p, na.rm = TRUE))
  }

  rows <- lapply(names(sets), function(nm) {
    dif <- ba[[nm]] - ba$spectral
    ci <- if (nm == "spectral") c(NA_real_, NA_real_) else bc_ci(dif)
    # Two sided p from the bootstrap, by the proportion of draws on the other
    # side of zero, doubled and bounded.
    # Folds where balanced accuracy is undefined are dropped here exactly as
    # the interval drops them, so the two cannot disagree.
    dfin <- dif[is.finite(dif)]
    p <- if (nm == "spectral" || length(dfin) < 2) NA_real_ else {
      bs <- replicate(BOOT_DRAWS, mean(sample(dfin, length(dfin), replace = TRUE)))
      min(1, 2 * min(mean(bs <= 0), mean(bs >= 0)))
    }
    data.frame(model = nm,
               balanced_accuracy = round(mean(ba[[nm]], na.rm = TRUE), 4),
               ba_sd = round(stats::sd(ba[[nm]], na.rm = TRUE), 4),
               diff_from_spectral = round(mean(dif, na.rm = TRUE), 4),
               ci_low = round(ci[1], 4), ci_high = round(ci[2], 4),
               p_raw = round(p, 4), stringsAsFactors = FALSE)
  })
  res <- do.call(rbind, rows)
  res$folds_used <- sum(is.finite(ba$spectral))

  # Secondary, and reported as such: every polygon predicted once from a model
  # that never saw it, pooled into a single confusion matrix. This uses the
  # units the single-class fold puts beyond the reach of the per-fold statistic.
  pooled <- lapply(names(sets), function(nm) {
    pr <- oof_predict(d, sets[[nm]], d$fold)
    ok <- !is.na(pr)
    tp <- sum(d$y[ok] == "IAB" & pr[ok] == "IAB")
    fn <- sum(d$y[ok] == "IAB" & pr[ok] != "IAB")
    tn <- sum(d$y[ok] != "IAB" & pr[ok] != "IAB")
    fp <- sum(d$y[ok] != "IAB" & pr[ok] == "IAB")
    data.frame(model = nm, n_predicted = sum(ok),
               sensitivity = round(tp / (tp + fn), 4),
               specificity = round(tn / (tn + fp), 4),
               balanced_accuracy = round(((tp / (tp + fn)) + (tn / (tn + fp))) / 2, 4),
               tp = tp, fn = fn, tn = tn, fp = fp, stringsAsFactors = FALSE)
  })
  write.csv(do.call(rbind, pooled),
            file.path(derived, "model-pooled.csv"), row.names = FALSE)
  # Holm across the two confirmatory tests, as pre-registered.
  k <- res$model != "spectral"
  res$p_holm <- NA_real_
  res$p_holm[k] <- round(stats::p.adjust(res$p_raw[k], method = "holm"), 4)
  res$verdict <- ifelse(res$model == "spectral", "baseline",
    ifelse(is.na(res$ci_low), "not estimable",
      ifelse(res$ci_low > 0, "confirmed",
        ifelse(res$ci_high < 0, "disconfirmed", "inconclusive"))))
  write.csv(res, fit_path, row.names = FALSE)
})
```

```{r}
#| label: model-inputs
#| echo: true
#| output: false
# Paths are built here rather than borrowed from the pipeline section, which
# appears later in the document than this one and has not run yet.
mres <- read.csv(file.path(derived, "model-results.csv"), stringsAsFactors = FALSE)
mpool <- read.csv(file.path(derived, "model-pooled.csv"), stringsAsFactors = FALSE)
mblk <- read.csv(file.path(derived, "spatial-blocks.csv"), stringsAsFactors = FALSE)
g <- function(tb, m, col) tb[[col]][tb$model == m]
f3 <- function(x) sprintf("%.3f", x)
sg <- function(x) sprintf("%+.3f", x)
ci <- function(m) sprintf("%+.3f to %+.3f", g(mres, m, "ci_low"), g(mres, m, "ci_high"))

ba_spec <- f3(g(mres, "spectral", "balanced_accuracy"))
ba_struct <- f3(g(mres, "structure", "balanced_accuracy"))
ba_all <- f3(g(mres, "all", "balanced_accuracy"))
d_struct <- sg(g(mres, "structure", "diff_from_spectral"))
d_all <- sg(g(mres, "all", "diff_from_spectral"))
ci_struct <- ci("structure"); ci_all <- ci("all")
pool_spec <- f3(g(mpool, "spectral", "balanced_accuracy"))
pool_struct <- f3(g(mpool, "structure", "balanced_accuracy"))
pool_all <- f3(g(mpool, "all", "balanced_accuracy"))
spec_min <- f3(min(mpool$specificity)); spec_max <- f3(max(mpool$specificity))
sens_min <- f3(min(mpool$sensitivity)); sens_max <- f3(max(mpool$sensitivity))
tp_all <- g(mpool, "all", "tp"); tn_all <- g(mpool, "all", "tn")
n_model <- nrow(mblk)
n_iab_model <- g(mpool, "all", "tp") + g(mpool, "all", "fn")
n_ibb_model <- g(mpool, "all", "tn") + g(mpool, "all", "fp")
imb_model <- sprintf("%.2f", max(n_iab_model, n_ibb_model) /
                       min(n_iab_model, n_ibb_model))
holm_struct <- sprintf("%.3f", g(mres, "structure", "p_holm"))
```

The comparison was run once, on the sample and folds fixed before it, and it is reported as it fell. Neither hypothesis is supported and one is contradicted (@tbl-primary).

The spectral baseline separates the two agents at a balanced accuracy of `r ba_spec` averaged over folds. Adding the structural family **reduces** that to `r ba_struct`, a paired difference of `r d_struct` whose 95 percent interval, `r ci_struct`, excludes zero in the negative direction. Under the pre-registered decision rule H2 is therefore disconfirmed rather than inconclusive: structure did not fail to help, it hurt, and it does so consistently, in three of the four evaluable folds and never helping in the fourth. Adding radar as well returns the model to `r ba_all`, a difference from the baseline of `r d_all` with an interval of `r ci_all` that spans zero, so H1 is inconclusive.

One fold could not be evaluated and the reason matters more than the arithmetic. Blocking is spatially determined and label-blind by design, so nothing prevented a fold from drawing only one agent, and one did: 13 adelgid polygons and no bark beetle. Balanced accuracy needs both classes to be defined, so that fold yields no value for any model and the primary statistic rests on the remaining four, covering 56 of the `r n_model` polygons. This is a deviation from the pre-registered analysis, which did not anticipate the case, and it is recorded as one. It also biases which polygons the primary statistic sees, since the excluded fold is entirely of the majority class.

Because that exclusion is uncomfortable, a secondary estimate is reported that uses every polygon. Each unit is predicted once by a model that never saw it, and the predictions are pooled into one confusion matrix rather than averaged over folds (@tbl-secondary). It agrees with the primary analysis in direction and is lower throughout: `r pool_spec` for the spectral baseline, `r pool_struct` with structure added, `r pool_all` with radar as well.

The per-class figures are where the modest headline numbers become legible. All three models identify adelgid reasonably and bark beetle poorly, with specificity between `r spec_min` and `r spec_max` against sensitivity between `r sens_min` and `r sens_max`. The multi-sensor model reaches its balanced accuracy by predicting adelgid more often, not by separating the classes better: it recovers `r tp_all` of the `r n_iab_model` adelgid polygons but only `r tn_all` of the `r n_ibb_model` bark beetle ones. On a sample imbalanced `r imb_model` to 1 in favour of adelgid, that is close to the behaviour the balanced accuracy metric was chosen to expose.

```{r}
#| label: tbl-secondary
#| tbl-cap: "Per-class performance from pooled out-of-fold predictions, every polygon predicted once by a model that never saw it."
local({
  p <- read.csv(file.path(derived, "model-pooled.csv"), stringsAsFactors = FALSE)
  nm <- c(spectral = "Spectral", structure = "Spectral and structure",
          all = "Spectral, structure and radar")
  fmt_tbl(data.frame(
    `Predictor set` = unname(nm[p$model]),
    Sensitivity = sprintf("%.3f", p$sensitivity),
    Specificity = sprintf("%.3f", p$specificity),
    `Balanced accuracy` = sprintf("%.3f", p$balanced_accuracy),
    `Adelgid correct` = paste0(p$tp, " of ", p$tp + p$fn),
    `Bark beetle correct` = paste0(p$tn, " of ", p$tn + p$fp),
    check.names = FALSE))
})
```

```{r}
#| label: tbl-primary
#| tbl-cap: "Separation of the two damage agents by predictor set, under spatially blocked cross-validation. The interval is the bias-corrected percentile bootstrap over folds on the paired difference from the spectral baseline."
local({
  r <- read.csv(file.path(derived, "model-results.csv"),
                stringsAsFactors = FALSE)
  nm <- c(spectral = "Spectral", structure = "Spectral and structure",
          all = "Spectral, structure and radar")
  fmt_tbl(data.frame(
    `Predictor set` = unname(nm[r$model]),
    `Balanced accuracy` = sprintf("%.3f", r$balanced_accuracy),
    `Difference` = ifelse(r$model == "spectral", "baseline",
                          sprintf("%+.3f", r$diff_from_spectral)),
    `95% interval` = ifelse(is.na(r$ci_low), "",
                            sprintf("%+.3f to %+.3f", r$ci_low, r$ci_high)),
    `Verdict` = r$verdict,
    check.names = FALSE))
})
```

# Discussion and conclusions

This study asks whether structural and radar predictors separate balsam woolly adelgid from western balsam bark beetle damage in a single host, where colour alone has been shown to struggle. The answer, on this sample and under the analysis fixed before it was run, is no. The spectral baseline reaches a balanced accuracy of `r ba_spec`, adding canopy structure lowers it to `r ba_struct`, and adding radar as well returns it to `r ba_all`, which is indistinguishable from the baseline. One hypothesis is inconclusive and the other is contradicted.

The contradicted one is the more informative. Structure did not simply fail to help: it made the classifier worse by `r d_struct`, with an interval excluding zero and a Holm-corrected p of `r holm_struct`. Sixteen structural variables carrying no signal would have left the baseline roughly where it was, since a random forest can ignore uninformative inputs. Degrading it means they carried something that competes with the spectral evidence, and on 69 units with `mtry` fixed at the square root of the predictor count, adding sixteen noisy candidates dilutes the chance that an informative spectral variable is offered at any given split. The structural family behaved as a source of variance rather than of information.

That is the outcome @Choi_2023 predicted and it was recorded in the pre-registration before the models ran: press disturbances, of which adelgid damage is one, produce structural impacts that "could not be clearly detected, likely because of compensatory growth". This study now supplies a quantitative instance of that claim in a second host genus, on a second continent's congener, at operational support and with the analysis registered in advance. It also supplies the boundary condition @Boucher_2020 did not have to test: the midstorey and permeability variables that carried 60 percent of the variation in hemlock mortality did not rescue the structural arm here.

Three cautions are owed before that reading is accepted. The severity of the sample is trace and light throughout, which is where any signal is weakest, so this is a test of whether structure separates *early* damage and not of whether it separates damage. The labels are one surveyor's judgement from an aircraft and are not independently verified, so every model is chasing a target that carries its own error. And the sample is 69 polygons in five spatially blocked folds, one of which held a single class and could not be evaluated at all, so the intervals are wide by construction and the primary statistic rests on 56 of the 69.

## Archive reliability

The most consequential result is that a coverage index and a coverage guarantee are different things. The LidarBC point cloud index is a footprint layer, and testing the delivered clouds against it shows `r n_dropped_struct` of `r n_set` polygons standing on ground the index claims and the data do not occupy. The footprints are not wrong about which tiles exist; they are wrong about where those tiles hold returns, and a polygon can intersect a footprint while lying entirely in the part of it that was never flown or never delivered.

This matters more than a lost twelfth of a sample. The same index defined the sample, and the same index supported the province-wide screen that fixed this study's scope, so a design was built on a layer that overstates what can be measured. The direction of that error is at least benign for the scope decision, since overstated coverage can only mean fewer polygons genuinely carry lidar, and the finding that no moderate-or-worse adelgid polygon falls under lidar in British Columbia is if anything firmer than when it was made. It is not benign for sample size, and it would not be benign for a study that reported coverage statistics from the index without opening a file. The practical lesson is narrow and cheap to apply: verify coverage against the point clouds, polygon by polygon, before treating an index intersection as data in hand.

## Density stratification

The second finding is that an archive assembled from many acquisitions is not one instrument. Every tile in this index is advertised at `r adv_dens` points per square metre. The delivered density ranges from `r dens_tile_min` to `r dens_tile_max`, a factor of more than four hundred, and it rises with acquisition year, so the campaigns that flew most of this sample are also the sparsest. Since every canopy metric responds to sampling density, an unthinned analysis would offer a classifier the aircraft as a predictor, and where campaign is even partly associated with agent that is a route to a confident and meaningless result.

Thinning to a common density is the obvious control and it is only a partial one, because it can discard points and cannot create them. `r n_below_target` of the `r n_struct` polygons never reach the target, and the consequence surfaces in exactly the metric where it should be expected: gap fraction correlates with achieved density at `r gap_dens_r`. The rule that an empty canopy cell marks an opening is sound where a cell expects several returns and false where it expects fewer than one, in which case the cell records the pulse spacing rather than the canopy. A study using a heterogeneous archive therefore needs a density floor as an inclusion criterion, not merely a thinning step, and the floor should be set by what the metric definitions require rather than by what preserves sample size.

## Terrain and radar

The radar result is a confirmation rather than a discovery, but it is a quantified one. Over this terrain the uncorrected product is very largely a description of the hillside: `r flat_r_sigma` correlation with the slope in the range direction, falling to `r flat_r_gamma` after radiometric flattening, with between-polygon spread contracting from `r flat_sd_sigma` to `r flat_sd_gamma` dB. Any study that puts Sentinel-1 backscatter into a classifier over mountainous ground without flattening is offering the model topography and calling it canopy condition.

The residue is the more useful detail. Flattening leaves a negative association with slope, `r flat_r_gamma_slope`, where none existed before, which follows from the convexity of the volumetric model rather than from any error in applying it, and it is the reason terrain remains a covariate here rather than being treated as solved.

## Limitations

The defining limitation is the severity of the sample, and it follows from data availability rather than from choice. No adelgid polygon rated moderate or worse in British Columbia falls under lidar coverage at any useful temporal offset, so a study with a structural arm is necessarily a study of trace and light damage. The analysis is therefore concentrated in the damage classes where any spectral or structural signal is weakest, and a null result would be the expected outcome of the severity distribution as much as a statement about the sensors.

The coverage and density findings above are limitations as well as results. The structural arm rests on `r n_struct` polygons rather than `r n_set`, and the polygons that remain are not a random subset: all of the losses fall in one acquisition year, so the surviving sample is shifted towards the later and denser campaigns. The density floor that the control family requires would shift it further in the same direction. Neither shift is towards the damage classes of interest, and both narrow the ground over which any result can be claimed to hold.

Other limitations apply irrespective of the result. The labels are sketch-mapped from an aircraft and their spatial accuracy is acknowledged by the custodian to be variable, so polygon boundaries are approximate and some damage is missed entirely. That concentration may fall below the detection floor of moderate-resolution optical sensing altogether: @Ye_2021 report that their framework detected mountain pine beetle only where red-stage mortality occupied at least 56 percent of a Landsat pixel, and spruce beetle only above 55 percent grey-stage occupancy. A trace-rated polygon will rarely approach those fractions, so a null result for the spectral baseline would be consistent with the severity distribution rather than with the agents being spectrally inseparable, and the two explanations cannot be distinguished with these labels alone. Agent attribution in the source data is the surveyor's judgement from the air and is not independently verified on the ground, so the models can at best reproduce that judgement and any systematic error in it propagates into the results. Field confirmation addresses this only partly, because the diagnostic signs of adelgid infestation, the woolly wax masses and stem gout, occur on living and declining trees and are lost from those long dead.

Restricting the sample to Vancouver Island buys internal validity at the price of external validity, and the trade is stated rather than concealed. Any result reported here describes the island. The adjacent mainland coast is where it would most naturally be tested, and it is precisely the region that cannot serve as a test, because its record is almost entirely one agent under a single lidar acquisition and offers no independent variation to check the transfer against.

## Detection and attribution

This study evaluates attribution rather than detection, and the distinction is maintained throughout. Establishing that a canopy has changed is a different problem from identifying the agent responsible, and a model can perform well at the first while failing the second [@Campbell_2023]. Accuracy figures are therefore reported with their support, their class balance and their validation design.

## Conclusions

Four conclusions follow, and the first is the answer to the question the study was built to ask.

Canopy structure from airborne lidar does not help separate balsam woolly adelgid from western balsam bark beetle damage in Pacific silver fir at trace and light severity. It hurts, measurably and repeatably across folds, and adding radar recovers only the ground that structure lost. A spectral baseline at `r ba_spec` is what these data support, and even that rests on identifying the majority class well and the minority class poorly.

Open lidar archives should be treated as a collection of surveys rather than as a survey. Coverage must be verified against the point clouds and not against the index, which here claimed ground the data did not occupy for twelve polygons, and density must be equalised and then floored, because the delivered product varies by more than two orders of magnitude inside a single advertised specification and varies systematically with acquisition.

Radar over mountainous terrain requires radiometric flattening before it can be interpreted as anything other than topography, and flattening leaves a residual association with steepness that keeps terrain in the model as a covariate.

And a null obtained this way is worth more than a positive obtained otherwise. The sample, the folds, the predictor sets, the hyperparameters and the decision rule were fixed and committed before the first model ran, the estimator was shown on synthetic data to recover a planted signal and to return one half on noise, and the one deviation forced by the data is reported rather than absorbed. The result is negative, and it is the result.

# Analysis pipeline

Every derived file the sections above read is produced by the code in this section, in the order given. Nothing here is a description of code kept elsewhere: this is the code, and `05.scripts/` is generated from it by `knitr::purl()` rather than maintained alongside it, so the two cannot diverge.

Each block is guarded on the file it writes. A render with the derived files present executes none of them and the sections above read what is already on disk; delete a file and the block that makes it runs. The guard is what allows a twenty minute Earth Engine export and a 79 GB lidar retrieval to sit in a document that renders in seconds. Each block is wrapped in `local()` so that a pipeline object cannot silently replace one the manuscript is already using.

Two steps were shell and Python until 2026-08-14 and are R here, so that everything the study runs can be read in one language and in one place.

## Label extraction

The survey archive is a 908 MB file geodatabase of 1,172,716 polygons. It is never read into memory: `sf::gdal_utils()` drives the same GDAL routine that `ogr2ogr` does, so the selection and reprojection happen inside GDAL and only the extracted records reach R.

```{r}
#| label: pipe-labels
#| echo: true
#| output: false
# Source:    Pest Infestation Polygons, BC Data Catalogue, record
#            450b67bb-02d5-4526-8bc0-ac7924125a1e
# Licence:   Open Government Licence - British Columbia
# Custodian: BC Ministry of Forests. Not GeoBC.
#
# The delivered geodatabase carries codes only, no common or latin names, so IAB
# is resolved from the Forest Health Aerial Overview Survey Standards rather than
# from the delivery.
#
# NOTE ON TESTING. The archive is not on this machine, so this block has not been
# run end to end since it was translated from shell. The GDAL call and the layer
# and field names were checked against the committed extracts on 2026-08-14; the
# WHERE clauses were not re-executed. Set ADELGID_GDB to the archive to run it.
gdb <- Sys.getenv("ADELGID_GDB", unset = "")
iab_path <- file.path(derived, "adelgid-iab-polygons.geojson")
vi_path <- file.path(derived, "vi-fir-agents.geojson")

if (!file.exists(iab_path) || !file.exists(vi_path)) local({
  if (!nzchar(gdb) || !file.exists(gdb))
    stop("set ADELGID_GDB to the pest_infestation_poly.gdb archive")

  # 1. Province-wide balsam woolly adelgid, for the extent and time series.
  # WGS84 so the file is readable without a projection library. The attributes
  # kept are the ones needed to judge label quality.
  sf::gdal_utils(
    util = "vectortranslate", source = gdb, destination = iab_path,
    options = c("-f", "GeoJSON", "-t_srs", "EPSG:4326", "-sql",
                paste("SELECT CAPTURE_YEAR, PEST_SEVERITY_CODE,",
                      "PEST_CAPTURE_METHOD_CODE, TREE_SPECIES_CODE, AREA_HA",
                      "FROM pest_infestation_poly",
                      "WHERE PEST_SPECIES_CODE = 'IAB'")))

  # 2. The two agents that share a true-fir host around Vancouver Island. IAB is
  # the target class and IBB, western balsam bark beetle, the confusable one:
  # both are mapped on Pacific silver fir in the same survey, so separating them
  # is attribution, not detection. The envelope is a rectangle and therefore also
  # catches the adjacent mainland; the sample is restricted to the island later,
  # and both counts are reported.
  sf::gdal_utils(
    util = "vectortranslate", source = gdb, destination = vi_path,
    options = c("-f", "GeoJSON", "-t_srs", "EPSG:4326",
                "-spat", "-128.5", "48.3", "-123.0", "51.0",
                "-spat_srs", "EPSG:4326", "-sql",
                paste("SELECT CAPTURE_YEAR, PEST_SPECIES_CODE,",
                      "PEST_SEVERITY_CODE, TREE_SPECIES_CODE, AREA_HA",
                      "FROM pest_infestation_poly",
                      "WHERE PEST_SPECIES_CODE IN ('IAB','IBB')",
                      "AND TREE_SPECIES_CODE IN ('B','BA','BG','BL')")))
})
```

```{r}
#| label: pipe-labels-out
fmt_tbl(data.frame(
  File = c("adelgid-iab-polygons.geojson", "vi-fir-agents.geojson"),
  Contents = c("Balsam woolly adelgid, province-wide",
               "Both agents on true fir, study envelope"),
  Polygons = c(nrow(iab), nrow(vi)),
  Years = c(paste(range(iab$CAPTURE_YEAR), collapse = " to "),
            paste(range(vi$CAPTURE_YEAR), collapse = " to "))))
```

## Tile index

The LidarBC point cloud index is queried once over the study envelope and saved, because every coverage decision downstream depends on the same tile geometry and acquisition years. The service returns at most 1,000 records per request, so the query is paged to exhaustion rather than truncated.

Until 2026-08-14 this file existed in `02.inputs/derived/` with no code in the repository that produced it, while four scripts read it and the sample depended on it. The block below was written to close that gap and reproduces the committed file exactly: 18,840 tiles, identical attributes and identical total area.

```{r}
#| label: pipe-tiles
#| echo: true
#| output: false
# Source:    LidarBC Open LiDAR Data Portal, ArcGIS FeatureServer
#            LiDAR_BC_S3_Public, layer 4, the point cloud index.
# Licence:   Open Government Licence - British Columbia.
# Custodian: GeoBC.
LIDAR_SERVICE <- paste0("https://services6.arcgis.com/ubm4tcTYICKBpist/ArcGIS/",
                        "rest/services/LiDAR_BC_S3_Public/FeatureServer")
PC_LAYER <- 4L
STUDY_ENV <- c(-128.5, 48.3, -123.0, 51.0)
PAGE <- 1000L

# One URL builder for every query this manuscript makes against the service.
# Caller arguments replace the defaults rather than being appended to them: the
# service answers a repeated parameter with HTTP 400, so a caller asking for
# f = "geojson" must overwrite f = "json" and not sit beside it.
lidar_url <- function(layer, env, ...) {
  gj <- sprintf(paste0('{"xmin":%f,"ymin":%f,"xmax":%f,"ymax":%f,',
                       '"spatialReference":{"wkid":4326}}'),
                env[1], env[2], env[3], env[4])
  p <- list(where = "1=1", geometry = gj,
            geometryType = "esriGeometryEnvelope", inSR = "4326",
            spatialRel = "esriSpatialRelIntersects", f = "json")
  extra <- list(...)
  p[names(extra)] <- extra
  paste0(LIDAR_SERVICE, "/", layer, "/query?",
         paste(names(p), vapply(p, function(v)
           URLencode(as.character(v), reserved = TRUE), ""),
           sep = "=", collapse = "&"))
}

tiles_path <- file.path(derived, "lidar-tiles.geojson")
if (!file.exists(tiles_path)) local({
  acc <- list(); off <- 0L
  repeat {
    u <- lidar_url(PC_LAYER, STUDY_ENV, outFields = "year,density,maptile",
                   returnGeometry = "true", outSR = "4326", f = "geojson",
                   resultOffset = off, resultRecordCount = PAGE)
    ch <- try(sf::st_read(u, quiet = TRUE), silent = TRUE)
    if (inherits(ch, "try-error") || nrow(ch) == 0L) break
    acc[[length(acc) + 1L]] <- ch
    if (nrow(ch) < PAGE) break
    off <- off + PAGE
  }
  sf::st_write(do.call(rbind, acc), tiles_path, quiet = TRUE)
})
```

```{r}
#| label: pipe-tiles-out
local({
  tl <- sf::st_read(tiles_path, quiet = TRUE)
  fmt_tbl(data.frame(
    Tiles = nrow(tl),
    `Acquisition years` = paste(range(tl$year, na.rm = TRUE), collapse = " to "),
    `Point density` = paste(unique(tl$density), collapse = ", "),
    `Area covered (km2)` = round(sum(as.numeric(
      sf::st_area(sf::st_transform(tl, BC_ALBERS)))) / 1e6),
    check.names = FALSE))
})
```

## Coverage summary

Tile counts over the study envelope and over the Comox field window, taken with `returnCountOnly` rather than by paging, so a count is exact even where it exceeds the record cap. Where the count exceeds the cap the attribute summaries are drawn from a capped sample and are flagged as such, because they are then what the sample happened to contain rather than the full range.

```{r}
#| label: pipe-coverage
#| echo: true
#| output: false
cov_path <- file.path(derived, "lidar-coverage.csv")
if (!file.exists(cov_path)) local({
  envelopes <- list("study area" = STUDY_ENV,
                    "Comox field window" = c(-125.3, 49.4, -124.5, 49.9))
  rows <- lapply(names(envelopes), function(nm) {
    env <- envelopes[[nm]]
    n <- jsonlite::fromJSON(lidar_url(PC_LAYER, env, returnCountOnly = "true"),
                            simplifyVector = FALSE)$count
    s <- jsonlite::fromJSON(
      lidar_url(PC_LAYER, env, outFields = "year,density,projection,classes",
                returnGeometry = "false", resultRecordCount = 1000),
      simplifyVector = FALSE)
    at <- lapply(s$features, `[[`, "attributes")
    pull <- function(k) sort(unique(unlist(lapply(at, function(a) a[[k]]))))
    data.frame(envelope = nm, tiles = n,
               attrs_from_full_index = n <= 1000,
               acquisition_years_sampled = paste(pull("year"), collapse = ";"),
               density_pts_m2 = paste(pull("density"), collapse = ";"),
               point_classes = paste(pull("classes"), collapse = ";"),
               stringsAsFactors = FALSE)
  })
  write.csv(do.call(rbind, rows), cov_path, row.names = FALSE)
})
```

## Basemap layers

Natural Earth vectors and an elevation hillshade, clipped and reprojected to BC Albers once so that the study area figure draws a metrically true scale bar. Natural Earth is public domain.

```{r}
#| label: pipe-basemap
#| echo: true
#| output: false
bm_path <- file.path(derived, "basemap.gpkg")
hs_path <- file.path(derived, "hillshade.tif")

# Map frame, slightly wider than the analysis envelope so the island sits inside
# a margin rather than against the neatline. This is the view the figures show.
FRAME <- c(xmin = -129.2, ymin = 48.0, xmax = -122.3, ymax = 51.4)
# Data are clipped wider than the view. The figures plot at a fixed aspect ratio,
# which expands the shorter axis past the view rectangle; land cropped at the
# view would stop short of the neatline and leave a band of sea where the
# mainland belongs.
PAD <- 1.2
DATA_FRAME <- FRAME + c(-PAD, -PAD, PAD, PAD)

if (!file.exists(bm_path) || !file.exists(hs_path)) local({
  library(rnaturalearth)
  library(elevatr)

  clip <- function(x) {
    x <- sf::st_make_valid(sf::st_transform(x, 4326))
    suppressWarnings(sf::st_crop(x, DATA_FRAME))
  }

  land <- ne_download(scale = 10, type = "land", category = "physical",
                      returnclass = "sf")
  states <- ne_states(country = c("Canada", "United States of America"),
                      returnclass = "sf")
  # The global rivers layer resolves only one river inside this frame, which is
  # useless for orientation. The North America supplement resolves 32.
  rivers <- ne_download(scale = 10, type = "rivers_north_america",
                        category = "physical", returnclass = "sf")
  # Several lake polygons carry self-intersecting rings. s2 rejects them
  # outright, so validity is repaired in planar mode.
  lakes <- tryCatch({
    old <- sf::sf_use_s2(FALSE); on.exit(sf::sf_use_s2(old), add = TRUE)
    ne_download(scale = 10, type = "lakes_north_america",
                category = "physical", returnclass = "sf")
  }, error = function(e) NULL)

  layers <- list(
    land = clip(land["featurecla"]),
    rivers = clip(rivers["name"]),
    # The view rectangle, saved so the figures set their limits from it rather
    # than from the bounding box of the land, which is deliberately wider.
    frame = sf::st_sf(name = "Map frame",
                      geometry = sf::st_as_sfc(sf::st_bbox(FRAME, crs = 4326))))

  # Only the shared land border is kept. Drawing the full provincial outline
  # would retrace the coastline in a second style, which is confusing at this
  # extent; the international boundary is the only administrative line inside
  # this frame.
  bc <- states[states$name == "British Columbia", ]
  us <- states[states$admin == "United States of America", ]
  border <- tryCatch({
    old <- sf::sf_use_s2(FALSE); on.exit(sf::sf_use_s2(old), add = TRUE)
    b <- sf::st_intersection(
      sf::st_boundary(sf::st_make_valid(sf::st_union(bc))),
      sf::st_buffer(sf::st_boundary(sf::st_make_valid(sf::st_union(us))), 0.02))
    sf::st_sf(name = "International boundary", geometry = sf::st_geometry(b))
  }, error = function(e) NULL)
  if (!is.null(border)) layers$admin <- clip(border)
  if (!is.null(lakes)) {
    old <- sf::sf_use_s2(FALSE); layers$lakes <- clip(lakes["name"])
    sf::sf_use_s2(old)
  }

  if (file.exists(bm_path)) file.remove(bm_path)
  for (nm in names(layers))
    sf::st_write(sf::st_transform(layers[[nm]], BC_ALBERS), bm_path,
                 layer = nm, quiet = TRUE, append = FALSE)

  # Place labels that orient a reader unfamiliar with the coast. Coordinates are
  # from the BC Geographical Names Office listings, entered by hand because the
  # Natural Earth populated-places layer omits the smaller island communities.
  places <- sf::st_as_sf(data.frame(
    name = c("Comox", "Campbell River", "Port Alberni", "Nanaimo",
             "Victoria", "Vancouver", "Tofino", "Port Hardy"),
    lon = c(-124.928, -125.247, -124.805, -123.936,
            -123.365, -123.121, -125.906, -127.418),
    lat = c(49.673, 50.024, 49.234, 49.166,
            48.428, 49.283, 49.153, 50.720)),
    coords = c("lon", "lat"), crs = 4326)
  sf::st_write(sf::st_transform(places, BC_ALBERS), bm_path, layer = "places",
               quiet = TRUE, append = FALSE)

  # Hillshade, so the reader sees the mountain spine rather than a flat outline.
  # AWS Terrain Tiles via elevatr. Zoom 7 is about 1 km ground resolution, ample
  # at this extent and small enough to commit.
  frame_sf <- sf::st_as_sf(sf::st_as_sfc(sf::st_bbox(DATA_FRAME, crs = 4326)))
  dem <- elevatr::get_elev_raster(frame_sf, z = 7, clip = "bbox",
                                  verbose = FALSE)
  dem <- terra::project(terra::rast(dem), paste0("EPSG:", BC_ALBERS))
  # Terrain tiles return zero or small positive values over water rather than
  # negative ones, so an elevation threshold does not separate sea from shore.
  # Masking by the coastline polygon does.
  dem <- terra::mask(dem, terra::vect(sf::st_geometry(
    sf::st_transform(layers$land, BC_ALBERS))))
  hs <- terra::shade(terra::terrain(dem, "slope", unit = "radians"),
                     terra::terrain(dem, "aspect", unit = "radians"),
                     angle = 40, direction = 315)
  terra::writeRaster(hs, hs_path, overwrite = TRUE,
                     gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2"))
})
```

```{r}
#| label: pipe-basemap-out
local({
  ly <- sf::st_layers(bm_path)
  fmt_tbl(data.frame(Layer = ly$name,
                     Features = as.integer(ly$features),
                     Geometry = vapply(ly$geomtype, function(g) g[1], "")))
})
```

## Site candidates

Every adelgid polygon in the province, with the lidar tiles over it, its acquisition years and the offset from the survey year. Coverage is queried per polygon envelope rather than over one rectangle, because the province-wide record runs from the coast to the Kootenays and a single envelope would pull far more tile geometry than this machine should hold.

```{r}
#| label: pipe-candidates
#| echo: true
#| output: false
cand_path <- file.path(derived, "site-candidates-province.csv")
if (!file.exists(cand_path)) local({
  sev <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
           V = "Very severe")
  host <- c(B = "True fir unspecified", BA = "Pacific silver fir",
            BG = "Grand fir", BL = "Subalpine fir")

  tiles_over <- function(bb) {
    n <- jsonlite::fromJSON(lidar_url(PC_LAYER, bb, returnCountOnly = "true"),
                            simplifyVector = FALSE)$count
    if (!length(n) || n == 0) return(list(n = 0L, years = integer(0)))
    d <- jsonlite::fromJSON(
      lidar_url(PC_LAYER, bb, outFields = "year", returnGeometry = "false",
                resultRecordCount = 1000), simplifyVector = FALSE)
    list(n = n, years = sort(unique(unlist(
      lapply(d$features, function(f) f$attributes$year)))))
  }

  src <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                     quiet = TRUE)
  rows <- lapply(seq_len(nrow(src)), function(i) {
    bb <- as.numeric(sf::st_bbox(sf::st_geometry(src)[i]))
    tv <- tiles_over(bb)
    yr <- src$CAPTURE_YEAR[i]
    near <- if (length(tv$years) && !is.na(yr))
      tv$years[which.min(abs(tv$years - yr))] else NA_integer_
    Sys.sleep(0.25)
    data.frame(
      capture_year = yr,
      severity = unname(sev[src$PEST_SEVERITY_CODE[i]]),
      severity_code = src$PEST_SEVERITY_CODE[i],
      host = unname(host[src$TREE_SPECIES_CODE[i]]),
      host_code = ifelse(is.na(src$TREE_SPECIES_CODE[i]), "",
                         src$TREE_SPECIES_CODE[i]),
      area_ha = round(src$AREA_HA[i], 1),
      lidar_tiles = tv$n,
      lidar_years = paste(tv$years, collapse = ";"),
      nearest_lidar_year = if (is.na(near)) "" else near,
      year_offset = if (is.na(near)) "" else abs(near - yr),
      lon = round(mean(bb[c(1, 3)]), 5), lat = round(mean(bb[c(2, 4)]), 5),
      on_vancouver_island = bb[3] < -123.2 && bb[4] < 51.2,
      stringsAsFactors = FALSE)
  })
  out <- do.call(rbind, rows)
  out <- out[order(-out$lidar_tiles, out$severity_code), ]
  write.csv(out, cand_path, row.names = FALSE)
})
```

## Availability screen

The study is located by where all three requirements already coexist, rather than by preference. Earlier work asked whether a chosen set of polygons had lidar, which is the wrong order and produced a design requiring lidar that does not exist. A region is only a candidate if it can supply both agents on the shared host, since adelgid polygons under lidar with no bark beetle polygons under the same lidar cannot support a two-class attribution test at all.

```{r}
#| label: pipe-availability
#| echo: true
#| output: false
avail_path <- file.path(derived, "availability-regions.csv")
if (!file.exists(avail_path)) local({
  MAX_OFFSET <- 3          # years between survey and lidar acquisition
  GRID <- 25000            # 25 km analysis cells

  v <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)), BC_ALBERS)
  tl <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)), BC_ALBERS)

  hit <- sf::st_intersects(v, tl)
  v$lidar_tiles <- lengths(hit)
  v$nearest_year <- vapply(seq_len(nrow(v)), function(k) {
    if (!v$lidar_tiles[k]) return(NA_integer_)
    ys <- tl$year[hit[[k]]]; ys <- ys[!is.na(ys)]
    if (!length(ys)) return(NA_integer_)
    ys[which.min(abs(ys - v$CAPTURE_YEAR[k]))]
  }, integer(1))
  v$offset <- abs(v$nearest_year - v$CAPTURE_YEAR)
  usable <- v[!is.na(v$offset) & v$offset <= MAX_OFFSET, ]

  cells <- sf::st_make_grid(usable, cellsize = GRID)
  cells <- sf::st_sf(cell = seq_along(cells), geometry = cells)
  ix <- sf::st_intersects(cells, usable)
  sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)

  rows <- lapply(seq_len(nrow(cells)), function(k) {
    sub <- usable[ix[[k]], ]
    if (!nrow(sub)) return(NULL)
    a <- sub[sub$PEST_SPECIES_CODE == "IAB", ]
    b <- sub[sub$PEST_SPECIES_CODE == "IBB", ]
    ctr <- sf::st_coordinates(sf::st_centroid(sf::st_transform(cells[k, ], 4326)))
    data.frame(
      cell = k, lon = round(ctr[1, 1], 3), lat = round(ctr[1, 2], 3),
      iab_n = nrow(a), ibb_n = nrow(b),
      # Shared host is what makes the comparison an attribution test.
      iab_ba = sum(a$TREE_SPECIES_CODE %in% "BA"),
      ibb_ba = sum(b$TREE_SPECIES_CODE %in% "BA"),
      iab_best_sev = if (nrow(a)) names(which.max(
        sev_rank[a$PEST_SEVERITY_CODE])) else NA_character_,
      iab_mod_plus = sum(a$PEST_SEVERITY_CODE %in% c("M", "S", "V")),
      iab_ha = round(sum(a$AREA_HA)),
      yrs = paste(range(sub$CAPTURE_YEAR), collapse = "-"),
      med_offset = median(sub$offset))
  })
  tab <- do.call(rbind, rows)
  tab$viable <- tab$iab_ba > 0 & tab$ibb_ba > 0
  tab <- tab[order(-tab$viable, -(tab$iab_ba + tab$ibb_ba), -tab$iab_mod_plus), ]
  write.csv(tab, avail_path, row.names = FALSE)
})
```

## Sample freeze

The five inclusion conditions, applied in order. This block writes the file that is the sample, and its size is the study's sample size. Nothing downstream filters these polygons again.

```{r}
#| label: pipe-sample
#| echo: true
#| output: false
set_path <- file.path(derived, "analysis-set.geojson")
restr_path <- file.path(derived, "analysis-set-restriction.csv")

if (!file.exists(set_path) || !file.exists(restr_path)) local({
  MAX_OFFSET <- 5
  # A point in Strathcona Park, well inland, used to pick Vancouver Island out of
  # the land layer by containment rather than by area, so the selection cannot
  # silently attach to a different landmass.
  VI_SEED <- c(-125.60, 49.60)

  v <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)), BC_ALBERS)
  tl <- sf::st_transform(sf::st_make_valid(sf::st_read(
    file.path(derived, "lidar-tiles.geojson"), quiet = TRUE)), BC_ALBERS)

  # 2. Host held constant, so the comparison is attribution rather than host
  # discrimination.
  v <- v[v$TREE_SPECIES_CODE %in% "BA", ]

  # 3 and 4. Lidar coverage, and an acquisition within MAX_OFFSET years.
  hit <- sf::st_intersects(v, tl)
  v$lidar_tiles <- lengths(hit)
  v$lidar_year <- vapply(seq_len(nrow(v)), function(k) {
    if (!v$lidar_tiles[k]) return(NA_integer_)
    ys <- tl$year[hit[[k]]]; ys <- ys[!is.na(ys)]
    if (!length(ys)) return(NA_integer_)
    ys[which.min(abs(ys - v$CAPTURE_YEAR[k]))]
  }, integer(1))
  v$year_offset <- abs(v$lidar_year - v$CAPTURE_YEAR)
  aset <- v[!is.na(v$year_offset) & v$year_offset <= MAX_OFFSET, ]

  # 5. Vancouver Island only. Off the island the record is 43 bark beetle to 5
  # adelgid and 38 of 48 polygons come from one 2020 acquisition, so landmass and
  # campaign each predict the agent almost perfectly. Blocking does not answer a
  # confound that holds within folds as well as across them, so region is
  # neutralised by restriction, the instrument already used for host.
  land_l <- sf::st_transform(sf::st_make_valid(
    sf::st_read(bm_path, "land", quiet = TRUE)), BC_ALBERS)
  seed <- sf::st_transform(sf::st_sfc(sf::st_point(VI_SEED), crs = 4326),
                           BC_ALBERS)
  island <- land_l[lengths(sf::st_intersects(land_l, seed)) > 0, ]
  stopifnot(nrow(island) == 1)
  on_island <- lengths(sf::st_intersects(
    sf::st_centroid(sf::st_geometry(aset)), island)) > 0

  # The counts that motivate the restriction are saved rather than quoted from a
  # terminal session, because the manuscript reports them.
  restriction <- as.data.frame(table(
    landmass = ifelse(on_island, "Vancouver Island", "Elsewhere"),
    agent = aset$PEST_SPECIES_CODE,
    lidar_year = aset$lidar_year), stringsAsFactors = FALSE)
  restriction <- restriction[restriction$Freq > 0, ]
  names(restriction)[names(restriction) == "Freq"] <- "n"
  write.csv(restriction, restr_path, row.names = FALSE)

  aset <- aset[on_island, ]
  if (file.exists(set_path)) invisible(file.remove(set_path))
  sf::st_write(sf::st_transform(aset, 4326), set_path, quiet = TRUE)
})
```

```{r}
#| label: pipe-sample-out
fmt_tbl(data.frame(
  Quantity = c("Polygons", "Adelgid", "Bark beetle", "Imbalance",
               "Survey years", "Lidar years", "Median offset (years)",
               "Total area (ha)"),
  Value = c(as.character(n_set), as.character(n_set_iab),
            as.character(n_set_ibb), paste(set_imbalance, "to 1"),
            paste(set_years, collapse = " to "),
            paste(sort(unique(aset$lidar_year)), collapse = ", "),
            as.character(set_offset_med),
            format(set_ha, big.mark = ","))))
```

## Optical predictors

Sentinel-2 Level 2A over the leaf-on window of the label epoch, masked from the scene classification layer, corrected scene by scene for topographic illumination by SCS+C, reduced to a median composite and summarised to the inward-buffered polygon. No label enters this block: the analysis set is stripped to geometry and a positional identifier before anything is sent to Earth Engine, and the output is asserted to carry no agent, severity or host column.

The empirical `c` parameter is fitted per scene and per band by regression of reflectance on the illumination condition, over the sample polygons padded outward, so the correction is calibrated on the terrain it is applied to. Pixels facing away from the sun are masked rather than corrected, since the correction is undefined there. A tasselled cap transform is deliberately absent: the Sentinel-2 coefficients have not been read from their source publication, and unverified coefficients are not written into code.

```{r}
#| label: pipe-optical
#| echo: true
#| output: false
opt_path <- file.path(derived, "predictors-optical.csv")
opt_log <- file.path(derived, "predictors-optical-log.csv")

EE_PROJECT <- "murphys-deforisk"
S2_START <- "2019-06-01"
S2_END <- "2021-09-30"
BUFFER_M <- 20            # inward erosion of the survey polygon, provisional
SUMM_SCALE <- 20          # metres, scale of the polygon summary

if (!file.exists(opt_path)) local({
  library(rgee)
  MAX_CLOUD  <- 40        # scene cloud cover, percent
  FIT_PAD_M  <- 5000      # outward pad of the region the c parameter is fitted on
  FIT_SCALE  <- 100       # metres, scale of the c parameter fit
  FIT_MIN_PX <- 500       # minimum valid pixels in the fit region to keep a scene

  BANDS <- c("B2", "B3", "B4", "B5", "B6", "B7", "B8", "B8A", "B11", "B12")
  # Scene classification classes retained: 4 vegetation, 5 not vegetated,
  # 7 unclassified. Masked: 0 no data, 1 saturated, 2 dark, 3 cloud shadow,
  # 6 water, 8 and 9 cloud, 10 cirrus, 11 snow.
  SCL_KEEP <- c(4L, 5L, 7L)

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  # poly_id is positional in the frozen file, and every predictor block derives
  # it the same way, so the three tables join on a common key.
  poly_id <- sprintf("P%03d", seq_len(n_all))
  # Everything except geometry is dropped here. No label reaches Earth Engine.
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)

  # Sketch-mapped boundaries are approximate, so the polygon is eroded before it
  # is summarised. Polygons too small to survive erosion are dropped and counted.
  eroded <- sf::st_buffer(geom, -BUFFER_M)
  alive <- !sf::st_is_empty(eroded)
  n_dropped <- sum(!alive)
  eroded <- eroded[alive]; poly_id <- poly_id[alive]
  polys <- sf::st_sf(poly_id = poly_id,
                     geometry = sf::st_transform(eroded, 4326))
  stopifnot(identical(setdiff(names(polys), "geometry"), "poly_id"))

  ee_Initialize(project = EE_PROJECT, quiet = TRUE)
  fc <- sf_as_ee(polys)
  # The c parameter is fitted over the sample padded outward, rather than over
  # the whole scene, so the regression describes the terrain the correction is
  # applied to and the computation stays bounded.
  fit_region <- sf_as_ee(sf::st_transform(
    sf::st_union(sf::st_buffer(sf::st_union(geom), FIT_PAD_M)), 4326))

  dem <- ee$ImageCollection("COPERNICUS/DEM/GLO30")$select("DEM")$mosaic()$
    setDefaultProjection("EPSG:3857", NULL, 30)
  terrain <- ee$Terrain$products(dem)
  slope_rad <- terrain$select("slope")$multiply(pi / 180)
  aspect_rad <- terrain$select("aspect")$multiply(pi / 180)

  mask_scl <- function(img) {
    scl <- img$select("SCL")
    keep <- scl$eq(SCL_KEEP[1])
    for (k in SCL_KEEP[-1]) keep <- keep$Or(scl$eq(k))
    img$updateMask(keep)
  }

  # Illumination condition, the cosine of the solar incidence angle on the local
  # surface. A non-positive value faces away from the sun and is masked.
  illumination <- function(img) {
    sz <- ee$Number(img$get("MEAN_SOLAR_ZENITH_ANGLE"))$multiply(pi / 180)
    sa <- ee$Number(img$get("MEAN_SOLAR_AZIMUTH_ANGLE"))$multiply(pi / 180)
    il <- ee$Image$constant(sz$cos())$multiply(slope_rad$cos())$add(
      ee$Image$constant(sz$sin())$multiply(slope_rad$sin())$
        multiply(aspect_rad$subtract(sa)$cos()))
    il$rename("IL")
  }

  prepare <- function(img) {
    refl <- mask_scl(img)$select(BANDS)$divide(10000)
    il <- illumination(img)
    refl$updateMask(il$gt(0))$addBands(il)$
      copyProperties(img, list("system:time_start",
                               "MEAN_SOLAR_ZENITH_ANGLE",
                               "MEAN_SOLAR_AZIMUTH_ANGLE"))
  }

  # SCS+C. rho_corrected = rho * (cos(slope) cos(zenith) + c) / (IL + c), with c
  # the ratio of intercept to slope from the per-band regression of reflectance
  # on IL. A non-positive fitted slope is physically wrong, so c is forced large
  # and the correction collapses to the identity for that band.
  correct_scsc <- function(img) {
    sz <- ee$Number(img$get("MEAN_SOLAR_ZENITH_ANGLE"))$multiply(pi / 180)
    il <- img$select("IL")
    design <- ee$Image$constant(1)$rename("constant")$addBands(il)$
      addBands(img$select(BANDS))$updateMask(il$mask())
    fit <- design$reduceRegion(
      reducer = ee$Reducer$linearRegression(numX = 2L, numY = length(BANDS)),
      geometry = fit_region, scale = FIT_SCALE, maxPixels = 1e9, tileScale = 4L)
    coefs <- ee$Array(fit$get("coefficients"))
    intercept <- ee$Image$constant(
      coefs$slice(0L, 0L, 1L)$project(list(1L))$toList())$rename(BANDS)
    slope <- ee$Image$constant(
      coefs$slice(0L, 1L, 2L)$project(list(1L))$toList())$rename(BANDS)
    cimg <- intercept$divide(slope)
    cimg <- cimg$where(slope$lte(0), 1e6)
    cimg <- cimg$where(cimg$lte(0), 1e6)
    numer <- ee$Image$constant(sz$cos())$multiply(slope_rad$cos())$add(cimg)
    img$select(BANDS)$multiply(numer$divide(il$add(cimg)))$
      copyProperties(img, list("system:time_start"))
  }

  # Scenes with too few valid pixels cannot support the regression and are
  # dropped rather than corrected from a degenerate fit.
  count_fit_px <- function(img) {
    img$set("n_fit", img$select("B8")$reduceRegion(
      reducer = ee$Reducer$count(), geometry = fit_region, scale = FIT_SCALE,
      maxPixels = 1e9, tileScale = 4L)$get("B8"))
  }

  raw <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(fit_region)$filterDate(S2_START, S2_END)$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", MAX_CLOUD))
  n_raw <- raw$size()$getInfo()

  prepared <- raw$map(ee_utils_pyfunc(prepare))$
    map(ee_utils_pyfunc(count_fit_px))$
    filter(ee$Filter$gt("n_fit", FIT_MIN_PX))
  n_fit <- prepared$size()$getInfo()

  composite <- prepared$map(ee_utils_pyfunc(correct_scsc))$median()

  # Indices selected for sensitivity to canopy water content and to visible and
  # shortwave change. Provisional; the set is fixed at pre-registration.
  nd <- function(x, y, nm) composite$normalizedDifference(c(x, y))$rename(nm)
  indices <- nd("B8", "B4", "NDVI")$
    addBands(nd("B8", "B11", "NDMI"))$
    addBands(nd("B8", "B12", "NBR"))$
    addBands(nd("B8", "B5", "NDRE1"))$
    addBands(nd("B8", "B3", "GNDVI"))$
    addBands(composite$select("B11")$divide(composite$select("B8"))$rename("MSI"))$
    addBands(composite$select("B11")$divide(composite$select("B12"))$rename("SWIRR"))$
    addBands(composite$expression(
      "2.5 * (nir - red) / (nir + 6 * red - 7.5 * blue + 1)",
      list(nir = composite$select("B8"), red = composite$select("B4"),
           blue = composite$select("B2")))$rename("EVI"))

  stack <- composite$addBands(indices)
  VARS <- c(BANDS, "NDVI", "NDMI", "NBR", "NDRE1", "GNDVI", "MSI", "SWIRR", "EVI")

  reducer <- ee$Reducer$mean()$
    combine(reducer2 = ee$Reducer$stdDev(), sharedInputs = TRUE)$
    combine(reducer2 = ee$Reducer$count(), sharedInputs = TRUE)
  summaries <- stack$reduceRegions(collection = fc, reducer = reducer,
                                   scale = SUMM_SCALE, tileScale = 4L)

  # Drive credentials are not set on this machine, so the export goes to an
  # Earth Engine asset and is read back. Polling is to completion rather than
  # for a fixed number of attempts: reading the asset before the task has
  # written it fails with a misleading "not found".
  asset_id <- sprintf("projects/%s/assets/adelgid-optical-%s", EE_PROJECT,
                      format(Sys.Date(), "%Y%m%d"))
  try(ee_manage_delete(asset_id), silent = TRUE)
  task <- ee_table_to_asset(collection = summaries,
                            description = "adelgid_optical", assetId = asset_id)
  task$start()
  repeat {
    st <- task$status()
    if (st$state %in% c("COMPLETED", "FAILED", "CANCELLED", "CANCEL_REQUESTED"))
      break
    Sys.sleep(30)
  }
  if (st$state != "COMPLETED") stop("Earth Engine export ", st$state)

  got <- ee$FeatureCollection(asset_id)$getInfo()
  rows <- lapply(got$features, function(f)
    as.data.frame(f$properties, stringsAsFactors = FALSE))
  tab <- do.call(rbind, lapply(rows, function(r) {
    wanted <- c("poly_id", unlist(lapply(VARS, function(v)
      paste0(v, c("_mean", "_stdDev", "_count")))))
    for (m in setdiff(wanted, names(r))) r[[m]] <- NA_real_
    r[, wanted, drop = FALSE]
  }))
  tab <- tab[order(tab$poly_id), ]

  # The guard that matters: no label may leave this block.
  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE|lidar|offset|AREA",
                        names(tab), ignore.case = TRUE)) == 0)
  write.csv(tab, opt_path, row.names = FALSE)
  write.csv(data.frame(
    parameter = c("collection", "start", "end", "max_scene_cloud_pct",
                  "scl_kept", "scenes_matched", "scenes_corrected",
                  "topographic_correction", "dem", "c_fit_scale_m",
                  "c_fit_pad_m", "c_fit_min_pixels", "inward_buffer_m",
                  "summary_scale_m", "polygons_in", "polygons_dropped_by_buffer",
                  "polygons_out", "built"),
    value = c("COPERNICUS/S2_SR_HARMONIZED", S2_START, S2_END, MAX_CLOUD,
              paste(SCL_KEEP, collapse = " "), n_raw, n_fit, "SCS+C",
              "COPERNICUS/DEM/GLO30", FIT_SCALE, FIT_PAD_M, FIT_MIN_PX,
              BUFFER_M, SUMM_SCALE, n_all, n_dropped, nrow(tab),
              as.character(Sys.Date())), stringsAsFactors = FALSE),
    opt_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-optical-out
local({
  o <- read.csv(opt_path)
  v <- c("NDVI", "NDMI", "NBR", "EVI")
  fmt_tbl(data.frame(
    Index = v,
    Minimum = round(vapply(v, function(k) min(o[[paste0(k, "_mean")]]), 0), 3),
    Median = round(vapply(v, function(k) median(o[[paste0(k, "_mean")]]), 0), 3),
    Maximum = round(vapply(v, function(k) max(o[[paste0(k, "_mean")]]), 0), 3),
    `Within-polygon SD` = round(vapply(v, function(k)
      median(o[[paste0(k, "_stdDev")]]), 0), 3),
    check.names = FALSE))
})
```

## Radar predictors

Sentinel-1 Ground Range Detected in interferometric wide swath mode, both polarisations, over the same window as the optical build. The archive is delivered thermal-noise-removed, radiometrically calibrated and Range Doppler terrain corrected. Border noise removal and orbit refinement are not documented for the collection, so border noise is handled here by eroding each scene footprint and masking a low-intensity floor, and no orbit refinement is claimed.

Backscatter is flattened radiometrically against a digital elevation model by the angular volumetric model of @Vollrath_2020, which is appropriate to a vegetated surface. Layover and shadow are masked on the published conditions and the retained fraction of every polygon is written out, so the reduction is reported rather than absorbed. Both temporal statistics are formed within orbit pass before the passes are combined: ascending and descending views of the same slope differ by more than flattening removes, and pooling them would write that difference into the dispersion as a terrain effect. The level is averaged in power and converted to decibels afterwards, because averaging decibels averages logarithms and biases the result low; the dispersion is the standard deviation of the decibel series, since speckle is multiplicative and so additive in the log.

```{r}
#| label: pipe-radar
#| echo: true
#| output: false
rad_path <- file.path(derived, "predictors-radar.csv")
rad_log <- file.path(derived, "predictors-radar-log.csv")

if (!file.exists(rad_path)) local({
  library(rgee)
  BORDER_ERODE_M <- 1000   # inward erosion of the scene footprint, border noise
  BORDER_FLOOR_DB <- -30   # below this is border artefact, not target
  MIN_PASS_SCENES <- 5     # a pass with fewer gives no usable dispersion
  HEADING_SCALE <- 1000    # metres, scale of the look direction estimate
  POLS <- c("VV", "VH")
  PASSES <- c("ASCENDING", "DESCENDING")
  VARS <- c("VV_gamma0", "VH_gamma0", "VV_tsd", "VH_tsd")

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  poly_id <- sprintf("P%03d", seq_len(n_all))
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)
  eroded <- sf::st_buffer(geom, -BUFFER_M)
  alive <- !sf::st_is_empty(eroded)
  n_dropped <- sum(!alive)
  polys <- sf::st_sf(poly_id = poly_id[alive],
                     geometry = sf::st_transform(eroded[alive], 4326))
  stopifnot(identical(setdiff(names(polys), "geometry"), "poly_id"))

  ee_Initialize(project = EE_PROJECT, quiet = TRUE)
  fc <- sf_as_ee(polys)
  # sf_as_ee on an sfc returns an ee$Geometry, not a Feature.
  region <- sf_as_ee(sf::st_transform(sf::st_union(geom), 4326))

  dem <- ee$ImageCollection("COPERNICUS/DEM/GLO30")$select("DEM")$mosaic()$
    setDefaultProjection("EPSG:3857", NULL, 30)
  ninety_rad <- ee$Image$constant(90)$multiply(pi / 180)

  # Angular radiometric slope correction, volumetric model. The equations were
  # read from Vollrath, Mullissa and Reiche (2020) and the implementation checked
  # against the authors' own module:
  #   gamma0   = sigma0 / cos(theta_i)
  #   alpha_r  = arctan(tan(alpha_s) cos(phi_r))
  #   gamma0_f = gamma0 * tan(90 - theta_i) / tan(90 - theta_i + alpha_r)
  slope_correct <- function(img) {
    scene <- img$geometry()
    proj <- img$select("VV")$projection()
    # The look direction is the aspect of the incidence angle surface, taken
    # coarsely because it is near constant across the swath.
    heading <- ee$Terrain$aspect(img$select("angle"))$
      reduceRegion(reducer = ee$Reducer$mean(), geometry = scene,
                   scale = HEADING_SCALE, maxPixels = 1e9, bestEffort = TRUE)$
      get("aspect")
    sigma0_pow <- ee$Image$constant(10)$pow(img$select(POLS)$divide(10))
    theta_i <- img$select("angle")$multiply(pi / 180)$clip(scene)
    phi_i <- ee$Image$constant(heading)$multiply(pi / 180)
    # Terrain geometry, evaluated in the projection of the radar image.
    alpha_s <- ee$Terrain$slope(dem)$select("slope")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    phi_s <- ee$Terrain$aspect(dem)$select("aspect")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    alpha_r <- alpha_s$tan()$multiply(phi_i$subtract(phi_s)$cos())$atan()
    gamma0 <- sigma0_pow$divide(theta_i$cos())
    scf <- ninety_rad$subtract(theta_i)$add(alpha_r)$tan()$
      divide(ninety_rad$subtract(theta_i)$tan())
    g_db <- ee$Image$constant(10)$multiply(gamma0$divide(scf)$log10())$
      select(POLS)$rename(POLS)
    # Valid where neither in active layover nor in active shadow.
    valid <- alpha_r$lt(theta_i)$
      And(alpha_r$gt(ninety_rad$subtract(theta_i)$multiply(-1)))
    g_db$updateMask(valid)$updateMask(g_db$gt(BORDER_FLOOR_DB))$
      clip(scene$buffer(-BORDER_ERODE_M, 100))$
      copyProperties(img, list("system:time_start", "orbitProperties_pass"))
  }

  to_power <- function(img) {
    ee$Image$constant(10)$pow(img$select(POLS)$divide(10))$
      copyProperties(img, list("system:time_start"))
  }

  raw <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(region)$filterDate(S2_START, S2_END)$
    filter(ee$Filter$eq("instrumentMode", "IW"))$
    filter(ee$Filter$eq("resolution_meters", 10))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VV"))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VH"))
  n_raw <- raw$size()$getInfo()
  corrected <- raw$map(ee_utils_pyfunc(slope_correct))

  pow_list <- list(); var_list <- list()
  pass_n <- stats::setNames(integer(length(PASSES)), PASSES)
  for (p in PASSES) {
    cp <- corrected$filter(ee$Filter$eq("orbitProperties_pass", p))
    n_p <- cp$size()$getInfo()
    pass_n[p] <- n_p
    if (n_p < MIN_PASS_SCENES) next
    pow_list[[p]] <- cp$map(ee_utils_pyfunc(to_power))$mean()
    var_list[[p]] <- cp$select(POLS)$reduce(ee$Reducer$stdDev())$pow(2)
  }
  n_used <- sum(pass_n[names(pow_list)])
  stopifnot(length(pow_list) > 0)

  gamma_db <- ee$ImageCollection$fromImages(unname(pow_list))$mean()$
    log10()$multiply(10)$rename(c("VV_gamma0", "VH_gamma0"))
  pooled_sd <- ee$ImageCollection$fromImages(unname(var_list))$mean()$sqrt()$
    rename(c("VV_tsd", "VH_tsd"))
  # An unmasked copy of one band gives the polygon's full pixel count on the
  # same grid, so the layover and shadow loss is reported as a retained fraction
  # rather than left implicit in a smaller count.
  all_band <- gamma_db$select("VV_gamma0")$unmask(-999)$rename("ALL")
  stack <- gamma_db$addBands(pooled_sd)$addBands(all_band)

  reducer <- ee$Reducer$mean()$
    combine(reducer2 = ee$Reducer$stdDev(), sharedInputs = TRUE)$
    combine(reducer2 = ee$Reducer$count(), sharedInputs = TRUE)
  summaries <- stack$reduceRegions(collection = fc, reducer = reducer,
                                   scale = SUMM_SCALE, tileScale = 4L)

  asset_id <- sprintf("projects/%s/assets/adelgid-radar-%s", EE_PROJECT,
                      format(Sys.Date(), "%Y%m%d"))
  try(ee_manage_delete(asset_id), silent = TRUE)
  task <- ee_table_to_asset(collection = summaries,
                            description = "adelgid_radar", assetId = asset_id)
  task$start()
  repeat {
    st <- task$status()
    if (st$state %in% c("COMPLETED", "FAILED", "CANCELLED", "CANCEL_REQUESTED"))
      break
    Sys.sleep(30)
  }
  if (st$state != "COMPLETED") stop("Earth Engine export ", st$state)

  got <- ee$FeatureCollection(asset_id)$getInfo()
  rows <- lapply(got$features, function(f)
    as.data.frame(f$properties, stringsAsFactors = FALSE))
  tab <- do.call(rbind, lapply(rows, function(r) {
    wanted <- c("poly_id", unlist(lapply(VARS, function(v)
      paste0(v, c("_mean", "_stdDev", "_count")))), "ALL_count")
    for (m in setdiff(wanted, names(r))) r[[m]] <- NA_real_
    r[, wanted, drop = FALSE]
  }))
  tab <- tab[order(tab$poly_id), ]
  # The reduction the methods require to be reported.
  tab$retained_frac <- round(tab$VV_gamma0_count / tab$ALL_count, 4)

  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE|lidar|offset|AREA",
                        names(tab), ignore.case = TRUE)) == 0)
  write.csv(tab, rad_path, row.names = FALSE)
  write.csv(data.frame(
    parameter = c("collection", "start", "end", "instrument_mode",
                  "polarisations", "resolution_m", "scenes_matched",
                  "scenes_ascending", "scenes_descending", "scenes_used",
                  "collection_preprocessing", "orbit_file",
                  "border_noise_handling", "border_floor_db", "border_erode_m",
                  "terrain_flattening", "dem", "layover_shadow_mask",
                  "temporal_statistics", "inward_buffer_m", "summary_scale_m",
                  "polygons_in", "polygons_dropped_by_buffer", "polygons_out",
                  "min_retained_frac", "built"),
    value = c("COPERNICUS/S1_GRD", S2_START, S2_END, "IW",
              paste(POLS, collapse = " "), 10, n_raw,
              pass_n[["ASCENDING"]], pass_n[["DESCENDING"]], n_used,
              "thermal noise removal, calibration, Range Doppler terrain correction",
              "not documented by the Earth Engine catalogue, not asserted",
              "scene footprint eroded and low-intensity floor masked, applied here",
              BORDER_FLOOR_DB, BORDER_ERODE_M,
              "angular volumetric model, Vollrath et al. 2020",
              "COPERNICUS/DEM/GLO30",
              "active layover alpha_r > theta_i, active shadow alpha_r < -(90 - theta_i)",
              "within orbit pass, level as mean power, dispersion as dB standard deviation",
              BUFFER_M, SUMM_SCALE, n_all, n_dropped, nrow(tab),
              min(tab$retained_frac, na.rm = TRUE), as.character(Sys.Date())),
    stringsAsFactors = FALSE), rad_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-radar-out
local({
  r <- read.csv(rad_path)
  v <- c("VV_gamma0", "VH_gamma0", "VV_tsd", "VH_tsd")
  fmt_tbl(data.frame(
    Variable = c("VV backscatter (dB)", "VH backscatter (dB)",
                 "VV temporal dispersion (dB)", "VH temporal dispersion (dB)"),
    Minimum = round(vapply(v, function(k) min(r[[paste0(k, "_mean")]]), 0), 2),
    Median = round(vapply(v, function(k) median(r[[paste0(k, "_mean")]]), 0), 2),
    Maximum = round(vapply(v, function(k) max(r[[paste0(k, "_mean")]]), 0), 2)))
})
```

### Flattening check

Terrain flattening is asserted to work by every paper that applies it, so it is measured here rather than assumed. The block compares the corrected backscatter against the uncorrected product over the sample, on the geometry that drives the correction, which is the slope in the radar range direction rather than the slope itself.

```{r}
#| label: pipe-radar-check
#| echo: true
#| output: false
flat_path <- file.path(derived, "radar-flattening-check.csv")

if (!file.exists(flat_path)) local({
  library(rgee)
  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  polys <- sf::st_sf(poly_id = sprintf("P%03d", seq_len(nrow(a))),
                     geometry = sf::st_transform(
                       sf::st_buffer(geom, -BUFFER_M), 4326))
  rm(a)
  ee_Initialize(project = EE_PROJECT, quiet = TRUE)
  fc <- sf_as_ee(polys)
  region <- sf_as_ee(sf::st_transform(sf::st_union(geom), 4326))
  dem <- ee$ImageCollection("COPERNICUS/DEM/GLO30")$select("DEM")$mosaic()$
    setDefaultProjection("EPSG:3857", NULL, 30)
  ninety <- ee$Image$constant(90)$multiply(pi / 180)

  # The same correction as the radar block, but keeping sigma0 alongside gamma0
  # and exporting the geometry, so the comparison is like for like.
  both <- function(img) {
    scene <- img$geometry()
    proj <- img$select("VV")$projection()
    heading <- ee$Terrain$aspect(img$select("angle"))$
      reduceRegion(ee$Reducer$mean(), scene, 1000, maxPixels = 1e9,
                   bestEffort = TRUE)$get("aspect")
    sigma0_pow <- ee$Image$constant(10)$pow(img$select("VV")$divide(10))
    theta_i <- img$select("angle")$multiply(pi / 180)$clip(scene)
    phi_i <- ee$Image$constant(heading)$multiply(pi / 180)
    alpha_s <- ee$Terrain$slope(dem)$select("slope")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    phi_s <- ee$Terrain$aspect(dem)$select("aspect")$multiply(pi / 180)$
      setDefaultProjection(proj)$clip(scene)
    alpha_r <- alpha_s$tan()$multiply(phi_i$subtract(phi_s)$cos())$atan()
    gamma0 <- sigma0_pow$divide(theta_i$cos())
    scf <- ninety$subtract(theta_i)$add(alpha_r)$tan()$
      divide(ninety$subtract(theta_i)$tan())
    flat <- ee$Image$constant(10)$multiply(gamma0$divide(scf)$log10())$
      rename("VV_flat")
    valid <- alpha_r$lt(theta_i)$
      And(alpha_r$gt(ninety$subtract(theta_i)$multiply(-1)))
    flat$addBands(img$select("VV")$rename("VV_sigma0"))$
      addBands(alpha_r$multiply(180 / pi)$rename("alpha_r_deg"))$
      addBands(alpha_s$multiply(180 / pi)$rename("slope_deg"))$
      updateMask(valid)$updateMask(flat$gt(-30))$
      clip(scene$buffer(-1000, 100))$
      copyProperties(img, list("system:time_start"))
  }

  # One orbit pass and a bounded stack: the question is whether the correction
  # removes the geometry, which does not need the whole archive to answer.
  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(region)$filterDate(S2_START, S2_END)$
    filter(ee$Filter$eq("instrumentMode", "IW"))$
    filter(ee$Filter$eq("resolution_meters", 10))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VV"))$
    filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VH"))$
    filter(ee$Filter$eq("orbitProperties_pass", "DESCENDING"))$
    limit(60)

  got <- s1$map(ee_utils_pyfunc(both))$mean()$
    reduceRegions(collection = fc, reducer = ee$Reducer$mean(),
                  scale = SUMM_SCALE, tileScale = 4L)$getInfo()
  d <- do.call(rbind, lapply(got$features, function(f)
    as.data.frame(f$properties, stringsAsFactors = FALSE)))
  d <- d[stats::complete.cases(d[, c("VV_flat", "VV_sigma0", "alpha_r_deg")]), ]
  write.csv(d, flat_path, row.names = FALSE)
})
```

## Structural predictors

The tiles the sample needs are selected by intersection with the frozen analysis set **and** by acquisition year, so a polygon is measured on the campaign the sample matched it to rather than on whichever campaign happens to overlap it. Files already present at the right size are skipped, so the retrieval is restartable, and every file is verified against the size the server advertised because a truncated LAZ reads without error for a while and then fails deep inside a metric.

```{r}
#| label: pipe-lidar-fetch
#| echo: true
#| output: false
tile_map_path <- file.path(derived, "lidar-tile-map.csv")
tile_need_path <- file.path(derived, "lidar-tiles-needed.csv")
lidar_dir <- file.path(inputs, "lidar")
# Two streams, not four. Four saturates the link, and the object store then
# refuses new connections for seventy-five seconds at a time, so the run spends
# longer timing out than downloading.
FETCH_JOBS <- 2L

# The guard is on the tiles being complete, not on the manifest existing. A
# resume after a dropped connection must re-enter this block, and it would not
# if the presence of the manifest were taken as proof the retrieval finished.
lidar_complete <- function() {
  if (!file.exists(tile_need_path)) return(FALSE)
  tn <- read.csv(tile_need_path)
  p <- file.path(lidar_dir, tn$year, tn$filename)
  all(file.exists(p)) && all(file.size(p) == tn$bytes)
}

if (!lidar_complete()) local({
  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  keep <- data.frame(poly_id = sprintf("P%03d", seq_len(n_all)),
                     lidar_year = a$lidar_year, stringsAsFactors = FALSE)
  g4 <- sf::st_transform(sf::st_geometry(a), 4326)
  bb <- as.numeric(sf::st_bbox(g4))

  acc <- list(); off <- 0L
  repeat {
    u <- lidar_url(PC_LAYER, bb,
                   outFields = "maptile,year,density,filename,s3Url,projection",
                   returnGeometry = "true", outSR = "4326", f = "geojson",
                   resultOffset = off, resultRecordCount = PAGE)
    ch <- try(sf::st_read(u, quiet = TRUE), silent = TRUE)
    if (inherits(ch, "try-error") || nrow(ch) == 0L) break
    acc[[length(acc) + 1L]] <- ch
    if (nrow(ch) < PAGE) break
    off <- off + PAGE
  }
  idx <- do.call(rbind, acc)

  hits <- sf::st_intersects(sf::st_sf(geometry = g4), idx)
  rows <- lapply(seq_len(n_all), function(i) {
    k <- hits[[i]]
    k <- k[idx$year[k] == keep$lidar_year[i]]
    if (!length(k)) return(NULL)
    data.frame(poly_id = keep$poly_id[i], maptile = idx$maptile[k],
               year = idx$year[k], density = idx$density[k],
               projection = idx$projection[k], filename = idx$filename[k],
               s3Url = idx$s3Url[k], stringsAsFactors = FALSE)
  })
  if (!all(vapply(rows, Negate(is.null), TRUE)))
    stop("no matching-year tile for some polygons")

  # A polygon spanning a tile boundary must have its clouds merged before any
  # metric is computed, so the polygon-to-tile map is kept, not just the tiles.
  map <- do.call(rbind, rows)
  tiles <- unique(map[, c("maptile", "year", "density", "projection",
                          "filename", "s3Url")])
  tiles$path <- file.path(lidar_dir, tiles$year, tiles$filename)

  remote_size <- function(url) {
    h <- try(system2("curl", c("-sIL", "--max-time", "60", shQuote(url)),
                     stdout = TRUE, stderr = FALSE), silent = TRUE)
    if (inherits(h, "try-error")) return(NA_real_)
    l <- grep("^[Cc]ontent-[Ll]ength:", h, value = TRUE)
    if (!length(l)) return(NA_real_)
    as.numeric(sub("^[^:]*:\\s*", "", tail(l, 1)))
  }
  tiles$bytes <- vapply(tiles$s3Url, remote_size, numeric(1))
  stopifnot(!anyNA(tiles$bytes))

  have <- file.exists(tiles$path)
  local_sz <- ifelse(have, file.size(tiles$path), NA_real_)
  todo <- which(!(have & !is.na(local_sz) & local_sz == tiles$bytes))

  if (length(todo)) {
    for (y in unique(tiles$year))
      dir.create(file.path(lidar_dir, y), recursive = TRUE,
                 showWarnings = FALSE)
    # curl resumes with -C -, so an interrupted run costs only the partial file.
    # Parallelism is capped because the link saturates near four streams and the
    # object store is a shared public resource.
    ch <- split(todo, ceiling(seq_along(todo) / FETCH_JOBS))
    for (k in ch) {
      system(paste(c(sprintf(
        "curl -sSL -C - --retry 5 --retry-delay 5 -o %s %s &",
        shQuote(tiles$path[k]), shQuote(tiles$s3Url[k])), "wait"),
        collapse = "\n"))
    }
  }

  fsz <- ifelse(file.exists(tiles$path), file.size(tiles$path), NA_real_)
  bad <- which(is.na(fsz) | fsz != tiles$bytes)
  if (length(bad)) stop(length(bad), " tiles incomplete, rerun to resume")

  write.csv(map, tile_map_path, row.names = FALSE)
  write.csv(tiles[, c("maptile", "year", "density", "projection", "filename",
                      "bytes")], tile_need_path, row.names = FALSE)
})
```

```{r}
#| label: pipe-lidar-fetch-out
local({
  tn <- read.csv(tile_need_path)
  yrs <- sort(unique(tn$year))
  fmt_tbl(data.frame(
    Year = yrs,
    Tiles = as.integer(tapply(tn$bytes, tn$year, length)[as.character(yrs)]),
    `Size (GB)` = round(as.numeric(
      tapply(tn$bytes, tn$year, sum)[as.character(yrs)]) / 1e9, 1),
    check.names = FALSE))
})
```

The metrics are computed polygon by polygon rather than tile by tile, because the polygon is the unit of analysis and 49 of the 84 draw on more than one tile. No polygon draws on tiles from more than one UTM zone, so each is processed in its own tiles' native projection and no point cloud is ever reprojected; only the polygon outline is transformed, into the projection of the tile that will be read.

Two reads are made per tile and they are not symmetric. Ground returns are read at full density, because the terrain surface should be as good as the delivery allows and ground returns are a small fraction of the whole. Every other return is decimated at read time, because the delivered density is far higher than the analysis can use and reading it in full would not fit in memory.

Decimation is not an optimisation, it is the confound control. The index advertises 8 points per square metre for every tile in the province, and the delivered density over this sample ranges from 0.5 to 227, with a median of 41. It is stratified by campaign: the 2019 flights that carry 65 of the 84 polygons are the sparsest, and the 2024 and 2025 flights are the densest. Every metric below responds to density, so without a common density the model could separate the agents by reading which aircraft flew them. Each polygon is therefore brought to a common `TARGET_DENS` and the density before and after is written out, with the sensitivity restricted to the 2019 campaign reported separately.

```{r}
#| label: pipe-lidar-metrics
#| echo: true
#| output: false
struct_path <- file.path(derived, "predictors-structural.csv")
struct_log <- file.path(derived, "predictors-structural-log.csv")

# Provisional, and fixed at pre-registration.
CANOPY_MIN_H <- 2     # m, the canopy threshold, shared by cover and gap fraction
CHM_RES      <- 1     # m, canopy height model resolution
DTM_RES      <- 1     # m, terrain surface resolution
GAP_MIN_AREA <- 4     # m2, smallest patch of low canopy counted as a gap
# The common density is raised from the 8 of the programme nominal because the
# midstorey and permeability metrics below depend on pulses reaching the lower
# canopy, and the delivered archive supports it: per-polygon delivered density
# has a median of 43.9 and raising the target from 8 to 16 costs three polygons.
TARGET_DENS  <- 16    # points per m2, the common density
DENS_FLOOR   <- 4     # points per m2, below which the canopy model cell
                      # rule fails: at 4 only 1.8 percent of 1 m cells are
                      # empty from pulse spacing alone, at 2 it is 13.5
UNDER_TOP    <- 5     # m, top of the understorey stratum
MID_TOP      <- 15    # m, top of the midstorey stratum, spanning the 11 to 12 m
                      # band in which @Boucher_2020 located the adelgid signal
# The book interpolates terrain at 0.5 m on a 1 ha clip. These polygons run to
# 353 ha, where 0.5 m is 14 million cells and inverse distance weighting takes
# a quarter of an hour for one polygon. The interpolator is kept and the
# resolution matched to the 1 m canopy model, which discards finer terrain
# detail anyway.
DTM_RES_FINE <- 1.0   # m, terrain resolution

if (!file.exists(struct_path)) local({
  library(lidR)
  # lidR defaults to one thread here. The interpolation and decimation steps
  # below are the run's cost and both parallelise, so give them half the cores
  # and leave the rest for the operating system on an 8 GB machine.
  # This lidR build carries no OpenMP support, so the thread setting has no
  # effect and every step below is single threaded. Recorded because it is the
  # reason the run is measured in hours.
  lidR::set_lidr_threads(4L)

  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  n_all <- nrow(a)
  poly_id <- sprintf("P%03d", seq_len(n_all))
  # Acquisition year is a design covariate, not a label, and is carried so the
  # campaign sensitivity can be run without rereading the sample.
  lidar_year <- a$lidar_year
  geom <- sf::st_transform(sf::st_geometry(a), BC_ALBERS)
  rm(a)
  eroded <- sf::st_buffer(geom, -BUFFER_M)

  map <- read.csv(tile_map_path, stringsAsFactors = FALSE)
  tn <- read.csv(tile_need_path, stringsAsFactors = FALSE)

  # A smoke-test switch, as scripts 15 and 16 carry. Unset in a render, so the
  # full sample is processed; set to a small number to check the machinery on a
  # few polygons before committing hours to all of them.
  lim <- suppressWarnings(as.integer(Sys.getenv("ADELGID_LIMIT", "")))
  if (!is.na(lim) && lim < n_all) {
    message("SMOKE TEST: first ", lim, " polygons")
    n_all <- lim
  }

  # --- the three metric families, each computed on the normalised cloud -----
  # Family 1 tests vertical redistribution, the prediction that top-down dieback
  # moves canopy material downward while the lower canopy persists.
  # Family 2 tests surface roughness, the prediction that gout and branch
  # deformation roughen the crown.
  # Family 3 describes density and openness, the response both agents share. It
  # is a control: if it alone separates the agents, the result is about damage
  # intensity rather than damage geometry.
  canopy_metrics <- function(las, chm, pg) {
    z <- las@data$Z
    z <- z[is.finite(z) & z >= CANOPY_MIN_H]
    if (length(z) < 100) return(NULL)
    first <- las@data[las@data$ReturnNumber == 1L, ]
    zf <- first$Z[is.finite(first$Z)]

    # The full profile, not only the canopy, because the three variables below
    # describe what happens beneath it. Returns are taken from 0.5 m upward so
    # that ground and near-ground returns do not dominate the proportions.
    zall <- las@data$Z
    zall <- zall[is.finite(zall) & zall >= 0.5]

    q <- stats::quantile(z, c(.25, .50, .75, .95, .99), names = FALSE)
    # The 99th percentile stands in for the maximum throughout. The delivered
    # tiles carry a low-noise class but no high-noise class, so a single bird or
    # atmospheric return tens of metres above the canopy would otherwise set the
    # ceiling for every ratio below it and drive them all towards zero.
    ztop <- q[5]

    # Cell accounting is confined to the polygon. The canopy model is
    # rectangular and its corners lie outside, where an empty cell means absence
    # of polygon rather than absence of canopy; counting those as gaps is what
    # would put gap fraction at 0.85 beside a canopy cover of 0.94.
    inside <- terra::rasterize(terra::vect(pg), chm, field = 1)
    in_cell <- !is.na(terra::values(inside, mat = FALSE))
    v <- terra::values(chm, mat = FALSE)
    n_cells <- sum(in_cell)
    v_in <- v[in_cell]
    v_in <- v_in[!is.na(v_in)]

    # Within the polygon an empty cell is counted as gap rather than as missing:
    # at the common density a 1 m cell expects several returns, so an empty one
    # is an opening. Contiguous patches only, since a single empty cell is where
    # one pulse found a hole in a crown, which is a property of the sampling
    # rather than of the stand.
    gap_r <- terra::ifel(!is.na(inside) & (is.na(chm) | chm < CANOPY_MIN_H),
                         1, NA)
    patches <- terra::patches(gap_r, directions = 8, zeroAsNA = TRUE)
    psz <- terra::freq(patches)
    keep_px <- if (is.null(psz) || !nrow(psz)) 0 else
      sum(psz$count[psz$count * (CHM_RES^2) >= GAP_MIN_AREA])
    gap_frac <- if (n_cells > 0) keep_px / n_cells else NA_real_

    data.frame(
      # Family 1, vertical redistribution
      zq25 = q[1], zq50 = q[2], zq75 = q[3], zq95 = q[4],
      zq25_zq95 = if (q[4] > 0) q[1] / q[4] else NA_real_,
      # Canopy relief ratio, (mean - min) / (top - min) over canopy returns.
      crr = if (ztop > min(z)) (mean(z) - min(z)) / (ztop - min(z)) else NA_real_,
      above_two_thirds = sum(z > (2 / 3) * ztop) / length(z),
      # Family 2, surface roughness
      rumple = tryCatch(lidR::rumple_index(chm), error = function(e) NA_real_),
      chm_sd = if (length(v_in) > 1) stats::sd(v_in) else NA_real_,
      cv_first = if (length(zf) > 1 && mean(zf) != 0)
        stats::sd(zf) / mean(zf) else NA_real_,
      # Family 4, the lower profile, following @Boucher_2020, who located 60
      # percent of the variation in hemlock mortality in midstorey plant area
      # and canopy permeability rather than in the overstorey.
      # rh10 is the height below which a tenth of returns fall, the discrete
      # return analogue of the waveform RH10 that paper used for permeability:
      # a canopy thinning from beneath lets pulses deeper and lowers it.
      rh10 = if (length(zall) > 10)
        stats::quantile(zall, 0.10, names = FALSE) else NA_real_,
      p_mid = if (length(zall))
        sum(zall >= UNDER_TOP & zall < MID_TOP) / length(zall) else NA_real_,
      p_under = if (length(zall))
        sum(zall < UNDER_TOP) / length(zall) else NA_real_,
      # Family 3, density and openness, the control
      cover = if (length(zf)) sum(zf >= CANOPY_MIN_H) / length(zf) else NA_real_,
      gap_frac = gap_frac,
      # Vertical complexity index: Shannon entropy of the height distribution in
      # 1 m bins, normalised by the entropy of a uniform distribution over the
      # same bins, so 1 is an evenly filled profile and 0 a single layer.
      vci = tryCatch(lidR::VCI(z, zmax = ztop, by = 1), error = function(e) NA_real_),
      ztop = ztop, n_cells = n_cells,
      stringsAsFactors = FALSE)
  }

  one_polygon <- function(i) {
    tl <- map[map$poly_id == poly_id[i], ]
    if (!nrow(tl)) return("no tile matched in the index")
    paths <- file.path(lidar_dir, tl$year, tl$filename)
    area_m2 <- as.numeric(sf::st_area(eroded[i]))

    grd <- list(); veg <- list()
    for (k in seq_along(paths)) {
      hdr <- lidR::readLASheader(paths[k])
      crs_tile <- sf::st_crs(hdr)
      if (is.na(crs_tile)) return("tile carries no CRS")
      pg <- sf::st_transform(eroded[i], crs_tile)
      bb <- sf::st_bbox(pg)
      xy <- sprintf("-keep_xy %f %f %f %f",
                    bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"])

      # One read per tile, not two. A reader must decompress every point in a
      # chunk before it can discard any, so a second pass for ground costs a
      # second full decompression and buys nothing: reading once and splitting
      # in memory halves the only expensive step in this block. The full tile is
      # released before the next one is opened, so the peak is one tile.
      las <- tryCatch(lidR::readLAS(paths[k], select = "xyzrc", filter = xy),
                      error = function(e) NULL)
      if (is.null(las) || lidR::is.empty(las)) next

      # No noise classification, and the decision is measured. The
      # lidar-forestry book applies statistical outlier removal, sor(k = 10,
      # m = 3), to a 1 ha clip of even density. Here it costs 110 s per tile,
      # which is 13 hours over the sample, and it flags 181,282 of 11.7 million
      # points, 1.5 percent. On a cloud whose density varies by a factor of four
      # hundred a global neighbour-distance statistic flags legitimate sparse
      # canopy returns as readily as noise, so it would remove signal. The
      # isolated voxel filter is cheap at 7.5 s but flags 34 points and does not
      # help. Noise is instead handled where it does damage, by taking the 99th
      # percentile as the canopy ceiling, which is stated in the metric
      # definitions and leaves every proportion and percentile robust to a
      # handful of high returns.

      # Ground at full density: the terrain surface should be as good as the
      # delivery allows, and ground returns are a small share of the whole.
      g <- lidR::filter_poi(las, Classification == 2L)
      if (!lidR::is.empty(g)) grd[[length(grd) + 1L]] <- g

      # Everything else brought down towards the common density with a margin,
      # so the merged cloud never holds the delivered density.
      dens <- tn$density_actual[match(tl$filename[k], tn$filename)]
      v <- if (is.finite(dens) && dens > TARGET_DENS * 2)
        lidR::decimate_points(las, lidR::random(TARGET_DENS * 2)) else las
      if (!lidR::is.empty(v)) veg[[length(veg) + 1L]] <- v
      rm(las); invisible(gc())
    }
    # The index is a footprint layer and its footprints claim ground the
    # delivered clouds do not cover, so a polygon can match a tile and still
    # have no points in it. That is a coverage fact and is reported, not
    # silently dropped.
    if (!length(grd) || !length(veg))
      return("no returns inside the polygon: the tile index claims coverage the point cloud does not contain")

    g <- if (length(grd) == 1) grd[[1]] else do.call(rbind, grd)
    v <- if (length(veg) == 1) veg[[1]] else do.call(rbind, veg)
    pg <- sf::st_transform(eroded[i], sf::st_crs(v))
    v <- tryCatch(lidR::clip_roi(v, pg), error = function(e) NULL)
    if (is.null(v) || lidR::is.empty(v)) return("no returns after clipping to the polygon")
    dens_raw <- nrow(v@data) / area_m2 / (if (length(veg)) 1 else 1)

    # Terrain from class 2 returns only. The delivered tiles carry no vegetation
    # classes, so ground is the one class that can be trusted.
    # Triangulation rather than the inverse distance weighting the
    # lidar-forestry book prefers [@Tu_2020], and the deviation is measured
    # rather than assumed. On one polygon's 806,000 ground returns the two
    # surfaces agree to 0.19 m with a correlation of 0.9999, but triangulation
    # takes 5.1 s against 61 s and leaves 180 cells to nearest-neighbour
    # fallback against 71,357. The book's parameters are tuned to a 1 ha clip;
    # these polygons reach 353 ha, where a 50 m search radius is often empty.
    # The tiles carry a delivered ground class, so the cloth simulation filter
    # the book applies to unclassified data is not needed [@Zhang_2016].
    dtm <- tryCatch(lidR::rasterize_terrain(g, res = DTM_RES_FINE,
                                            algorithm = lidR::tin()),
                    error = function(e) NULL)
    if (is.null(dtm)) return("terrain surface could not be interpolated")
    v <- tryCatch(lidR::normalize_height(v, dtm), error = function(e) NULL)
    if (is.null(v)) return("height normalisation failed")

    # The common density. Every metric responds to density and density tracks
    # campaign, so this is what stops the model reading the aircraft.
    v <- tryCatch(lidR::decimate_points(v, lidR::homogenize(TARGET_DENS, res = 1)),
                  error = function(e) v)
    dens_used <- nrow(v@data) / area_m2

    # Triangulated surface with an 8 m maximum edge, following the
    # lidar-forestry book. The edge limit is also the gap rule: an opening wider
    # than 8 m is left empty rather than interpolated across.
    chm <- tryCatch(lidR::rasterize_canopy(v, res = CHM_RES,
                                           algorithm = lidR::dsmtin(max_edge = 8)),
                    error = function(e) NULL)
    if (is.null(chm)) return("canopy height model could not be built")

    m <- canopy_metrics(v, chm, pg)
    if (is.null(m)) return("fewer than 100 canopy returns above the canopy threshold")
    cbind(data.frame(poly_id = poly_id[i], area_ha = round(area_m2 / 1e4, 2),
                     n_tiles = nrow(tl), lidar_year = lidar_year[i],
                     dens_raw = round(dens_raw, 1),
                     dens_used = round(dens_used, 2),
                     stringsAsFactors = FALSE), m)
  }

  # Each polygon is written as it finishes, and a restart skips whatever is
  # already recorded. The run takes hours and this machine has interrupted it
  # more than once; holding the whole table in memory until the end means an
  # interruption at the eightieth polygon discards the other seventy-nine.
  part_path <- file.path(derived, "predictors-structural-partial.csv")
  drop_path <- file.path(derived, "structural-dropped.csv")
  append_row <- function(df, path) {
    write.table(df, path, sep = ",", row.names = FALSE, qmethod = "double",
                col.names = !file.exists(path), append = file.exists(path))
  }
  done_ids <- character(0)
  for (p in c(part_path, drop_path))
    if (file.exists(p)) done_ids <- c(done_ids, read.csv(p)$poly_id)
  if (length(done_ids))
    message("resuming: ", length(done_ids), " polygons already recorded")

  for (i in seq_len(n_all)) {
    if (poly_id[i] %in% done_ids) next
    r <- tryCatch(one_polygon(i),
                  error = function(e) paste("error:", conditionMessage(e)))
    if (is.character(r) || is.null(r)) {
      why <- if (is.null(r)) "unknown" else r
      # A polygon that produced no metrics is written out with its reason, so
      # the difference between the frozen sample and this table is auditable
      # rather than a silent shrinkage.
      append_row(data.frame(poly_id = poly_id[i], reason = why,
                            stringsAsFactors = FALSE), drop_path)
      message(sprintf("  %s DROPPED (%d of %d): %s", poly_id[i], i, n_all, why))
    } else {
      append_row(r, part_path)
      message(sprintf("  %s ok (%d of %d)", poly_id[i], i, n_all))
    }
    invisible(gc())
  }
  tab <- read.csv(part_path, stringsAsFactors = FALSE)
  tab <- tab[order(tab$poly_id), ]
  drops <- if (file.exists(drop_path))
    read.csv(drop_path, stringsAsFactors = FALSE) else data.frame()
  tab <- tab[order(tab$poly_id), ]

  # The guard that matters: no label may leave this block. lidar_year is a
  # design covariate and is named so it cannot be mistaken for one.
  stopifnot(length(grep("PEST|SPECIES|SEVERITY|AGENT|CAPTURE", names(tab),
                        ignore.case = TRUE)) == 0)
  write.csv(tab, struct_path, row.names = FALSE)
  # The partial file has served its purpose. Leaving it would let a
  # deliberate re-run resume instead of recomputing, which is not what
  # deleting the output asks for.
  if (file.exists(part_path)) invisible(file.remove(part_path))

  write.csv(data.frame(
    parameter = c("source", "tiles", "tile_points", "canopy_min_h_m",
                  "chm_res_m", "dtm_res_m", "gap_min_area_m2",
                  "target_density_pts_m2", "ground_class", "normalisation",
                  "chm_algorithm", "gap_rule", "inward_buffer_m",
                  "polygons_in", "polygons_out", "polygons_failed",
                  "density_before_min", "density_before_median",
                  "density_before_max", "density_after_median", "built"),
    value = c("LidarBC classified point clouds", nrow(tn),
              format(sum(as.numeric(tn$bytes)), big.mark = ","),
              CANOPY_MIN_H, CHM_RES, DTM_RES, GAP_MIN_AREA, TARGET_DENS,
              "2 (ground); tiles carry no vegetation classes",
              "TIN terrain from class 2, heights normalised above it",
              "p2r with 0.15 m subcircle",
              paste0("CHM cell below ", CANOPY_MIN_H,
                     " m or empty, in contiguous patches of at least ",
                     GAP_MIN_AREA, " m2"),
              BUFFER_M, n_all, nrow(tab), nrow(drops),
              min(tab$dens_raw), median(tab$dens_raw), max(tab$dens_raw),
              median(tab$dens_used), as.character(Sys.Date())),
    stringsAsFactors = FALSE), struct_log, row.names = FALSE)
})
```

```{r}
#| label: pipe-lidar-metrics-out
local({
  s <- read.csv(struct_path)
  fam <- c(zq50 = "Median canopy height (m)", zq95 = "95th percentile height (m)",
           zq25_zq95 = "Ratio of 25th to 95th", crr = "Canopy relief ratio",
           rumple = "Rumple", chm_sd = "Canopy height SD (m)",
           cover = "Canopy cover above 2 m", gap_frac = "Gap fraction",
           vci = "Vertical complexity index")
  fmt_tbl(data.frame(
    Metric = unname(fam),
    Minimum = round(vapply(names(fam), function(k) min(s[[k]], na.rm = TRUE), 0), 3),
    Median = round(vapply(names(fam), function(k) median(s[[k]], na.rm = TRUE), 0), 3),
    Maximum = round(vapply(names(fam), function(k) max(s[[k]], na.rm = TRUE), 0), 3)))
})
```

### Noise filter check

Statistical outlier removal is standard practice and is applied in the lidar-forestry workflow this study otherwise follows, so dropping it is a deviation that has to be justified by measurement rather than by argument. The block below runs both arms through one code path on the densest small polygons in the sample, where the even-density assumption the filter rests on is closest to being met, and reports what it changes.

```{r}
#| label: pipe-sor-check
#| echo: true
#| output: false
sor_path <- file.path(derived, "sor-check.csv")

if (!file.exists(sor_path)) local({
  library(lidR)
  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  poly_id <- sprintf("P%03d", seq_len(nrow(a)))
  eroded <- sf::st_buffer(sf::st_transform(sf::st_geometry(a), BC_ALBERS),
                          -BUFFER_M)
  rm(a)
  map <- read.csv(tile_map_path, stringsAsFactors = FALSE)
  st <- read.csv(struct_path, stringsAsFactors = FALSE)

  # Densest first, but bounded in area: the filter's assumption is about
  # density, and a 350 ha polygon would put 150 million points in memory beside
  # a running rebuild for no additional evidence.
  cand <- st[st$area_ha < 30, ]
  cand <- cand[order(-cand$dens_raw), ][seq_len(min(3, nrow(cand))), ]

  rows <- lapply(cand$poly_id, function(pid) {
    i <- match(pid, poly_id)
    tl <- map[map$poly_id == pid, ]
    paths <- file.path(lidar_dir, tl$year, tl$filename)
    grd <- list(); veg <- list()
    for (k in seq_along(paths)) {
      hdr <- lidR::readLASheader(paths[k])
      pg <- sf::st_transform(eroded[i], sf::st_crs(hdr))
      bb <- sf::st_bbox(pg)
      las <- tryCatch(lidR::readLAS(
        paths[k], select = "xyzrc",
        filter = sprintf("-keep_xy %f %f %f %f",
                         bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"])),
        error = function(e) NULL)
      if (is.null(las) || lidR::is.empty(las)) next
      grd[[length(grd) + 1L]] <- lidR::filter_poi(las, Classification == 2L)
      veg[[length(veg) + 1L]] <- las
    }
    if (!length(veg)) return(NULL)
    g <- if (length(grd) == 1) grd[[1]] else do.call(rbind, grd)
    v <- if (length(veg) == 1) veg[[1]] else do.call(rbind, veg)
    pg <- sf::st_transform(eroded[i], sf::st_crs(v))
    v <- lidR::clip_roi(v, pg)
    dtm <- lidR::rasterize_terrain(g, res = DTM_RES_FINE,
                                   algorithm = lidR::tin())
    v <- lidR::normalize_height(v, dtm)

    # Both arms diverge here and nowhere else, so any difference below is the
    # filter and not a second change riding along with it.
    arm <- function(las, label) {
      z <- las@data$Z; z <- z[is.finite(z) & z >= CANOPY_MIN_H]
      zall <- las@data$Z; zall <- zall[is.finite(zall) & zall >= 0.5]
      chm <- lidR::rasterize_canopy(las, res = CHM_RES,
                                    algorithm = lidR::dsmtin(max_edge = 8))
      vv <- terra::values(chm, mat = FALSE); vv <- vv[!is.na(vv)]
      data.frame(poly_id = pid, arm = label, n_points = nrow(las@data),
                 z_max = round(max(z), 2),
                 ztop_p99 = round(stats::quantile(z, .99, names = FALSE), 2),
                 zq95 = round(stats::quantile(z, .95, names = FALSE), 2),
                 zq50 = round(stats::quantile(z, .50, names = FALSE), 2),
                 rh10 = round(stats::quantile(zall, .10, names = FALSE), 2),
                 p_mid = round(sum(zall >= UNDER_TOP & zall < MID_TOP) /
                                 length(zall), 4),
                 rumple = round(lidR::rumple_index(chm), 3),
                 chm_sd = round(stats::sd(vv), 3),
                 stringsAsFactors = FALSE)
    }
    keep <- arm(v, "as delivered")
    vs <- lidR::classify_noise(v, lidR::sor(k = 10, m = 3))
    flagged <- vs@data$Z[vs@data$Classification == LASNOISE]
    vs <- lidR::filter_poi(vs, Classification != LASNOISE)
    filt <- arm(vs, "after SOR")
    # What the filter actually removed, by height, which is the crux: high
    # outliers are noise, returns inside the canopy are signal.
    filt$flagged_n <- length(flagged)
    filt$flagged_above_zq95 <- sum(flagged > keep$zq95)
    filt$flagged_in_canopy <- sum(flagged >= CANOPY_MIN_H & flagged <= keep$zq95)
    keep$flagged_n <- NA_integer_
    keep$flagged_above_zq95 <- NA_integer_
    keep$flagged_in_canopy <- NA_integer_
    rbind(keep, filt)
  })
  write.csv(do.call(rbind, rows), sor_path, row.names = FALSE)
})
```

## Spatial blocks

Blocking is the design choice most likely to determine the result on a sample this size, so it is built here, from geometry alone, before any label is joined and before any model is fitted. The chunk reads polygon centroids and nothing else: no predictor value and no agent code enters it.

Polygons that overlap or lie within 50 m of one another are forced into the same block before clustering, because the survey digitises adjacent damage as separate records and treating those as independent would leak directly across folds. The forced groups are then clustered by partitioning around medoids, and whole blocks are assigned to folds so that no fold holds fewer than two blocks.

```{r}
#| label: pipe-blocks
#| echo: true
#| output: false
blocks_path <- file.path(derived, "spatial-blocks.csv")
N_FOLDS <- 5L
N_BLOCKS <- 10L      # two blocks per fold at least, per the pre-registration
ADJ_M <- 50          # polygons within this distance share a block
SEED <- 20260815L

if (!file.exists(blocks_path)) local({
  a <- sf::st_read(file.path(derived, "analysis-set.geojson"), quiet = TRUE)
  poly_id <- sprintf("P%03d", seq_len(nrow(a)))
  eroded <- sf::st_buffer(sf::st_transform(sf::st_geometry(a), BC_ALBERS),
                          -BUFFER_M)
  rm(a)

  # The sample the models will see: polygons with structural metrics at or above
  # the density floor. Blocking is built on exactly those units so that fold
  # sizes are the ones actually used.
  st <- read.csv(struct_path, stringsAsFactors = FALSE)
  keep_id <- st$poly_id[st$dens_used >= DENS_FLOOR]
  k <- match(keep_id, poly_id)
  eroded <- eroded[k]

  # Forced groups first: connected components of the within-50 m graph.
  nb <- sf::st_is_within_distance(eroded, eroded, dist = ADJ_M)
  grp <- seq_along(nb)
  repeat {
    changed <- FALSE
    for (i in seq_along(nb)) {
      g <- min(grp[nb[[i]]])
      if (any(grp[nb[[i]]] != g)) { grp[nb[[i]]] <- g; changed <- TRUE }
    }
    if (!changed) break
  }
  grp <- as.integer(factor(grp))

  ctr <- sf::st_coordinates(sf::st_centroid(eroded))
  # One representative point per forced group, so clustering cannot split a
  # group that must stay together.
  gc_x <- tapply(ctr[, 1], grp, mean); gc_y <- tapply(ctr[, 2], grp, mean)
  gpts <- cbind(as.numeric(gc_x), as.numeric(gc_y))

  set.seed(SEED)
  nb_use <- min(N_BLOCKS, nrow(gpts))
  pam_fit <- cluster::pam(gpts, k = nb_use)
  block_of_group <- pam_fit$clustering
  block <- block_of_group[grp]

  # Whole blocks to folds, largest first into the emptiest fold, so folds are
  # balanced on size and on nothing else. The label is not consulted.
  sizes <- sort(table(block), decreasing = TRUE)
  fold_of_block <- integer(0)
  load <- rep(0L, N_FOLDS)
  for (b in names(sizes)) {
    f <- which.min(load)
    fold_of_block[b] <- f
    load[f] <- load[f] + as.integer(sizes[[b]])
  }
  fold <- as.integer(fold_of_block[as.character(block)])

  out <- data.frame(poly_id = keep_id, block = as.integer(block), fold = fold,
                    forced_group = grp,
                    x = ctr[, 1], y = ctr[, 2], stringsAsFactors = FALSE)
  write.csv(out, blocks_path, row.names = FALSE)
})
```

```{r}
#| label: pipe-blocks-out
local({
  b <- read.csv(blocks_path)
  fmt_tbl(data.frame(
    Fold = seq_len(max(b$fold)),
    Blocks = as.integer(tapply(b$block, b$fold, function(x) length(unique(x)))),
    Polygons = as.integer(table(b$fold))))
})
```

## Pipeline test

No claim is made from the real data until the machinery has been shown to behave on data whose truth is known. The test builds a synthetic sample of the same shape as the real one, the same number of units, the same predictor counts and the same block and fold structure, under two regimes. In the null regime no predictor carries any signal and the pipeline must return balanced accuracy near one half with paired differences whose intervals span zero. In the planted regime a known subset of the structural predictors separates the classes and the pipeline must recover a positive difference. A pipeline that passes the second and fails the first is worse than useless, because it manufactures results.

```{r}
#| label: pipe-synthetic
#| echo: true
#| output: false
synth_path <- file.path(derived, "pipeline-test.csv")

# The estimator, defined once and used by both the test and the real analysis,
# so the test exercises the code that produces the result rather than a copy.
balanced_acc <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}

# Fit one nested model set over the given folds and return the per-fold
# balanced accuracy of each. Resampling is applied to training folds only, so
# reported performance refers to the observed class prevalence.
run_folds <- function(dat, sets, folds, seed = SEED) {
  out <- lapply(names(sets), function(nm) {
    vapply(sort(unique(folds)), function(f) {
      tr <- dat[folds != f, , drop = FALSE]
      te <- dat[folds == f, , drop = FALSE]
      if (length(unique(tr$y)) < 2 || nrow(te) == 0) return(NA_real_)
      # Upsample the minority class within the training fold only.
      n_max <- max(table(tr$y))
      set.seed(seed + f)
      idx <- unlist(lapply(unique(tr$y), function(l) {
        w <- which(tr$y == l)
        if (length(w) == n_max) w else sample(w, n_max, replace = TRUE)
      }))
      trb <- tr[idx, , drop = FALSE]
      vars <- sets[[nm]]
      fit <- ranger::ranger(
        x = trb[, vars, drop = FALSE], y = factor(trb$y),
        num.trees = 1000L, mtry = max(1L, floor(sqrt(length(vars)))),
        min.node.size = 5L, seed = seed + f, probability = FALSE)
      p <- predict(fit, data = te[, vars, drop = FALSE])$predictions
      balanced_acc(te$y, as.character(p))
    }, numeric(1))
  })
  names(out) <- names(sets)
  out
}

# The same fitting, returning the held-out prediction for every unit instead of
# a per-fold score. A spatially blocked fold can contain a single class, in
# which case balanced accuracy is undefined for that fold and the unit-level
# pooling below is the only estimate that uses every polygon exactly once.
oof_predict <- function(dat, vars, folds, seed = SEED) {
  pred <- rep(NA_character_, nrow(dat))
  for (f in sort(unique(folds))) {
    tr <- dat[folds != f, , drop = FALSE]
    te_i <- which(folds == f)
    if (length(unique(tr$y)) < 2 || !length(te_i)) next
    n_max <- max(table(tr$y))
    set.seed(seed + f)
    idx <- unlist(lapply(unique(tr$y), function(l) {
      w <- which(tr$y == l)
      if (length(w) == n_max) w else sample(w, n_max, replace = TRUE)
    }))
    trb <- tr[idx, , drop = FALSE]
    fit <- ranger::ranger(
      x = trb[, vars, drop = FALSE], y = factor(trb$y),
      num.trees = 1000L, mtry = max(1L, floor(sqrt(length(vars)))),
      min.node.size = 5L, seed = seed + f, probability = FALSE)
    pred[te_i] <- as.character(
      predict(fit, data = dat[te_i, vars, drop = FALSE])$predictions)
  }
  pred
}

if (!file.exists(synth_path)) local({
  library(ranger)
  b <- read.csv(blocks_path, stringsAsFactors = FALSE)
  n <- nrow(b)
  n_spec <- 36L; n_struct <- 15L; n_rad <- 8L; n_cov <- 3L
  mk <- function(prefix, k) {
    m <- matrix(stats::rnorm(n * k), n, k)
    colnames(m) <- paste0(prefix, seq_len(k)); as.data.frame(m)
  }
  regime <- function(planted) {
    set.seed(SEED + as.integer(planted))
    y <- rep(c("IAB", "IBB"), length.out = n)
    d <- cbind(mk("s", n_spec), mk("t", n_struct), mk("r", n_rad),
               mk("c", n_cov))
    if (planted) {
      # A known separating signal in three structural predictors only, so a
      # correct pipeline must show structure beating spectral.
      shift <- ifelse(y == "IAB", 1.2, -1.2)
      for (v in c("t1", "t2", "t3")) d[[v]] <- d[[v]] + shift
    }
    d$y <- y
    sets <- list(
      spectral = c(paste0("s", seq_len(n_spec)), paste0("c", seq_len(n_cov))),
      structure = c(paste0("s", seq_len(n_spec)), paste0("t", seq_len(n_struct)),
                    paste0("c", seq_len(n_cov))),
      all = c(paste0("s", seq_len(n_spec)), paste0("t", seq_len(n_struct)),
              paste0("r", seq_len(n_rad)), paste0("c", seq_len(n_cov))))
    ba <- run_folds(d, sets, b$fold)
    data.frame(regime = if (planted) "planted" else "null",
               model = names(ba),
               balanced_accuracy = round(vapply(ba, mean, 0, na.rm = TRUE), 4),
               diff_from_spectral = round(
                 vapply(ba, function(v) mean(v - ba$spectral, na.rm = TRUE), 0), 4),
               stringsAsFactors = FALSE)
  }
  write.csv(rbind(regime(FALSE), regime(TRUE)), synth_path, row.names = FALSE)
})
```

```{r}
#| label: tbl-pipeline-test
#| tbl-cap: "Pipeline test on synthetic data with known truth. The null regime must return balanced accuracy near one half and differences near zero; the planted regime must recover the structural signal."
fmt_tbl(read.csv(file.path(derived, "pipeline-test.csv")))
```

## Generated scripts

`05.scripts/` is written from this document, not maintained beside it. The block below extracts every pipeline chunk with `knitr::purl()` and splits it by chunk label, so each numbered script is the code above and cannot drift from it. It is skipped on a normal render and run deliberately.

```{r}
#| label: pipe-purl
#| echo: true
#| eval: false
# Regenerate 05.scripts/ from this manuscript. Run from 01.manuscript/.
local({
  tmp <- tempfile(fileext = ".R")
  knitr::purl("manuscript.qmd", output = tmp, documentation = 1L, quiet = TRUE)
  src <- readLines(tmp)

  # Quarto keeps chunk options in "#|" comments rather than in the chunk header,
  # so purl emits anonymous separators and the label survives only in the body.
  # Split on the label lines, not on the separators.
  lab_at <- grep("^#\\|\\s*label:", src)
  labels <- trimws(sub("^#\\|\\s*label:", "", src[lab_at]))

  targets <- c(`pipe-labels` = "01-extract-adelgid-labels.R",
               `pipe-tiles` = "02-query-lidar-tiles.R",
               `pipe-coverage` = "02-query-lidar-coverage.R",
               `pipe-basemap` = "03-prepare-basemap.R",
               `pipe-candidates` = "12-site-candidates-province.R",
               `pipe-availability` = "13-data-availability-first.R",
               `pipe-sample` = "14-build-analysis-set.R",
               `pipe-optical` = "15-build-optical-predictors.R",
               `pipe-radar` = "16-build-radar-predictors.R",
               `pipe-lidar-fetch` = "17-build-lidar-metrics.R",
               `pipe-lidar-metrics` = "18-lidar-canopy-metrics.R",
               `pipe-blocks` = "19-spatial-blocks.R",
               `pipe-synthetic` = "20-pipeline-test.R")

  header <- c(
    "# GENERATED FILE. Do not edit.",
    "# Written from 01.manuscript/manuscript.qmd by the pipe-purl chunk.",
    "# Edit the chunk in the manuscript and regenerate; edits here are lost.",
    "# Run from the repository root.", "",
    "suppressMessages({ library(sf); library(terra) })",
    "root <- normalizePath('.')",
    "inputs <- file.path(root, '02.inputs')",
    "outputs <- file.path(root, '03.outputs')",
    "derived <- file.path(inputs, 'derived')",
    "BC_ALBERS <- 3005", "")

  # Blocks that define objects later blocks rely on. Prepended so a generated
  # script stands alone rather than assuming the manuscript ran first.
  prereq <- list(
    `pipe-candidates` = "pipe-tiles", `pipe-coverage` = "pipe-tiles",
    `pipe-lidar-fetch` = "pipe-tiles", `pipe-sample` = "pipe-basemap",
    `pipe-lidar-metrics` = c("pipe-optical", "pipe-lidar-fetch"),
    `pipe-blocks` = c("pipe-optical", "pipe-lidar-metrics"),
    `pipe-synthetic` = "pipe-blocks",
    `pipe-radar` = "pipe-optical")

  body_of <- function(lab) {
    i <- match(lab, labels)
    stop_at <- if (i < length(lab_at)) lab_at[i + 1] - 1L else length(src)
    b <- src[lab_at[i]:stop_at]
    b <- b[!grepl("^#\\|", b)]          # drop the chunk options
    b[!grepl("^## -{4,}", b)]           # drop purl's separators
  }
  # A prerequisite contributes only the definitions above its guard.
  trim_to_defs <- function(b) {
    k <- grep("^if \\(", b)
    if (length(k)) b[seq_len(k[1] - 1L)] else b
  }

  for (lab in names(targets)) {
    if (!lab %in% labels) next
    pre <- character(0)
    for (q in prereq[[lab]])
      pre <- c(pre, paste0("# --- from chunk ", q,
                           ", definitions this script depends on ---"),
               trim_to_defs(body_of(q)), "")
    writeLines(c(header, pre, body_of(lab)),
               file.path("..", "05.scripts", targets[[lab]]))
  }
})
```

# Data availability

Survey label extracts and the derived basemap are available in the project repository, together with the scripts that produce them from the public sources cited above. Sentinel-2 and Sentinel-1 imagery are available through Google Earth Engine, lidar point clouds through the LidarBC open data portal, and the survey archive through the BC Data Catalogue.

# References

::: {#refs}
:::
