---
title: "Separating balsam woolly adelgid from bark beetle damage in Pacific silver fir on Vancouver Island: a multi-sensor attribution test [TITLE PROPOSED, NOT FIXED. Seamus fixes the title; this placeholder is a description of the study, not a decision.]"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: |
  [ABSTRACT PENDING. Written last, from the rendered numbers, never from
  expectation. Every result sentence stands as a bracketed pending marker until
  the pre-registered analysis executes, so this abstract reads as a record of
  what was promised and what was delivered.]

  Balsam woolly adelgid (*Adelges piceae*) and western balsam bark beetle
  (*Dryocoetes confusus*) are both mapped on Pacific silver fir (*Abies
  amabilis*) in the British Columbia aerial overview survey, so a canopy-damage
  detector that cannot separate them measures neither. [PENDING: whether the
  addition of structural and radar predictors to a spectral baseline improves
  separation of the two agents, and by how much.] [PENDING: the confusion matrix,
  its support, its class balance and its spatial validation design.] [PENDING:
  the decision under the pre-registered rule, stated as confirmed, disconfirmed
  or inconclusive.]
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
# Chunk order is load-bearing: never move a chunk above one that defines an
# object it reads. Load project libraries here, not scattered through the file.
library(sf)
library(flextable)

root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
outputs <- file.path(root, "03.outputs")
derived <- file.path(inputs, "derived")

# House table format: 9 pt, centred headers, numeric right / text left,
# single line spacing.
fmt_tbl <- function(df, right = NULL) {
  right_cols <- union(names(df)[vapply(df, is.numeric, TRUE)], right)
  left_cols <- setdiff(names(df), right_cols)
  ft <- flextable::flextable(df)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "header")
  if (length(right_cols))
    ft <- flextable::align(ft, j = right_cols, align = "right", part = "body")
  if (length(left_cols))
    ft <- flextable::align(ft, j = left_cols, align = "left", part = "body")
  ft <- flextable::line_spacing(ft, space = 1, part = "all")
  ft <- flextable::padding(ft, padding.top = 1, padding.bottom = 1,
                           part = "all")
  flextable::set_table_properties(ft, layout = "autofit", width = 1)
}
```

```{r}
#| label: load-labels
#| echo: true
#| output: false
# The two derived label layers, both written by 05.scripts/01-extract-adelgid-labels.sh
# from the BC aerial overview survey archive. The 908 MB source geodatabase is
# never read here: only these extracts, which are small enough to commit.
iab <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                   quiet = TRUE)
vi <- sf::st_read(file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)

# Host codes used by the survey, resolved from the survey standards rather than
# from the delivery, which carries codes only. BL and BA are the two true firs
# that matter here; B is an unspecified true fir and BG is grand fir.
host_label <- c(B = "True fir, unspecified", BA = "Pacific silver fir",
                BG = "Grand fir", BL = "Subalpine fir")
agent_label <- c(IAB = "Balsam woolly adelgid",
                 IBB = "Western balsam bark beetle")
sev_label <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
               V = "Very severe")

# Counts quoted in the prose are computed here, once, so prose and numbers
# cannot drift. Nothing below is typed by hand into the text.
n_iab_bc <- nrow(iab)
yr_iab <- range(iab$CAPTURE_YEAR)
n_vi <- nrow(vi)
n_vi_iab <- sum(vi$PEST_SPECIES_CODE == "IAB")
n_vi_ibb <- sum(vi$PEST_SPECIES_CODE == "IBB")
ha_vi_iab <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IAB"]))
ha_vi_ibb <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IBB"]))

# The shared host is what makes this an attribution problem rather than a
# detection problem.
ba <- vi[vi$TREE_SPECIES_CODE == "BA", ]
n_ba_iab <- sum(ba$PEST_SPECIES_CODE == "IAB")
n_ba_ibb <- sum(ba$PEST_SPECIES_CODE == "IBB")

# Severity is heavily weighted to the two lowest classes, which bounds every
# claim the study can make.
sev_iab <- table(factor(vi$PEST_SEVERITY_CODE[vi$PEST_SPECIES_CODE == "IAB"],
                        levels = names(sev_label)))
n_iab_ms <- sum(sev_iab[c("M", "S", "V")])
```

# Introduction

Balsam woolly adelgid, *Adelges piceae* (Ratzeburg), is an introduced insect that
damages and kills North American true firs of all age classes, having entered the
continent in the early twentieth century through separate introductions on the east
and west coasts [@Campbell_2023]. Its congener the hemlock woolly adelgid, *Adelges
tsugae* Annand, is the better-studied insect and the one that drives the North
American research literature, but the two behave in opposite ways on this coast.
In eastern North America *A. tsugae* is invasive and lethal, and infested eastern
hemlock die within four to fifteen years with within-stand mortality that can exceed
ninety percent [@MacQuarrie_2025]. In western North America the same species is a
different lineage and a native one, and @MacQuarrie_2025 record that it "exists at
low densities in forests and rarely kills trees". The Canadian Food Inspection
Agency states the consequence plainly: damage to western hemlock in British Columbia
"has been minor, due also to the combined action of natural enemies and host
resistance."

That asymmetry sets the scope of this study. In British Columbia the woolly adelgid
that damages forests is *A. piceae* on true fir, not *A. tsugae* on hemlock, and the
provincial operational record confirms it. The aerial overview survey archive
carries no damage-agent code for hemlock woolly adelgid at all, while balsam woolly
adelgid appears under code IAB in `r n_iab_bc` polygons spanning
`r yr_iab[1]` to `r yr_iab[2]`. A remote sensing study of adelgid damage in this
province is therefore a study of *A. piceae*, and its host on the coast is Pacific
silver fir, *Abies amabilis* (Douglas) Forbes.

The mechanism that licenses a remote measurement is specific and it is not simple
mortality. Adelgid feeding injects saliva that is toxic to the host, inhibiting bud
formation, deforming stems and branches into the swellings known as gout, and
killing the crown from the top down over several years. Bark beetle attack on the
same host kills the whole crown comparatively quickly and without the preceding
structural deformation. Two agents that share a host but differ in the geometry and
the tempo of the damage they cause are, in principle, separable by an instrument
sensitive to canopy structure, and not separable by an instrument sensitive only to
foliage colour.

The literature bears out the second half of that proposition directly.
@Campbell_2023 mapped balsam woolly adelgid in subalpine fir in northern Utah using
three model families, and reported that spectral data alone "successfully detected
forest degradation" but "failed to distinguish between causal agents of change",
with a mean coefficient of determination of 0.482. Terrain and climate predictors
reached 0.746 but, in the authors' words, "lacked the ability to characterize
local-scale tree damage". Only the combination reached 0.836. The failure mode of
the spectral-only approach is precisely attribution, and it is the failure mode this
study addresses.

Work on the eastern congener points the same way. @Jones_2015 improved species
distribution model predictions of hemlock woolly adelgid by supplying Landsat
difference indices rather than single-date indices, gaining three percent in
accuracy while halving the predicted infested area, from 9.6 percent of the study
area to 5.1 percent, which the authors read as a large reduction in errors of
commission. @Bright_2020 and @Bost_2019 establish that Landsat time series map
insect disturbance across large regions, and @Cornelsen_2024 model where the eastern
adelgid may reach under future climate. None of these studies uses canopy structure
from lidar or backscatter from synthetic aperture radar to separate one agent from
another in a shared host.

That is the gap. The objectives follow from it. First, to quantify how far a
spectral-only classifier can separate balsam woolly adelgid damage from western
balsam bark beetle damage in Pacific silver fir on Vancouver Island. Second, to test
whether adding lidar-derived canopy structure and Sentinel-1 backscatter improves
that separation. Third, to report the improvement against a validation design that
respects the spatial dependence of the labels and the support at which those labels
were drawn.

The hypotheses are stated here and their decision thresholds are fixed at
pre-registration, before any predictor value is joined to any label. H0 is that a
model with structural and radar predictors separates the two agents no better than a
spectral-only model of equal complexity. H1, directional, is that the multi-sensor
model achieves a higher balanced accuracy, because gout and top-kill alter canopy
geometry in a way that red-attack mortality does not. A result in which the
multi-sensor model fails to beat the spectral baseline is reported as a
disconfirmation and not reframed.

# Materials and methods

## Study area

The study area is Vancouver Island and the immediately adjacent mainland coast of
British Columbia, within a rectangular envelope from 128.5 to 123.0 degrees west and
48.3 to 51.0 degrees north. The envelope is a selection device only; it also catches
part of the Sunshine Coast, and the analysis clips to a coastline before any
modelling, reporting counts inside and outside the island. Pacific silver fir occurs
through the Coastal Western Hemlock and Mountain Hemlock biogeoclimatic zones here.

[PENDING: the cartographic standard for every spatial figure, declared once here,
naming projection, graticule, representative fraction and coordinate reference
system, once the first map is drawn. The working projection is NAD83 / BC Albers,
EPSG 3005, which is the projection of the source archive.]

## Input datasets

@tbl-inputs lists every dataset, its role and its access terms. A dataset whose
licence has not been read from its own source is recorded as unread, not as
permissive.

```{r}
#| label: tbl-inputs
#| tbl-cap: "Input datasets, their role in the analysis, and access terms. Licence for the survey archive was read from the BC Data Catalogue record on 2026-08-13. Imagery rows record the intended source and are marked pending until retrieval is scripted and the licence is read from the source."
inputs_tbl <- data.frame(
  Dataset = c("Pest Infestation Polygons",
              "Sentinel-2 L2A",
              "Sentinel-1 GRD",
              "LidarBC open lidar"),
  Role = c("Labels, both agent classes",
           "Optical predictors",
           "Radar predictors",
           "Canopy structure predictors"),
  Source = c("BC Data Catalogue, 450b67bb-02d5-4526-8bc0-ac7924125a1e",
             "Google Earth Engine, COPERNICUS/S2_SR_HARMONIZED",
             "Google Earth Engine, COPERNICUS/S1_GRD",
             "LidarBC open lidar portal"),
  Licence = c("Open Government Licence - British Columbia",
              "pending, read from source",
              "pending, read from source",
              "pending, read from source"),
  Retrieved = c("2026-08-13", "pending", "pending", "pending")
)
fmt_tbl(inputs_tbl)
```

One entry in @tbl-inputs carries a caveat that belongs in the text rather than a
footnote. The custodian of the survey archive is the BC Ministry of Forests, not
GeoBC; GeoBC operates the lidar programme and the geospatial services, and the two
are separate.

The delivered geodatabase carries damage-agent codes without names, so the codes
were resolved against the *Forest Health Aerial Overview Survey Standards for
British Columbia*, the Ministry of Forests adaptation of Canadian Forest Service
FHN Report 97-1, retrieved 2026-08-13. Its code list gives "IAB Balsam Woolly
Adelgid" under the heading "IA APHIDS" and "IBB Western balsam bark beetle" under
"IB BARK BEETLES". The same document's host and timing table lists IAB against
sub-alpine true firs, "amabilis fir" and "grand fir", which matches the host codes
present in the extract and is an independent check on the code assignment. That
document carries a 2000 date and may be superseded by a later revision that could
not be retrieved, so the codes are confirmed as of that edition rather than as of
the current survey year.

The same code list is the strongest evidence for the scope decision recorded in the
introduction: it contains no code for hemlock woolly adelgid, and no entry for
*Adelges tsugae* anywhere. An agent with no code cannot be mapped, so the absence of
that species from the archive is a property of the survey design and not an artefact
of the years sampled.

## Reference data

Labels are polygons from the provincial aerial overview survey, sketch-mapped from a
fixed-wing aircraft. Their limitations are documented by the custodian and they
bound every claim below. The catalogue record states that "there are errors of
omission and commission and general spatial accuracy can be poor depending on a
number of factors but is accurate enough to meet the main purpose of the data which
is to provide general trends in disturbance size and frequency", and the survey
methods page states that "the survey is not normally accurate enough to direct
operational activities." A minimum of eighty percent of the province is flown each
year, with gaps caused by weather and smoke.

The support mismatch is explicit in the digitising rule. A clump of one to thirty
dead or dying trees is delineated as 0.25 ha and a clump of thirty-one to fifty as
0.5 ha, and either is assigned a severe intensity rating on digitising. A polygon is
therefore not a delineation of a damaged canopy and must not be treated as a
pixel-level mask. The analysis unit is fixed accordingly in the design below.

Severity is an ordinal judgement with a defined but coarse basis. The survey
standards define the mortality classes as "Light = 1-10% of trees recently killed",
"Moderate = 11-29% of trees recently killed" and "Severe = 30%+ of trees recently
killed", and state that "all dots (point sources) are classified as severe". Two
consequences follow. A severe rating can mean a stand with a third of its trees
killed or a single small clump recorded as a point, and those are not the same
observation. And the trace class that dominates the adelgid record does not appear
in that mortality scale at all, which is consistent with adelgid damage being
recorded as a condition rather than as mortality. This is a further reason severity
is excluded as a response variable.

@tbl-labels reports the label inventory actually available, computed from the
extract rather than asserted.

```{r}
#| label: tbl-labels
#| tbl-cap: "Aerial overview survey labels available on Vancouver Island and the adjacent mainland coast, by damage agent and host tree species. Counts are polygons; area is the survey-recorded polygon area. Computed from the derived extract, not asserted."
lab <- as.data.frame(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Host  = host_label[vi$TREE_SPECIES_CODE]
), stringsAsFactors = FALSE)
names(lab)[3] <- "Polygons"
lab <- lab[lab$Polygons > 0, ]

area <- tapply(vi$AREA_HA,
               list(agent_label[vi$PEST_SPECIES_CODE],
                    host_label[vi$TREE_SPECIES_CODE]),
               function(v) round(sum(v)))
lab$`Area (ha)` <- mapply(function(a, h) area[a, h], lab$Agent, lab$Host)
lab <- lab[order(lab$Agent, -lab$Polygons), ]
fmt_tbl(lab)
```

Two features of @tbl-labels set the design. Both agents are mapped on Pacific silver
fir, `r n_ba_iab` adelgid polygons against `r n_ba_ibb` bark beetle polygons, which
is what makes this an attribution problem rather than a detection problem. And the
adelgid class is small: `r n_vi_iab` polygons covering `r format(ha_vi_iab, big.mark = ",")`
ha against `r n_vi_ibb` bark beetle polygons covering
`r format(ha_vi_ibb, big.mark = ",")` ha.

@tbl-severity gives the severity distribution, which bounds the study harder than
the sample size does.

```{r}
#| label: tbl-severity
#| tbl-cap: "Severity class of the mapped polygons by agent. Severity is the surveyor's ordinal rating, not a measured quantity, and the digitising rule assigns severe to any small clump regardless of condition."
sev <- as.data.frame.matrix(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Severity = factor(sev_label[vi$PEST_SEVERITY_CODE],
                    levels = unname(sev_label))
))
sev <- sev[, colSums(sev) > 0, drop = FALSE]
sev <- cbind(Agent = rownames(sev), sev)
fmt_tbl(sev)
```

Only `r n_iab_ms` adelgid polygons are rated moderate or above. Any claim this study
makes about severe damage rests on that number, and it is too small to support a
severity-graded model. The pre-registered analysis is therefore a two-class
attribution test at polygon level, not a severity regression, and the contrast with
@Campbell_2023, who modelled severity as a continuous response, is a difference in
what the labels can carry rather than a difference in ambition.

```{r}
#| label: fig-extent
#| fig-cap: "Balsam woolly adelgid polygons recorded in the British Columbia aerial overview survey, 1996 to 2024, and the Vancouver Island study envelope. Coordinates are geographic, WGS84."
#| fig-width: 7
#| fig-height: 6
op <- par(mar = c(4, 4, 1, 1))
plot(sf::st_geometry(iab), col = "grey20", border = NA,
     axes = TRUE, xlab = "Longitude", ylab = "Latitude")
rect(-128.5, 48.3, -123.0, 51.0, border = "firebrick", lwd = 2)
legend("topright", bty = "n", cex = 0.8,
       legend = c("IAB polygon", "Study envelope"),
       col = c("grey20", "firebrick"), pch = c(15, NA), lty = c(NA, 1), lwd = 2)
par(op)
```

```{r}
#| label: fig-timeseries
#| fig-cap: "Polygons mapped per survey year on Vancouver Island true-fir hosts, by agent. The adelgid record is concentrated in a few years, which constrains the choice of imagery epoch."
#| fig-width: 7
#| fig-height: 4
tt <- table(vi$CAPTURE_YEAR, agent_label[vi$PEST_SPECIES_CODE])
op <- par(mar = c(4, 4, 1, 1))
matplot(as.numeric(rownames(tt)), tt, type = "l", lty = 1, lwd = 2,
        col = c("firebrick", "steelblue"),
        xlab = "Survey year", ylab = "Polygons mapped")
legend("topleft", colnames(tt), col = c("firebrick", "steelblue"),
       lty = 1, lwd = 2, bty = "n", cex = 0.8)
par(op)
```

## Imagery availability

The optical and radar predictors are drawn from Google Earth Engine. The chunk below
establishes the connection and reports scene availability over the study envelope
for the epoch the label record concentrates in. It deliberately stops there. No
predictor value is extracted at any label location in this document, because the
sample and the fold assignment are not yet locked, and joining an outcome to a
predictor before that point is the specific thing the project's conventions forbid.

```{r}
#| label: gee-availability
#| echo: true
# Earth Engine is reached through rgee under the registered project. The call is
# wrapped so that a render on a machine without credentials degrades to a
# recorded status rather than a failure, but the status is printed either way:
# a silent skip would let a broken connection pass as a working one.
gee_status <- tryCatch({
  library(rgee)
  ee_Initialize(project = "murphys-deforisk", quiet = TRUE)

  aoi <- ee$Geometry$Rectangle(c(-128.5, 48.3, -123.0, 51.0))

  # Leaf-on window, the epoch in which the adelgid label record concentrates.
  s2 <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))

  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$eq("instrumentMode", "IW"))

  data.frame(
    Collection = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
    Window = "2019-06-01 to 2021-09-30",
    Scenes = c(s2$size()$getInfo(), s1$size()$getInfo()),
    Status = "connected"
  )
}, error = function(e) {
  data.frame(Collection = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
             Window = "2019-06-01 to 2021-09-30",
             Scenes = NA_integer_,
             Status = paste("unavailable:", substr(conditionMessage(e), 1, 60)))
})
```

```{r}
#| label: tbl-imagery
#| tbl-cap: "Imagery scene availability over the study envelope for the label epoch, queried live from Google Earth Engine at render time."
fmt_tbl(gee_status)
```

Lidar coverage is confirmed and is reported in @tbl-lidar from the saved output of
`05.scripts/02-query-lidar-coverage.py`, which queries the LidarBC point cloud tile
index. Point clouds are not retrieved here; the index carries direct object-store
URLs and the tiles stay off this machine until the analysis design fixes which are
needed.

```{r}
#| label: tbl-lidar
#| tbl-cap: "LidarBC point cloud tile coverage over the study envelope and over the Comox field window. Tile counts are exact. Acquisition years and point classes are drawn from a sample capped at 1000 records by the service, so for the study area they are what the sample contained and not the full range."
lidar <- read.csv(file.path(derived, "lidar-coverage.csv"),
                  check.names = FALSE)
names(lidar) <- c("Envelope", "Tiles", "Attributes from full index",
                  "Acquisition years (sampled)", "Density (pts/m2)",
                  "Point classes")
fmt_tbl(lidar)
```

Three features of @tbl-lidar bear on the design. Coverage exists over the Comox
field window, which is where ground confirmation is practical. The acquisition years
sampled there include 2019, which falls inside the epoch the adelgid label record
concentrates in, so a structural predictor can be matched to a label without a long
temporal offset. And the point classes are limited: the tiles are classified to
ground, noise and overlap but carry no vegetation classes, so canopy metrics must be
derived as heights above a ground surface built from class 2 returns rather than
read from a vegetation classification.

## Data preparation

[PENDING. Compositing, cloud and terrain masking, radiometric terrain correction for
the Sentinel-1 stack, speckle filtering, harmonisation of the lidar acquisitions to a
common canopy height model resolution, and the grid on which everything is aligned.
Each choice states its threshold and its reason. Where a preparation choice moves the
signal at a magnitude comparable to the effect being measured, that is stated and
tested. Written at the analysis design gate, which is two gates downstream of where
this manuscript currently sits.]

## Analysis design

[PENDING, and deliberately so. This section is fixed at the designing-the-analysis
gate and then frozen at pre-registration. What is already determined by the labels
rather than by preference is recorded here so the design cannot quietly drift:]

The unit of analysis is the survey polygon, not the pixel, because the labels are
sketch-mapped and the digitising rule manufactures area for small clumps. The
response is the two-class agent label, adelgid against bark beetle, restricted to
Pacific silver fir where both agents occur, because a comparison across different
hosts would confound agent with host spectra. Cross-validation is spatially blocked,
since polygons cluster and a random split would leak. Class imbalance is approximately
`r round(n_ba_ibb / n_ba_iab, 1)` to 1 against the adelgid class and is handled in the
resampling scheme, with balanced accuracy as the headline metric rather than raw
accuracy.

## Validation harness

Metric functions are unit-tested against cases whose answers are known by
construction, and the test table renders with the manuscript, so a metric that
silently breaks cannot pass unnoticed. @tbl-checks records the checks.

```{r}
#| label: eval-harness
#| echo: true
#| output: false
# Balanced accuracy is the headline metric, so it is defined once here and tested
# below against cases whose value is known analytically. A guard that cannot fail
# is not a guard: each check names its expected value and why that value is right.
balanced_accuracy <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}
```

```{r}
#| label: tbl-checks
#| tbl-cap: "Unit tests of the evaluation metric, each against a case whose value is known by construction. A failing row renders as FALSE rather than stopping the document, so a broken metric is visible in the output."
t_perfect <- c("IAB", "IAB", "IBB", "IBB")
t_inverted <- c("IBB", "IBB", "IAB", "IAB")
# An all-majority-class prediction: 100% specificity, 0% sensitivity, so 0.5.
t_majority <- rep("IBB", 4)
# Imbalanced case: 1 adelgid, 3 beetle, all called beetle. Still 0.5, which is
# the whole reason balanced accuracy replaces raw accuracy here (raw = 0.75).
imb_truth <- c("IAB", "IBB", "IBB", "IBB")

checks <- data.frame(
  Check = c("Perfect prediction",
            "Fully inverted prediction",
            "All predicted majority class",
            "Imbalanced, all majority (raw accuracy would read 0.75)"),
  Expected = c(1, 0, 0.5, 0.5),
  Observed = c(
    balanced_accuracy(t_perfect, t_perfect),
    balanced_accuracy(t_perfect, t_inverted),
    balanced_accuracy(t_perfect, t_majority),
    balanced_accuracy(imb_truth, rep("IBB", 4))
  )
)
checks$Pass <- abs(checks$Expected - checks$Observed) < 1e-9
fmt_tbl(checks)
```

## Pipeline validation

[PENDING, and mandatory before any claim from real data. A synthetic signal of known
magnitude and known timing is injected into real null cases and the pipeline must
recover it. The report states what the pipeline recovers and, more importantly, the
level below which it recovers nothing, because that is the floor of every downstream
claim. Given a positive class of `r n_ba_iab` polygons, establishing that floor
matters more here than it would in a large-sample study.]

# Results

[No value in this section is estimated, assumed or illustrative. Every table below
carries its final structure and displays pending until the pre-registered analysis
executes. The analysis has not executed: no predictor value has been joined to any
label, by design, because the framing gate is the current gate and pre-registration
lies two gates downstream.]

## Primary result

[PENDING. Prose reports the pre-registered primary test and nothing else: the
difference in balanced accuracy between the multi-sensor model and the spectral
baseline, its interval, and the decision under the pre-registered rule stated as
confirmed, disconfirmed or inconclusive.]

```{r}
#| label: tbl-primary
#| tbl-cap: "Primary pre-registered test: multi-sensor against spectral-only separation of the two agents. Values pending execution of the pre-registered analysis."
primary <- data.frame(
  Comparison = "Multi-sensor vs spectral-only",
  `Balanced accuracy` = "pending",
  `Difference` = "pending",
  CI95 = "pending",
  Decision = "pending",
  check.names = FALSE
)
fmt_tbl(primary)
```

## Secondary results

[PENDING. Per-class sensitivity and specificity, the confusion matrix with its
support stated, and predictor importance grouped by sensor family. Anything not named
in the pre-registration is exploratory and says so in its own sentence.]

```{r}
#| label: tbl-secondary
#| tbl-cap: "Per-class performance and sensor contribution. Values pending execution of the pre-registered analysis."
secondary <- data.frame(
  Model = c("Spectral only", "Spectral + structure", "Spectral + structure + radar"),
  Sensitivity = "pending",
  Specificity = "pending",
  `Balanced accuracy` = "pending",
  check.names = FALSE
)
fmt_tbl(secondary)
```

# Discussion

## Principal finding

[PENDING. Written in both directions before the numbers exist, so that a null is
reportable rather than embarrassing. If the multi-sensor model beats the spectral
baseline, the finding extends @Campbell_2023 by supplying the attribution capability
their spectral models lacked. If it does not, the finding is that canopy structure
does not carry separable information about these two agents at the support the
provincial survey labels provide, which is a statement about the labels as much as
about the sensors, and it should be written as such.]

## Limitations

The limitations are known now and are not contingent on the result. The labels are
sketch-mapped from aircraft and the custodian states that their spatial accuracy can
be poor. The adelgid class is small, `r n_ba_iab` polygons on the shared host, and
only `r n_iab_ms` polygons across the whole envelope are rated moderate or worse. The
digitising rule assigns a severe rating to any clump of fifty trees or fewer on
digitising, which means severity is partly an artefact of the recording procedure.
Agent attribution in the source data is the surveyor's judgement from the air and is
not independently verified on the ground, so the study can at most reproduce that
judgement, never exceed it. Field confirmation by the author addresses this only
partly, since the diagnostic signs of adelgid infestation, the woolly masses and stem
gout, are found on living and declining trees and are lost from long-dead ones.

## Detection and attribution

This study is an attribution test, not a detection study, and the distinction is
kept in every claim. Detecting that a canopy has changed is not the same as
identifying what changed it, and @Campbell_2023 demonstrate that a model can do the
first well while failing the second. Accuracy figures reported here state their
support, which is the survey polygon, their class balance, and their validation
design, or they are not reported.

# Conclusion

[PENDING. One paragraph, written from the rendered numbers, stating what was asked,
what was found, and what the finding licenses. It is written after the results
exist and not before.]

# Data availability

Label extracts are committed at `02.inputs/derived/`. The extraction is scripted at
`05.scripts/01-extract-adelgid-labels.sh` and reads the Pest Infestation Polygons
archive, which is not committed because it is 908 MB. Reference verification is
scripted at `05.scripts/verify-references-crossref.py` and literature text
extraction at `05.scripts/extract-literature-text.py`. Every number in this document
is computed by the chunks above at render time.

# Conflicts of interest

[PENDING declaration. Ecuador does not arise in this study.]

# References

::: {#refs}
:::
