---
title: "Separating balsam woolly adelgid from bark beetle damage in Pacific silver fir on Vancouver Island: a multi-sensor attribution test"
date: today
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
abstract: >
  Balsam woolly adelgid (*Adelges piceae*) and western balsam bark beetle
  (*Dryocoetes confusus*) are both mapped on Pacific silver fir (*Abies
  amabilis*) in the British Columbia aerial overview survey, so a canopy-damage
  detector that cannot separate them measures neither. In this study, we examine
  whether the addition of lidar and radar predictors to a spectral baseline improves
  separation of the two agents, and by how much.
keywords:
  - balsam woolly adelgid
  - Adelges piceae
  - Abies amabilis
  - forest health monitoring
  - lidar
  - synthetic aperture radar
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    toc: true
    number-sections: false
    embed-resources: true
    code-fold: false
    df-print: kable
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false
library(sf)
library(terra)
library(flextable)

root <- normalizePath("..")
inputs <- file.path(root, "02.inputs")
outputs <- file.path(root, "03.outputs")
derived <- file.path(inputs, "derived")

BC_ALBERS <- 3005

# Figures are written to 03.outputs/figures as well as embedded, so the README
# and any submission package reference committed files rather than extracting
# images from a rendered document.
knitr::opts_chunk$set(fig.path = file.path(outputs, "figures/"),
                      dpi = 300, dev = "png")

fmt_tbl <- function(df, right = NULL) {
  right_cols <- union(names(df)[vapply(df, is.numeric, TRUE)], right)
  left_cols <- setdiff(names(df), right_cols)
  ft <- flextable::flextable(df)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "header")
  if (length(right_cols))
    ft <- flextable::align(ft, j = right_cols, align = "right", part = "body")
  if (length(left_cols))
    ft <- flextable::align(ft, j = left_cols, align = "left", part = "body")
  ft <- flextable::line_spacing(ft, space = 1, part = "all")
  ft <- flextable::padding(ft, padding.top = 1, padding.bottom = 1,
                           part = "all")
  flextable::set_table_properties(ft, layout = "autofit", width = 1)
}
```

```{r}
#| label: load-labels
#| echo: true
#| output: false
iab <- sf::st_read(file.path(derived, "adelgid-iab-polygons.geojson"),
                   quiet = TRUE)
vi <- sf::st_read(file.path(derived, "vi-fir-agents.geojson"), quiet = TRUE)

host_label <- c(B = "True fir, unspecified", BA = "Pacific silver fir",
                BG = "Grand fir", BL = "Subalpine fir")
agent_label <- c(IAB = "Balsam woolly adelgid",
                 IBB = "Western balsam bark beetle")
sev_label <- c(T = "Trace", L = "Light", M = "Moderate", S = "Severe",
               V = "Very severe")

n_iab_bc <- nrow(iab)
yr_iab <- range(iab$CAPTURE_YEAR)
n_vi_iab <- sum(vi$PEST_SPECIES_CODE == "IAB")
n_vi_ibb <- sum(vi$PEST_SPECIES_CODE == "IBB")
ha_vi_iab <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IAB"]))
ha_vi_ibb <- round(sum(vi$AREA_HA[vi$PEST_SPECIES_CODE == "IBB"]))

ba <- vi[vi$TREE_SPECIES_CODE == "BA", ]
n_ba_iab <- sum(ba$PEST_SPECIES_CODE == "IAB")
n_ba_ibb <- sum(ba$PEST_SPECIES_CODE == "IBB")
imbalance <- round(n_ba_ibb / n_ba_iab, 1)

sev_iab <- table(factor(vi$PEST_SEVERITY_CODE[vi$PEST_SPECIES_CODE == "IAB"],
                        levels = names(sev_label)))
n_iab_ms <- sum(sev_iab[c("M", "S", "V")])
```

# Introduction

Balsam woolly adelgid, *Adelges piceae* (Ratzeburg), is an introduced insect that damages and kills North American true firs of all age classes, having entered the continent in the early twentieth century through separate introductions on the east and west coasts [@Campbell_2023]. Its congener the hemlock woolly adelgid, *Adelges tsugae* Annand, is the better-studied insect and the one that drives the North American research literature, but the two behave differently on this coast. In eastern North America *A. tsugae* is invasive and lethal, and infested eastern hemlock die within four to fifteen years with within-stand mortality that can exceed ninety percent [@MacQuarrie_2025]. In western North America the same species is a separate and native lineage that "exists at low densities in forests and rarely kills trees" [@MacQuarrie_2025]. Damage to western hemlock in British Columbia has accordingly been minor, and the provincial forest health survey reflects the asymmetry directly: it carries a damage-agent code for balsam woolly adelgid and none for hemlock woolly adelgid. In British Columbia the woolly adelgid of management concern is therefore *A. piceae* on true fir, and its principal host on the coast is Pacific silver fir, *Abies amabilis* (Douglas) Forbes.

The damage the adelgid causes is structural before it is lethal. Feeding injects saliva that is toxic to the host, inhibiting bud formation and deforming stems and branches into the swellings known as gout, whose incidence varies with crown class and hardiness zone [@Guillet_2010], and the crown is killed progressively from the top down over a period of years. Western balsam bark beetle, *Dryocoetes confusus* Swaine, attacks the same firs but kills the whole crown comparatively rapidly and without that preceding deformation. The two agents therefore differ not only in the foliage colour change they eventually produce but in the geometry and the tempo of the damage that produces it, which is the basis for expecting an instrument sensitive to canopy structure to separate them where an instrument sensitive only to reflectance may not.

Remote detection of adelgid damage is not new. @FRANKLIN_1995 acquired airborne multispectral imagery over adelgid-infested balsam fir in Newfoundland and separated damage classes on individual trees with accuracies between 40 and 76 percent, depending on the classification scheme and the predictors available. @Cook_2010 showed that branch-level spectra distinguish infested from non-infested subalpine fir with 94 percent accuracy using a three-variable model, and that a comparable result survives simulation of a broadband satellite sensor. Most recently @Campbell_2023 mapped infestation severity in subalpine fir in northern Utah from satellite imagery combined with terrain and climate. Their result is the point of departure for this study: spectral data alone "successfully detected forest degradation" but "failed to distinguish between causal agents of change", returning a mean coefficient of determination of 0.482, while terrain and climate reached 0.746 but "lacked the ability to characterize local-scale tree damage", and only the combination reached 0.836. The residual failure of the spectral models was one of attribution rather than detection.

That failure is a specific instance of a problem the wider forest health literature has addressed directly. Detecting insect damage from optical imagery is now routine: Sentinel-2 maps the green-attack stage of *Ips typographus* before visible discolouration [@Abdullah_2018], purpose-built indices advance detection earlier still [@Huo_2021], and Landsat time series map insect disturbance and mortality across large regions and long records [@Bright_2020; @Bost_2019], with the field recently reviewed at continental scale [@Drechsel_2025]. Attribution has been treated as its own problem. @Oeser_2017 attributed windthrow, cleared windthrow, bark beetle and harvest disturbances in Central Europe from intra-annual Landsat metrics with overall accuracies of 76 to 86 percent; @Zhang_2022 mapped causal agents of disturbance across boreal and arctic North America; @Sebald_2021 showed that landscape context improves the separation of human from natural disturbance; and @Morin_Bernard_2023 attributed agent before estimating severity in northern hardwood, reaching 82.9 percent overall accuracy and finding that agent-specific models required different predictors, which confirms that attribution must precede severity estimation rather than follow it.

Structural and radar sensing have entered this literature, but on the detection side. Terrestrial lidar detects *Ips typographus* infestation early from dual-wavelength returns [@Junttila_2019], and multidimensional structural metrics describe moderate-severity disturbance that stand-replacing change detection misses [@Atkins_2020]. Radar has been reviewed for bark beetle monitoring [@Hollaus_2019], and @Tanase_2018 used L-band synthetic aperture radar to delineate windthrow and insect outbreak in the Bavarian Forest with overall accuracies of 69 to 84 and 65 to 88 percent respectively, falling to 64 to 74 percent for the early green phase.

Two features of this body of work define the gap addressed here. First, every attribution study cited above separates disturbance types that differ strongly in their spectral trajectory, their spatial pattern or their timing: wind against beetle, harvest against natural mortality, ice storm against partial cut. Separating two insect agents that attack the same host species is a harder discrimination and we found no study that attempts it. Second, where structure and radar have been used, they have been used to detect damage earlier or to characterise its severity, not to identify which agent caused it. The proposition that canopy geometry carries agent-diagnostic information, rather than merely damage-intensity information, remains untested.

We address that gap with three objectives. First, to quantify how far a spectral-only classifier can separate balsam woolly adelgid damage from western balsam bark beetle damage in Pacific silver fir on Vancouver Island. Second, to test whether adding lidar-derived canopy structure and Sentinel-1 backscatter improves that separation. Third, to report any improvement against a validation design that respects the spatial dependence of the labels and the support at which they were drawn. We hypothesise that the multi-sensor model achieves higher balanced accuracy than the spectral baseline, because gout and progressive top-kill alter canopy geometry in a way that rapid whole-crown mortality does not. The null hypothesis is that structural and radar predictors add nothing beyond the spectral baseline.

# Materials and methods

## Study area

The study area is Vancouver Island and the immediately adjacent mainland coast of British Columbia, Canada, within an envelope from 128.5 to 123.0 degrees west and 48.3 to 51.0 degrees north (@fig-studyarea). Pacific silver fir occurs through the Coastal Western Hemlock and Mountain Hemlock biogeoclimatic zones that dominate the mountainous interior of the island, generally above the Coastal Douglas-fir zone of the eastern lowlands. All spatial data are handled in NAD83 / BC Albers, EPSG 3005, which is the projection of the source survey archive, and every map is drawn in that projection.

```{r}
#| label: fig-studyarea
#| fig-cap: "Study area on Vancouver Island and the adjacent mainland coast of British Columbia, Canada, showing aerial overview survey polygons attributed to balsam woolly adelgid and to western balsam bark beetle on true fir hosts. Relief is shaded from a 1 km digital elevation model illuminated from the northwest. Projection NAD83 / BC Albers (EPSG 3005); axis labels in degrees."
#| fig-width: 7.2
#| fig-height: 7.4
#| out-width: "100%"
bm <- file.path(derived, "basemap.gpkg")
land   <- sf::st_read(bm, "land",   quiet = TRUE)
admin  <- sf::st_read(bm, "admin",  quiet = TRUE)
rivers <- sf::st_read(bm, "rivers", quiet = TRUE)
lakes  <- sf::st_read(bm, "lakes",  quiet = TRUE)
places <- sf::st_read(bm, "places", quiet = TRUE)
hs <- terra::rast(file.path(derived, "hillshade.tif"))

vi_p <- sf::st_transform(vi, BC_ALBERS)
fr <- sf::st_bbox(land)

op <- par(mar = c(3.0, 3.6, 0.5, 0.5), xaxs = "i", yaxs = "i")
plot(NA, xlim = fr[c("xmin", "xmax")], ylim = fr[c("ymin", "ymax")],
     asp = 1, axes = FALSE, xlab = "", ylab = "")

# Sea, then relief-shaded land, then hydrography, then boundaries. The sea is
# painted across the whole plot region rather than the land bounding box, since
# the fixed aspect ratio leaves the region taller than the data.
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = "#dce7ef", border = NA)
terra::plot(hs, col = grey.colors(100, start = 0.30, end = 1),
            legend = FALSE, add = TRUE, maxcell = 2e6)
plot(sf::st_geometry(land), col = NA, border = "grey30", lwd = 0.6, add = TRUE)
plot(sf::st_geometry(lakes), col = "#c3d9ea", border = "#6f9ab5", lwd = 0.3,
     add = TRUE)
plot(sf::st_geometry(rivers), col = "#5b8db0", lwd = 0.8, add = TRUE)
# Only the international boundary, drawn dashed. The provincial outline is not
# drawn because within this frame it coincides with the coastline.
plot(sf::st_geometry(admin), col = "grey20", lwd = 1.1, lty = 2, add = TRUE)

# Bark beetle first: the far less numerous adelgid class must stay legible.
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IBB", ]),
     col = "#2c6f9e", border = "#2c6f9e", add = TRUE)
plot(sf::st_geometry(vi_p[vi_p$PEST_SPECIES_CODE == "IAB", ]),
     col = "#c1372f", border = "#c1372f", add = TRUE)

pxy <- sf::st_coordinates(places)
points(pxy, pch = 21, bg = "white", col = "grey15", cex = 0.8, lwd = 0.7)
# Label side chosen per settlement so no label runs off the frame or across a
# neighbouring one.
lab_pos <- c(Comox = 4, `Campbell River` = 2, `Port Alberni` = 2,
             Nanaimo = 4, Victoria = 2, Vancouver = 4, Tofino = 2,
             `Port Hardy` = 4)
text(pxy[, 1], pxy[, 2], places$name, pos = lab_pos[places$name],
     offset = 0.4, cex = 0.6, col = "grey10")

# Axes labelled in degrees although the plot is projected, so the reader gets
# geographic coordinates without the distortion of plotting in degrees.
xt <- pretty(fr[c("xmin", "xmax")], 5)
yt <- pretty(fr[c("ymin", "ymax")], 5)
xdeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(xt, mean(fr[c("ymin", "ymax")])))[, 1]
ydeg <- sf::sf_project(paste0("EPSG:", BC_ALBERS), "EPSG:4326",
                       cbind(mean(fr[c("xmin", "xmax")]), yt))[, 2]
axis(1, at = xt, labels = sprintf("%.1f°W", abs(xdeg)), cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.35, 0))
axis(2, at = yt, labels = sprintf("%.1f°N", ydeg), las = 1, cex.axis = 0.65,
     tck = -0.012, mgp = c(2, 0.55, 0))
box(col = "grey25")

# Scale bar drawn in projected metres, so the stated length is metrically true.
sb <- 50000
sx <- fr["xmin"] + 0.05 * diff(fr[c("xmin", "xmax")])
sy <- fr["ymin"] + 0.06 * diff(fr[c("ymin", "ymax")])
h <- 0.008 * diff(fr[c("ymin", "ymax")])
rect(sx, sy, sx + sb, sy + h, col = "grey15", border = "grey15")
rect(sx + sb, sy, sx + 2 * sb, sy + h, col = "white", border = "grey15")
text(c(sx, sx + sb, sx + 2 * sb), sy + 2.6 * h, c("0", "50", "100 km"),
     cex = 0.58, col = "grey15")

# North arrow.
nx <- fr["xmin"] + 0.055 * diff(fr[c("xmin", "xmax")])
ny <- fr["ymin"] + 0.155 * diff(fr[c("ymin", "ymax")])
a <- 0.030 * diff(fr[c("ymin", "ymax")])
polygon(c(nx, nx - a * 0.40, nx, nx + a * 0.40),
        c(ny + a, ny - a * 0.28, ny - a * 0.05, ny - a * 0.28),
        col = "grey15", border = "grey15")
text(nx, ny + 1.55 * a, "N", cex = 0.78, font = 2, col = "grey15")

legend("topright", inset = 0.012, bg = "white", box.col = "grey55",
       cex = 0.62, pt.cex = 1.1, pch = 22,
       pt.bg = c("#c1372f", "#2c6f9e"), col = c("#c1372f", "#2c6f9e"),
       legend = c("Balsam woolly adelgid", "Western balsam bark beetle"))
par(op)
```

## Data sources

@tbl-inputs lists each dataset and its access terms. The custodian of the survey archive is the BC Ministry of Forests; GeoBC operates the lidar programme and the two are separate agencies.

```{r}
#| label: tbl-inputs
#| tbl-cap: "Datasets used in the analysis, with their role and access terms."
inputs_tbl <- data.frame(
  Dataset = c("Pest Infestation Polygons", "Sentinel-2 L2A", "Sentinel-1 GRD",
              "LidarBC open lidar"),
  Role = c("Damage agent labels", "Optical predictors", "Radar predictors",
           "Canopy structure predictors"),
  Source = c("BC Data Catalogue", "Google Earth Engine",
             "Google Earth Engine", "LidarBC open data portal"),
  Licence = c("Open Government Licence - British Columbia",
              "Copernicus open access", "Copernicus open access",
              "Open Government Licence - British Columbia")
)
fmt_tbl(inputs_tbl)
```

Damage agents in the delivered geodatabase are recorded as three-letter codes without names. Codes were resolved against the *Forest Health Aerial Overview Survey Standards for British Columbia*, the Ministry of Forests adaptation of Canadian Forest Service FHN Report 97-1, which gives IAB for balsam woolly adelgid and IBB for western balsam bark beetle. The same document lists IAB against sub-alpine true firs, amabilis fir and grand fir, matching the host codes present in the extract. That code list contains no entry for hemlock woolly adelgid, so the absence of that species from the archive is a property of the survey design rather than of the years sampled.

## Survey data

Damage labels are polygons from the British Columbia aerial overview survey. Surveyors sketch-map visible forest damage from a fixed-wing aircraft, recording host species, damage agent and an ordinal severity class for each polygon. A minimum of eighty percent of the province is flown annually, with gaps caused by weather and smoke.

Survey polygons carry limitations that constrain the analysis and that the custodian states explicitly: errors of omission and commission occur, spatial accuracy can be poor, and the survey is not intended to be accurate enough to direct operational activities. Its stated purpose is to record general trends in disturbance size and frequency. Polygon area is in part a product of the digitising rule rather than a measurement, since a clump of one to thirty dead or dying trees is delineated as 0.25 ha and a clump of thirty-one to fifty as 0.5 ha. Severity is likewise an ordinal judgement: the standards define light, moderate and severe as 1 to 10, 11 to 29 and 30 percent or more of trees recently killed, and classify all point sources as severe on digitising. The unit of analysis here is therefore the survey polygon rather than the pixel, and severity is not modelled.

Across the study area the survey records `r n_vi_iab` adelgid polygons covering `r format(ha_vi_iab, big.mark = ",")` ha and `r n_vi_ibb` bark beetle polygons covering `r format(ha_vi_ibb, big.mark = ",")` ha on true fir hosts (@tbl-labels). Province-wide the adelgid record comprises `r n_iab_bc` polygons spanning `r yr_iab[1]` to `r yr_iab[2]`. Both agents are recorded on Pacific silver fir, `r n_ba_iab` adelgid against `r n_ba_ibb` bark beetle polygons, and it is this shared host that makes the separation an attribution problem rather than a detection problem. Analysis is restricted to that host so that agent is not confounded with differences in host reflectance.

```{r}
#| label: tbl-labels
#| tbl-cap: "Aerial overview survey polygons in the study area by damage agent and host tree species. Area is the survey-recorded polygon area."
lab <- as.data.frame(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Host  = host_label[vi$TREE_SPECIES_CODE]
), stringsAsFactors = FALSE)
names(lab)[3] <- "Polygons"
lab <- lab[lab$Polygons > 0, ]

area <- tapply(vi$AREA_HA,
               list(agent_label[vi$PEST_SPECIES_CODE],
                    host_label[vi$TREE_SPECIES_CODE]),
               function(v) round(sum(v)))
lab$`Area (ha)` <- mapply(function(a, h) area[a, h], lab$Agent, lab$Host)
lab <- lab[order(lab$Agent, -lab$Polygons), ]
fmt_tbl(lab)
```

The severity distribution is strongly weighted towards the two lowest classes (@tbl-severity). Only `r n_iab_ms` adelgid polygons in the study area are rated moderate or worse, too few to support the severity regression used by @Campbell_2023, and a further reason the response here is the two-class agent label.

```{r}
#| label: tbl-severity
#| tbl-cap: "Severity class of survey polygons by damage agent. Severity is the surveyor's ordinal rating."
sev <- as.data.frame.matrix(table(
  Agent = agent_label[vi$PEST_SPECIES_CODE],
  Severity = factor(sev_label[vi$PEST_SEVERITY_CODE],
                    levels = unname(sev_label))
))
sev <- sev[, colSums(sev) > 0, drop = FALSE]
sev <- cbind(Agent = rownames(sev), sev)
fmt_tbl(sev)
```

The two agents also peak in different survey years (@fig-timeseries). The adelgid record concentrates in 2019 to 2021 while the bark beetle record is distributed more evenly across the past decade. A classifier trained on imagery drawn from each polygon's own survey year would therefore risk learning the year rather than the agent, and a single common imagery epoch is used for both classes.

```{r}
#| label: fig-timeseries
#| fig-cap: "Survey polygons mapped per year on true fir hosts in the study area, by damage agent."
#| fig-width: 7
#| fig-height: 3.6
tt <- table(vi$CAPTURE_YEAR, agent_label[vi$PEST_SPECIES_CODE])
op <- par(mar = c(4, 4.2, 0.6, 0.6))
matplot(as.numeric(rownames(tt)), tt, type = "l", lty = 1, lwd = 2,
        col = c("#c1372f", "#2c6f9e"), las = 1,
        xlab = "Survey year", ylab = "Polygons mapped")
legend("topleft", colnames(tt), col = c("#c1372f", "#2c6f9e"),
       lty = 1, lwd = 2, bty = "n", cex = 0.75)
par(op)
```

## Remote sensing data

Optical predictors are drawn from Sentinel-2 Level 2A surface reflectance and radar predictors from Sentinel-1 Ground Range Detected imagery in interferometric wide swath mode, both accessed through Google Earth Engine. Scenes are restricted to the leaf-on window of the epoch in which the adelgid record concentrates, 1 June 2019 to 30 September 2021, and optical scenes to those with less than forty percent cloud cover. Scene counts available over the study area are given in @tbl-imagery.

```{r}
#| label: gee-availability
#| echo: true
#| output: false
gee_status <- tryCatch({
  library(rgee)
  ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  aoi <- ee$Geometry$Rectangle(c(-128.5, 48.3, -123.0, 51.0))

  s2 <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))

  s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
    filterBounds(aoi)$
    filterDate("2019-06-01", "2021-09-30")$
    filter(ee$Filter$eq("instrumentMode", "IW"))

  data.frame(
    Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
    Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
    Period = "2019-06-01 to 2021-09-30",
    Scenes = c(s2$size()$getInfo(), s1$size()$getInfo())
  )
}, error = function(e) {
  data.frame(Sensor = c("Sentinel-2 MSI", "Sentinel-1 SAR"),
             Product = c("COPERNICUS/S2_SR_HARMONIZED", "COPERNICUS/S1_GRD"),
             Period = "2019-06-01 to 2021-09-30",
             Scenes = NA_integer_)
})
```

```{r}
#| label: tbl-imagery
#| tbl-cap: "Satellite scenes available over the study area for the analysis period."
fmt_tbl(gee_status)
```

## Lidar data

Canopy structure is derived from the LidarBC open lidar programme, which publishes classified point clouds for public download. Coverage over the study area comprises 18,840 tiles at a nominal density of eight points per square metre (@tbl-lidar), including acquisitions in 2019 that coincide with the concentration of the adelgid survey record. Tiles are classified to ground, low point and overlap but carry no vegetation classes, so canopy height is computed as height above a ground surface interpolated from ground returns rather than read from an existing classification.

```{r}
#| label: tbl-lidar
#| tbl-cap: "Lidar point cloud coverage over the study area and over the Comox field window. Acquisition years and point classes are summarised from a sample of up to 1000 tiles."
lidar <- read.csv(file.path(derived, "lidar-coverage.csv"), check.names = FALSE)
names(lidar) <- c("Extent", "Tiles", "Full index", "Acquisition years",
                  "Density (pts/m2)", "Point classes")
fmt_tbl(lidar)
```

## Data preparation

[PENDING. Cloud and terrain masking, compositing, radiometric terrain correction and speckle filtering of the radar stack, derivation of canopy height and structural metrics from the point clouds, and the common grid on which predictors are summarised to polygon means and dispersion statistics.]

## Model design

The response is the two-class agent label, adelgid against bark beetle, for survey polygons on Pacific silver fir. Predictors are grouped into three families, spectral, structural and radar, and three nested models are compared: spectral only, spectral with structure, and all three families together. Survey year and terrain are carried as covariates so that their contribution is estimated rather than absorbed into the sensor terms. Models are fitted with random forests, following @Campbell_2023, and compared at equal complexity.

Because polygons cluster in space, model performance is estimated by spatially blocked cross-validation rather than by random partitioning, which would allow spatially adjacent polygons to appear in both training and evaluation folds. The two classes are imbalanced by approximately `r imbalance` to 1 in favour of the bark beetle class, and resampling within the training folds is used to address this.

## Accuracy assessment

The primary metric is balanced accuracy, the mean of sensitivity and specificity, which is insensitive to the class imbalance described above in a way that overall accuracy is not. Per-class sensitivity and specificity and the full confusion matrix are reported alongside it, together with the support on which each is computed. The metric implementation is verified against cases whose values are known by construction (@tbl-checks).

```{r}
#| label: eval-metric
#| echo: true
#| output: false
balanced_accuracy <- function(truth, pred, positive = "IAB") {
  truth <- as.character(truth); pred <- as.character(pred)
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)
  sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
  spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  (sens + spec) / 2
}
```

```{r}
#| label: tbl-checks
#| tbl-cap: "Verification of the balanced accuracy implementation against cases with analytically known values."
t_perfect <- c("IAB", "IAB", "IBB", "IBB")
t_inverted <- c("IBB", "IBB", "IAB", "IAB")
imb_truth <- c("IAB", "IBB", "IBB", "IBB")

checks <- data.frame(
  Case = c("Perfect prediction",
           "Fully inverted prediction",
           "All assigned to majority class",
           "Imbalanced, all majority (overall accuracy 0.75)"),
  Expected = c(1, 0, 0.5, 0.5),
  Observed = c(
    balanced_accuracy(t_perfect, t_perfect),
    balanced_accuracy(t_perfect, t_inverted),
    balanced_accuracy(t_perfect, rep("IBB", 4)),
    balanced_accuracy(imb_truth, rep("IBB", 4))
  )
)
checks$Agrees <- abs(checks$Expected - checks$Observed) < 1e-9
fmt_tbl(checks)
```

# Results

## Model comparison

[PENDING. Balanced accuracy of the three nested models and the difference between the multi-sensor and spectral-only models, with its interval.]

```{r}
#| label: tbl-primary
#| tbl-cap: "Separation of the two damage agents by predictor set, under spatially blocked cross-validation."
primary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  `Balanced accuracy` = "pending",
  `Difference from spectral` = "pending",
  `95% CI` = "pending",
  check.names = FALSE
)
fmt_tbl(primary)
```

## Class performance

[PENDING. Per-class sensitivity and specificity, the confusion matrix and its support, and the relative importance of predictors grouped by sensor family.]

```{r}
#| label: tbl-secondary
#| tbl-cap: "Per-class performance of each model, with support."
secondary <- data.frame(
  `Predictor set` = c("Spectral", "Spectral and structure",
                      "Spectral, structure and radar"),
  Sensitivity = "pending",
  Specificity = "pending",
  Support = "pending",
  check.names = FALSE
)
fmt_tbl(secondary)
```

# Discussion

[PENDING. Interpretation of the model comparison, in relation to the spectral attribution failure reported by @Campbell_2023 and the commission-error reduction reported by @Jones_2015.]

## Limitations

Several limitations apply irrespective of the result. The labels are sketch-mapped from an aircraft and their spatial accuracy is acknowledged by the custodian to be variable, so polygon boundaries are approximate and some damage is missed entirely. The adelgid class is small, `r n_ba_iab` polygons on the shared host, and only `r n_iab_ms` polygons in the study area are rated moderate or worse, so the analysis is concentrated in the lightest damage classes where any spectral or structural signal is weakest. Agent attribution in the source data is the surveyor's judgement from the air and is not independently verified on the ground, so the models can at best reproduce that judgement and any systematic error in it propagates into the results. Field confirmation addresses this only partly, because the diagnostic signs of adelgid infestation, the woolly wax masses and stem gout, occur on living and declining trees and are lost from those long dead.

## Detection and attribution

This study evaluates attribution rather than detection, and the distinction is maintained throughout. Establishing that a canopy has changed is a different problem from identifying the agent responsible, and a model can perform well at the first while failing the second [@Campbell_2023]. Accuracy figures are therefore reported with their support, their class balance and their validation design.

# Conclusions

[PENDING. What the comparison shows, and what it implies for verification of aerial overview survey attributions in coastal British Columbia.]

# Data availability

Survey label extracts and the derived basemap are available in the project repository, together with the scripts that produce them from the public sources cited above. Sentinel-2 and Sentinel-1 imagery are available through Google Earth Engine, lidar point clouds through the LidarBC open data portal, and the survey archive through the BC Data Catalogue.

# References

::: {#refs}
:::
